---
title: "Data Filtration and Sanity Checks"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  cache.path = "cache/",
  fig.retina = 2
)

# Reproducibility
set.seed(1)

############################################################
# 0) Clear R "brain" + libraries
############################################################
rm(list = ls())
gc()

suppressPackageStartupMessages({
  library(tidyr)
  library(lubridate)
  library(readr)
  library(stringr)
  library(ggplot2)
  library(dplyr)
})

############################################################
# 1) Set working directory (analysis folder)
############################################################
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_04/Data_Filtration/")

############################################################
# 2) Load exported tables
#    (these should be created by your preprocessing script)
############################################################
abundance_table <- read_csv("../../Figure_03/LakeOverview/Species_Abundance_Table.csv")
gene_table <- read_csv("../../Figure_03/LakeOverview/Species_Gene_Table.csv")
protein_table <- read_csv("../../Figure_03/LakeOverview/Species_Protein_Table.csv")
meta_table <- read_csv("../../Figure_03/LakeOverview/Species_Meta_Table.csv")

############################################################
# 3) Quick inspection
############################################################
cat("=== Dimensions ===\n")
cat("species_abundance:", paste(dim(abundance_table), collapse = " x "), "\n")
cat("species_meta     :", paste(dim(meta_table), collapse = " x "), "\n")
cat("gene_table       :", paste(dim(gene_table), collapse = " x "), "\n")
cat("protein_table    :", paste(dim(protein_table), collapse = " x "), "\n\n")

cat("=== Glimpse ===\n")
glimpse(abundance_table)
glimpse(gene_table)
glimpse(protein_table)
glimpse(meta_table)

############################################################
# 4) Minimal key checks (optional but recommended)
############################################################
cat("\n=== Key checks ===\n")
cat("Unique species (abundance):", n_distinct(abundance_table$Species_ID), "\n")
cat("Unique species (meta):     ", n_distinct(meta_table$Species_ID), "\n")
cat("Unique species (protein):  ", n_distinct(protein_table$Species_ID), "\n")
cat("Unique dates (abundance):  ", n_distinct(abundance_table$Date), "\n")
cat("Unique dates (protein):    ", n_distinct(protein_table$Date), "\n")

############################################################
# 5) Keep only annotated proteins
############################################################
gene_table <- gene_table %>%
  filter(KEGG_ID != "Unclassified")

protein_table <- protein_table %>%
  filter(Protein_ID %in% gene_table$Protein_ID)
```

```{r}
############################################################
# PREP — Define proteomics dataset for transporter investment
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
})

# ----------------------------
# Parameters
# ----------------------------
groups_to_analyze      <- c("Oligo_like", "Copio_like", "Intermediate")
kegg_multi_delim_regex <- ";"

# ----------------------------
# 0) Build Step-1 universe robustly
#    - If protein_step1 exists, use it
#    - Else derive from protein_table (which is already loaded)
# ----------------------------
if (exists("protein_step1")) {
  protein_step1_local <- protein_step1
} else {
  stopifnot(exists("protein_table"))
  protein_step1_local <- protein_table %>%
    filter(Group_For_Plots %in% groups_to_analyze)
}

# Apply the Step-1 QC restrictions only if columns exist
prot_universe <- protein_step1_local

if ("Species_Specific_ProteinGroup" %in% names(prot_universe)) {
  prot_universe <- prot_universe %>% filter(Species_Specific_ProteinGroup)
}
if ("any_missing_taxonomy" %in% names(prot_universe)) {
  prot_universe <- prot_universe %>% filter(!any_missing_taxonomy)
}

# Safety checks
cat("\n================ PREP: PROTEOMICS FILTERING ================\n")
cat("Trophic groups included: ",
    paste(sort(unique(prot_universe$Group_For_Plots)), collapse = ", "), "\n")

# ----------------------------
# 1) Join KEGG_Subcategory and split multi-annotations proportionally
#    - NA subcategory is retained as "Unannotated" (counts as non-transporter)
# ----------------------------
stopifnot(exists("gene_table"))

prot_annot <- prot_universe %>%
  left_join(
    gene_table %>%
      select(Species_ID, Protein_ID, KEGG_Subcategory) %>%
      distinct(),
    by = c("Species_ID", "Protein_ID")
  ) %>%
  mutate(
    KEGG_Subcategory = as.character(KEGG_Subcategory),
    KEGG_Subcategory = str_replace_all(KEGG_Subcategory, "\\s+", " "),
    KEGG_Subcategory = na_if(str_trim(KEGG_Subcategory), ""),
    KEGG_Subcategory = if_else(is.na(KEGG_Subcategory), "Unannotated", KEGG_Subcategory)
  )

prot_split <- prot_annot %>%
  mutate(subcat_list = str_split(KEGG_Subcategory, kegg_multi_delim_regex)) %>%
  unnest(subcat_list) %>%
  mutate(
    subcat_list = str_trim(subcat_list),
    subcat_list = if_else(is.na(subcat_list) | subcat_list == "", "Unannotated", subcat_list)
  ) %>%
  group_by(Species_ID, Date, Protein_ID) %>%
  mutate(
    n_cats = n_distinct(subcat_list),
    weight = 1 / n_cats
  ) %>%
  ungroup() %>%
  mutate(
    Class = if_else(subcat_list == "Transporters", "Transporter", "Non-transporter")
  )

# ----------------------------
# 2) Structural validity per species×date (fractional, reviewer-proof)
#    - require ≥1 transporter equivalent AND ≥1 non-transporter equivalent
# ----------------------------
valid_dates <- prot_split %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    n_proteins           = n_distinct(Protein_ID),
    transporter_equiv    = sum(weight[Class == "Transporter"],     na.rm = TRUE),
    nontransporter_equiv = sum(weight[Class == "Non-transporter"], na.rm = TRUE),
    Has_Transporter      = transporter_equiv    >= 1,
    Has_NonTransporter   = nontransporter_equiv >= 1,
    Structurally_Valid   = Has_Transporter & Has_NonTransporter,
    .groups = "drop"
  )

valid_dates_only <- valid_dates %>%
  filter(Structurally_Valid)

# ----------------------------
# 3) Filter the *protein-level* table to valid species×date
#    (use prot_annot here so you keep one row per detected protein)
# ----------------------------
protein_filtered <- prot_annot %>%
  semi_join(valid_dates_only %>% select(Species_ID, Date),
            by = c("Species_ID", "Date"))

# ----------------------------
# 4) Reporting (your original section can stay as-is)
# ----------------------------
n_sd_before <- nrow(valid_dates)
n_sd_after  <- nrow(valid_dates_only)

n_species_before <- n_distinct(valid_dates$Species_ID)
n_species_after  <- n_distinct(valid_dates_only$Species_ID)

n_dates_before <- n_distinct(valid_dates$Date)
n_dates_after  <- n_distinct(valid_dates_only$Date)

n_protein_rows_before <- nrow(prot_universe)
n_protein_rows_after  <- nrow(protein_filtered)

cat("\n--- Overall retention ---\n")
cat("Species×date observations before:", n_sd_before, "\n")
cat("Species×date observations after : ", n_sd_after, "\n")
cat("Retained fraction              : ",
    round(n_sd_after / n_sd_before, 3), "\n\n")

cat("Unique species before:", n_species_before, "\n")
cat("Unique species after : ", n_species_after, "\n")
cat("Species retained     : ",
    round(n_species_after / n_species_before, 3), "\n\n")

cat("Unique dates before:", n_dates_before, "\n")
cat("Unique dates after : ", n_dates_after, "\n\n")

cat("Protein-level rows before:", n_protein_rows_before, "\n")
cat("Protein-level rows after : ", n_protein_rows_after, "\n")
cat("Protein-row retained      : ",
    round(n_protein_rows_after / n_protein_rows_before, 3), "\n")

group_retention <- valid_dates %>%
  group_by(Group_For_Plots) %>%
  summarise(
    SpeciesDate_Before = n(),
    SpeciesDate_After  = sum(Structurally_Valid),
    Retained_Fraction  = SpeciesDate_After / SpeciesDate_Before,
    Species_Before     = n_distinct(Species_ID),
    Species_After      = n_distinct(Species_ID[Structurally_Valid]),
    Dates_Before       = n_distinct(Date),
    Dates_After        = n_distinct(Date[Structurally_Valid]),
    .groups = "drop"
  ) %>%
  arrange(desc(SpeciesDate_Before))

cat("\n--- Retention by trophic group ---\n")
print(group_retention)

failure_modes <- valid_dates %>%
  mutate(
    Failure = case_when(
      Structurally_Valid ~ "Valid (has both)",
      !Has_Transporter & Has_NonTransporter ~ "Missing transporter",
      Has_Transporter & !Has_NonTransporter ~ "Missing non-transporter",
      TRUE ~ "Missing both / no proteins"
    )
  ) %>%
  count(Group_For_Plots, Failure) %>%
  group_by(Group_For_Plots) %>%
  mutate(Fraction = n / sum(n)) %>%
  ungroup()

cat("\n--- Why species×dates were removed (by group) ---\n")
print(failure_modes)

# Outputs:
#   protein_filtered  : protein-level table restricted to valid species×date
#   valid_dates_only  : species×date table with validity flags
############################################################

```

### Proteomics filtering for transporter investment analyses

- **Analysis universe:** We restricted the dataset to the three trophic groups used throughout the study (**Oligo_like, Copio_like, Intermediate**).
- **Structural validity filter (non-circular):** For each *species × date*, we retained only observations with evidence for both functional classes required to define a transporter fraction:  
  **≥ 1 transporter protein AND ≥ 1 non-transporter protein** detected.  
  *Rationale:* Without both classes present, the transporter fraction is undefined or trivially 0/1 and therefore not interpretable as regulation.

#### Overall retention
- **Species × date observations:** 9,239 → 6,633 retained (**71.8%**; 2,606 removed).
- **Species represented:** 265 → 241 retained (**90.9%**; 24 species removed entirely).
- **Sampling dates retained:** 36 → 36 (**no dates removed globally**; filtering acts at the species×date level).
- **Protein-level detections (rows):** 198,005 → 183,424 retained (**92.6%**).

#### Retention by trophic group (species × date level)
- **Oligo_like:** 3,618 → 3,069 retained (**84.8%**).
- **Intermediate:** 5,050 → 3,224 retained (**63.8%**).
- **Copio_like:** 571 → 340 retained (**59.5%**).

#### Why observations were removed (species × date level)
- The dominant failure mode was **missing transporter detections** (rather than missing non-transporters):
  - **Intermediate:** 34.9% missing transporter; 1.27% missing non-transporter.
  - **Copio_like:** 39.8% missing transporter; 0.70% missing non-transporter.
  - **Oligo_like:** 10.8% missing transporter; 4.39% missing non-transporter.

#### Interpretation
- The filtering step is **biologically justified and conservative**: it removes only cases where transporter investment cannot be meaningfully defined (absence of one class).
- The approach is **non-circular with respect to downstream investment analyses**: it does not optimize or threshold based on transporter fractions, only on the presence of both protein classes.
- The retained dataset preserves **broad taxonomic and temporal coverage** (**241 species across all 36 sampling dates**) and keeps the vast majority of protein detections (**92.6%**), indicating that the filter primarily removes sparse, uninformative species×date observations rather than discarding substantial proteomic signal.
- A detectable **group-wise imbalance** is present (highest retention in Oligo_like; lower retention in Intermediate and Copio_like), suggesting that subsequent analyses should (and will) include an explicit **coverage-based comparability QC** to ensure that inferred temporal trends are not driven by differential detectability a

```{r}
############################################################
# QC FIGURE — Proteome comparability vs coverage
# Two separate panels (side-by-side):
#   (1) Jaccard similarity (presence/absence overlap)
#   (2) Spearman correlation (abundance stability)
#
# Both panels restricted to y ∈ [0.3, 1.0]
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(patchwork)
})

abundance_col <- "Protein_Abundance_IBAQ_Norm"

# ----------------------------------------------------------
# 1) Quantitative proteome table
# ----------------------------------------------------------
prot_abund <- protein_filtered %>%
  group_by(Species_ID, Date, Protein_ID) %>%
  summarise(
    Abund = sum(.data[[abundance_col]], na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(is.finite(Abund), Abund > 0)

counts_df <- valid_dates_only %>%
  select(Species_ID, Date, n_proteins)

# ----------------------------------------------------------
# 2) Consecutive-date similarity metrics
# ----------------------------------------------------------
pair_metrics <- prot_abund %>%
  arrange(Species_ID, Date) %>%
  group_by(Species_ID) %>%
  group_modify(~{
    df <- .x
    dates <- sort(unique(df$Date))
    if (length(dates) < 2) return(tibble())

    out <- vector("list", length(dates) - 1)

    for (i in seq_len(length(dates) - 1)) {
      d1 <- dates[i]
      d2 <- dates[i + 1]

      x <- df %>% filter(Date == d1)
      y <- df %>% filter(Date == d2)

      s1 <- x$Protein_ID
      s2 <- y$Protein_ID

      inter_ids <- intersect(s1, s2)
      uni_ids   <- union(s1, s2)

      jaccard <- if (length(uni_ids) == 0) NA_real_
                 else length(inter_ids) / length(uni_ids)

      shared <- inner_join(x, y, by = "Protein_ID",
                           suffix = c("_1", "_2"))

      rho <- if (nrow(shared) < 3) NA_real_
             else suppressWarnings(
               cor(shared$Abund_1, shared$Abund_2,
                   method = "spearman")
             )

      out[[i]] <- tibble(
        Date1 = d1,
        Date2 = d2,
        Jaccard = jaccard,
        rho = rho
      )
    }
    bind_rows(out)
  }) %>%
  ungroup() %>%
  left_join(counts_df, by = c("Species_ID", "Date1" = "Date")) %>%
  rename(n1 = n_proteins) %>%
  left_join(counts_df, by = c("Species_ID", "Date2" = "Date")) %>%
  rename(n2 = n_proteins) %>%
  mutate(min_n_proteins = pmin(n1, n2))

# ----------------------------------------------------------
# 3) Summaries (coverage 1–20)
# ----------------------------------------------------------
summ_j <- pair_metrics %>%
  filter(is.finite(Jaccard), min_n_proteins >= 1, min_n_proteins <= 20) %>%
  group_by(min_n_proteins) %>%
  summarise(
    median = median(Jaccard),
    q25 = quantile(Jaccard, 0.25),
    q75 = quantile(Jaccard, 0.75),
    .groups = "drop"
  )

summ_rho <- pair_metrics %>%
  filter(is.finite(rho), min_n_proteins >= 1, min_n_proteins <= 20) %>%
  group_by(min_n_proteins) %>%
  summarise(
    median = median(rho),
    q25 = quantile(rho, 0.25),
    q75 = quantile(rho, 0.75),
    .groups = "drop"
  )

# ----------------------------------------------------------
# 4) Nature-style plots (y-axis restricted to 0.3–1.0)
# ----------------------------------------------------------
base_theme <- theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

p_jaccard <- ggplot(summ_j, aes(min_n_proteins, median)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75),
                width = 0.30, linewidth = 0.5) +
  scale_x_continuous(breaks = seq(1, 20, 1)) +
  scale_y_continuous(limits = c(0.3, 1), breaks = seq(0.3, 1, 0.1)) +
  labs(
    title = "Proteome overlap (Jaccard)",
    subtitle = "shared proteins / union",
    x = "Detected proteins per species×date\n(min of consecutive dates)",
    y = "Similarity (median ± IQR)"
  ) +
  base_theme

p_spearman <- ggplot(summ_rho, aes(min_n_proteins, median)) +
  geom_line(linewidth = 0.7) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = q25, ymax = q75),
                width = 0.30, linewidth = 0.5) +
  scale_x_continuous(breaks = seq(1, 20, 1)) +
  scale_y_continuous(limits = c(0.3, 1), breaks = seq(0.3, 1, 0.1)) +
  labs(
    title = "Quantitative stability (Spearman)",
    subtitle = "abundances of shared proteins",
    x = "Detected proteins per species×date\n(min of consecutive dates)",
    y = "Spearman ρ (median ± IQR)"
  ) +
  base_theme

(p_jaccard | p_spearman) +
  plot_annotation(
    title = "Proteome comparability increases with protein coverage",
    subtitle = "Consecutive sampling dates within species; stable overlap and abundance correlations emerge with increasing protein coverage"
  )

```

# final filter application
```{r}
############################################################
# FINAL QC FILTERING PIPELINE
#
# Filters applied in order:
#   Step 1) Define analysis universe (Oligo_like, Copio_like, Intermediate)
#           and keep only high-confidence proteomics rows.
#   Step 2) Structural validity (annotation-weighted):
#           retain species×date with ≥1 transporter-equivalent AND
#           ≥1 non-transporter-equivalent protein.
#           (Multi-annotation proteins are split 1/n across categories.)
#   Step 3) Coverage cutoff:
#           retain species×date with ≥10 detected proteins (unique Protein_ID).
#   Step 4) Temporal depth cutoff:
#           retain species with ≥5 valid dates remaining.
#
# Outputs:
#   - protein_final           : protein-level table after all filters
#   - valid_sd_final          : species×date table after all filters
#   - summary tables + absolute & fraction plots for retained vs removed
#
# Requires:
#   protein_table, gene_table (already loaded)
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(scales)
})

# ----------------------------
# User parameters
# ----------------------------
groups_to_analyze <- c("Oligo_like", "Copio_like", "Intermediate")
abundance_col <- "Protein_Abundance_IBAQ_Norm"

min_proteins_per_date <- 10
min_dates_per_species <- 5

# ----------------------------
# Helper: stacked plots (ABS + FRACTION)
# ----------------------------
plot_stacked_abs <- function(df_long, title, subtitle, ylab="Count") {
  ggplot(df_long, aes(x = Group_For_Plots, y = value, fill = Status)) +
    geom_col() +
    scale_fill_manual(values = c("Removed"="grey80", "Retained"="grey40")) +
    scale_y_continuous(labels = label_comma()) +
    theme_classic(base_size = 12) +
    labs(title = title, subtitle = subtitle, x = "Trophic group", y = ylab) +
    theme(plot.title = element_text(face="bold"),
          axis.title = element_text(face="bold"))
}

plot_stacked_frac <- function(df_long, title, subtitle, ylab="Fraction") {
  df_long %>%
    group_by(Group_For_Plots) %>%
    mutate(frac = value / sum(value)) %>%
    ungroup() %>%
    ggplot(aes(x = Group_For_Plots, y = frac, fill = Status)) +
    geom_col() +
    scale_fill_manual(values = c("Removed"="grey80", "Retained"="grey40")) +
    scale_y_continuous(limits = c(0,1), labels = percent_format(accuracy = 1)) +
    theme_classic(base_size = 12) +
    labs(title = title, subtitle = subtitle, x = "Trophic group", y = ylab) +
    theme(plot.title = element_text(face="bold"),
          axis.title = element_text(face="bold"))
}

# ----------------------------
# Helper: χ² vs Fisher robust selection
# ----------------------------
run_test <- function(wide_tbl) {
  # wide_tbl must have: Group_For_Plots, Removed, Retained
  mat <- as.matrix(wide_tbl[, c("Removed","Retained")])
  rownames(mat) <- wide_tbl$Group_For_Plots

  chi <- suppressWarnings(chisq.test(mat))
  if (any(is.na(chi$expected)) || any(chi$expected < 5)) {
    list(method="Fisher's exact test (expected < 5)", out=fisher.test(mat))
  } else {
    list(method="Pearson's χ² test", out=chi)
  }
}

# ==========================================================
# STEP 1 — Analysis universe + high-confidence proteomics
# ==========================================================
prot_base <- protein_table %>%
  filter(
    Group_For_Plots %in% groups_to_analyze,
    Species_Specific_ProteinGroup,
    !any_missing_taxonomy
  )

# Attach KEGG_Subcategory (protein-level annotation)
prot_annot <- prot_base %>%
  left_join(
    gene_table %>%
      select(Species_ID, Protein_ID, KEGG_Subcategory, KEGG_ID, KEGG_Annotation, KEGG_Category, Transporter_Family, Substracte_Group, Substrate_Category) %>%
      distinct(),
    by = c("Species_ID", "Protein_ID")
  )

# ==========================================================
# STEP 2 — Structural validity using weighted splitting
#   - split multi-annotations (e.g., "Transporters~Enzymes")
#   - each protein contributes 1/n to each listed category
#   - transporter_equiv and nontransporter_equiv are sums of weights
# ==========================================================
prot_split <- prot_annot %>%
  mutate(
    KEGG_Subcategory = as.character(KEGG_Subcategory),
    KEGG_Subcategory = str_replace_all(KEGG_Subcategory, "\\s+", " "),
    # split on "~" (adjust here if your delimiter differs)
    subcat_list = str_split(KEGG_Subcategory, "~")
  ) %>%
  unnest(subcat_list) %>%
  mutate(
    subcat_list = str_trim(subcat_list)
  ) %>%
  group_by(Species_ID, Date, Protein_ID) %>%
  mutate(
    n_cats = n_distinct(subcat_list),
    weight = 1 / n_cats
  ) %>%
  ungroup() %>%
  mutate(
    IsTransporter = subcat_list == "Transporters"
  )

# Species×date summaries
sd_summary <- prot_split %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    n_proteins = n_distinct(Protein_ID),
    transporter_equiv = sum(weight[IsTransporter], na.rm = TRUE),
    nontransporter_equiv = sum(weight[!IsTransporter], na.rm = TRUE),
    Has_Transporter = transporter_equiv >= 1,
    Has_NonTransporter = nontransporter_equiv >= 1,
    Pass_Structural = Has_Transporter & Has_NonTransporter,
    .groups = "drop"
  )

# ==========================================================
# STEP 3 — Coverage cutoff at species×date level (≥10 proteins)
# ==========================================================
sd_summary <- sd_summary %>%
  mutate(
    Pass_Coverage10 = Pass_Structural & (n_proteins >= min_proteins_per_date)
  )

# ==========================================================
# STEP 4 — Temporal depth cutoff at species level (≥5 valid dates)
# ==========================================================
species_depth <- sd_summary %>%
  group_by(Species_ID, Group_For_Plots) %>%
  summarise(
    Dates_Valid = sum(Pass_Coverage10),
    .groups = "drop"
  ) %>%
  mutate(Pass_Dates5 = Dates_Valid >= min_dates_per_species)

# Final retained species list
species_keep <- species_depth %>%
  filter(Pass_Dates5) %>%
  select(Species_ID) %>%
  distinct()

# Final retained species×date
valid_sd_final <- sd_summary %>%
  filter(Pass_Coverage10) %>%
  semi_join(species_keep, by = "Species_ID")

# Final retained protein-level table
protein_final <- prot_annot %>%
  semi_join(valid_sd_final %>% select(Species_ID, Date), by = c("Species_ID","Date"))

# ==========================================================
# SUMMARY + PLOTS (retained vs removed) — absolute + fraction
#   “Removed” is always relative to the SAME base universe:
#     prot_annot (protein rows with annotations, within 3 groups)
#     sd_summary (species×date within that universe)
#     species_depth (species within that universe)
# ==========================================================

# ---------- A) Species×date retained vs removed ----------
sd_status <- sd_summary %>%
  mutate(
    Status = ifelse(
      Pass_Coverage10 & (Species_ID %in% species_keep$Species_ID),
      "Retained", "Removed"
    )
  ) %>%
  count(Group_For_Plots, Status, name = "value") %>%
  tidyr::pivot_wider(names_from = Status, values_from = value, values_fill = 0) %>%
  # back to long for plotting (but keep wide for test)
  mutate(Total = Removed + Retained)

sd_wide <- sd_status %>%
  select(Group_For_Plots, Removed, Retained)

sd_long <- sd_wide %>%
  pivot_longer(cols = c(Removed, Retained), names_to = "Status", values_to = "value")

sd_test <- run_test(sd_wide)

cat("\n=== Species×date retention after ALL filters ===\n")
print(sd_wide)
cat("\nTest used:", sd_test$method, "\n")
print(sd_test$out)

# ---------- B) Species retained vs removed ----------
sp_wide <- species_depth %>%
  mutate(Status = ifelse(Pass_Dates5, "Retained", "Removed")) %>%
  count(Group_For_Plots, Status, name="value") %>%
  pivot_wider(names_from = Status, values_from = value, values_fill = 0) %>%
  select(Group_For_Plots, Removed, Retained)

sp_long <- sp_wide %>%
  pivot_longer(cols = c(Removed, Retained), names_to = "Status", values_to = "value")

sp_test <- run_test(sp_wide)

cat("\n=== Species retention (≥5 valid dates) ===\n")
print(sp_wide)
cat("\nTest used:", sp_test$method, "\n")
print(sp_test$out)

# ---------- C) Protein rows retained vs removed ----------
prot_rows_wide <- prot_annot %>%
  mutate(
    Status = ifelse(
      paste(Species_ID, Date) %in% paste(valid_sd_final$Species_ID, valid_sd_final$Date),
      "Retained", "Removed"
    )
  ) %>%
  count(Group_For_Plots, Status, name="value") %>%
  pivot_wider(names_from = Status, values_from = value, values_fill = 0) %>%
  select(Group_For_Plots, Removed, Retained)

prot_rows_long <- prot_rows_wide %>%
  pivot_longer(cols = c(Removed, Retained), names_to = "Status", values_to = "value")

prot_rows_test <- run_test(prot_rows_wide)

cat("\n=== Protein-row retention (after ALL filters) ===\n")
print(prot_rows_wide)
cat("\nTest used:", prot_rows_test$method, "\n")
print(prot_rows_test$out)

# ---------- D) Protein abundance signal retained vs removed ----------
# (sum of iBAQ; show absolute + fraction; no χ² because signal is continuous)
signal_wide <- prot_annot %>%
  mutate(
    Status = ifelse(
      paste(Species_ID, Date) %in% paste(valid_sd_final$Species_ID, valid_sd_final$Date),
      "Retained", "Removed"
    )
  ) %>%
  group_by(Group_For_Plots, Status) %>%
  summarise(value = sum(.data[[abundance_col]], na.rm = TRUE), .groups = "drop")

cat("\n=== Protein abundance signal (summed iBAQ) retained vs removed ===\n")
print(signal_wide)

# ==========================================================
# FIGURES: 2×2 FRACTION and 2×2 ABSOLUTE
# ==========================================================
# Fractions
p_sd_f  <- plot_stacked_frac(sd_long, "Species×date observations", "After structural + ≥10 proteins + ≥5 dates")
p_sp_f  <- plot_stacked_frac(sp_long, "Species", "After ≥5 valid dates per species")
p_pr_f  <- plot_stacked_frac(prot_rows_long, "Protein detections (rows)", "After all filters")
p_sig_f <- plot_stacked_frac(signal_wide, "Protein abundance signal", "Fraction of summed iBAQ retained")

fig_frac <- (p_sd_f | p_sp_f) / (p_pr_f | p_sig_f) +
  plot_annotation(
    title = "Filtering impact across trophic groups (fractions)",
    subtitle = paste0(
      "Structural validity (≥1 transporter & ≥1 non-transporter; annotation-weighted) → ",
      "coverage (≥", min_proteins_per_date, " proteins/date) → ",
      "temporal depth (≥", min_dates_per_species, " dates/species)"
    )
  )

# Absolute values
p_sd_a  <- plot_stacked_abs(sd_long, "Species×date observations", "After all filters", ylab="Species×date count")
p_sp_a  <- plot_stacked_abs(sp_long, "Species", "After ≥5 valid dates per species", ylab="Species count")
p_pr_a  <- plot_stacked_abs(prot_rows_long, "Protein detections (rows)", "After all filters", ylab="Protein rows")
p_sig_a <- plot_stacked_abs(signal_wide, "Protein abundance signal", "After all filters", ylab="Summed iBAQ (norm)")

fig_abs <- (p_sd_a | p_sp_a) / (p_pr_a | p_sig_a) +
  plot_annotation(
    title = "Filtering impact across trophic groups (absolute values)",
    subtitle = paste0(
      "Structural validity → ≥", min_proteins_per_date, " proteins/date → ≥",
      min_dates_per_species, " dates/species"
    )
  )

print(fig_frac)
print(fig_abs)

# ==========================================================
# Return final objects you will use downstream
# ==========================================================
cat("\n=== FINAL DATASETS ===\n")
cat("Final retained species×date:", nrow(valid_sd_final), "\n")
cat("Final retained species:", n_distinct(valid_sd_final$Species_ID), "\n")
cat("Final protein rows:", nrow(protein_final), "\n")

```



### Proteomics quality control and filtering strategy: rationale, expectations, and statistical evaluation

#### Overview and motivation
To robustly quantify transporter investment over time, we applied a **stepwise, non-circular filtering strategy** designed to exclude species–date observations that are biologically or statistically uninformative, while preserving as much quantitative proteomic signal as possible. Each filter addresses a specific source of instability (annotation incompleteness, low proteome coverage, insufficient temporal replication) and is evaluated explicitly for potential bias across trophic groups.

---

### Stepwise filtering logic (expectations *a priori*)

1. **Structural validity (functional presence filter)**  
   *Criterion:* retain species×date observations with evidence for **both transporters and non-transporters** (≥1 equivalent protein of each class, annotation-weighted).  
   *Expectation:*  
   - Removes dates where transporter fractions are undefined or trivially 0/1 due to detection limits.  
   - Disproportionately affects low-coverage observations but should not strongly bias species-level retention.

2. **Proteome coverage cutoff (≥10 detected proteins per date)**  
   *Criterion:* retain species×date observations with ≥10 distinct proteins detected.  
   *Expectation:*  
   - Stabilizes estimates of relative protein investment by reducing small-N effects.  
   - Removes a substantial fraction of species×date observations, especially for rarer taxa, but should retain the majority of **protein abundance signal**.

3. **Temporal depth cutoff (≥5 valid dates per species)**  
   *Criterion:* retain only species represented by ≥5 filtered sampling dates.  
   *Expectation:*  
   - Ensures that means and temporal trends are based on minimally sufficient replication.  
   - May remove entire species, but should do so **proportionally across trophic groups**, if the cutoff is not biologically biased.

---

### Observed outcomes and interpretation

#### Species×date observations
- A large fraction of species×date observations is removed after all filters, as expected when enforcing both functional presence and coverage constraints.
- **Statistical test:** Pearson’s χ² test shows a significant difference in retention fractions across trophic groups (*p* < 2.2×10⁻¹⁶).  
- **Interpretation:**  
  This result reflects known differences in detectability and temporal persistence between trophic strategies (e.g. Copio-like vs. Intermediate taxa), rather than a methodological artifact. Importantly, this filter acts at the *date* level and does not imply systematic loss of entire species.

#### Species-level retention (≥5 valid dates)
- The majority of species in all trophic groups are retained (168 species total).
- **Statistical test:** Pearson’s χ² test shows **no significant difference** in species retention across trophic groups (*p* = 0.156).  
- **Interpretation:**  
  The temporal depth cutoff does **not** preferentially exclude species from any trophic group. This directly addresses concerns of circularity or group-specific bias introduced by QC thresholds.

#### Protein detections (rows)
- Most protein detections are retained across all trophic groups.
- **Statistical test:** Pearson’s χ² test indicates a significant difference (*p* ≈ 3.3×10⁻¹³).  
- **Interpretation:**  
  While statistically detectable, these differences are small in relative terms and reflect differences in total proteome size and sampling depth between groups, not selective loss of specific functional categories.

#### Protein abundance signal (summed iBAQ)
- **Key result:** Across all trophic groups, **the vast majority of total protein abundance signal is retained** (>85–90%).  
- **Interpretation:**  
  Even though many low-coverage observations are removed, quantitative proteomic signal is strongly preserved. This demonstrates that filtering primarily removes noise-dominated observations rather than biologically meaningful signal.

---

### Overall assessment

- The filtering strategy is **biologically motivated, conservative, and non-circular** with respect to downstream transporter investment analyses.
- Species-level retention is **statistically unbiased across trophic groups**, satisfying a central reviewer concern.
- Differences observed at the species×date and protein-row levels are expected consequences of ecological and detectability differences, not artifacts of the filtering logic.
- Crucially, **protein abundance signal is largely retained**, ensuring that subsequent analyses are driven by robust quantitative data rather than detection limits.

**Conclusion:**  
The final dataset (3,958 species×date observations, 168 species, 129,110 protein detections) provides a statistically and biologically sound basis for analyzing temporal patterns in transporter versus non-transporter investment across trophic strategies.

```{r}
############################################################
# Expected dataframes in your environment:
#   protein_table  (must contain: Species_ID, Date, Protein_ID, Group_For_Plots,
#                   Protein_Abundance_IBAQ_Norm)
#   gene_table     (must contain: Species_ID, Protein_ID, KEGG_Subcategory)
#
# Optional (used only if present):
#   protein_table$Species_Specific_ProteinGroup (logical)
#   protein_table$any_missing_taxonomy (logical)
#
# Filters applied:
#   1) per Species_ID × Date: require ≥1 transporter AND ≥1 non-transporter (after proportional splitting)
#   2) per Species_ID × Date: require ≥10 proteins detected
#   3) per Species_ID: require ≥5 remaining dates
#
# Multi-annotated KEGG_Subcategory (e.g. "Transporters~Enzymes"):
#   - protein counts and abundance are split proportionally (1/n) across subcategories
#   - then added to transporter vs non-transporter buckets
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(patchwork)
  library(scales)
})

# ----------------------------
# Parameters (edit if needed)
# ----------------------------
groups_to_analyze        <- c("Oligo_like", "Copio_like", "Intermediate")
min_proteins_per_date    <- 10
min_dates_per_species    <- 5
kegg_multi_delim_regex   <- "~"                 # delimiter inside KEGG_Subcategory
abundance_col            <- "Protein_Abundance_IBAQ_Norm"

# ----------------------------
# Basic checks
# ----------------------------
stopifnot(exists("protein_table"), exists("gene_table"))

req_prot <- c("Species_ID", "Date", "Protein_ID", "Group_For_Plots", abundance_col)
req_gene <- c("Species_ID", "Protein_ID", "KEGG_Subcategory")

miss_prot <- setdiff(req_prot, names(protein_table))
miss_gene <- setdiff(req_gene, names(gene_table))

if (length(miss_prot) > 0) stop("protein_table missing: ", paste(miss_prot, collapse = ", "))
if (length(miss_gene) > 0) stop("gene_table missing: ", paste(miss_gene, collapse = ", "))

# Normalize Date
protein_table <- protein_table %>% mutate(Date = as.Date(Date))

# ----------------------------
# 1) Define analysis universe
# ----------------------------
prot_base <- protein_table %>%
  filter(Group_For_Plots %in% groups_to_analyze)

if ("Species_Specific_ProteinGroup" %in% names(prot_base)) {
  prot_base <- prot_base %>% filter(Species_Specific_ProteinGroup)
}
if ("any_missing_taxonomy" %in% names(prot_base)) {
  prot_base <- prot_base %>% filter(!any_missing_taxonomy)
}

# Attach KEGG_Subcategory at protein level
prot_annot <- prot_base %>%
  left_join(
    gene_table %>%
      select(Species_ID, Protein_ID, KEGG_Subcategory) %>%
      distinct(),
    by = c("Species_ID", "Protein_ID")
  ) %>%
  mutate(
    KEGG_Subcategory = as.character(KEGG_Subcategory),
    KEGG_Subcategory = str_replace_all(KEGG_Subcategory, "\\s+", " "),
    KEGG_Subcategory = na_if(str_trim(KEGG_Subcategory), ""),
    KEGG_Subcategory = if_else(is.na(KEGG_Subcategory), "Unannotated", KEGG_Subcategory)
  )

# ----------------------------
# 2) Proportional splitting of multi-annotated KEGG_Subcategory
#    -> weights sum to 1 per (Species_ID, Date, Protein_ID)
# ----------------------------
prot_split <- prot_annot %>%
  mutate(subcat_list = str_split(KEGG_Subcategory, kegg_multi_delim_regex)) %>%
  unnest(subcat_list) %>%
  mutate(
    subcat_list = str_trim(subcat_list),
    subcat_list = if_else(is.na(subcat_list) | subcat_list == "", "Unannotated", subcat_list)
  ) %>%
  group_by(Species_ID, Date, Protein_ID) %>%
  mutate(
    n_cats = n_distinct(subcat_list),
    weight = 1 / n_cats
  ) %>%
  ungroup() %>%
  mutate(
    Class = if_else(subcat_list == "Transporters", "Transporter", "Non-transporter")
  )

# ----------------------------
# 3) Species×date filters:
#    (a) structural validity: ≥1 transporter AND ≥1 non-transporter equivalent
#    (b) coverage: ≥10 proteins detected
# ----------------------------
sd_summary <- prot_split %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    n_proteins           = n_distinct(Protein_ID),
    transporter_equiv    = sum(weight[Class == "Transporter"],     na.rm = TRUE),
    nontransporter_equiv = sum(weight[Class == "Non-transporter"], na.rm = TRUE),
    Pass_Structural      = transporter_equiv >= 1 & nontransporter_equiv >= 1,
    Pass_Coverage10      = Pass_Structural & (n_proteins >= min_proteins_per_date),
    .groups = "drop"
  )

# Keep only passing species×date
keep_sd <- sd_summary %>%
  filter(Pass_Coverage10) %>%
  select(Species_ID, Date, Group_For_Plots)

# ----------------------------
# 4) Species-level filter: ≥5 dates remaining
# ----------------------------
keep_species <- keep_sd %>%
  count(Species_ID, Group_For_Plots, name = "n_dates") %>%
  filter(n_dates >= min_dates_per_species) %>%
  select(Species_ID)

keep_sd_final <- keep_sd %>%
  semi_join(keep_species, by = "Species_ID")

# ----------------------------
# 5) Restrict split table to retained (Species_ID, Date)
# ----------------------------
prot_retained <- prot_split %>%
  semi_join(keep_sd_final, by = c("Species_ID", "Date"))

# ----------------------------
# 6) Aggregations for figure panels
# ----------------------------

# A) retained species per trophic group (stacked single bar)
species_counts <- keep_sd_final %>%
  distinct(Species_ID, Group_For_Plots) %>%
  count(Group_For_Plots, name = "n_species") %>%
  mutate(panel = "Retained species")

# B) total proteins (equivalent counts) + stacked transporter vs non
protein_equiv_counts <- prot_retained %>%
  group_by(Group_For_Plots, Class) %>%
  summarise(value = sum(weight, na.rm = TRUE), .groups = "drop") %>%
  mutate(panel = "Protein detections (equiv. count)")

# C) total protein abundance (sum iBAQ) stacked transporter vs non
protein_abund <- prot_retained %>%
  mutate(abund = .data[[abundance_col]]) %>%
  group_by(Group_For_Plots, Class) %>%
  summarise(value = sum(weight * abund, na.rm = TRUE), .groups = "drop") %>%
  mutate(panel = "Protein abundance (sum iBAQ)")

# D) transporter abundance only (absolute scale)
transporter_abund <- prot_retained %>%
  filter(Class == "Transporter") %>%
  mutate(abund = .data[[abundance_col]]) %>%
  group_by(Group_For_Plots) %>%
  summarise(value = sum(weight * abund, na.rm = TRUE), .groups = "drop") %>%
  mutate(panel = "Transporter abundance (sum iBAQ)")

# order groups as in your workflow
group_levels <- groups_to_analyze
species_counts$Group_For_Plots      <- factor(species_counts$Group_For_Plots, levels = group_levels)
protein_equiv_counts$Group_For_Plots<- factor(protein_equiv_counts$Group_For_Plots, levels = group_levels)
protein_abund$Group_For_Plots       <- factor(protein_abund$Group_For_Plots, levels = group_levels)
transporter_abund$Group_For_Plots   <- factor(transporter_abund$Group_For_Plots, levels = group_levels)

# ----------------------------
# 7) Plot styling helpers
# ----------------------------
theme_pub <- function() {
  theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major.x = element_blank(),
      axis.title.x = element_blank(),
      legend.title = element_blank(),
      plot.title = element_text(face = "bold", size = 12),
      strip.text = element_text(face = "bold")
    )
}

# ----------------------------
# 8) Build panels
# ----------------------------

# Panel A: single stacked bar per trophic group
pA <- ggplot(species_counts, aes(x = Group_For_Plots, y = n_species, fill = Group_For_Plots)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  geom_text(aes(label = n_species), vjust = -0.3, size = 3.5) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.08))) +
  labs(title = "Retained species (n)") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Panel B: stacked transporter vs non (protein equivalent counts)
pB <- ggplot(protein_equiv_counts, aes(x = Group_For_Plots, y = value, fill = Class)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = comma) +
  labs(title = "Protein detections (equivalent counts)", y = "Equivalent protein count") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Panel C: stacked transporter vs non (sum iBAQ)
pC <- ggplot(protein_abund, aes(x = Group_For_Plots, y = value, fill = Class)) +
  geom_col(width = 0.7) +
  scale_y_continuous(labels = scientific) +
  labs(title = "Protein abundance (sum iBAQ)", y = "Summed iBAQ") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Panel D: transporter abundance only (sum iBAQ)
pD <- ggplot(transporter_abund, aes(x = Group_For_Plots, y = value, fill = Group_For_Plots)) +
  geom_col(width = 0.7, show.legend = FALSE) +
  scale_y_continuous(labels = scientific) +
  labs(title = "Transporter abundance (sum iBAQ)", y = "Summed iBAQ (transporters)") +
  theme_pub() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

# Combine (2×2)
final_fig <- (pA | pB) / (pC | pD) +
  plot_annotation(
    title = "Dataset retained after filtering (≥10 proteins/date; ≥5 dates/species; transporter + non-transporter present)",
    theme = theme(plot.title = element_text(face = "bold", size = 14))
  )

# Display
final_fig

# ----------------------------
# 9) Save (publication-ready)
# ----------------------------
ggsave(
  filename = "Figure_Retained_Dataset_Overview.pdf",
  plot = final_fig,
  width = 12, height = 8, units = "in", dpi = 300
)

# ----------------------------
# 10) Optional: export the retained keys for downstream analyses
# ----------------------------
retained_species <- keep_sd_final %>% distinct(Species_ID, Group_For_Plots)
retained_species_dates <- keep_sd_final %>% distinct(Species_ID, Date, Group_For_Plots)



```
# valid dates retained
```{r}
############################################################
# HEATMAP: How many valid species×dates per sampling date
# (36 dates on x-axis; trophic groups on y-axis)
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
})

stopifnot(exists("valid_sd_final"))

# Ensure Date is Date
valid_sd_final <- valid_sd_final %>% mutate(Date = as.Date(Date))

date_group_counts <- valid_sd_final %>%
  distinct(Species_ID, Date, Group_For_Plots) %>%
  count(Group_For_Plots, Date, name = "n_valid_species") %>%
  # make sure missing combinations become 0
  tidyr::complete(Group_For_Plots, Date, fill = list(n_valid_species = 0)) %>%
  mutate(
    Group_For_Plots = factor(Group_For_Plots, levels = groups_to_analyze)
  )

p_counts <- ggplot(date_group_counts, aes(x = Date, y = Group_For_Plots, fill = n_valid_species)) +
  geom_tile(color = NA) +
  scale_x_date(date_breaks = "1 week", date_labels = "%b %d") +
  scale_fill_viridis_c(option = "C", labels = comma) +
  labs(
    title = "Valid species×date coverage over time",
    subtitle = "Counts after all filters (structural validity + ≥10 proteins/date + ≥5 dates/species)",
    x = "Sampling date",
    y = "Trophic group",
    fill = "Valid species"
  ) +
  theme_classic(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold")
  )

p_counts

```


```{r}
dates_per_species <- valid_sd_final %>%
  distinct(Species_ID, Date, Group_For_Plots) %>%
  count(Group_For_Plots, Species_ID, name = "n_valid_dates")

ggplot(dates_per_species, aes(x = n_valid_dates)) +
  geom_histogram(binwidth = 1, boundary = 0, closed = "left",
                 fill = "grey40", color = "white") +
  facet_wrap(~ Group_For_Plots, nrow = 1, scales = "free_y") +
  scale_x_continuous(
    breaks = seq(0, 36, 2),
    limits = c(0, 36)
  ) +
  labs(
    title = "Temporal coverage per species after filtering",
    subtitle = "Distribution of valid sampling dates per species (≥5-date cutoff applied)",
    x = "Number of valid sampling dates per species",
    y = "Number of species"
  ) +
  theme_classic(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title = element_text(face = "bold"),
    strip.text = element_text(face = "bold")
  )


```

```{r}
############################################################
# Species persistence across dates
# Panel A: "survival" curves = #species with >= X valid dates
# Panel B: compact distribution (box + jitter)
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(tidyr)
  library(patchwork)
  library(scales)
})

stopifnot(exists("valid_sd_final"))

# --- 1) Dates-per-species
dps <- valid_sd_final %>%
  distinct(Species_ID, Date, Group_For_Plots) %>%
  count(Group_For_Plots, Species_ID, name = "n_valid_dates") %>%
  mutate(Group_For_Plots = factor(Group_For_Plots, levels = groups_to_analyze))

# --- 2) Persistence curve data: for each group and threshold X, how many species have >= X dates
curve_df <- dps %>%
  group_by(Group_For_Plots) %>%
  summarise(max_dates = max(n_valid_dates), .groups = "drop") %>%
  tidyr::expand(Group_For_Plots, threshold = 1:36) %>%
  left_join(dps, by = "Group_For_Plots") %>%
  group_by(Group_For_Plots, threshold) %>%
  summarise(n_species_ge = sum(n_valid_dates >= threshold), .groups = "drop")

# --- Styling helpers (Nature-ish: clean, no heavy grids, subtle axes)
theme_nature <- theme_classic(base_size = 12) +
  theme(
    axis.title = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    legend.title = element_blank(),
    legend.position = "top",
    legend.justification = "left"
  )

# Panel A: persistence (survival) curves
pA <- ggplot(curve_df, aes(x = threshold, y = n_species_ge, linetype = Group_For_Plots)) +
  geom_step(linewidth = 0.9) +
  geom_vline(xintercept = 5, linewidth = 0.5, linetype = "dashed") +
  annotate("text", x = 5, y = Inf, label = "≥5 dates cutoff", vjust = 1.4, hjust = -0.05, size = 3) +
  scale_x_continuous(breaks = seq(0, 36, 6), limits = c(1, 36)) +
  scale_y_continuous(labels = comma) +
  labs(
    title = "Species persistence across sampling dates",
    subtitle = "Number of species retained with ≥X valid dates",
    x = "Minimum number of valid dates (X)",
    y = "Number of species (≥X dates)"
  ) +
  theme_nature +
  theme(
    panel.grid = element_blank()
  )

# Panel B: compact distribution (box + jitter)
# (Shows “how many dates each species spans”, without an ugly histogram)
pB <- ggplot(dps, aes(x = Group_For_Plots, y = n_valid_dates)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, linewidth = 0.6) +
  geom_jitter(width = 0.12, height = 0, size = 0.9, alpha = 0.35) +
  geom_hline(yintercept = 5, linewidth = 0.5, linetype = "dashed") +
  scale_y_continuous(breaks = seq(0, 36, 6), limits = c(0, 36)) +
  labs(
    title = "Valid-date span per species",
    subtitle = "Each point is a species",
    x = NULL,
    y = "Valid dates per species"
  ) +
  theme_nature +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "none"
  )

# Combine: A big, B compact
final_fig <- pA + pB + plot_layout(widths = c(1.7, 1)) +
  plot_annotation(tag_levels = "A")

final_fig

# Optional save
ggsave("Figure_SpeciesPersistence_NatureStyle.pdf", final_fig, width = 11, height = 4.2, units = "in")


```
```{r}
############################################################
# CSV EXPORT — retained proteins, species, and dates only
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
})

stopifnot(exists("protein_final"), exists("valid_sd_final"))

# Normalize Date
protein_final  <- protein_final  %>% mutate(Date = as.Date(Date))
valid_sd_final <- valid_sd_final %>% mutate(Date = as.Date(Date))

# ----------------------------
# 1) Retained protein-level table
#    (main input for downstream analyses)
# ----------------------------
write_csv(
  protein_final,
  "retained_protein_table.csv"
)

# ----------------------------
# 2) Retained species × date table
#    (authoritative analysis universe)
# ----------------------------
retained_species_dates <- valid_sd_final %>%
  distinct(Species_ID, Date, Group_For_Plots) %>%
  arrange(Group_For_Plots, Species_ID, Date)

write_csv(
  retained_species_dates,
  "retained_species_dates.csv"
)

# ----------------------------
# 3) Retained species list
# ----------------------------
retained_species <- retained_species_dates %>%
  distinct(Species_ID, Group_For_Plots) %>%
  arrange(Group_For_Plots, Species_ID)

write_csv(
  retained_species,
  "retained_species.csv"
)

# ----------------------------
# 4) OPTIONAL but recommended:
#    Weighted transporter vs non-transporter split
#    (ready-to-use for investment calculations)
# ----------------------------
if (exists("prot_split")) {

  prot_split_retained <- prot_split %>%
    semi_join(
      retained_species_dates,
      by = c("Species_ID", "Date", "Group_For_Plots")
    )

  write_csv(
    prot_split_retained,
    "retained_protein_split_weighted.csv"
  )
}

cat("\n=== CSV EXPORT COMPLETE ===\n")
cat("Files written:\n")
cat(" - retained_protein_table.csv\n")
cat(" - retained_species_dates.csv\n")
cat(" - retained_species.csv\n")
if (exists("prot_split")) {
  cat(" - retained_protein_split_weighted.csv\n")
}

```

```{r}
############################################################
# SUMMARY TABLE (no genes):
# - how many retained species per trophic group
# - valid dates per species (median/IQR/min/max)
# - retained proteins per species (median/IQR)
#
# Requires:
#   valid_sd_final  (Species_ID, Date, Group_For_Plots)
#   protein_final   (Species_ID, Date, Protein_ID)
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
})

stopifnot(exists("valid_sd_final"), exists("protein_final"))

# Normalize Date just in case
valid_sd_final <- valid_sd_final %>% mutate(Date = as.Date(Date))
protein_final  <- protein_final  %>% mutate(Date = as.Date(Date))

# If Protein_ID column name differs, adjust here
stopifnot("Protein_ID" %in% names(protein_final))

# ----------------------------
# 1) Valid dates per species (retained universe)
# ----------------------------
species_dates <- valid_sd_final %>%
  distinct(Species_ID, Date, Group_For_Plots) %>%
  count(Group_For_Plots, Species_ID, name = "n_valid_dates")

# ----------------------------
# 2) Retained proteins per species (count unique Protein_ID across retained rows)
#    (this is "how many proteins a species has" in the retained dataset)
# ----------------------------
species_proteins <- protein_final %>%
  distinct(Species_ID, Protein_ID) %>%
  count(Species_ID, name = "n_proteins_retained")

# Species-level table (optional to inspect/export)
species_summary <- species_dates %>%
  left_join(species_proteins, by = "Species_ID") %>%
  mutate(n_proteins_retained = replace_na(n_proteins_retained, 0L))

# ----------------------------
# 3) Group-level summary table
# ----------------------------
summary_by_group <- species_summary %>%
  group_by(Group_For_Plots) %>%
  summarise(
    n_species_retained = n_distinct(Species_ID),

    dates_median = median(n_valid_dates),
    dates_IQR    = paste0(quantile(n_valid_dates, 0.25), "–", quantile(n_valid_dates, 0.75)),
    dates_min    = min(n_valid_dates),
    dates_max    = max(n_valid_dates),

    proteins_median = median(n_proteins_retained),
    proteins_IQR    = paste0(quantile(n_proteins_retained, 0.25), "–", quantile(n_proteins_retained, 0.75)),

    .groups = "drop"
  ) %>%
  arrange(factor(Group_For_Plots, levels = c("Oligo_like","Copio_like","Intermediate")))

cat("\n=== Retained dataset summary by trophic group ===\n")
print(summary_by_group)

cat("\n=== Species-level summary (first 10) ===\n")
print(head(species_summary, 10))

# Optional exports
readr::write_csv(summary_by_group, "Summary_Retained_ByGroup_NoGenes.csv")
readr::write_csv(species_summary,  "Summary_Retained_BySpecies_NoGenes.csv")

```

