---
title: Proteomic Data Overview Contaminants
subtitle: SpringBloom2021 - Genome-resolved Metaproteomic Dataset
date: 09.04.2025
author: Cyrill Hofer
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    code_folding: hide
    thumbnails: true
    gallery: false
    fig_width: 15
    df_print: kable
    fig_caption: yes
    use_bookdown: true
    css: custom.css
---

# Data Preparation
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Clear R's Brain
rm(list = ls())

# Set the working directory
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Supplementary/SpringBloom_Metaproteomics/ProteomeContaminants/")

# Load the libraries
library(readr)
library(dplyr)
library(ggplot2)
library(DT)
library(plotly)
library(tidyr)
library(scales)
library(forcats)

# Load the color palette
source("ColorPalette.R")

# Load the data
dat <- read_csv(file = "Proteomic_Data_Overview_Contaminants.csv")
```

# Number of contaminant-peptides per sample per database ?
```{r}
# Prepare the data
plot_dat <- dat %>% 
  select(Date, Unique_Peptides, Unique_Peptides_In_One_Replicate, Unique_Peptides_In_Both_Replicates, Overall_Unique_Peptides_Detected) %>%
  distinct() %>%
  pivot_longer(cols = c(Unique_Peptides_In_One_Replicate, Unique_Peptides_In_Both_Replicates), names_to = "Replicate_Type", values_to = "Number_Of_Peptides") %>%
  mutate(Replicate_Type = factor(Replicate_Type, levels = c("Unique_Peptides_In_One_Replicate", "Unique_Peptides_In_Both_Replicates"), labels = c("One Replicate", "Both Replicates"))) %>%
  mutate(Date = as.factor(Date))

# Plot the number of peptides for the costumdb
Plot_Number_Of_Peptides <- ggplot(plot_dat, aes(x = Date, y = Number_Of_Peptides, fill = Replicate_Type)) +
  geom_col(width = 0.8, position = "stack", color = ColorPalette[16]) +
  scale_y_continuous(limits = c(0, 200), breaks = seq(0, 200, by = 20)) +
  scale_fill_manual(values = c("One Replicate"  = ColorPalette[4], "Both Replicates" = ColorPalette[2]), labels = c("One Replicate", "Both Replicates")) +
  labs(x = "Date", y = "Contaminant Peptides [n]", title = "Database: Costum", subtitle = paste("Total Unique Contaminant Peptides Detected:", plot_dat %>% select(Overall_Unique_Peptides_Detected) %>% distinct()), fill = "Detected within:") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title  = element_text(size = 14, face = "bold"))

Plot_Number_Of_Peptides
```

# Abundance of contaminant-peptides per sample per database ?
```{r}
# Prepare the data
plot_dat <- dat %>% 
  select(Date, Total_Peptide_Abundance, Abundance_Peptides_In_One_Replicate, Abundance_Peptides_In_Both_Replicates, Overall_Unique_Peptides_Detected) %>%
  distinct() %>%
  pivot_longer(cols = c(Abundance_Peptides_In_One_Replicate, Abundance_Peptides_In_Both_Replicates), names_to = "Replicate_Type", values_to = "Abundance_Of_Peptides") %>%
  mutate(Replicate_Type = factor(Replicate_Type, levels = c("Abundance_Peptides_In_One_Replicate", "Abundance_Peptides_In_Both_Replicates"), labels = c("One Replicate", "Both Replicates"))) %>%
  mutate(Date = as.factor(Date))

# Plot the number of peptides for the costumdb
Plot_Peptide_Abundance <- ggplot(plot_dat, aes(x = Date, y = Abundance_Of_Peptides, fill = Replicate_Type)) +
  geom_col(width = 0.8, position = "stack", color = ColorPalette[16]) +
  scale_fill_manual(values = c("One Replicate"  = ColorPalette[4], "Both Replicates" = ColorPalette[2]), labels = c("One Replicate", "Both Replicates")) +
  labs(x = "Date", y = "Contaminant Peptide [Signal]", title = "Database: Costum", subtitle = paste("Total Unique Contaminant Peptides Detected:", plot_dat %>% select(Overall_Unique_Peptides_Detected) %>% distinct()), fill = "Detected within:") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title  = element_text(size = 14, face = "bold"))

Plot_Peptide_Abundance
```


# Number of contaminants per sample per database ?
```{r}
# Prepare the data
plot_dat <- dat %>% 
  filter(Protein_Abundance_IBAQ > 0) %>%
  select(Date, Unique_Proteins, Unique_Proteins_In_One_Replicate, Unique_Proteins_In_Both_Replicates, Overall_Unique_Proteins_Detected) %>%
  distinct() %>%
  pivot_longer(cols = c(Unique_Proteins_In_One_Replicate, Unique_Proteins_In_Both_Replicates), names_to = "Replicate_Type", values_to = "Number_Of_Proteins") %>%
  mutate(Replicate_Type = factor(Replicate_Type, levels = c("Unique_Proteins_In_One_Replicate", "Unique_Proteins_In_Both_Replicates"), labels = c("One Replicate", "Both Replicates"))) %>%
  mutate(Date = as.factor(Date))

# Plot the number of Proteins for the costumdb
Plot_Number_Of_Proteins <- ggplot(plot_dat, aes(x = Date, y = Number_Of_Proteins, fill = Replicate_Type)) +
  geom_col(width = 0.8, position = "stack", color = ColorPalette[16]) +
  scale_y_continuous(limits = c(0, 40), breaks = seq(0, 40, by = 10)) +
  scale_fill_manual(values = c("One Replicate"  = ColorPalette[4], "Both Replicates" = ColorPalette[2]), labels = c("One Replicate", "Both Replicates")) +
  labs(x = "Date", y = "Contamintns [n]", title = "Database: Costum", subtitle = paste("Total Unique Contaminants Detected:", plot_dat %>% select(Overall_Unique_Proteins_Detected) %>% distinct()),fill = "Detected within:") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title  = element_text(size = 14, face = "bold"))

Plot_Number_Of_Proteins
```

# Abundance of contaminants per sample per database ?
```{r}
# Prepare the data
plot_dat <- dat %>% 
  filter(Protein_Abundance_IBAQ > 0) %>%
  select(Date, Total_Protein_Abundance, Abundance_Proteins_In_One_Replicate, Abundance_Proteins_In_Both_Replicates, Overall_Unique_Proteins_Detected) %>%
  distinct() %>%
  pivot_longer(cols = c(Abundance_Proteins_In_One_Replicate, Abundance_Proteins_In_Both_Replicates), names_to = "Replicate_Type", values_to = "Abundance_Of_Proteins") %>%
  mutate(Replicate_Type = factor(Replicate_Type, levels = c("Abundance_Proteins_In_One_Replicate", "Abundance_Proteins_In_Both_Replicates"), labels = c("One Replicate", "Both Replicates"))) %>%
  mutate(Date = as.factor(Date))

# Plot the number of Proteins for the costumdb
Plot_Protein_Abundance <- ggplot(plot_dat, aes(x = Date, y = Abundance_Of_Proteins, fill = Replicate_Type)) +
  geom_col(width = 0.8, position = "stack", color = ColorPalette[16]) +
  scale_fill_manual(values = c("One Replicate"  = ColorPalette[4], "Both Replicates" = ColorPalette[2]), labels = c("One Replicate", "Both Replicates")) +
  labs(x = "Date", y = "Contaminants [Signal]", title = "Database: Costum", subtitle = paste("Total Unique Contaminants Detected:", plot_dat %>% select(Overall_Unique_Proteins_Detected) %>% distinct()), fill = "Detected within:") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y = element_text(size = 10), 
    axis.title.x = element_text(size = 12),
    axis.title.y = element_text(size = 12),
    plot.title  = element_text(size = 14, face = "bold"))

Plot_Protein_Abundance
```

# Number of contaminants per sample per type of conataminant
```{r}
# Identify top‑8 species by total Protein_Abundance_IBAQ
Top8_Species_Highest_Abundance <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  group_by(Species_ID) %>%
  summarise(TotalIntensity = sum(Protein_Abundance_IBAQ, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalIntensity)) %>%
  slice_head(n = 8) %>%
  pull(Species_ID)

# Prepare plotting data (absolute abundance per sample)
plot_dat_abund <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  mutate(Species_Conc = if_else(
    Species_ID %in% Top8_Species_Highest_Abundance, 
    Species_ID, 
    "Rest"
  )) %>%
  group_by(Date, Species_Conc) %>%
  summarise(
    Protein_Abundance = sum(Protein_Abundance_IBAQ, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Date = as.factor(Date),
    Species_Conc = fct_relevel(
      Species_Conc, 
      Top8_Species_Highest_Abundance, 
      after = 0
    ) %>%
    fct_relevel("Rest", after = Inf)
  )

# plot the data
Plot_Abundance_Of_Proteins_Per_Species <- ggplot(plot_dat_abund,
    aes(x = Date, y = Protein_Abundance, fill = Species_Conc)
  ) +
  geom_col(
    width    = 0.8, 
    position = "stack", 
    color    = ColorPalette[16]
  )  +
  labs(
    x     = "Date",
    y     = "contaminant Abundance [IBAQ]",
    title = "Database: Costum",
    subtitle = "Top 8 species by total signal; others = Rest",
    fill  = "Species:"
  ) +
  theme_minimal() +
  theme(
    axis.text.x     = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y     = element_text(size = 10),
    axis.title.x    = element_text(size = 12),
    axis.title.y    = element_text(size = 12),
    plot.title      = element_text(size = 14, face = "bold"),
    plot.subtitle   = element_text(size = 10),
    legend.position = "right"
  ) +
  scale_fill_manual(
    values = ColorPalette[1:(length(Top8_Species_Highest_Abundance) + 1)]
  )

```


# Abundance of contaminants per sample per type of contaminant?
```{r}
# 1) Identify top-8 species by total Protein_Abundance_IBAQ (excluding contaminants)
Top8_Species_Highest_Abundance <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  group_by(Species_ID) %>%
  summarise(TotalIntensity = sum(Protein_Abundance_IBAQ, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(TotalIntensity)) %>%
  slice_head(n = 8) %>%
  pull(Species_ID)

# 2) Prepare plotting data with percent abundance per sample
plot_dat <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  # collapse species outside top 8 into "Rest"
  mutate(Species_Conc = if_else(Species_ID %in% Top8_Species_Highest_Abundance,
                                Species_ID, "Rest")) %>%
  # sum abundance per Date × Species_Conc
  group_by(Date, Species_Conc) %>%
  summarise(Protein_Abundance = sum(Protein_Abundance_IBAQ, na.rm = TRUE),
            .groups = "drop") %>%
  # compute percent per Date
  group_by(Date) %>%
  mutate(Percent_Protein_Abundance = Protein_Abundance / sum(Protein_Abundance) * 100) %>%
  ungroup() %>%
  mutate(
    Date = as.factor(Date),
    Species_Conc = fct_relevel(Species_Conc, Top8_Species_Highest_Abundance, after = 0) %>%
                   fct_relevel("Rest", after = Inf)
  )

# 3) Build the stacked 100%-bar plot
Plot_Percentage_Of_Proteins_Per_Species <- ggplot(plot_dat,
    aes(x = Date, y = Percent_Protein_Abundance, fill = Species_Conc)
  ) +
  geom_col(width = 0.8, position = "stack", color = ColorPalette[16]) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  labs(
    x        = "Date",
    y        = "Contaminant Abundance [%]",
    title    = "Database: Costum",
    subtitle = "Top 8 species by total signal; others = Rest",
    fill     = "Species:"
  ) +
  theme_minimal() +
  theme(
    axis.text.x    = element_text(angle = 45, hjust = 1, size = 10),
    axis.text.y    = element_text(size = 10),
    axis.title.x   = element_text(size = 12),
    axis.title.y   = element_text(size = 12),
    plot.title     = element_text(size = 14, face = "bold"),
    plot.subtitle  = element_text(size = 10),
    legend.position = "right"
  ) +
  # assign one color per species (8 + Rest)
  scale_fill_manual(values = ColorPalette[1:(length(Top8_Species_Highest_Abundance) + 1)])

# 4) Display interactively and save
Plot_Percentage_Of_Proteins_Per_Species
```

# Number of peptides per contaminant ?
```{r}
# Prepare the plotting data for Number of Peptides per Protein
plot_dat <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>% 
   mutate(Overall_Unique_Proteins_Detected = n_distinct(Protein_ID)) %>%
  select(Nr_Peptides, Protein_ID, Overall_Unique_Proteins_Detected) %>%
  distinct()

# Create the histogram plot
Plot_Number_Of_Peptides_Per_Protein <- ggplot(plot_dat, aes(x = factor(Nr_Peptides))) +
  geom_bar(fill = ColorPalette[2], color = "black") +
  scale_y_continuous(limits = c(0, 10), breaks = seq(0, 10, by = 2)) +
  labs(y = "Contaminants [n]", x = "Contaminant Peptides [n]", title = "Distribution of Number of Peptides per Contaminant", subtitle = paste("Total Unique Contaminants Detected:", unique(plot_dat$Overall_Unique_Proteins_Detected)), x = "Number of Peptides", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        plot.title  = element_text(size = 14, face = "bold"))


# display the plots
Plot_Number_Of_Peptides_Per_Protein
```

```{r}
library(dplyr)
library(ggplot2)
library(scales)
library(patchwork)

# -----------------------------------------------------------------------------
# Data summary
# -----------------------------------------------------------------------------
cont_summary <- dat %>%
  group_by(Date) %>%
  summarise(
    num_contaminants        = n_distinct(Protein_ID),
    cont_abundance          = unique(Total_Protein_Abundance),
    total_protein_count_all = unique(Total_Protein_Count_All_Proteins),
    total_protein_abund_all = unique(Total_Protein_Abundance_All_Proteins),
    percent_cont_signal     = 100 * cont_abundance / total_protein_abund_all,
    .groups = "drop"
  )

# -----------------------------------------------------------------------------
# Panel A: contaminant counts vs total proteins per sample
# -----------------------------------------------------------------------------
p1 <- ggplot(
  cont_summary,
  aes(x = total_protein_count_all, y = num_contaminants)
) +
  geom_point(
    shape = 21,
    fill = ColorPalette[1],
    color = "black",
    alpha = 0.75,
    size = 3.2,
    stroke = 0.4
  ) +
  geom_hline(
    yintercept = median(cont_summary$num_contaminants, na.rm = TRUE),
    linetype = "dashed",
    color = "grey40",
    linewidth = 0.5
  ) +
  scale_x_continuous(
    limits = c(4000, 12000),
    breaks = seq(4000, 12000, by = 2000),
    labels = comma,
    expand = expansion(mult = c(0.01, 0.02))
  ) +
  scale_y_continuous(
    limits = c(10, 40),
    breaks = seq(10, 40, by = 5),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = "Total proteins per sample (n)",
    y = "Contaminant proteins [n]"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88", linewidth = 0.3),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# -----------------------------------------------------------------------------
# Panel B: contaminant signal vs total protein abundance per sample
# -----------------------------------------------------------------------------
p2 <- ggplot(
  cont_summary,
  aes(x = total_protein_abund_all, y = percent_cont_signal)
) +
  geom_point(
    shape = 21,
    fill = ColorPalette[2],
    color = "black",
    alpha = 0.75,
    size = 3.2,
    stroke = 0.4
  ) +
  geom_hline(
    yintercept = median(cont_summary$percent_cont_signal, na.rm = TRUE),
    linetype = "dashed",
    color = "grey40",
    linewidth = 0.5
  ) +
  scale_x_continuous(
    labels = label_scientific(digits = 2),
    breaks = pretty_breaks(n = 6),
    expand = expansion(mult = c(0.02, 0.03))
  ) +
  scale_y_continuous(
    limits = c(0, 12),
    breaks = seq(0, 12, by = 2),
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0))
  ) +
  labs(
    x = "Total protein abundance per sample (iBAQ)",
    y = "Contaminant signal [% of total]"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(color = "grey88", linewidth = 0.3),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# -----------------------------------------------------------------------------
# Combine panels
# -----------------------------------------------------------------------------
combined <- p1 + p2 + plot_layout(ncol = 2)

print(combined)

ggsave(
  filename = "Contaminant_independence_plot_refined.png",
  plot = combined,
  width = 11,
  height = 5,
  units = "in",
  dpi = 600
)
```


```{r}
library(dplyr)
library(ggplot2)
library(forcats)
library(readr)
library(stringr)
library(patchwork)

# ----------------------------
# 3) Manual annotation lookup
# ----------------------------
top8_labels_df <- tibble(
  Species_ID = c(
    "Y-FGCZCont00051",
    "Y-FGCZCont00180",
    "Y-FGCZCont00100",
    "Y-FGCZCont00159",
    "Y-FGCZCont00257",
    "Y-FGCZCont00261",
    "Y-FGCZCont00147",
    "Y-FGCZCont00416"
  ),
  Annotation_short = c(
    "Human keratin (P04264)",
    "Trypsin, pig (P00761)",
    "Human keratin (P35527)",
    "Human keratin (P13645)",
    "Rubber elongation factor protein (P15252)",
    "iRT protein with AAAA spacer",
    "Human keratin (P35908)",
    "Human keratin (Q7Z794)"
  )
) %>%
  filter(Species_ID %in% Top8_Species_Highest_Abundance) %>%
  mutate(
    Pretty_Label = paste0(Species_ID, " = ", Annotation_short)
  )

plot_label_vec <- c(
  setNames(top8_labels_df$Annotation_short, top8_labels_df$Species_ID),
  "Rest" = "Rest"
)

# ----------------------------
# 4) Absolute abundance plot
# ----------------------------
plot_dat_abund <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  mutate(Species_Conc = if_else(
    Species_ID %in% Top8_Species_Highest_Abundance,
    Species_ID,
    "Rest"
  )) %>%
  group_by(Date, Species_Conc) %>%
  summarise(
    Protein_Abundance = sum(Protein_Abundance_IBAQ, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Date = as.factor(Date),
    Species_Conc = fct_relevel(
      Species_Conc,
      Top8_Species_Highest_Abundance,
      after = 0
    ) %>%
      fct_relevel("Rest", after = Inf)
  )

p_abs <- ggplot(
  plot_dat_abund,
  aes(x = Date, y = Protein_Abundance, fill = Species_Conc)
) +
  geom_col(
    width = 0.8,
    position = "stack",
    color = ColorPalette[16]
  ) +
  labs(
    x = "Date",
    y = "Contaminant abundance [IBAQ]",
    title = "Absolute abundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11),
    plot.title = element_text(size = 13, face = "bold"),
    legend.position = "none"
  ) +
  scale_fill_manual(
    values = ColorPalette[1:(length(Top8_Species_Highest_Abundance) + 1)],
    breaks = c(Top8_Species_Highest_Abundance, "Rest"),
    labels = plot_label_vec
  )

# ----------------------------
# 5) Relative abundance plot
# ----------------------------
plot_dat_rel <- dat %>%
  filter(Protein_Abundance_IBAQ > 0) %>%
  mutate(Species_Conc = if_else(
    Species_ID %in% Top8_Species_Highest_Abundance,
    Species_ID,
    "Rest"
  )) %>%
  group_by(Date, Species_Conc) %>%
  summarise(
    Protein_Abundance = sum(Protein_Abundance_IBAQ, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(Date) %>%
  mutate(
    Percent_Protein_Abundance = Protein_Abundance / sum(Protein_Abundance) * 100
  ) %>%
  ungroup() %>%
  mutate(
    Date = as.factor(Date),
    Species_Conc = fct_relevel(
      Species_Conc,
      Top8_Species_Highest_Abundance,
      after = 0
    ) %>%
      fct_relevel("Rest", after = Inf)
  )

p_rel <- ggplot(
  plot_dat_rel,
  aes(x = Date, y = Percent_Protein_Abundance, fill = Species_Conc)
) +
  geom_col(
    width = 0.8,
    position = "stack",
    color = ColorPalette[16]
  ) +
  scale_y_continuous(limits = c(0, 100), breaks = seq(0, 100, by = 20)) +
  labs(
    x = "Date",
    y = "Relative contaminant abundance [%]",
    title = "Relative abundance"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    axis.text.y = element_text(size = 10),
    axis.title.x = element_text(size = 11),
    axis.title.y = element_text(size = 11),
    plot.title = element_text(size = 13, face = "bold"),
    legend.position = "none"
  ) +
  scale_fill_manual(
    values = ColorPalette[1:(length(Top8_Species_Highest_Abundance) + 1)],
    breaks = c(Top8_Species_Highest_Abundance, "Rest"),
    labels = plot_label_vec
  )

# ----------------------------
# 6) Build annotation panel below
# ----------------------------
legend_text <- c(
  "Annotated contaminants:",
  paste0(seq_len(nrow(top8_labels_df)), ". ", top8_labels_df$Pretty_Label),
  paste0(length(Top8_Species_Highest_Abundance) + 1, ". Rest = all remaining contaminants")
)

legend_df <- tibble(
  x = 0,
  y = rev(seq_along(legend_text)),
  label = legend_text
)

p_annot <- ggplot(legend_df, aes(x = x, y = y, label = label)) +
  geom_text(hjust = 0, vjust = 1, size = 3.5, lineheight = 1.05) +
  xlim(0, 1) +
  ylim(0.5, length(legend_text) + 0.5) +
  theme_void()

# ----------------------------
# 7) Combine into one figure
# ----------------------------
final_plot <- (p_abs | p_rel) | p_annot +
  plot_layout(heights = c(3.2, 1.2)) +
  plot_annotation(
    title = "Database: Custom",
    subtitle = "Top 8 contaminants by total signal; others grouped as Rest"
  )

# display
final_plot

# save
ggsave(
  filename = "Contaminants_absolute_relative_annotated.png",
  plot = final_plot,
  width = 18,
  height = 6,
  units = "in",
  dpi = 300
)
```

