---
title: "Trophic Index Overview"
author: "Cyrill Hofer"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 2
    df_print: paged
  pdf_document:
    toc: true
    toc_depth: 2
---

```{r}
# =============================================================================
# BLOCK 0 — Setup and data loading
# =============================================================================

rm(list = ls())
gc()
options(stringsAsFactors = FALSE)

suppressPackageStartupMessages({
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(ggrepel)
  library(scales)
  library(patchwork)
  library(tidyr)
})

# ---- Working directory ----
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Supplementary/TrophicIndex/LZ-BITS/")

# ---- Load data ----
train_dat <- read_csv("IndicatorSpecies_GenomicProperties.csv")
lake_dat  <- read_csv("Species_GenomicProperties_ObservedVsEstimated.csv")
species_dat <- read_csv("../../../../Data/Species_Data/Final_Data/Species_Table.csv")
source(file = "ColorPalette.R")

# ---- Core metadata columns ----
meta_cols <- c("Species_ID", "Taxonomy", "Accession", "Classification")

# ---- Enforce trophic class order (training data only) ----
train_dat <- train_dat %>%
  mutate(Classification = factor(
    Classification,
    levels = c("Oligotroph", "Mesotroph", "Copiotroph")
  ))

# ---- Minimal sanity check ----
cat("\nTraining set:", nrow(train_dat), "species\n")
print(table(train_dat$Classification, useNA = "ifany"))

# ---- Consistent colors for all plots ----
trophic_cols <- c(
  Oligotroph = "#2166AC",
  Mesotroph  = "grey60",
  Copiotroph = "#B2182B"
)

# prepare species dat
species_dat <- species_dat %>%
  select(Species_ID, Phylum, Class, Order, Family, Genus, Species) %>%
  distinct()

# ---- seed for reproducibility of bootstraps ---- 
set.seed(1234)
glimpse(species_dat)
glimpse(train_dat)
glimpse(lake_dat)
```

# Motivation for the trophic index
```{r}
# =============================================================================
# Supplementary opening figure for trophic index rationale
# Clean 3-panel version: genome size, trait correlations, PCA
# =============================================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(ggrepel)
library(patchwork)

train_dat <- train_dat %>%
  mutate(
    Classification = recode(Classification,
      Mesotroph = "Intermediate"
    )
  )

train_dat <- train_dat %>%
  mutate(
    Classification = factor(
      Classification,
      levels = c("Oligotroph", "Intermediate", "Copiotroph")
    )
  )

# -----------------------------------------------------------------------------
# Settings
# -----------------------------------------------------------------------------
trophic_cols <- c(
  Oligotroph = "#2166AC",
  Intermediate  = "grey60",
  Copiotroph = "#B2182B"
)

size_cutoff_bp  <- 2.1e6
size_cutoff_mbp <- 2.1

theme_nat <- theme_bw(base_size = 10) +
  theme(
    panel.grid = element_blank(),
    panel.border = element_rect(colour = "black", linewidth = 0.5),
    axis.title = element_text(size = 10),
    axis.text = element_text(size = 9, colour = "black"),
    legend.title = element_text(size = 9),
    legend.text = element_text(size = 8),
    legend.key = element_blank(),
    plot.title = element_text(size = 11, face = "bold"),
    plot.margin = margin(6, 6, 6, 6)
  )

train_plot <- train_dat %>%
  mutate(
    Genome_Mbp = GENOME_LENGTH / 1e6,
    Size_Class = ifelse(GENOME_LENGTH < size_cutoff_bp, "< 2.1 Mbp", ">= 2.1 Mbp"),
    Size_Class = factor(Size_Class, levels = c("< 2.1 Mbp", ">= 2.1 Mbp"))
  )

# -----------------------------------------------------------------------------
# Panel a — genome size vs trophic class
# -----------------------------------------------------------------------------
p_a <- ggplot(
  train_plot,
  aes(x = Genome_Mbp, y = Classification, colour = Classification)
) +
  geom_vline(
    xintercept = size_cutoff_mbp,
    linetype = "dashed",
    linewidth = 0.5,
    colour = "black"
  ) +
  geom_jitter(height = 0.12, width = 0, size = 2.8, alpha = 0.95) +
  scale_colour_manual(values = trophic_cols) +
  scale_x_continuous(
    breaks = pretty_breaks(n = 6),
    expand = expansion(mult = c(0.02, 0.04))
  ) +
  labs(
    x = "Genome size (Mbp)",
    y = NULL,
    title = "a"
  ) +
  annotate(
    "text",
    x = size_cutoff_mbp,
    y = 3.35,
    label = "2.1 Mbp cutoff",
    size = 3,
    vjust = -0.4
  ) +
  theme_nat +
  theme(
    legend.position = "none",
    plot.title.position = "plot"
  )

# -----------------------------------------------------------------------------
# Panel b — lower-triangle correlation matrix of selected traits
# -----------------------------------------------------------------------------
corr_df <- train_plot %>%
  transmute(
    `Genome size`    = Genome_Mbp,
    `Coding density` = CODING_DENSITY,
    `tRNA count`     = TOTAL_TRNA,
    `Sigma factors`  = NUM_SIGMA
  )

corr_mat <- cor(corr_df, method = "pearson", use = "pairwise.complete.obs")

corr_long <- as.data.frame(as.table(corr_mat), stringsAsFactors = FALSE) %>%
  rename(Trait1 = Var1, Trait2 = Var2, r = Freq)

trait_order <- colnames(corr_mat)

corr_long$Trait1 <- factor(corr_long$Trait1, levels = trait_order)
corr_long$Trait2 <- factor(corr_long$Trait2, levels = trait_order)

corr_long <- corr_long %>%
  mutate(
    i = as.integer(Trait1),
    j = as.integer(Trait2)
  ) %>%
  filter(i >= j)

p_b <- ggplot(corr_long, aes(x = Trait1, y = Trait2, fill = r)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r)), size = 3.2) +
  scale_fill_gradient2(
    low = "grey80",
    mid = "white",
    high = "black",
    midpoint = 0,
    limits = c(-1, 1)
  ) +
  labs(
    x = NULL,
    y = NULL,
    fill = "r",
    title = "b"
  ) +
  theme_nat +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    axis.text.y = element_text(hjust = 1),
    legend.position = "right",
    plot.title.position = "plot"
  )

# -----------------------------------------------------------------------------
# Panel c — PCA biplot with taxonomy labels
# -----------------------------------------------------------------------------
pca_input <- train_plot %>%
  transmute(
    Genome_size    = Genome_Mbp,
    Coding_density = CODING_DENSITY,
    tRNA_count     = TOTAL_TRNA,
    Sigma_factors  = NUM_SIGMA
  )

pca_fit <- prcomp(pca_input, center = TRUE, scale. = TRUE)

scores <- as.data.frame(pca_fit$x[, 1:2])
scores$Classification <- train_plot$Classification
scores$Taxonomy <- train_plot$Taxonomy
scores$Size_Class <- train_plot$Size_Class

loadings <- as.data.frame(pca_fit$rotation[, 1:2])
loadings$Trait <- rownames(loadings)

loadings <- loadings %>%
  mutate(
    Trait = recode(
      Trait,
      Genome_size    = "Genome size",
      Coding_density = "Coding density",
      tRNA_count     = "tRNA",
      Sigma_factors  = "Sigma factors"
    )
  )

var_exp <- round(100 * summary(pca_fit)$importance[2, 1:2], 1)

arrow_scale <- 2.3
loadings$PC1 <- loadings$PC1 * arrow_scale
loadings$PC2 <- loadings$PC2 * arrow_scale

# optional: shorten long taxonomy strings a bit if needed
# scores$Taxonomy <- gsub("^.*;s__", "", scores$Taxonomy)

p_c <- ggplot(
  scores,
  aes(x = PC1, y = PC2, colour = Classification, shape = Size_Class)
) +
  geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey75") +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey75") +
  geom_point(size = 2.8, alpha = 0.95) +
  geom_text_repel(
    aes(label = Taxonomy),
    size = 1,
    max.overlaps = Inf,
    box.padding = 0.25,
    point.padding = 0.18,
    segment.size = 0.2,
    min.segment.length = 0,
    show.legend = FALSE
  ) +
  geom_segment(
    data = loadings,
    aes(x = 0, y = 0, xend = PC1, yend = PC2),
    inherit.aes = FALSE,
    arrow = arrow(length = unit(0.18, "cm")),
    linewidth = 0.5,
    colour = "black"
  ) +
  geom_text_repel(
    data = loadings,
    aes(x = PC1, y = PC2, label = Trait),
    inherit.aes = FALSE,
    size = 3.2,
    colour = "black",
    box.padding = 0.35,
    point.padding = 0.2,
    segment.colour = NA
  ) +
  scale_colour_manual(values = trophic_cols) +
  scale_shape_manual(values = c(16, 17)) +
  labs(
    x = paste0("PC1 (", var_exp[1], "%)"),
    y = paste0("PC2 (", var_exp[2], "%)"),
    colour = "Classification",
    shape = "Size class",
    title = "c"
  ) +
  coord_cartesian(clip = "off") +
  theme_nat +
  theme(
    legend.position = "right",
    plot.title.position = "plot",
    plot.margin = margin(8, 30, 8, 8)
  )

# -----------------------------------------------------------------------------
# Assemble figure
# Give the PCA row clearly more height
# -----------------------------------------------------------------------------
final_fig <- (p_a | p_b) / p_c +
  plot_layout(
    heights = c(0.85, 1.55),
    widths = c(1.05, 1.0)
  )

print(final_fig)

# -----------------------------------------------------------------------------
# Save
# -----------------------------------------------------------------------------
ggsave(
  filename = "Supplementary_TrophicIndex_OpeningFigure.png",
  plot = final_fig,
  width = 12,
  height = 8.5,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

# Highlight from discrete genome size based cutoff towards continuous trophic axis
```{r}
# ============================================================
# Supplementary figure:
# Continuous trophic index vs genome size
#
# Panels:
#   a) Trophic index by trophic class
#      - raw genomes (points)
#      - class IQR (horizontal bars)
#      - class median (point)
#      - global Q1 / Q3
#
#   b) Genome size vs trophic index
#      - points colored by class
#      - regression line
#      - 2.1 Mbp cutoff (vertical)
#      - TI Q1 / Q3 (horizontal)
# ============================================================

library(dplyr)
library(ggplot2)
library(patchwork)

# ------------------------------------------------------------
# Theme
# ------------------------------------------------------------
theme_set(
  theme_classic(base_size = 10) +
    theme(
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black"),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      plot.title = element_text(face = "bold", hjust = 0)
    )
)

# ------------------------------------------------------------
# Prepare data
# ------------------------------------------------------------
train_plot <- train_dat %>%
  mutate(
    Classification = case_when(
      Classification %in% c("Oligotroph","Oligotrophic") ~ "Oligotroph",
      Classification %in% c("Mesotroph","Mesotrophic","Intermediate") ~ "Intermediate",
      Classification %in% c("Copiotroph","Copiotrophic") ~ "Copiotroph"
    )
  ) %>%
  filter(Classification %in% c("Oligotroph","Intermediate","Copiotroph")) %>%
  filter(if_all(c(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA), ~ !is.na(.))) %>%
  mutate(
    Classification = factor(Classification,
      levels = c("Oligotroph","Intermediate","Copiotroph")),
    GenomeSize_Mbp = GENOME_LENGTH / 1e6
  )

# ------------------------------------------------------------
# Colors
# ------------------------------------------------------------
cols <- c(
  "Oligotroph"   = ColorPalette[2],
  "Intermediate" = "grey65",
  "Copiotroph"   = ColorPalette[7]
)

# ------------------------------------------------------------
# Trophic index (PCA)
# ------------------------------------------------------------
pca_fit <- prcomp(
  train_plot %>% select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA),
  center = TRUE, scale. = TRUE
)

ti <- predict(pca_fit)[,1]
if (cor(train_plot$GENOME_LENGTH, ti) < 0) ti <- -ti

train_plot$TrophicIndex <- ti

# ------------------------------------------------------------
# Stats
# ------------------------------------------------------------
stats_df <- train_plot %>%
  group_by(Classification) %>%
  summarise(
    q1 = quantile(TrophicIndex, 0.25),
    med = median(TrophicIndex),
    q3 = quantile(TrophicIndex, 0.75),
    .groups = "drop"
  )

ti_q1 <- quantile(train_plot$TrophicIndex, 0.25)
ti_q3 <- quantile(train_plot$TrophicIndex, 0.75)

r_val <- cor(train_plot$GenomeSize_Mbp, train_plot$TrophicIndex)

# ------------------------------------------------------------
# Panel a — clean distribution (NO violins)
# ------------------------------------------------------------
p_a <- ggplot(train_plot,
              aes(x = TrophicIndex, y = Classification, color = Classification)) +
  
  # global quartiles
  geom_vline(xintercept = ti_q1, linetype = "dashed", color = "grey40") +
  geom_vline(xintercept = ti_q3, linetype = "dashed", color = "grey40") +
  
   # IQR bars
  geom_segment(data = stats_df,
               aes(x = q1, xend = q3,
                   y = Classification, yend = Classification),
               linewidth = 1,
               inherit.aes = FALSE,
               color = "black") +
  
  # median
  geom_point(data = stats_df,
             aes(x = med, y = Classification),
             size = 3.2,
             shape = 21,
             fill = "white",
             color = "black",
             stroke = 0.5,
             inherit.aes = FALSE) +
  
  # raw points
  geom_jitter(height = 0.08, size = 2.4, alpha = 0.95) +
  

  scale_color_manual(values = cols) +
  
  annotate("text", x = ti_q1, y = 3.4,
           label = paste0("Q1=", round(ti_q1,2)),
           angle = 90, vjust = -0.3, size = 3) +
  
  annotate("text", x = ti_q3, y = 3.4,
           label = paste0("Q3=", round(ti_q3,2)),
           angle = 90, vjust = -0.3, size = 3) +
  
  labs(x = "Trophic index", y = NULL, title = "a") +
  
  theme(legend.position = "none")

# ------------------------------------------------------------
# Panel b — relationship
# ------------------------------------------------------------
p_b <- ggplot(train_plot,
              aes(x = GenomeSize_Mbp, y = TrophicIndex, color = Classification)) +
  
  geom_vline(xintercept = 2.1, color = "grey40", linewidth = 0.6) +
  
  geom_hline(yintercept = ti_q1, linetype = "dashed", color = "grey40") +
  geom_hline(yintercept = ti_q3, linetype = "dashed", color = "grey40") +
  
  geom_point(size = 2.8, alpha = 0.95) +
  
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  
  scale_color_manual(values = cols) +
  
  annotate("text", x = 2.1, y = max(train_plot$TrophicIndex),
           label = "2.1 Mbp", angle = 90, vjust = -0.3, size = 3) +
  
  annotate("text",
           x = min(train_plot$GenomeSize_Mbp),
           y = ti_q1,
           label = paste0("Q1=", round(ti_q1,2)),
           hjust = 0, vjust = -0.4, size = 3) +
  
  annotate("text",
           x = min(train_plot$GenomeSize_Mbp),
           y = ti_q3,
           label = paste0("Q3=", round(ti_q3,2)),
           hjust = 0, vjust = -0.4, size = 3) +
  
  annotate("text",
           x = min(train_plot$GenomeSize_Mbp),
           y = max(train_plot$TrophicIndex),
           label = paste0("r = ", round(r_val,2)),
           hjust = 0, vjust = 1, size = 3.2) +
  
  labs(x = "Genome size (Mbp)", y = "Trophic index", title = "b") +
  
  theme(legend.position = "right")

# ------------------------------------------------------------
# Combine
# ------------------------------------------------------------
final_fig <- p_a | p_b

ggsave("TrophicIndex_Final_Clean.png", final_fig,
       width = 12, height = 7, dpi = 600)

final_fig
```

# show how LZ-BTIS taxa >80% complete, parasites, cyanobacteria and archea removed were estimated to 100% for trophic index fitting only
```{r}
# ============================================================
# Supplementary figure:
# completion-corrected trophic-index traits and changed taxa
#
# Assumptions:
#   - train_dat is already loaded
#   - lake_dat is already loaded
#   - ColorPalette is already loaded
#
# Output:
#   - One multi-panel figure with:
#       a) observed vs estimated genome size
#       b) observed vs estimated total tRNA count
#       c) observed vs estimated sigma factor count
#       d) dumbbell plot of taxa whose trophic-group assignment
#          changed after completion correction
# ============================================================

library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
library(scales)
library(forcats)
library(stringr)

# ------------------------------------------------------------
# Minimal plotting theme
# ------------------------------------------------------------
theme_set(
  theme_classic(base_size = 10) +
    theme(
      axis.text = element_text(color = "black"),
      axis.title = element_text(color = "black"),
      strip.background = element_blank(),
      strip.text = element_text(face = "bold"),
      legend.title = element_text(size = 9),
      legend.text = element_text(size = 8),
      plot.title = element_text(face = "bold", hjust = 0)
    )
)

# ------------------------------------------------------------
# Check required columns
# ------------------------------------------------------------
req_train <- c(
  "Species_ID",
  "GENOME_LENGTH",
  "CODING_DENSITY",
  "TOTAL_TRNA",
  "NUM_SIGMA"
)

req_lake <- c(
  "Species_ID",
  "Phylum",
  "Genus",
  "Completeness",
  "CODING_DENSITY",
  "GENOME_LENGTH",
  "Estimated_Genome_Size",
  "Obs_TOTAL_TRNA",
  "Est_TOTAL_TRNA",
  "Obs_NUM_SIGMA",
  "Est_NUM_SIGMA"
)

missing_train <- setdiff(req_train, colnames(train_dat))
missing_lake  <- setdiff(req_lake, colnames(lake_dat))

if (length(missing_train) > 0) {
  stop("train_dat is missing required columns: ",
       paste(missing_train, collapse = ", "))
}

if (length(missing_lake) > 0) {
  stop("lake_dat is missing required columns: ",
       paste(missing_lake, collapse = ", "))
}

# ------------------------------------------------------------
# Completeness classes and colors
# Adjust indices if you want slightly different palette entries
# ------------------------------------------------------------
lake_comp_cols <- c(
  "80-85%"  = ColorPalette[7],
  "85-90%"  = ColorPalette[5],
  "90-95%"  = ColorPalette[14],
  "95-100%" = ColorPalette[10]
)

bin_completeness <- function(x) {
  cut(
    x,
    breaks = c(80, 85, 90, 95, 100),
    include.lowest = TRUE,
    right = TRUE,
    labels = c("80-85%", "85-90%", "90-95%", "95-100%")
  )
}

# ------------------------------------------------------------
# Build trophic axis from the training genomes
# Uses the 4 trophic-index traits:
#   - genome size
#   - coding density
#   - total tRNA count
#   - sigma factor count
# ------------------------------------------------------------
train_traits <- train_dat %>%
  select(
    Species_ID,
    GENOME_LENGTH,
    CODING_DENSITY,
    TOTAL_TRNA,
    NUM_SIGMA
  ) %>%
  filter(
    if_all(c(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA), ~ !is.na(.))
  )

pca_fit <- prcomp(
  train_traits %>% select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA),
  center = TRUE,
  scale. = TRUE
)

train_scores <- predict(
  pca_fit,
  newdata = train_traits %>% select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA)
)[, 1]

# ------------------------------------------------------------
# Reverse PC1 if needed so that:
#   small-genome oligotroph-like taxa = lower trophic index
#   large-genome copiotroph-like taxa = higher trophic index
# ------------------------------------------------------------
if (cor(train_traits$GENOME_LENGTH, train_scores, use = "complete.obs") < 0) {
  pca_fit$rotation[, 1] <- -pca_fit$rotation[, 1]
  train_scores <- -train_scores
}

train_traits <- train_traits %>%
  mutate(TI_train = train_scores)

# ------------------------------------------------------------
# Define trophic-group cutoffs from the training distribution
# ------------------------------------------------------------
ti_q25 <- quantile(train_traits$TI_train, 0.25, na.rm = TRUE)
ti_q75 <- quantile(train_traits$TI_train, 0.75, na.rm = TRUE)

classify_trophic_group <- function(x, q25, q75) {
  case_when(
    x <= q25 ~ "Oligotroph",
    x >= q75 ~ "Copiotroph",
    TRUE ~ "Intermediate"
  )
}

# ------------------------------------------------------------
# Prepare lake data
# Only MAGs with >=80% completeness are considered
# ------------------------------------------------------------
lake_proj <- lake_dat %>%
  mutate(Species_ID = as.character(Species_ID)) %>%
  filter(!is.na(Completeness), Completeness >= 80) %>%
  mutate(
    GS_obs = GENOME_LENGTH,
    GS_est = Estimated_Genome_Size,
    TRNA_obs = Obs_TOTAL_TRNA,
    TRNA_est = Est_TOTAL_TRNA,
    SIGMA_obs = Obs_NUM_SIGMA,
    SIGMA_est = Est_NUM_SIGMA,
    completeness_bin = bin_completeness(Completeness)
  ) %>%
  filter(
    if_all(c(
      GS_obs, GS_est,
      TRNA_obs, TRNA_est,
      SIGMA_obs, SIGMA_est,
      CODING_DENSITY
    ), ~ !is.na(.))
  )

# ------------------------------------------------------------
# Project observed lake trait values onto the training PCA axis
# Coding density is retained as observed in both projections
# ------------------------------------------------------------
obs_mat <- lake_proj %>%
  transmute(
    GENOME_LENGTH = GS_obs,
    CODING_DENSITY = CODING_DENSITY,
    TOTAL_TRNA = TRNA_obs,
    NUM_SIGMA = SIGMA_obs
  )

# ------------------------------------------------------------
# Project completion-corrected lake trait values onto the same
# training PCA axis
# ------------------------------------------------------------
est_mat <- lake_proj %>%
  transmute(
    GENOME_LENGTH = GS_est,
    CODING_DENSITY = CODING_DENSITY,
    TOTAL_TRNA = TRNA_est,
    NUM_SIGMA = SIGMA_est
  )

# ------------------------------------------------------------
# Calculate observed and estimated trophic index for each lake
# taxon, then assign trophic groups using training cutoffs
# ------------------------------------------------------------
lake_proj <- lake_proj %>%
  mutate(
    TI_obs = predict(pca_fit, newdata = obs_mat)[, 1],
    TI_est = predict(pca_fit, newdata = est_mat)[, 1]
  ) %>%
  mutate(
    Group_obs = classify_trophic_group(TI_obs, ti_q25, ti_q75),
    Group_est = classify_trophic_group(TI_est, ti_q25, ti_q75),
    Changed   = if_else(Group_obs != Group_est, "Changed", "Unchanged"),
    delta_TI  = TI_est - TI_obs
  )

plot_dat <- lake_proj

# ------------------------------------------------------------
# Create labels for taxa whose trophic group changed
# y-axis label format:
#   Phylum | Genus | Species_ID
# ------------------------------------------------------------
changed_dat <- plot_dat %>%
  filter(Changed == "Changed") %>%
  mutate(
    y_label = paste(
      if_else(is.na(Phylum) | Phylum == "", "NA", Phylum),
      if_else(is.na(Genus)  | Genus  == "", "NA", Genus),
      if_else(is.na(Species_ID) | Species_ID == "", "NA", Species_ID),
      sep = " | "
    ),
    y_label = fct_reorder(y_label, TI_est)
  )

# ------------------------------------------------------------
# Helper function for observed-vs-estimated trait panels
# Uses discrete completeness colors
# ------------------------------------------------------------
obs_est_panel_discrete <- function(df, xvar, yvar, xlab, ylab, mbp = FALSE) {
  xr <- range(c(df[[xvar]], df[[yvar]]), na.rm = TRUE)

  p <- ggplot(df, aes(x = .data[[xvar]], y = .data[[yvar]], color = completeness_bin)) +
    geom_abline(
      intercept = 0,
      slope = 1,
      linetype = "dashed",
      linewidth = 0.4,
      color = "grey55"
    ) +
    geom_point(size = 1.7, alpha = 0.9) +
    scale_color_manual(
      values = lake_comp_cols,
      drop = FALSE,
      name = "Completeness"
    ) +
    coord_cartesian(xlim = xr, ylim = xr, expand = TRUE) +
    labs(x = xlab, y = ylab) +
    theme(aspect.ratio = 1)

  if (mbp) {
    p <- p +
      scale_x_continuous(labels = label_number(scale = 1e-6, accuracy = 0.1)) +
      scale_y_continuous(labels = label_number(scale = 1e-6, accuracy = 0.1))
  }

  p
}

# ------------------------------------------------------------
# Top row:
# completion-corrected trophic-index traits
# ------------------------------------------------------------
p_genome <- obs_est_panel_discrete(
  plot_dat,
  xvar = "GS_obs",
  yvar = "GS_est",
  xlab = "Observed genome size (Mbp)",
  ylab = "Estimated genome size (Mbp)",
  mbp = TRUE
) + ggtitle("a")

p_trna <- obs_est_panel_discrete(
  plot_dat,
  xvar = "TRNA_obs",
  yvar = "TRNA_est",
  xlab = "Observed tRNA count",
  ylab = "Estimated tRNA count"
) + ggtitle("b")

p_sigma <- obs_est_panel_discrete(
  plot_dat,
  xvar = "SIGMA_obs",
  yvar = "SIGMA_est",
  xlab = "Observed sigma factors",
  ylab = "Estimated sigma factors"
) + ggtitle("c")

top_row <- (p_genome + p_trna + p_sigma) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# ------------------------------------------------------------
# Bottom panel:
# taxa whose trophic-group assignment changed after correction
#
# open circles  = TI based on observed incomplete MAG values
# filled points = TI based on completion-corrected values
# dashed lines  = training-set Q25 and Q75 thresholds
# ------------------------------------------------------------
if (nrow(changed_dat) > 0) {

  p_changed <- ggplot(changed_dat) +
    geom_segment(
      aes(
        y = y_label,
        yend = y_label,
        x = TI_obs,
        xend = TI_est,
        color = completeness_bin
      ),
      linewidth = 0.7,
      alpha = 0.9
    ) +
    geom_point(
      aes(
        x = TI_obs,
        y = y_label,
        fill = completeness_bin
      ),
      size = 2.0,
      shape = 21,
      color = "black",
      stroke = 0.35
    ) +
    geom_point(
      aes(
        x = TI_est,
        y = y_label,
        color = completeness_bin
      ),
      size = 2.0,
      shape = 16
    ) +
    geom_vline(
      xintercept = ti_q25,
      linetype = "dashed",
      linewidth = 0.4,
      color = "grey45"
    ) +
    geom_vline(
      xintercept = ti_q75,
      linetype = "dashed",
      linewidth = 0.4,
      color = "grey45"
    ) +
    scale_color_manual(
      values = lake_comp_cols,
      drop = FALSE,
      name = "Completeness"
    ) +
    scale_fill_manual(
      values = lake_comp_cols,
      drop = FALSE,
      guide = "none"
    ) +
    labs(
      x = "Trophic index",
      y = NULL
    ) +
    ggtitle("d") +
    theme(
      axis.text.y = element_text(size = 6.8),
      legend.position = "right"
    )

} else {

  p_changed <- ggplot() +
    annotate(
      "text",
      x = 0.5,
      y = 0.5,
      label = "No taxa changed group",
      size = 3
    ) +
    xlim(0, 1) +
    ylim(0, 1) +
    ggtitle("d") +
    theme_void()
}

# ------------------------------------------------------------
# Combine final figure
# ------------------------------------------------------------
final_fig <- top_row / p_changed +
  plot_layout(heights = c(1, 1.15), guides = "collect") &
  theme(legend.position = "right")

# ------------------------------------------------------------
# Save figure
# Bottom panel height scales with number of changed taxa
# ------------------------------------------------------------
fig_height <- if (nrow(changed_dat) > 0) {
  max(6.8, 3.4 + 0.16 * nrow(changed_dat))
} else {
  6.8
}

ggsave(
  filename = "Supplement_TrophicIndex_CompletionAndChangedTaxa_Multipanel.png",
  plot = final_fig,
  width = 10.5,
  height = fig_height,
  dpi = 600,
  bg = "white"
)

# ------------------------------------------------------------
# Optional summary output
# ------------------------------------------------------------
summary_tbl <- plot_dat %>%
  summarise(
    n_taxa = n(),
    n_changed = sum(Changed == "Changed", na.rm = TRUE),
    frac_changed = n_changed / n_taxa,
    median_delta_ti = median(delta_TI, na.rm = TRUE)
  )

print(summary_tbl)

# ------------------------------------------------------------
# Preview figure
# ------------------------------------------------------------
final_fig
```

# fit the lz-bits estimated to 100% complentess on trophic index axis based on estimated genome traits
```{r}
# =============================================================================
# Final summary plot:
# Lake Zurich taxa fitted onto the training-defined trophic index
# - lake taxa: estimated traits, colored by phylum
# - training taxa: larger points, colored by trophic class, labeled
# - horizontal dashed lines: training-based Q1 / Q3 cutoffs
# =============================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(ggrepel)
  library(scales)
})

# -----------------------------------------------------------------------------
# 1) Define consistent trophic class colors for training taxa
# -----------------------------------------------------------------------------
trophic_cols <- c(
  Oligotroph  = "#2166AC",
  Intermediate = "grey60",
  Copiotroph  = "#B2182B"
)

# -----------------------------------------------------------------------------
# 2) Prepare training data and rebuild trophic index from PCA
# -----------------------------------------------------------------------------
train_scored <- train_dat %>%
  mutate(
    Classification = case_when(
      Classification %in% c("Oligotroph", "Oligotrophic") ~ "Oligotroph",
      Classification %in% c("Mesotroph", "Mesotrophic", "Intermediate") ~ "Intermediate",
      Classification %in% c("Copiotroph", "Copiotrophic") ~ "Copiotroph",
      TRUE ~ as.character(Classification)
    )
  ) %>%
  filter(Classification %in% c("Oligotroph", "Intermediate", "Copiotroph")) %>%
  filter(if_all(c(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA), ~ !is.na(.))) %>%
  mutate(
    Classification = factor(
      Classification,
      levels = c("Oligotroph", "Intermediate", "Copiotroph")
    ),
    Label = coalesce(Taxonomy, Species_ID)
  )

# PCA on training traits
pca_fit <- prcomp(
  train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA),
  center = TRUE,
  scale. = TRUE
)

# PC1 scores for training species
train_pc1 <- predict(
  pca_fit,
  newdata = train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA)
)[, 1]

# orient axis so small genomes = lower TI, large genomes = higher TI
if (cor(train_scored$GENOME_LENGTH, train_pc1, use = "complete.obs") < 0) {
  train_pc1 <- -train_pc1
  pca_fit$rotation[, 1] <- -pca_fit$rotation[, 1]
}

train_scored <- train_scored %>%
  mutate(TrophicIndex = train_pc1)

# -----------------------------------------------------------------------------
# 3) Define training-based cutoffs
# -----------------------------------------------------------------------------
ti_q25 <- unname(quantile(train_scored$TrophicIndex, 0.25, na.rm = TRUE))
ti_q75 <- unname(quantile(train_scored$TrophicIndex, 0.75, na.rm = TRUE))

# -----------------------------------------------------------------------------
# 4) Prepare lake data using ESTIMATED trait values only
#    Exclude taxa you said should not be considered for this fitting plot
# -----------------------------------------------------------------------------
lake_scored_est <- lake_dat %>%
  mutate(
    Species_ID = as.character(Species_ID),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", as.character(Phylum))
  ) %>%
  
  # completeness filter
  filter(!is.na(Completeness), Completeness >= 80) %>%
  
  # keep only rows with full estimated trait set
  filter(
    if_all(
      c(Estimated_Genome_Size, CODING_DENSITY, Est_TOTAL_TRNA, Est_NUM_SIGMA),
      ~ !is.na(.)
    )
  )

# project estimated lake traits onto training PCA
lake_pc1_est <- predict(
  pca_fit,
  newdata = lake_scored_est %>%
    transmute(
      GENOME_LENGTH   = Estimated_Genome_Size,
      CODING_DENSITY  = CODING_DENSITY,
      TOTAL_TRNA      = Est_TOTAL_TRNA,
      NUM_SIGMA       = Est_NUM_SIGMA
    )
)[, 1]

lake_scored_est <- lake_scored_est %>%
  mutate(TrophicIndex = lake_pc1_est)

# -----------------------------------------------------------------------------
# 5) Build one shared ranked axis across lake + training taxa
# -----------------------------------------------------------------------------
combined_rank_df <- bind_rows(
  lake_scored_est %>%
    transmute(
      Species_ID,
      Label = Species_ID,
      Phylum,
      Classification = NA_character_,
      TrophicIndex,
      Source = "Lake"
    ),
  train_scored %>%
    transmute(
      Species_ID,
      Label,
      Phylum = NA_character_,
      Classification = as.character(Classification),
      TrophicIndex,
      Source = "Training"
    )
) %>%
  arrange(TrophicIndex, Source) %>%
  mutate(Rank = row_number())

lake_ranked <- combined_rank_df %>% filter(Source == "Lake")
train_ranked <- combined_rank_df %>% filter(Source == "Training")

# -----------------------------------------------------------------------------
# 6) Phylum colors from ColorPalette
# -----------------------------------------------------------------------------
phylum_levels <- sort(unique(lake_ranked$Phylum))

phylum_cols <- setNames(
  colorRampPalette(ColorPalette)(length(phylum_levels)),
  phylum_levels
)

# optional: keep Unclassified at end
if ("Unclassified" %in% names(phylum_cols)) {
  phylum_cols <- c(
    phylum_cols[names(phylum_cols) != "Unclassified"],
    Unclassified = phylum_cols["Unclassified"]
  )
}

# -----------------------------------------------------------------------------
# 7) Final plot
# -----------------------------------------------------------------------------
p_final_trophic <- ggplot() +
  
  # light background bands for operational trophic zones
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf,
    ymin = -Inf, ymax = ti_q25,
    fill = alpha("#2166AC", 0.05), color = NA
  ) +
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf,
    ymin = ti_q25, ymax = ti_q75,
    fill = alpha("grey70", 0.06), color = NA
  ) +
  annotate(
    "rect",
    xmin = -Inf, xmax = Inf,
    ymin = ti_q75, ymax = Inf,
    fill = alpha("#B2182B", 0.05), color = NA
  ) +
  
  # lake taxa by phylum
  geom_point(
    data = lake_ranked,
    aes(x = Rank, y = TrophicIndex, color = Phylum),
    size = 2.0,
    alpha = 0.9
  ) +
  
  # training cutoffs
  geom_hline(
    yintercept = ti_q25,
    linetype = "dashed",
    linewidth = 0.5,
    color = "#2166AC"
  ) +
  geom_hline(
    yintercept = ti_q75,
    linetype = "dashed",
    linewidth = 0.5,
    color = "#B2182B"
  ) +
  
  # training species, larger and on top
  geom_point(
    data = train_ranked,
    aes(x = Rank, y = TrophicIndex, fill = Classification),
    size = 4.2,
    shape = 21,
    color = "black",
    stroke = 0.35
  ) +
  
  # geom_text_repel(
  #   data = train_ranked,
  #   aes(x = Rank, y = TrophicIndex, label = Label, color = Classification),
  #   size = 3.8,
  #   max.overlaps = Inf,
  #   box.padding = 0.3,
  #   point.padding = 0.2,
  #   segment.size = 0.35,
  #   min.segment.length = 0
  # ) +
  
  scale_fill_manual(
    values = trophic_cols,
    name = "Training class"
  ) +
  scale_color_manual(
    values = c(phylum_cols, trophic_cols),
    breaks = names(phylum_cols),
    name = "Lake phylum"
  ) +
  
  labs(
    title = "Lake Zurich taxa fitted onto the training-defined trophic index",
    subtitle = "Lake taxa are projected using estimated genomic traits; labeled larger points indicate training indicator species",
    x = "Rank along combined trophic index (low to high)",
    y = "Trophic index (training PC1 score)"
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    panel.border = element_rect(color = "black", linewidth = 0.5),
    plot.title = element_text(face = "bold", size = 16),
    plot.subtitle = element_text(size = 11),
    axis.title = element_text(size = 13),
    axis.text = element_text(size = 10, color = "black"),
    legend.position = "right",
    legend.title = element_text(size = 10),
    legend.text = element_text(size = 8)
  )

print(p_final_trophic)

# -----------------------------------------------------------------------------
# 8) Save
# -----------------------------------------------------------------------------
ggsave(
  filename = "LakeZurich_TrophicIndex_Estimated_FinalSummary.png",
  plot = p_final_trophic,
  width = 14,
  height = 8,
  units = "in",
  dpi = 600,
  bg = "white"
)
```
# where did all the species end up according to the phylum ? 
```{r}
# =============================================================================
# Stacked barplot of trophic classes across LZ-BITS phyla
# - final plot contains ONLY lake MAGs
# - training data are used only to define PCA axis + thresholds
# - x axis: Phylum
# - y axis: absolute number of MAGs
# - fill: Oligotroph / Intermediate / Copiotroph
# - labels: within-phylum percentages
# =============================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
})

# -----------------------------------------------------------------------------
# 1) Colors
# -----------------------------------------------------------------------------
trophic_cols <- c(
  "Oligotroph"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copiotroph"   = "#BA5205"
)

# -----------------------------------------------------------------------------
# 2) Build trophic index from TRAINING data
# -----------------------------------------------------------------------------
train_scored <- train_dat %>%
  mutate(
    Classification = case_when(
      Classification %in% c("Oligotroph", "Oligotrophic") ~ "Oligotroph",
      Classification %in% c("Mesotroph", "Mesotrophic", "Intermediate") ~ "Intermediate",
      Classification %in% c("Copiotroph", "Copiotrophic") ~ "Copiotroph",
      TRUE ~ as.character(Classification)
    )
  ) %>%
  filter(Classification %in% c("Oligotroph", "Intermediate", "Copiotroph")) %>%
  filter(if_all(c(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA), ~ !is.na(.)))

pca_fit <- prcomp(
  train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA),
  center = TRUE,
  scale. = TRUE
)

train_pc1 <- predict(
  pca_fit,
  newdata = train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA)
)[, 1]

# orient axis once and reuse that orientation everywhere
flip_axis <- cor(train_scored$GENOME_LENGTH, train_pc1, use = "complete.obs") < 0

if (flip_axis) {
  train_pc1 <- -train_pc1
}

train_scored <- train_scored %>%
  mutate(TrophicIndex = train_pc1)

# -----------------------------------------------------------------------------
# 3) Define cutoffs from training distribution
# -----------------------------------------------------------------------------
ti_q25 <- unname(quantile(train_scored$TrophicIndex, 0.25, na.rm = TRUE))
ti_q75 <- unname(quantile(train_scored$TrophicIndex, 0.75, na.rm = TRUE))

classify_trophic_group <- function(x, q25, q75) {
  case_when(
    x <= q25 ~ "Oligotroph",
    x >= q75 ~ "Copiotroph",
    TRUE     ~ "Intermediate"
  )
}

# -----------------------------------------------------------------------------
# 4) Project ESTIMATED LZ-BITS traits onto the training-defined PCA axis
# -----------------------------------------------------------------------------
lake_scored <- lake_dat %>%
  mutate(
    Species_ID = as.character(Species_ID),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", as.character(Phylum))
  ) %>%
  filter(!is.na(Completeness), Completeness >= 80) %>%
  filter(
    if_all(
      c(Estimated_Genome_Size, CODING_DENSITY, Est_TOTAL_TRNA, Est_NUM_SIGMA),
      ~ !is.na(.)
    )
  )

lake_pc1 <- predict(
  pca_fit,
  newdata = lake_scored %>%
    transmute(
      GENOME_LENGTH  = Estimated_Genome_Size,
      CODING_DENSITY = CODING_DENSITY,
      TOTAL_TRNA     = Est_TOTAL_TRNA,
      NUM_SIGMA      = Est_NUM_SIGMA
    )
)[, 1]

# apply the SAME orientation as for the training axis
if (flip_axis) {
  lake_pc1 <- -lake_pc1
}

lake_scored <- lake_scored %>%
  mutate(
    TrophicIndex = lake_pc1,
    TrophicClass = classify_trophic_group(TrophicIndex, ti_q25, ti_q75),
    TrophicClass = factor(
      TrophicClass,
      levels = c("Oligotroph", "Intermediate", "Copiotroph")
    )
  )

# -----------------------------------------------------------------------------
# 5) Summarize ONLY LZ-BITS MAGs by phylum
# -----------------------------------------------------------------------------
plot_df <- lake_scored %>%
  count(Phylum, TrophicClass, name = "n") %>%
  group_by(Phylum) %>%
  mutate(
    phylum_total = sum(n),
    pct = 100 * n / phylum_total,
    pct_label = ifelse(pct >= 6, paste0(round(pct), "%"), NA_character_)
  ) %>%
  ungroup()

# add missing class combinations so stacks are consistent
plot_df <- plot_df %>%
  tidyr::complete(
    Phylum,
    TrophicClass = factor(c("Oligotroph", "Intermediate", "Copiotroph"),
                          levels = c("Oligotroph", "Intermediate", "Copiotroph")),
    fill = list(n = 0)
  ) %>%
  group_by(Phylum) %>%
  mutate(
    phylum_total = max(phylum_total, na.rm = TRUE),
    pct = ifelse(phylum_total > 0, 100 * n / phylum_total, 0),
    pct_label = ifelse(pct >= 6 & n > 0, paste0(round(pct), "%"), NA_character_)
  ) %>%
  ungroup()

# order phyla by total abundance
phylum_order <- plot_df %>%
  distinct(Phylum, phylum_total) %>%
  arrange(desc(phylum_total)) %>%
  pull(Phylum)

plot_df <- plot_df %>%
  mutate(Phylum = factor(Phylum, levels = phylum_order))

# -----------------------------------------------------------------------------
# 6) Plot
# -----------------------------------------------------------------------------

# dataframe for totals per phylum
totals_df <- plot_df %>%
  distinct(Phylum, phylum_total)

p_phylum_stack <- ggplot(plot_df, aes(x = Phylum, y = n, fill = TrophicClass)) +
  
  geom_col(width = 0.8, color = "black", linewidth = 0.25) +
  
  # percentage labels inside bars
  geom_text(
    aes(label = pct_label),
    position = position_stack(vjust = 0.5),
    size = 2.2,
    color = "black",
    na.rm = TRUE
  ) +
  
  # total counts ABOVE bars
  geom_text(
    data = totals_df,
    aes(x = Phylum, y = phylum_total, label = phylum_total),
    inherit.aes = FALSE,
    vjust = -0.4,
    size = 3.2,
    fontface = "bold"
  ) +
  
  scale_fill_manual(
    values = trophic_cols,
    breaks = c("Oligotroph", "Intermediate", "Copiotroph"),
    name = NULL
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.1)),  # extra space for labels on top
    breaks = pretty_breaks(n = 8)
  ) +
  
  labs(
    x = "Phylum",
    y = "LZ-BITS MAGs [n]"
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    panel.border = element_rect(color = "black", linewidth = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    legend.position = "top",
    legend.text = element_text(size = 9)
  )

print(p_phylum_stack)
# -----------------------------------------------------------------------------
# 7) Save
# -----------------------------------------------------------------------------
ggsave(
  filename = "LZBITS_Phylum_TrophicClass_StackedBarplot.png",
  plot = p_phylum_stack,
  width = 10.5,
  height = 6.5,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

```{r}
# =============================================================================
# Stacked barplot of trophic classes across LZ-BITS phyla
# - final plot contains ONLY retained lake MAGs
# - training data are used only to define PCA axis + thresholds
# - x axis: Phylum
# - y axis: absolute number of retained MAGs
# - fill: Oligotroph / Intermediate / Copiotroph
# - labels inside bars: within-phylum percentages among retained MAGs
# - labels above bars: retained / total per phylum, where total comes from species_dat
# =============================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
  library(tidyr)
})

# -----------------------------------------------------------------------------
# 1) Colors
# -----------------------------------------------------------------------------
trophic_cols <- c(
  "Oligotroph"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copiotroph"   = "#BA5205"
)

# -----------------------------------------------------------------------------
# 2) Build trophic index from TRAINING data
# -----------------------------------------------------------------------------
train_scored <- train_dat %>%
  mutate(
    Classification = case_when(
      Classification %in% c("Oligotroph", "Oligotrophic") ~ "Oligotroph",
      Classification %in% c("Mesotroph", "Mesotrophic", "Intermediate") ~ "Intermediate",
      Classification %in% c("Copiotroph", "Copiotrophic") ~ "Copiotroph",
      TRUE ~ as.character(Classification)
    )
  ) %>%
  filter(Classification %in% c("Oligotroph", "Intermediate", "Copiotroph")) %>%
  filter(if_all(c(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA), ~ !is.na(.)))

pca_fit <- prcomp(
  train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA),
  center = TRUE,
  scale. = TRUE
)

train_pc1 <- predict(
  pca_fit,
  newdata = train_scored %>%
    select(GENOME_LENGTH, CODING_DENSITY, TOTAL_TRNA, NUM_SIGMA)
)[, 1]

# orient axis once and reuse that orientation everywhere
flip_axis <- cor(train_scored$GENOME_LENGTH, train_pc1, use = "complete.obs") < 0

if (flip_axis) {
  train_pc1 <- -train_pc1
}

train_scored <- train_scored %>%
  mutate(TrophicIndex = train_pc1)

# -----------------------------------------------------------------------------
# 3) Define cutoffs from training distribution
# -----------------------------------------------------------------------------
ti_q25 <- unname(quantile(train_scored$TrophicIndex, 0.25, na.rm = TRUE))
ti_q75 <- unname(quantile(train_scored$TrophicIndex, 0.75, na.rm = TRUE))

classify_trophic_group <- function(x, q25, q75) {
  case_when(
    x <= q25 ~ "Oligotroph",
    x >= q75 ~ "Copiotroph",
    TRUE     ~ "Intermediate"
  )
}

# -----------------------------------------------------------------------------
# 4) Project ESTIMATED LZ-BITS traits onto the training-defined PCA axis
# -----------------------------------------------------------------------------
lake_scored <- lake_dat %>%
  mutate(
    Species_ID = as.character(Species_ID),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", as.character(Phylum))
  ) %>%
  filter(!is.na(Completeness), Completeness >= 80) %>%
  filter(
    if_all(
      c(Estimated_Genome_Size, CODING_DENSITY, Est_TOTAL_TRNA, Est_NUM_SIGMA),
      ~ !is.na(.)
    )
  )

lake_pc1 <- predict(
  pca_fit,
  newdata = lake_scored %>%
    transmute(
      GENOME_LENGTH  = Estimated_Genome_Size,
      CODING_DENSITY = CODING_DENSITY,
      TOTAL_TRNA     = Est_TOTAL_TRNA,
      NUM_SIGMA      = Est_NUM_SIGMA
    )
)[, 1]

# apply the SAME orientation as for the training axis
if (flip_axis) {
  lake_pc1 <- -lake_pc1
}

lake_scored <- lake_scored %>%
  mutate(
    TrophicIndex = lake_pc1,
    TrophicClass = classify_trophic_group(TrophicIndex, ti_q25, ti_q75),
    TrophicClass = factor(
      TrophicClass,
      levels = c("Oligotroph", "Intermediate", "Copiotroph")
    )
  )

# -----------------------------------------------------------------------------
# 5) Build denominator table from ALL species_dat taxa
#    and numerator table from retained lake_scored taxa
# -----------------------------------------------------------------------------
species_totals <- species_dat %>%
  mutate(
    Species_ID = as.character(Species_ID),
    Phylum = ifelse(is.na(Phylum) | Phylum == "", "Unclassified", as.character(Phylum))
  ) %>%
  distinct(Species_ID, Phylum) %>%
  count(Phylum, name = "total_species")

retained_totals <- lake_scored %>%
  distinct(Species_ID, Phylum) %>%
  count(Phylum, name = "retained_species")

retention_df <- species_totals %>%
  full_join(retained_totals, by = "Phylum") %>%
  mutate(
    total_species = replace_na(total_species, 0L),
    retained_species = replace_na(retained_species, 0L),
    retain_label = paste0(retained_species, "/", total_species)
  )

# -----------------------------------------------------------------------------
# 6) Summarize retained LZ-BITS MAGs by phylum and trophic class
# -----------------------------------------------------------------------------
plot_df <- lake_scored %>%
  count(Phylum, TrophicClass, name = "n") %>%
  left_join(retention_df, by = "Phylum") %>%
  group_by(Phylum) %>%
  mutate(
    phylum_total_retained = sum(n),
    pct = 100 * n / phylum_total_retained,
    pct_label = ifelse(pct >= 6, paste0(round(pct), "%"), NA_character_)
  ) %>%
  ungroup()

# add missing class combinations so stacks are consistent
plot_df <- lake_scored %>%
  count(Phylum, TrophicClass, name = "n") %>%
  complete(
    Phylum,
    TrophicClass = factor(
      c("Oligotroph", "Intermediate", "Copiotroph"),
      levels = c("Oligotroph", "Intermediate", "Copiotroph")
    ),
    fill = list(n = 0)
  ) %>%
  left_join(retention_df, by = "Phylum") %>%
  group_by(Phylum) %>%
  mutate(
    phylum_total_retained = sum(n),
    total_species = max(total_species, na.rm = TRUE),
    retained_species = max(retained_species, na.rm = TRUE),
    retain_label = paste0(retained_species, "/", total_species),
    pct = ifelse(phylum_total_retained > 0, 100 * n / phylum_total_retained, 0),
    pct_label = ifelse(pct >= 6 & n > 0, paste0(round(pct), "%"), NA_character_)
  ) %>%
  ungroup()

# -----------------------------------------------------------------------------
# 7) Order phyla by retained abundance
# -----------------------------------------------------------------------------
phylum_order <- plot_df %>%
  distinct(Phylum, phylum_total_retained) %>%
  arrange(desc(phylum_total_retained)) %>%
  pull(Phylum)

plot_df <- plot_df %>%
  mutate(Phylum = factor(Phylum, levels = phylum_order))

totals_df <- plot_df %>%
  distinct(Phylum, phylum_total_retained, total_species, retained_species, retain_label)

# -----------------------------------------------------------------------------
# 8) Plot
# -----------------------------------------------------------------------------
p_phylum_stack <- ggplot(plot_df, aes(x = Phylum, y = n, fill = TrophicClass)) +
  
  geom_col(width = 0.8, color = "black", linewidth = 0.25) +
  
  # percentage labels inside bars
  geom_text(
    aes(label = pct_label),
    position = position_stack(vjust = 0.5),
    size = 2.2,
    color = "black",
    na.rm = TRUE
  ) +
  
  # retained / total labels above bars
  geom_text(
    data = totals_df,
    aes(x = Phylum, y = phylum_total_retained, label = retain_label),
    inherit.aes = FALSE,
    vjust = -0.4,
    size = 3.0,
    fontface = "bold"
  ) +
  
  scale_fill_manual(
    values = trophic_cols,
    breaks = c("Oligotroph", "Intermediate", "Copiotroph"),
    name = NULL
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.12)),
    breaks = pretty_breaks(n = 8)
  ) +
  
  labs(
    x = "Phylum",
    y = "Retained LZ-BITS MAGs [n]"
  ) +
  
  theme_bw(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(color = "grey90", linewidth = 0.3),
    panel.border = element_rect(color = "black", linewidth = 0.5),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1, color = "black"),
    axis.text.y = element_text(color = "black"),
    axis.title = element_text(color = "black"),
    legend.position = "top",
    legend.text = element_text(size = 9)
  )

print(p_phylum_stack)

# -----------------------------------------------------------------------------
# 9) Save
# -----------------------------------------------------------------------------
ggsave(
  filename = "LZBITS_Phylum_TrophicClass_StackedBarplot.png",
  plot = p_phylum_stack,
  width = 10.5,
  height = 6.5,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

