---
title: "Rhodopsin Expression Analysis"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
# Reproducibility
set.seed(1)

############################################################
# 0) Clear R "brain" + libraries
############################################################
rm(list = ls())
gc()

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(lubridate)
  library(tibble)   # for glimpse()
  library(ggplot2)
})

############################################################
# 1) Set working directory (analysis folder)
############################################################
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_05/")

############################################################
# 2) Load exported analysis tables (from the preparation script)
############################################################

species_meta      <- readr::read_csv("../Figure_03/LakeOverview/Species_Meta_Table.csv", show_col_types = FALSE)
species_abundance <- readr::read_csv("../Figure_03/LakeOverview/Species_Abundance_Table.csv", show_col_types = FALSE)
gene_table        <- readr::read_csv("../Figure_03/LakeOverview/Species_Gene_Table.csv", show_col_types = FALSE)
protein_table     <- readr::read_csv("../Figure_03/LakeOverview/Species_Protein_Table.csv", show_col_types = FALSE)

############################################################
# 3) Quick look
############################################################
glimpse(species_meta)
glimpse(species_abundance)
glimpse(gene_table)
glimpse(protein_table)

############################################################
# 4) Ensure Date columns are Date class (robust to CSV parsing)
############################################################
if ("Date" %in% names(species_abundance)) {
  species_abundance <- species_abundance %>% mutate(Date = as.Date(Date))
}
if ("Date" %in% names(protein_table)) {
  protein_table <- protein_table %>% mutate(Date = as.Date(Date))
}
```


# fetch rhodopsins
```{r}
# Rhodopsin KEGG Orthologs of interest
rhodo_kos <- c("K04643", "K04641")

# ------------------------------------------------------------------------------
# 1) Keep only species that encode at least one rhodopsin KO (GENE TABLE)
#    - This retains ALL rows for those species (not only the rhodopsin rows).
# ------------------------------------------------------------------------------
gene_table_rhodopsin <- gene_table %>%
  group_by(Species_ID) %>%
  filter(any(KEGG_ID %in% rhodo_kos, na.rm = TRUE)) %>%
  ungroup()

# Species universe that is rhodopsin-positive in genes
rhodo_species <- gene_table_rhodopsin %>%
  distinct(Species_ID)

# ------------------------------------------------------------------------------
# 2) Add KEGG info to protein_table and keep only species with rhodopsin KOs
#    - left_join keeps all protein rows and adds KEGG columns where available
#    - then we filter species-level (keep all protein rows for species that have
#      at least one rhodopsin KO mapped somewhere)
# ------------------------------------------------------------------------------
protein_table_rhodopsin <- protein_table %>%
  left_join(
    gene_table %>% select(Protein_ID, KEGG_ID, KEGG_Annotation),
    by = "Protein_ID"
  ) %>%
  group_by(Species_ID) %>%
  filter(any(KEGG_ID %in% rhodo_kos, na.rm = TRUE)) %>%
  ungroup()

# ------------------------------------------------------------------------------
# 3) Filter other tables to the same rhodopsin-positive species set (from genes)
# ------------------------------------------------------------------------------
species_meta_rhodopsin <- species_meta %>%
  filter(Species_ID %in% rhodo_species$Species_ID)

species_abundance_rhodopsin <- species_abundance %>%
  filter(Species_ID %in% rhodo_species$Species_ID)

# ------------------------------------------------------------------------------
# 4) Overview: How many rhodopsin-positive species per TrophicGroup?
#    - "Genes": species that encode rhodopsin KOs (from gene_table_rhodopsin)
#    - "Proteins": species with rhodopsin KOs detected in proteomics
#      (from protein_table_rhodopsin)
#    IMPORTANT: We count unique species per trophic group (not rows).
# ------------------------------------------------------------------------------
gene_counts <- gene_table_rhodopsin %>%
  distinct(Species_ID, Group_For_Plots) %>%
  count(Group_For_Plots, name = "n_species") %>%
  mutate(Source = "Genes (encode rhodopsin)")

protein_counts <- protein_table_rhodopsin %>%
  distinct(Species_ID, Group_For_Plots) %>%   # removes multiple dates / proteins
  count(Group_For_Plots, name = "n_species") %>%
  mutate(Source = "Proteins (detect rhodopsin)")

overview_counts <- bind_rows(gene_counts, protein_counts) %>%
  # optional: make missing trophic group explicit
  mutate(Group_For_Plots = if_else(is.na(Group_For_Plots), "Unknown", Group_For_Plots))

# ------------------------------------------------------------------------------
# 5) Plot (side-by-side bars per trophic group)
# ------------------------------------------------------------------------------
ggplot(overview_counts, aes(x = Group_For_Plots, y = n_species, fill = Source)) +
  geom_col(
    position = position_dodge(width = 0.8),
    width = 0.7
  ) +
  geom_text(
    aes(label = n_species),
    position = position_dodge(width = 0.8),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    title = "Rhodopsin-positive species per trophic group",
    x = "Trophic group",
    y = "Number of species"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_blank()
  )


# ------------------------------------------------------------------------------
# 6) Export rhodopsin-filtered tables to disk
#    (written to the current working directory)
# ------------------------------------------------------------------------------

readr::write_csv(
  gene_table_rhodopsin,
  "Gene_Table_Rhodopsin.csv"
)

readr::write_csv(
  protein_table_rhodopsin,
  "Protein_Table_Rhodopsin.csv"
)

readr::write_csv(
  species_meta_rhodopsin,
  "Species_Meta_Table_Rhodopsin.csv"
)

readr::write_csv(
  species_abundance_rhodopsin,
  "Species_Abundance_Table_Rhodopsin.csv"
)

glimpse(gene_table_rhodopsin)
glimpse(protein_table_rhodopsin)
glimpse(species_meta_rhodopsin)
glimpse(species_abundance_rhodopsin)

```

# how many species have rhodopsins expressed and how abundant are they ? 
```{r}
############################################################
# RHODOPSIN MULTIPANEL FIGURE
# FINAL REFACTORED VERSION
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(patchwork)
})

############################################################
# 1) DEFINE RHODOPSIN SPECIES SETS
############################################################

rhodo_gene_species <- gene_table %>%
  filter(KEGG_ID %in% rhodo_kos) %>%
  distinct(Species_ID) %>%
  mutate(has_rhodo_gene = TRUE)

rhodo_detected_species <- protein_table %>%
  left_join(
    gene_table %>%
      select(Protein_ID, KEGG_ID) %>%
      distinct(),
    by = "Protein_ID"
  ) %>%
  filter(KEGG_ID %in% rhodo_kos) %>%
  distinct(Species_ID) %>%
  mutate(has_rhodo_detected = TRUE)

############################################################
# 2) BUILD MASTER SPECIES TABLE
############################################################

species_status <- species_meta %>%
  select(Species_ID, Group_For_Plots) %>%
  left_join(rhodo_gene_species, by = "Species_ID") %>%
  left_join(rhodo_detected_species, by = "Species_ID") %>%
  mutate(
    has_rhodo_gene     = ifelse(is.na(has_rhodo_gene), FALSE, has_rhodo_gene),
    has_rhodo_detected = ifelse(is.na(has_rhodo_detected), FALSE, has_rhodo_detected),

    genome_class = case_when(
      has_rhodo_gene & Group_For_Plots == "Oligo_like" ~ "Oligo-like",
      has_rhodo_gene                                   ~ "Rest",
      TRUE                                             ~ "Non-rhodopsin"
    ),

    detect_class = case_when(
      has_rhodo_gene & has_rhodo_detected & Group_For_Plots == "Oligo_like" ~ "Oligo-like",
      has_rhodo_gene & has_rhodo_detected                                   ~ "Rest",
      has_rhodo_gene & !has_rhodo_detected                                  ~ "Not detected",
      TRUE                                                                  ~ NA_character_
    ),

    point_class = case_when(
      has_rhodo_gene & has_rhodo_detected & Group_For_Plots == "Oligo_like" ~ "Oligo-like",
      has_rhodo_gene & has_rhodo_detected                                   ~ "Rest",
      has_rhodo_gene & !has_rhodo_detected                                  ~ "Not detected",
      TRUE                                                                  ~ NA_character_
    )
  )

############################################################
# 3) CALCULATE MEAN RELATIVE ABUNDANCE PER SPECIES
############################################################

species_rel_abundance <- species_abundance %>%
  group_by(Date) %>%
  mutate(
    rel_abundance = Species_Abundance / sum(Species_Abundance, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  group_by(Species_ID) %>%
  summarise(
    mean_rel_abundance = mean(rel_abundance, na.rm = TRUE),
    .groups = "drop"
  )

############################################################
# 4) PANEL A DATA
# IMPORTANT:
# ggplot stacks in reverse of factor order for fill,
# so to get Oligo-like on TOP we use:
# levels = c("Oligo-like", "Rest", "Non-rhodopsin")
############################################################

panelA_data <- species_status %>%
  count(genome_class, name = "n") %>%
  complete(
    genome_class = c("Oligo-like", "Rest", "Non-rhodopsin"),
    fill = list(n = 0)
  ) %>%
  mutate(
    genome_class = factor(
      genome_class,
      levels = c("Oligo-like", "Rest", "Non-rhodopsin")
    ),
    prop = n / sum(n)
  )

############################################################
# 5) PANEL B DATA
# To get Not detected on BOTTOM and Oligo-like on TOP:
# levels = c("Oligo-like", "Rest", "Not detected")
############################################################

panelB_data <- species_status %>%
  filter(has_rhodo_gene) %>%
  count(detect_class, name = "n") %>%
  complete(
    detect_class = c("Oligo-like", "Rest", "Not detected"),
    fill = list(n = 0)
  ) %>%
  mutate(
    detect_class = factor(
      detect_class,
      levels = c("Oligo-like", "Rest", "Not detected")
    ),
    prop = n / sum(n)
  )

############################################################
# 6) PANEL C DATA
############################################################

panelC_data <- species_status %>%
  filter(has_rhodo_gene) %>%
  left_join(species_rel_abundance, by = "Species_ID") %>%
  mutate(
    mean_rel_abundance = ifelse(is.na(mean_rel_abundance), 0, mean_rel_abundance),
    detect_binary = ifelse(has_rhodo_detected, "Detected", "Not detected"),
    detect_binary = factor(detect_binary, levels = c("Detected", "Not detected")),
    point_class = case_when(
      has_rhodo_detected & Group_For_Plots == "Oligo_like" ~ "Oligo-like",
      has_rhodo_detected ~ "Rest",
      TRUE ~ "Not detected"
    ),
    point_class = factor(point_class, levels = c("Oligo-like", "Rest", "Not detected"))
  ) %>%
  filter(mean_rel_abundance > 0)

############################################################
# 7) STATISTICS
############################################################

wilcox_res <- wilcox.test(
  mean_rel_abundance ~ detect_binary,
  data = panelC_data,
  exact = FALSE
)

p_val <- wilcox_res$p.value

sig_label <- dplyr::case_when(
  is.na(p_val)    ~ "ns",
  p_val < 1e-4    ~ "****",
  p_val < 1e-3    ~ "***",
  p_val < 1e-2    ~ "**",
  p_val < 5e-2    ~ "*",
  TRUE            ~ "ns"
)

############################################################
# 8) COLORS AND THEME
############################################################

fill_values <- c(
  "Oligo-like"    = "#9ecae1",
  "Rest"          = "#4d4d4d",
  "Non-rhodopsin" = "#bdbdbd",
  "Not detected"  = "#f0ad6b"
)

base_theme <- theme_bw(base_size = 12) +
  theme(
    plot.title      = element_text(face = "bold", size = 13),
    axis.title      = element_text(size = 12),
    axis.text       = element_text(size = 11),
    legend.title    = element_blank(),
    legend.position = "bottom",
    legend.box      = "horizontal",
    legend.text     = element_text(size = 10),
    plot.margin     = margin(6, 8, 6, 8)
  )

############################################################
# 9) PANEL A
############################################################

pA <- ggplot(
  panelA_data,
  aes(x = "Genome prevalence", y = prop, fill = genome_class)
) +
  geom_col(width = 0.72, colour = "black", linewidth = 0.35) +
  geom_text(
    aes(label = ifelse(n > 0, n, "")),
    position = position_stack(vjust = 0.5),
    colour = "white",
    fontface = "bold",
    size = 3.6
  ) +
  scale_fill_manual(
    values = fill_values,
    breaks = c("Oligo-like", "Rest", "Non-rhodopsin")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = NULL,
    y = "Species (%)",
    title = "A. Genome prevalence"
  ) +
  base_theme +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank()
  )

############################################################
# 10) PANEL B
############################################################

pB <- ggplot(
  panelB_data,
  aes(x = "Expression prevalence", y = prop, fill = detect_class)
) +
  geom_col(width = 0.72, colour = "black", linewidth = 0.35) +
  geom_text(
    aes(label = ifelse(n > 0, n, "")),
    position = position_stack(vjust = 0.5),
    colour = "white",
    fontface = "bold",
    size = 3.6
  ) +
  scale_fill_manual(
    values = fill_values,
    breaks = c("Oligo-like", "Rest", "Not detected")
  ) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = NULL,
    y = NULL,
    title = "B. Expression prevalence"
  ) +
  base_theme +
  theme(
    axis.text.x  = element_blank(),
    axis.ticks.x = element_blank(),
    axis.text.y  = element_blank(),
    axis.ticks.y = element_blank(),
    axis.line.y  = element_blank()
  )

############################################################
# 11) PANEL C
############################################################

y_max_C   <- max(panelC_data$mean_rel_abundance, na.rm = TRUE)
y_bracket <- y_max_C * 1.8
y_star    <- y_max_C * 2.15

pC <- ggplot(
  panelC_data,
  aes(x = detect_binary, y = mean_rel_abundance)
) +
  geom_boxplot(
    aes(group = detect_binary),
    width = 0.5,
    outlier.shape = NA,
    fill = "white",
    colour = "black",
    linewidth = 0.45
  ) +
  geom_jitter(
    aes(color = point_class),
    width = 0.23,
    size = 5,
    alpha = 0.75
  ) +
  annotate("segment", x = 1, xend = 2, y = y_bracket, yend = y_bracket, linewidth = 0.4) +
  annotate("segment", x = 1, xend = 1, y = y_bracket / 1.15, yend = y_bracket, linewidth = 0.4) +
  annotate("segment", x = 2, xend = 2, y = y_bracket / 1.15, yend = y_bracket, linewidth = 0.4) +
  annotate(
    "text",
    x = 1.5,
    y = y_star,
    label = sig_label,
    size = 5.5,
    fontface = "bold"
  ) +
   scale_color_manual(
    limits = c("Oligo-like", "Rest", "Not detected"),
    values = c("#9ecae1", "#4d4d4d", "#f0ad6b"),
    drop = FALSE
  ) +
  scale_y_log10(
    labels = function(x) paste0(scales::number(x * 100, accuracy = 0.01), "%")
  ) +
  labs(
    x = NULL,
    y = "Mean relative abundance (%)",
    color = NULL,
    title = "C. Detected carriers are more abundant"
  ) +
  base_theme

############################################################
# 12) COMBINE
############################################################

p_final <- (pA | pB | pC) +
  plot_layout(widths = c(1, 1, 1.35), guides = "collect") &
  theme(
    legend.position = "bottom",
    legend.box = "horizontal"
  )

print(p_final)

ggsave(
  filename = "Rhodopsin_Multipanel_OneRow_Refactored.pdf",
  plot = p_final,
  width = 9,
  height = 9,
  device = pdf,
  bg = "white"
)

############################################################
# 13) OPTIONAL CONSOLE OUTPUT
############################################################

cat("\n========================================\n")
cat("Rhodopsin multipanel figure exported\n")
cat("File: Rhodopsin_Multipanel_OneRow_Refactored.pdf\n")
cat("========================================\n\n")

cat("Panel C Wilcoxon test:\n")
print(wilcox_res)

cat("\nPanel C significance label:\n")
cat(sig_label, "\n")

############################################################
# 14) EXPORT SUMMARY TABLES AND STATS REPORT
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(readr)
})

############################################################
# A) PANEL A SUMMARY
#    Genome prevalence among all species
############################################################

panelA_summary <- panelA_data %>%
  transmute(
    panel = "A_genome_prevalence",
    category = as.character(genome_class),
    n_species = n,
    proportion = prop
  )

############################################################
# B) PANEL B SUMMARY
#    Expression prevalence among rhodopsin-genome species
############################################################

panelB_summary <- panelB_data %>%
  transmute(
    panel = "B_expression_prevalence",
    category = as.character(detect_class),
    n_species = n,
    proportion = prop
  )

############################################################
# C) PANEL C SUMMARY
#    Mean relative abundance statistics by detection group
############################################################

panelC_summary_full <- panelC_data %>%
  group_by(detect_binary) %>%
  summarise(
    n_species = n(),
    min = min(mean_rel_abundance, na.rm = TRUE),
    q1 = quantile(mean_rel_abundance, 0.25, na.rm = TRUE),
    median = median(mean_rel_abundance, na.rm = TRUE),
    mean = mean(mean_rel_abundance, na.rm = TRUE),
    q3 = quantile(mean_rel_abundance, 0.75, na.rm = TRUE),
    max = max(mean_rel_abundance, na.rm = TRUE),
    sd = sd(mean_rel_abundance, na.rm = TRUE),
    iqr = IQR(mean_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    panel = "C_mean_relative_abundance"
  ) %>%
  select(
    panel, detect_binary, n_species,
    min, q1, median, mean, q3, max, sd, iqr
  )

############################################################
# D) PANEL C SUMMARY BY POINT CLASS
#    Optional: Oligo-like / Rest / Not detected
############################################################

panelC_pointclass_summary <- panelC_data %>%
  group_by(point_class) %>%
  summarise(
    n_species = n(),
    min = min(mean_rel_abundance, na.rm = TRUE),
    q1 = quantile(mean_rel_abundance, 0.25, na.rm = TRUE),
    median = median(mean_rel_abundance, na.rm = TRUE),
    mean = mean(mean_rel_abundance, na.rm = TRUE),
    q3 = quantile(mean_rel_abundance, 0.75, na.rm = TRUE),
    max = max(mean_rel_abundance, na.rm = TRUE),
    sd = sd(mean_rel_abundance, na.rm = TRUE),
    iqr = IQR(mean_rel_abundance, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    panel = "C_mean_relative_abundance_point_class"
  ) %>%
  select(
    panel, point_class, n_species,
    min, q1, median, mean, q3, max, sd, iqr
  )

############################################################
# E) COMBINED MANUSCRIPT SUMMARY TABLE
############################################################

summary_table_counts <- bind_rows(
  panelA_summary,
  panelB_summary
)

summary_file_counts <- "Rhodopsin_Figure_Summary_Counts.csv"
summary_file_abund  <- "Rhodopsin_Figure_Summary_Abundance.csv"

write_csv(summary_table_counts, summary_file_counts)
write_csv(bind_rows(panelC_summary_full, panelC_pointclass_summary), summary_file_abund)

############################################################
# F) EFFECT SIZE / FOLD CHANGE HELPERS FOR MANUSCRIPT TEXT
############################################################

detected_stats <- panelC_data %>%
  filter(detect_binary == "Detected") %>%
  summarise(
    n = n(),
    mean = mean(mean_rel_abundance, na.rm = TRUE),
    median = median(mean_rel_abundance, na.rm = TRUE),
    sd = sd(mean_rel_abundance, na.rm = TRUE),
    iqr = IQR(mean_rel_abundance, na.rm = TRUE)
  )

not_detected_stats <- panelC_data %>%
  filter(detect_binary == "Not detected") %>%
  summarise(
    n = n(),
    mean = mean(mean_rel_abundance, na.rm = TRUE),
    median = median(mean_rel_abundance, na.rm = TRUE),
    sd = sd(mean_rel_abundance, na.rm = TRUE),
    iqr = IQR(mean_rel_abundance, na.rm = TRUE)
  )

median_fold_change <- detected_stats$median / not_detected_stats$median
mean_fold_change   <- detected_stats$mean / not_detected_stats$mean

############################################################
# G) STATS REPORT AS TXT
############################################################

stats_file <- "Rhodopsin_Figure_Stats_Report.txt"

sink(stats_file)

cat("============================================================\n")
cat("Rhodopsin multipanel figure: statistical report\n")
cat("============================================================\n\n")

cat("Figure files\n")
cat("------------\n")
cat("Plot PDF: Rhodopsin_Multipanel_OneRow_Refactored.pdf\n")
cat("Counts summary CSV: ", summary_file_counts, "\n", sep = "")
cat("Abundance summary CSV: ", summary_file_abund, "\n", sep = "")
cat("Stats report TXT: ", stats_file, "\n\n", sep = "")

cat("Panel A: Genome prevalence\n")
cat("--------------------------\n")
print(panelA_summary)
cat("\n")

cat("Panel B: Expression prevalence\n")
cat("------------------------------\n")
print(panelB_summary)
cat("\n")

cat("Panel C: Mean relative abundance summary by detection group\n")
cat("-----------------------------------------------------------\n")
print(panelC_summary_full)
cat("\n")

cat("Panel C: Mean relative abundance summary by point class\n")
cat("-------------------------------------------------------\n")
print(panelC_pointclass_summary)
cat("\n")

cat("Wilcoxon rank-sum test\n")
cat("----------------------\n")
print(wilcox_res)
cat("\n")

cat("Significance label used in figure: ", sig_label, "\n\n", sep = "")

cat("Detected vs Not detected manuscript-ready numbers\n")
cat("-------------------------------------------------\n")
cat("Detected species (n): ", detected_stats$n, "\n", sep = "")
cat("Not detected species (n): ", not_detected_stats$n, "\n", sep = "")
cat("Detected mean relative abundance: ", format(detected_stats$mean, scientific = TRUE), "\n", sep = "")
cat("Not detected mean relative abundance: ", format(not_detected_stats$mean, scientific = TRUE), "\n", sep = "")
cat("Detected median relative abundance: ", format(detected_stats$median, scientific = TRUE), "\n", sep = "")
cat("Not detected median relative abundance: ", format(not_detected_stats$median, scientific = TRUE), "\n", sep = "")
cat("Detected SD: ", format(detected_stats$sd, scientific = TRUE), "\n", sep = "")
cat("Not detected SD: ", format(not_detected_stats$sd, scientific = TRUE), "\n", sep = "")
cat("Detected IQR: ", format(detected_stats$iqr, scientific = TRUE), "\n", sep = "")
cat("Not detected IQR: ", format(not_detected_stats$iqr, scientific = TRUE), "\n", sep = "")
cat("Median fold-change (Detected / Not detected): ", round(median_fold_change, 3), "\n", sep = "")
cat("Mean fold-change (Detected / Not detected): ", round(mean_fold_change, 3), "\n\n", sep = "")

cat("Percent-format helpers for manuscript text\n")
cat("------------------------------------------\n")
cat("Detected mean relative abundance (%): ", round(detected_stats$mean * 100, 4), "\n", sep = "")
cat("Not detected mean relative abundance (%): ", round(not_detected_stats$mean * 100, 4), "\n", sep = "")
cat("Detected median relative abundance (%): ", round(detected_stats$median * 100, 4), "\n", sep = "")
cat("Not detected median relative abundance (%): ", round(not_detected_stats$median * 100, 4), "\n", sep = "")

sink()

############################################################
# H) CONSOLE MESSAGE
############################################################

cat("\n========================================\n")
cat("Summary tables and stats report exported\n")
cat("========================================\n")
cat("Counts summary: ", summary_file_counts, "\n", sep = "")
cat("Abundance summary: ", summary_file_abund, "\n", sep = "")
cat("Stats report: ", stats_file, "\n", sep = "")
```


# which proteins are covarrying with rhodopsins ? 
```{r}
############################################################
# Spearman correlations between pooled rhodopsin dynamics
# and non-rhodopsin proteins within species
#
# LOGIC
#   1) Retain species with:
#        - at least 8 sampled dates
#        - at least 4 unique pooled rhodopsin abundance values
#   2) Compute CLR-transformed protein abundances within each
#      Species × Date proteome using a small pseudocount
#   3) Exclude rhodopsin proteins from the tested universe
#   4) Compute Spearman correlations across dates, within species,
#      between pooled rhodopsin abundance and each non-rhodopsin protein
#   5) Test only protein–species pairs with at least 5 paired dates
#   6) Apply BH correction within species
#   7) Retain positive significant hits:
#        rho >= 0.6 and BH-adjusted p <= 0.05
#   8) Summarize functional composition of positive hits
#   9) Export one final figure + one final stats file
#
# REQUIRED OBJECTS
#   - protein_table_rhodopsin
#   - species_meta_rhodopsin
#   - gene_table
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
  library(readr)
  library(grid)
})

############################################################
# PARAMETERS
############################################################
rhodo_kos          <- c("K04641", "K04643")
min_dates_total    <- 8
min_unique_values  <- 4
min_pairs_for_test <- 5
rho_cutoff         <- 0.6
fdr_cutoff         <- 0.05
lump_threshold     <- 0.02
pseudo             <- 1e-8

############################################################
# OUTPUT FILE NAMES
############################################################
suffix_tag <- paste0("rho_ge_", gsub("\\.", "_", as.character(rho_cutoff)))

plot_file <- paste0(
  "Spearman_correlations_with_rhodopsin_dynamics_",
  suffix_tag,
  ".pdf"
)

stats_file <- paste0(
  "Spearman_correlations_with_rhodopsin_dynamics_",
  suffix_tag,
  "_stats.csv"
)

############################################################
# 1) GROUP DEFINITIONS
############################################################
species_groups <- species_meta_rhodopsin %>%
  distinct(Species_ID, Group_For_Plots) %>%
  mutate(
    Group = if_else(Group_For_Plots == "Oligo_like", "Oligo-like", "Rest")
  ) %>%
  select(Species_ID, Group)

############################################################
# 2) KEGG MAPPING AND ANNOTATION TABLES
############################################################
protein_kegg_map <- gene_table %>%
  select(Protein_ID, KEGG_ID) %>%
  distinct()

gene_annot_min <- gene_table %>%
  select(Protein_ID, KEGG_Category, KEGG_Subcategory, Transporter_Family) %>%
  distinct()

############################################################
# 3) BUILD POOLED RHODOPSIN TIME SERIES PER SPECIES × DATE
############################################################
rhodo_ts <- protein_table_rhodopsin %>%
  left_join(protein_kegg_map, by = c("Protein_ID", "KEGG_ID")) %>%
  filter(KEGG_ID %in% rhodo_kos) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    rhodo_iBAQ = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE),
    .groups = "drop"
  )

all_species_dates <- protein_table_rhodopsin %>%
  distinct(Species_ID, Date)

# Fill absent rhodopsin observations as zero
rhodo_ts_full <- all_species_dates %>%
  left_join(rhodo_ts, by = c("Species_ID", "Date")) %>%
  mutate(
    rhodo_iBAQ = ifelse(is.na(rhodo_iBAQ), 0, rhodo_iBAQ)
  )

############################################################
# 4) SPECIES-LEVEL FILTERING
#    Species must have:
#      - >= 8 sampled dates
#      - >= 4 unique rhodopsin abundance values
############################################################
qc_species_filter <- all_species_dates %>%
  group_by(Species_ID) %>%
  summarise(
    n_dates_total = n_distinct(Date),
    .groups = "drop"
  ) %>%
  left_join(
    rhodo_ts_full %>%
      group_by(Species_ID) %>%
      summarise(
        n_unique_rhodo = n_distinct(rhodo_iBAQ),
        n_rhodo_pos_dates = sum(rhodo_iBAQ > 0, na.rm = TRUE),
        .groups = "drop"
      ),
    by = "Species_ID"
  ) %>%
  mutate(
    pass_filter = (n_dates_total >= min_dates_total) &
                  (n_unique_rhodo >= min_unique_values)
  ) %>%
  left_join(species_groups, by = "Species_ID")

filtered_species <- qc_species_filter %>%
  filter(pass_filter) %>%
  pull(Species_ID)

############################################################
# 5) CLR TRANSFORM WITHIN EACH SPECIES × DATE PROTEOME
#    Pseudocount is retained because zeros are present and
#    CLR requires finite log-values.
############################################################
protein_clr_long <- protein_table_rhodopsin %>%
  filter(Species_ID %in% filtered_species) %>%
  group_by(Species_ID, Date) %>%
  mutate(
    logx = log(Protein_Abundance_IBAQ_Norm + pseudo),
    clr  = logx - mean(logx, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(Species_ID, Date, Protein_ID, Protein_Abundance_IBAQ_Norm, clr) %>%
  left_join(species_groups, by = "Species_ID")

clr_means <- protein_table_rhodopsin %>%
  filter(Species_ID %in% filtered_species) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    mean_logx = mean(log(Protein_Abundance_IBAQ_Norm + pseudo), na.rm = TRUE),
    .groups = "drop"
  )

rhodo_clr <- rhodo_ts_full %>%
  filter(Species_ID %in% filtered_species) %>%
  left_join(clr_means, by = c("Species_ID", "Date")) %>%
  mutate(
    clr_rhodo = log(rhodo_iBAQ + pseudo) - mean_logx
  ) %>%
  select(Species_ID, Date, rhodo_iBAQ, clr_rhodo)

############################################################
# 6) EXCLUDE RHODOPSIN PROTEINS FROM CORRELATION TESTING
############################################################
protein_clr_no_rhodo <- protein_clr_long %>%
  left_join(protein_kegg_map, by = "Protein_ID") %>%
  filter(!(KEGG_ID %in% rhodo_kos)) %>%
  select(-KEGG_ID)

############################################################
# 7) ROBUST SPEARMAN FUNCTION
#    Correlations are computed within species across dates.
#    Only protein–species pairs with >= 5 paired dates are tested.
############################################################
safe_spearman <- function(df, min_pairs = 5) {
  df <- df[is.finite(df$clr) & is.finite(df$clr_rhodo), , drop = FALSE]
  n_pairs <- nrow(df)

  if (n_pairs < min_pairs) {
    return(tibble(
      n_pairs = n_pairs,
      rho = NA_real_,
      p_value = NA_real_
    ))
  }

  rho <- suppressWarnings(cor(df$clr, df$clr_rhodo, method = "spearman"))
  p_value <- suppressWarnings(cor.test(df$clr, df$clr_rhodo, method = "spearman")$p.value)

  tibble(
    n_pairs = n_pairs,
    rho = rho,
    p_value = p_value
  )
}

############################################################
# 8) COMPUTE CORRELATIONS AND APPLY BH WITHIN SPECIES
############################################################
corr_results <- protein_clr_no_rhodo %>%
  inner_join(rhodo_clr, by = c("Species_ID", "Date")) %>%
  group_by(Species_ID, Protein_ID, Group) %>%
  group_modify(~ safe_spearman(.x, min_pairs = min_pairs_for_test)) %>%
  ungroup() %>%
  group_by(Species_ID) %>%
  mutate(
    p_adj = ifelse(is.na(p_value), NA_real_, p.adjust(p_value, method = "BH"))
  ) %>%
  ungroup()

############################################################
# 9) QC CHECK: ENSURE NO RHODOPSIN SELF-CORRELATIONS
############################################################
qc_selfcorr_check <- corr_results %>%
  left_join(protein_kegg_map, by = "Protein_ID") %>%
  summarise(
    n_total_rows = n(),
    n_rhodopsin_proteins_in_tests = sum(KEGG_ID %in% rhodo_kos, na.rm = TRUE)
  )

print(qc_selfcorr_check)

############################################################
# 10) DEFINE POSITIVE SIGNIFICANT HITS
############################################################
corr_hits_pos <- corr_results %>%
  filter(
    !is.na(p_adj),
    rho > 0,
    rho >= rho_cutoff,
    p_adj <= fdr_cutoff
  )

############################################################
# 11) MEAN PROTEIN ABUNDANCE PER SPECIES × PROTEIN
#     Used for abundance-weighted functional summaries
############################################################
mean_iBAQ_map <- protein_table_rhodopsin %>%
  filter(Protein_Abundance_IBAQ_Norm > 0) %>%
  group_by(Species_ID, Protein_ID) %>%
  summarise(
    mean_iBAQ = mean(Protein_Abundance_IBAQ_Norm, na.rm = TRUE),
    .groups = "drop"
  )

############################################################
# 12) BUILD BASE TABLE FOR SUMMARIES
############################################################
df_base <- corr_results %>%
  filter(!is.na(rho), is.finite(rho)) %>%
  left_join(mean_iBAQ_map, by = c("Species_ID", "Protein_ID")) %>%
  left_join(gene_annot_min, by = "Protein_ID") %>%
  mutate(
    mean_iBAQ = ifelse(is.na(mean_iBAQ), 0, mean_iBAQ),
    KEGG_Category = ifelse(is.na(KEGG_Category) | KEGG_Category == "", "Unclassified", KEGG_Category),
    KEGG_Subcategory = ifelse(is.na(KEGG_Subcategory) | KEGG_Subcategory == "", "Unclassified", KEGG_Subcategory),
    Transporter_Family = ifelse(is.na(Transporter_Family) | Transporter_Family == "", "Unclassified", Transporter_Family),
    is_transporter = str_detect(KEGG_Subcategory, regex("Transporters", ignore_case = TRUE)),
    positive_sig = !is.na(p_adj) & p_adj <= fdr_cutoff & rho >= rho_cutoff & rho > 0
  ) %>%
  filter(mean_iBAQ > 0)

df_sig <- df_base %>%
  filter(positive_sig)

############################################################
# 13) ASSIGN FUNCTIONAL PLOTTING CATEGORIES
#     Transporters are split by transporter family
#     Non-transporters use KEGG_Category
############################################################
df_sig <- df_sig %>%
  mutate(
    Plot_Category_raw = ifelse(
      is_transporter,
      paste0("Transporter_", Transporter_Family),
      KEGG_Category
    )
  )

############################################################
# 14) SPLIT MULTI-ANNOTATION ENTRIES AND WEIGHT EQUALLY
############################################################
df_long <- df_sig %>%
  rowwise() %>%
  mutate(
    Plot_Category_list = list(str_split(Plot_Category_raw, "\\s*;\\s*|\\s*,\\s*|\\s*\\|\\s*")[[1]]),
    n_cats = length(Plot_Category_list)
  ) %>%
  ungroup() %>%
  unnest_longer(Plot_Category_list, values_to = "Plot_Category") %>%
  mutate(
    Plot_Category = str_replace_all(Plot_Category, "_", " "),
    Plot_Category = ifelse(Plot_Category == "" | is.na(Plot_Category), "Unclassified", Plot_Category),
    w_count = 1 / n_cats,
    w_abund = mean_iBAQ / n_cats
  ) %>%
  filter(Plot_Category != "Unclassified")

############################################################
# 15) LUMP SMALL CATEGORIES INTO "OTHER"
############################################################
category_sizes <- df_long %>%
  group_by(Plot_Category) %>%
  summarise(
    total_abund = sum(w_abund, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    prop_global = total_abund / sum(total_abund, na.rm = TRUE),
    Plot_Category2 = ifelse(prop_global < lump_threshold, "Other", Plot_Category)
  )

df_long <- df_long %>%
  left_join(
    category_sizes %>% select(Plot_Category, Plot_Category2),
    by = "Plot_Category"
  )

############################################################
# 16) SET CATEGORY ORDER
############################################################
category_order <- df_long %>%
  group_by(Plot_Category2) %>%
  summarise(
    total_abund = sum(w_abund, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_abund)) %>%
  pull(Plot_Category2)

df_long <- df_long %>%
  mutate(
    Group = factor(Group, levels = c("Oligo-like", "Rest")),
    Plot_Category2 = factor(Plot_Category2, levels = category_order)
  )

fill_vals <- setNames(
  grDevices::hcl.colors(length(category_order), "Dynamic"),
  category_order
)

############################################################
# 17) GROUP-LEVEL COUNTS FOR REPORTING
############################################################
tested_counts <- df_base %>%
  group_by(Group) %>%
  summarise(
    tested_proteins = n_distinct(paste(Species_ID, Protein_ID, sep = "__")),
    tested_species  = n_distinct(Species_ID),
    .groups = "drop"
  )

sig_counts <- df_sig %>%
  group_by(Group) %>%
  summarise(
    sig_proteins = n_distinct(paste(Species_ID, Protein_ID, sep = "__")),
    sig_species  = n_distinct(Species_ID),
    .groups = "drop"
  )

panel_info <- tested_counts %>%
  left_join(sig_counts, by = "Group") %>%
  mutate(
    sig_proteins = ifelse(is.na(sig_proteins), 0, sig_proteins),
    sig_species  = ifelse(is.na(sig_species), 0, sig_species)
  )

print(panel_info)

############################################################
# 18) COUNT-WEIGHTED FUNCTIONAL COMPOSITION
############################################################
plot_counts <- df_long %>%
  group_by(Group, Plot_Category2) %>%
  summarise(
    value = sum(w_count, na.rm = TRUE),
    n_species = n_distinct(Species_ID),
    n_proteins = n_distinct(paste(Species_ID, Protein_ID, sep = "__")),
    .groups = "drop"
  ) %>%
  group_by(Group) %>%
  mutate(
    prop = value / sum(value, na.rm = TRUE)
  ) %>%
  ungroup()

############################################################
# 19) ABUNDANCE-WEIGHTED FUNCTIONAL COMPOSITION
############################################################
plot_abund <- df_long %>%
  group_by(Group, Plot_Category2) %>%
  summarise(
    value = sum(w_abund, na.rm = TRUE),
    n_species = n_distinct(Species_ID),
    n_proteins = n_distinct(paste(Species_ID, Protein_ID, sep = "__")),
    .groups = "drop"
  ) %>%
  group_by(Group) %>%
  mutate(
    prop = value / sum(value, na.rm = TRUE)
  ) %>%
  ungroup()

############################################################
# 20) COMPLETE MISSING CATEGORIES
############################################################
plot_counts <- plot_counts %>%
  complete(
    Group = factor(c("Oligo-like", "Rest"), levels = c("Oligo-like", "Rest")),
    Plot_Category2 = factor(category_order, levels = category_order),
    fill = list(
      value = 0,
      n_species = 0,
      n_proteins = 0,
      prop = 0
    )
  )

plot_abund <- plot_abund %>%
  complete(
    Group = factor(c("Oligo-like", "Rest"), levels = c("Oligo-like", "Rest")),
    Plot_Category2 = factor(category_order, levels = category_order),
    fill = list(
      value = 0,
      n_species = 0,
      n_proteins = 0,
      prop = 0
    )
  )

############################################################
# 21) BUILD ONE FINAL FIGURE
############################################################
plot_counts2 <- plot_counts %>%
  mutate(Metric = "Proteins")

plot_abund2 <- plot_abund %>%
  mutate(Metric = "Protein abundance")

plot_all <- bind_rows(plot_counts2, plot_abund2) %>%
  mutate(
    Metric = factor(Metric, levels = c("Proteins", "Protein abundance")),
    Group = factor(Group, levels = c("Oligo-like", "Rest")),
    GroupLabel = factor(
      paste(Group, Metric, sep = "___"),
      levels = c(
        "Oligo-like___Proteins",
        "Oligo-like___Protein abundance",
        "Rest___Proteins",
        "Rest___Protein abundance"
      )
    )
  )

x_labels <- c(
  "Oligo-like___Proteins" = "Proteins",
  "Oligo-like___Protein abundance" = "Protein abundance",
  "Rest___Proteins" = "Proteins",
  "Rest___Protein abundance" = "Protein abundance"
)

p_final <- ggplot(
  plot_all,
  aes(x = GroupLabel, y = prop, fill = Plot_Category2)
) +
  geom_col(width = 0.72) +
  scale_fill_manual(values = fill_vals, drop = FALSE) +
  scale_x_discrete(labels = x_labels) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    title = "Spearman correlations with rhodopsin dynamics",
    subtitle = paste0(
      "Species retained if represented at >= ", min_dates_total,
      " dates and showing >= ", min_unique_values,
      " unique pooled rhodopsin abundance values; ",
      "proteins tested with >= ", min_pairs_for_test,
      " paired dates; positive significant hits defined as rho >= ",
      rho_cutoff, " and BH-adjusted p <= ", fdr_cutoff
    ),
    x = NULL,
    y = "Proportion",
    fill = "Category"
  ) +
  facet_grid(
    . ~ Group,
    scales = "free_x",
    space = "free_x",
    switch = "x"
  ) +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    strip.background = element_blank(),
    strip.placement = "outside",
    strip.text.x = element_text(face = "bold", size = 12),
    axis.text.x = element_text(face = "bold"),
    axis.title.x = element_blank(),
    panel.spacing.x = unit(1.2, "lines")
  )

print(p_final)

ggsave(
  filename = plot_file,
  plot = p_final,
  width = 8.2,
  height = 5.8,
  device = pdf,
  bg = "white"
)

############################################################
# 22) DIAGNOSTIC / STATS TABLE
#     Export one compact reviewer-safe CSV
############################################################

# Parameters
stats_parameters <- tibble(
  section = "parameters",
  metric = c(
    "rhodo_kos",
    "min_dates_total",
    "min_unique_values",
    "min_pairs_for_test",
    "rho_cutoff",
    "fdr_cutoff",
    "lump_threshold",
    "pseudo"
  ),
  group = NA_character_,
  category = NA_character_,
  value = c(
    paste(rhodo_kos, collapse = ";"),
    as.character(min_dates_total),
    as.character(min_unique_values),
    as.character(min_pairs_for_test),
    as.character(rho_cutoff),
    as.character(fdr_cutoff),
    as.character(lump_threshold),
    as.character(pseudo)
  )
)

# Species filter summary
stats_species_filter <- qc_species_filter %>%
  group_by(Group, pass_filter) %>%
  summarise(
    n_species = n(),
    .groups = "drop"
  ) %>%
  transmute(
    section = "species_filter",
    metric = ifelse(pass_filter, "passed_species", "failed_species"),
    group = Group,
    category = NA_character_,
    value = as.character(n_species)
  )

# Group counts
stats_group_counts <- panel_info %>%
  pivot_longer(
    cols = c(tested_proteins, tested_species, sig_proteins, sig_species),
    names_to = "metric",
    values_to = "value"
  ) %>%
  transmute(
    section = "group_counts",
    metric = metric,
    group = as.character(Group),
    category = NA_character_,
    value = as.character(value)
  )

# Rho diagnostics at rho >= 0.6
rho_plot_df <- corr_results %>%
  filter(!is.na(rho), !is.na(p_value)) %>%
  mutate(Group = factor(Group, levels = c("Oligo-like", "Rest")))

stats_rho_diag <- rho_plot_df %>%
  group_by(Group) %>%
  summarise(
    n_tested = n(),
    n_rho_ge_cutoff = sum(rho >= rho_cutoff, na.rm = TRUE),
    prop_rho_ge_cutoff = n_rho_ge_cutoff / n_tested,
    median_rho = median(rho, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(n_tested, n_rho_ge_cutoff, prop_rho_ge_cutoff, median_rho),
    names_to = "metric",
    values_to = "value"
  ) %>%
  transmute(
    section = "rho_diagnostics",
    metric = metric,
    group = as.character(Group),
    category = NA_character_,
    value = as.character(value)
  )

# Functional composition by counts
stats_functional_counts <- plot_counts %>%
  transmute(
    section = "functional_composition_detected_proteins",
    metric = "category_prop",
    group = as.character(Group),
    category = as.character(Plot_Category2),
    value = as.character(prop)
  )

# Functional composition by abundance
stats_functional_abund <- plot_abund %>%
  transmute(
    section = "functional_composition_protein_abundance",
    metric = "category_prop",
    group = as.character(Group),
    category = as.character(Plot_Category2),
    value = as.character(prop)
  )

stats_export <- bind_rows(
  stats_parameters,
  stats_species_filter,
  stats_group_counts,
  stats_rho_diag,
  stats_functional_counts,
  stats_functional_abund
)

write_csv(stats_export, stats_file)

############################################################
# 23) CONSOLE OUTPUT
############################################################
cat("\n========================================\n")
cat("Done\n")
cat("========================================\n")
cat("Final plot: ", plot_file, "\n", sep = "")
cat("Final stats: ", stats_file, "\n", sep = "")
cat("Species retained: ", length(filtered_species), "\n", sep = "")
cat("Positive significant hits: ", nrow(corr_hits_pos), "\n", sep = "")
```

# how abundant are rhodopsins vs the rest of the rhodopsin expressing proteomes
```{r}
# =====================================================================
# PROTEIN EXPRESSION SCALING ANALYSIS
# =====================================================================
#
# PURPOSE
# ---------------------------------------------------------------------
# This script quantifies how protein abundance scales with species
# abundance across time using a log–log linear mixed-effects model.
# The analysis compares scaling relationships among protein classes
# (Cytoplasmic, Transporter/Secreted, Rhodopsin) and between ecological
# species groups (Oligo-like vs Rest).
#
# The model evaluates whether different protein classes scale
# differently with species abundance and whether these scaling
# relationships differ between ecological groups.
#
#
# MODEL STRUCTURE
# ---------------------------------------------------------------------
# Response variable:
#   log10(Protein abundance iBAQ)
#
# Predictor:
#   log10(Species abundance)
#
# Fixed effects:
#   logA * ProteinClass * Group2
#
# Random effects:
#   (1 | Species_ID)  – controls for repeated measurements per species
#   (1 | Date)        – controls for shared environmental variation
#
# This log–log formulation corresponds to a power-law scaling:
#
#   Protein abundance = k * (Species abundance)^β
#
# where β is the scaling exponent estimated by the model.
#
#
# DATA FILTERING
# ---------------------------------------------------------------------
# Observations are retained only if:
#
#   - Protein abundance > 0
#   - Species abundance > 0
#   - KEGG annotation is present and valid
#   - Species group classification is available
#
# Protein classes are assigned with the following priority:
#
#   1. Rhodopsin
#   2. Transporter (or Secreted in the first version)
#   3. Cytoplasmic
#
# Proteins are assigned to only one class using this hierarchy.
#
#
# SLOPE ESTIMATION
# ---------------------------------------------------------------------
# Class-specific scaling slopes are estimated using:
#
#   emtrends()
#
# which extracts the derivative:
#
#   d log10(Protein) / d log10(Species)
#
# Pairwise slope differences within each ecological group are tested
# using Tukey-adjusted contrasts.
#
#
# FIGURE TYPES GENERATED
# ---------------------------------------------------------------------
# Two visualization strategies are produced without changing the
# underlying statistical model.
#
# 1) FITTED MODEL VISUALIZATION
#
#    - Shows only the mixed-model fitted relationships
#    - Includes 95% confidence intervals for the fitted mean
#    - Highlights model-estimated scaling relationships
#
# 2) EMPIRICAL SPREAD VISUALIZATION
#
#    - Shows the fitted mixed-model line
#    - Adds ribbons summarizing the observed data distribution
#
#      outer ribbon : 10–90% empirical quantiles
#      inner ribbon : 25–75% empirical quantiles
#      dashed line  : empirical median per abundance bin
#
#    - These ribbons describe the observed spread of the data and
#      are not model confidence intervals.
#
#
# OUTPUT FILES
# ---------------------------------------------------------------------
# Each pipeline produces three files:
#
#   1) Figure PDF
#        - stacked panels for Oligo-like and Rest groups
#
#   2) Statistics file
#        - model summary
#        - ANOVA table
#        - slope estimates
#        - pairwise slope contrasts
#        - QC statistics
#
#   3) Model diagnostics PDF
#        - residual vs fitted
#        - Q–Q plot
#        - residual histogram
#        - residual vs predictor
#
#
# INTERPRETATION
# ---------------------------------------------------------------------
# Slopes represent scaling exponents on the log–log scale.
#
#   β = 1   → protein abundance scales proportionally with species
#   β < 1   → protein increases more slowly than species abundance
#   β > 1   → protein increases faster than species abundance
#
# Differences in slopes among protein classes indicate differences
# in how cellular functions scale with species abundance across
# ecological groups.
#
# =====================================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(readr)
  library(patchwork)
})

# ============================================================
# PARAMETERS
# ============================================================
rhodo_kos <- c("K04641", "K04643")

PSEUDO_A <- 1e-12
PSEUDO_P <- 1e-6

# empirical spread settings
n_bins         <- 24
min_points_bin <- 5

# output prefixes
prefix_fitted <- "Protein_Expression_FittedOnly_Transporter_Cytoplasmic_Rhodopsin"
prefix_spread <- "Protein_Expression_EmpiricalSpread_Transporter_Cytoplasmic_Rhodopsin"

plot_file_fitted    <- paste0(prefix_fitted, ".pdf")
summary_file_fitted <- paste0(prefix_fitted, "_summary.csv")
stats_file_fitted   <- paste0(prefix_fitted, "_stats.txt")

plot_file_spread    <- paste0(prefix_spread, ".pdf")
summary_file_spread <- paste0(prefix_spread, "_summary.csv")
stats_file_spread   <- paste0(prefix_spread, "_stats.txt")

# ============================================================
# HELPER FUNCTIONS
# ============================================================
num_stats <- function(x) {
  tibble(
    min    = min(x, na.rm = TRUE),
    max    = max(x, na.rm = TRUE),
    mean   = mean(x, na.rm = TRUE),
    sd     = sd(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    iqr    = IQR(x, na.rm = TRUE)
  )
}

safe_p_line <- function(pairs_tbl, group_name, labelA, labelB) {
  x1 <- pairs_tbl %>%
    filter(Group2 == group_name, contrast == paste(labelA, "-", labelB))

  x2 <- pairs_tbl %>%
    filter(Group2 == group_name, contrast == paste(labelB, "-", labelA))

  if (nrow(x1) == 1) {
    return(sprintf("%s vs %s: p = %s", labelA, labelB, signif(x1$p.value[1], 3)))
  }
  if (nrow(x2) == 1) {
    return(sprintf("%s vs %s: p = %s", labelA, labelB, signif(x2$p.value[1], 3)))
  }
  sprintf("%s vs %s: p = NA", labelA, labelB)
}

safe_slope_line <- function(slopes_tbl, group_name, class_name) {
  x <- slopes_tbl %>%
    filter(Group2 == group_name, ProteinClass == class_name)

  if (nrow(x) == 1) {
    sprintf("%s slope = %.3f", class_name, x$logA.trend[1])
  } else {
    sprintf("%s slope = NA", class_name)
  }
}

make_prediction_grid <- function(model, dat, n = 200, with_ci = TRUE) {

  grid <- dat %>%
    group_by(Group2, ProteinClass) %>%
    summarise(
      logA_min = quantile(logA, 0.02, na.rm = TRUE),
      logA_max = quantile(logA, 0.98, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    do({
      tibble(
        Group2 = .$Group2,
        ProteinClass = .$ProteinClass,
        logA = seq(.$logA_min, .$logA_max, length.out = n)
      )
    }) %>%
    ungroup() %>%
    mutate(
      Group2 = factor(Group2, levels = levels(dat$Group2)),
      ProteinClass = factor(ProteinClass, levels = levels(dat$ProteinClass))
    )

  X <- model.matrix(~ logA * ProteinClass * Group2, data = grid)
  beta <- fixef(model)

  grid$fit <- as.numeric(X %*% beta)

  if (with_ci) {
    V <- vcov(model)
    grid$se  <- sqrt(diag(X %*% V %*% t(X)))
    grid$lwr <- grid$fit - 1.96 * grid$se
    grid$upr <- grid$fit + 1.96 * grid$se
  }

  grid
}

make_empirical_spread <- function(dat, n_bins = 24, min_points_bin = 5) {

  dat %>%
    group_by(Group2, ProteinClass) %>%
    mutate(
      x_bin = cut(logA, breaks = n_bins, include.lowest = TRUE)
    ) %>%
    group_by(Group2, ProteinClass, x_bin) %>%
    summarise(
      x_mid = mean(logA, na.rm = TRUE),
      x_min = min(logA, na.rm = TRUE),
      x_max = max(logA, na.rm = TRUE),
      y_q10 = quantile(logP, 0.10, na.rm = TRUE),
      y_q25 = quantile(logP, 0.25, na.rm = TRUE),
      y_med = quantile(logP, 0.50, na.rm = TRUE),
      y_q75 = quantile(logP, 0.75, na.rm = TRUE),
      y_q90 = quantile(logP, 0.90, na.rm = TRUE),
      n_bin = n(),
      .groups = "drop"
    ) %>%
    filter(n_bin >= min_points_bin) %>%
    arrange(Group2, ProteinClass, x_mid)
}

write_stats_file <- function(
  out_file,
  plot_file,
  summary_file,
  mode_label,
  qc_counts,
  qc_summary,
  model_summary,
  model_anova,
  slopes_tbl,
  pairs_tbl,
  ann_df,
  summary_csv,
  spread_bin_summary = NULL
) {
  sink(out_file)

  cat("============================================================\n")
  cat("Protein expression mixed model statistics\n")
  cat(mode_label, "\n")
  cat("============================================================\n\n")

  cat("Files written by this script:\n")
  cat(" -", plot_file, "\n")
  cat(" -", summary_file, "\n")
  cat(" -", out_file, "\n\n")

  cat("============================================================\n")
  cat("QC COUNTS\n")
  cat("============================================================\n\n")
  print(qc_counts)
  cat("\n")
  print(qc_summary)
  cat("\n\n")

  if (!is.null(spread_bin_summary)) {
    cat("============================================================\n")
    cat("EMPIRICAL SPREAD BIN SUMMARY\n")
    cat("============================================================\n\n")
    print(spread_bin_summary)
    cat("\n\n")
  }

  cat("============================================================\n")
  cat("MIXED MODEL SUMMARY\n")
  cat("============================================================\n\n")
  print(model_summary)
  cat("\n\n")

  cat("============================================================\n")
  cat("MIXED MODEL ANOVA\n")
  cat("============================================================\n\n")
  print(model_anova)
  cat("\n\n")

  cat("============================================================\n")
  cat("SLOPES (EMTRENDS)\n")
  cat("============================================================\n\n")
  print(slopes_tbl)
  cat("\n\n")

  cat("============================================================\n")
  cat("PAIRWISE SLOPE CONTRASTS WITHIN GROUP (TUKEY)\n")
  cat("============================================================\n\n")
  print(pairs_tbl)
  cat("\n\n")

  cat("============================================================\n")
  cat("ANNOTATION TEXT USED IN FIGURE\n")
  cat("============================================================\n\n")
  print(ann_df)
  cat("\n\n")

  cat("============================================================\n")
  cat("SUMMARY CSV CONTENT\n")
  cat("============================================================\n\n")
  print(summary_csv)
  cat("\n\n")

  sink()
}

# ------------------------------------------------------------
# plotting function
# mode = "fitted_only" or "empirical_spread"
# ------------------------------------------------------------
make_group_plot <- function(
  group_name,
  ann_df,
  grid,
  spread_dat = NULL,
  mode = c("fitted_only", "empirical_spread"),
  show_x_axis = TRUE
) {
  mode <- match.arg(mode)

  x_limits <- c(-1.2, 1.6)
  y_limits <- c(2, 8)

  x_breaks <- seq(-1, 1.5, by = 0.5)
  y_breaks <- seq(2, 8, by = 2)

  x_labels <- sprintf("%.1f", x_breaks)
  x_labels[x_labels == "-0.0"] <- "0.0"

  ann_x <- -1.18
  ann_y <- 7.95

  ann_sub <- ann_df %>%
    filter(Group2 == group_name)

  p <- ggplot()

  if (mode == "fitted_only") {
    p <- p +
      geom_ribbon(
        data = grid %>% filter(Group2 == group_name),
        aes(x = logA, ymin = lwr, ymax = upr, fill = ProteinClass),
        alpha = 0.18
      )
  }

  if (mode == "empirical_spread") {
    p <- p +
      geom_ribbon(
        data = spread_dat %>% filter(Group2 == group_name),
        aes(x = x_mid, ymin = y_q10, ymax = y_q90, fill = ProteinClass),
        alpha = 0.10
      ) +
      geom_ribbon(
        data = spread_dat %>% filter(Group2 == group_name),
        aes(x = x_mid, ymin = y_q25, ymax = y_q75, fill = ProteinClass),
        alpha = 0.22
      ) +
      geom_line(
        data = spread_dat %>% filter(Group2 == group_name),
        aes(x = x_mid, y = y_med, color = ProteinClass),
        linewidth = 0.7,
        linetype = "22"
      )
  }

  p <- p +
    geom_line(
      data = grid %>% filter(Group2 == group_name),
      aes(x = logA, y = fit, color = ProteinClass),
      linewidth = 1.1
    ) +
    geom_text(
      data = ann_sub,
      aes(label = ann),
      x = ann_x,
      y = ann_y,
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 1,
      size = 3.1
    ) +
    scale_color_manual(
      values = c(
        "Cytoplasmic" = "#1f77b4",
        "Transporter" = "#ff7f0e",
        "Rhodopsin"   = "#d62728"
      ),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = c(
        "Cytoplasmic" = "#1f77b4",
        "Transporter" = "#ff7f0e",
        "Rhodopsin"   = "#d62728"
      ),
      drop = FALSE
    ) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels
    ) +
    scale_y_continuous(
      breaks = y_breaks,
      labels = y_breaks
    ) +
    coord_cartesian(
      xlim = x_limits,
      ylim = y_limits,
      expand = FALSE
    ) +
    labs(
      title = group_name,
      x = if (show_x_axis) "log10(Species abundance)" else NULL,
      y = "log10(Protein abundance iBAQ)",
      color = "Protein class",
      fill  = "Protein class"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      panel.grid.major = element_line(color = "grey85", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "grey50", linewidth = 0.4),
      axis.title.x = if (show_x_axis) element_text() else element_blank(),
      axis.text.x  = if (show_x_axis) element_text() else element_blank(),
      axis.ticks.x = if (show_x_axis) element_line(color = "grey50") else element_blank()
    )

  p
}

# ============================================================
# 1) SPECIES GROUPS
# ============================================================
species_groups2 <- species_meta_rhodopsin %>%
  select(Species_ID, Group_For_Plots) %>%
  distinct() %>%
  mutate(Group2 = ifelse(Group_For_Plots == "Oligo_like", "Oligo-like", "Rest")) %>%
  select(Species_ID, Group2)

# ============================================================
# 2) PROTEIN ANNOTATION MAP
# ============================================================
prot_annot <- gene_table_rhodopsin %>%
  select(Protein_ID, KEGG_ID, KEGG_Subcategory, PHOBIUS_Annotation) %>%
  distinct()

# ============================================================
# 3) MAIN ANALYSIS TABLE
#     same logic as your current computation
# ============================================================
dat <- protein_table_rhodopsin %>%
  left_join(prot_annot, by = c("Protein_ID", "KEGG_ID")) %>%
  left_join(
    species_abundance_rhodopsin %>%
      select(Species_ID, Date, Species_Abundance),
    by = c("Species_ID", "Date")
  ) %>%
  left_join(species_groups2, by = "Species_ID") %>%
  filter(
    Protein_Abundance_IBAQ_Norm > 0,
    Species_Abundance > 0,
    !is.na(KEGG_ID),
    KEGG_ID != "",
    KEGG_ID != "Unclassified",
    !is.na(Group2)
  ) %>%
  mutate(
    ProteinClass = case_when(
      KEGG_ID %in% rhodo_kos ~ "Rhodopsin",
      grepl("Transporter", KEGG_Subcategory, ignore.case = TRUE) ~ "Transporter",
      PHOBIUS_Annotation == "CytoplasmaticProtein" ~ "Cytoplasmic",
      TRUE ~ NA_character_
    ),
    ProteinClass = factor(
      ProteinClass,
      levels = c("Cytoplasmic", "Transporter", "Rhodopsin")
    ),
    Group2 = factor(Group2, levels = c("Oligo-like", "Rest")),
    logA = log10(Species_Abundance + PSEUDO_A),
    logP = log10(Protein_Abundance_IBAQ_Norm + PSEUDO_P)
  ) %>%
  filter(!is.na(ProteinClass))

# ============================================================
# 4) QC
# ============================================================
qc_counts <- dat %>%
  count(Group2, ProteinClass)

qc_summary <- tibble(
  metric = c("N_obs", "N_species", "N_proteins", "N_dates"),
  value = c(
    nrow(dat),
    n_distinct(dat$Species_ID),
    n_distinct(dat$Protein_ID),
    n_distinct(dat$Date)
  )
)

cat("\n===== QC =====\n")
print(qc_counts)
print(qc_summary)

# ============================================================
# 5) MIXED MODEL
# ============================================================
m <- lmer(
  logP ~ logA * ProteinClass * Group2 + (1 | Species_ID) + (1 | Date),
  data = dat,
  REML = FALSE
)

model_summary <- summary(m)
model_anova   <- anova(m)

cat("\n===== MIXED MODEL SUMMARY =====\n")
print(model_summary)

cat("\n===== MIXED MODEL ANOVA =====\n")
print(model_anova)

# ============================================================
# 6) SLOPES PER CLASS WITHIN GROUP
# ============================================================
slopes <- emtrends(m, ~ ProteinClass | Group2, var = "logA")
slopes_tbl <- as.data.frame(slopes)

pairs_tbl <- as.data.frame(
  pairs(slopes, by = "Group2", adjust = "tukey")
)

cat("\n===== SLOPES (d logP / d logA) =====\n")
print(slopes_tbl)

cat("\n===== SLOPE COMPARISONS (within Group2; Tukey) =====\n")
print(pairs_tbl)

# ============================================================
# 7) PANEL ANNOTATION TEXT
# ============================================================
ann_df <- tibble(
  Group2 = levels(dat$Group2),
  ann = c(
    paste(
      safe_slope_line(slopes_tbl, "Oligo-like", "Cytoplasmic"),
      safe_slope_line(slopes_tbl, "Oligo-like", "Transporter"),
      safe_slope_line(slopes_tbl, "Oligo-like", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Transporter", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Cytoplasmic", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Cytoplasmic", "Transporter"),
      sep = "\n"
    ),
    paste(
      safe_slope_line(slopes_tbl, "Rest", "Cytoplasmic"),
      safe_slope_line(slopes_tbl, "Rest", "Transporter"),
      safe_slope_line(slopes_tbl, "Rest", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Transporter", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Cytoplasmic", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Cytoplasmic", "Transporter"),
      sep = "\n"
    )
  )
)

# ============================================================
# 8) SHARED GRIDS / SPREAD
# ============================================================
grid_fitted <- make_prediction_grid(m, dat, n = 200, with_ci = TRUE)

spread_dat <- make_empirical_spread(
  dat = dat,
  n_bins = n_bins,
  min_points_bin = min_points_bin
)

spread_bin_summary <- spread_dat %>%
  group_by(Group2, ProteinClass) %>%
  summarise(
    n_bins_retained = n(),
    min_bin_n       = min(n_bin, na.rm = TRUE),
    max_bin_n       = max(n_bin, na.rm = TRUE),
    mean_bin_n      = mean(n_bin, na.rm = TRUE),
    median_bin_n    = median(n_bin, na.rm = TRUE),
    .groups = "drop"
  )

# ============================================================
# 9) SUMMARY CSV
#     keep logic same, export twice under different names
# ============================================================
sum_dat <- protein_table_rhodopsin %>%
  left_join(prot_annot, by = c("Protein_ID", "KEGG_ID")) %>%
  left_join(
    species_abundance_rhodopsin %>%
      select(Species_ID, Date, Species_Abundance),
    by = c("Species_ID", "Date")
  ) %>%
  left_join(species_groups2, by = "Species_ID") %>%
  filter(
    Protein_Abundance_IBAQ_Norm > 0,
    Species_Abundance > 0,
    !is.na(Group2)
  ) %>%
  mutate(
    Category = case_when(
      is.na(KEGG_ID) | KEGG_ID == "" | KEGG_ID == "Unclassified" ~ "NoKEGGID",
      KEGG_ID %in% rhodo_kos ~ "Rhodopsin",
      grepl("Transporter", KEGG_Subcategory, ignore.case = TRUE) ~ "Transporter",
      PHOBIUS_Annotation == "CytoplasmaticProtein" ~ "Cytoplasmic",
      TRUE ~ "Other_Annotated"
    ),
    Category = factor(
      Category,
      levels = c("Rhodopsin", "Transporter", "Cytoplasmic", "NoKEGGID", "Other_Annotated")
    ),
    Group2 = factor(Group2, levels = c("Oligo-like", "Rest"))
  )

sum_dat_main <- sum_dat %>%
  filter(Category %in% c("Rhodopsin", "Transporter", "Cytoplasmic", "NoKEGGID"))

rhodopsin_species_by_group <- sum_dat_main %>%
  filter(Category == "Rhodopsin") %>%
  distinct(Group2, Species_ID) %>%
  count(Group2, name = "n_rhodopsin_containing_species") %>%
  mutate(section = "rhodopsin_species_by_group") %>%
  relocate(section)

protein_dates <- sum_dat_main %>%
  distinct(Category, Protein_ID, Date) %>%
  count(Category, Protein_ID, name = "n_dates_expressed")

protein_level_summary <- protein_dates %>%
  group_by(Category) %>%
  summarise(
    n_expressed_proteins = n_distinct(Protein_ID),
    n_total_species      = sum_dat_main %>% filter(Category == first(Category)) %>% pull(Species_ID) %>% n_distinct(),
    n_total_dates        = sum_dat_main %>% filter(Category == first(Category)) %>% pull(Date) %>% n_distinct(),
    .groups = "drop"
  ) %>%
  left_join(
    protein_dates %>%
      group_by(Category) %>%
      reframe(num_stats(n_dates_expressed)),
    by = "Category"
  ) %>%
  mutate(section = "protein_level_summary") %>%
  relocate(section)

species_category_counts <- sum_dat_main %>%
  group_by(Category, Species_ID) %>%
  summarise(
    n_proteins = n_distinct(Protein_ID),
    n_dates    = n_distinct(Date),
    .groups = "drop"
  )

species_contribution_summary <- species_category_counts %>%
  group_by(Category) %>%
  summarise(
    n_species_contributing = n_distinct(Species_ID),
    .groups = "drop"
  ) %>%
  left_join(
    species_category_counts %>%
      group_by(Category) %>%
      reframe(
        proteins_per_species_min    = min(n_proteins, na.rm = TRUE),
        proteins_per_species_max    = max(n_proteins, na.rm = TRUE),
        proteins_per_species_mean   = mean(n_proteins, na.rm = TRUE),
        proteins_per_species_sd     = sd(n_proteins, na.rm = TRUE),
        proteins_per_species_median = median(n_proteins, na.rm = TRUE),
        proteins_per_species_iqr    = IQR(n_proteins, na.rm = TRUE),
        dates_per_species_min       = min(n_dates, na.rm = TRUE),
        dates_per_species_max       = max(n_dates, na.rm = TRUE),
        dates_per_species_mean      = mean(n_dates, na.rm = TRUE),
        dates_per_species_sd        = sd(n_dates, na.rm = TRUE),
        dates_per_species_median    = median(n_dates, na.rm = TRUE),
        dates_per_species_iqr       = IQR(n_dates, na.rm = TRUE)
      ),
    by = "Category"
  ) %>%
  mutate(section = "species_contribution_summary") %>%
  relocate(section)

species_category_group_counts <- sum_dat_main %>%
  group_by(Group2, Category, Species_ID) %>%
  summarise(
    n_proteins = n_distinct(Protein_ID),
    n_dates    = n_distinct(Date),
    .groups = "drop"
  )

species_contribution_by_group <- species_category_group_counts %>%
  group_by(Group2, Category) %>%
  summarise(
    n_species_contributing = n_distinct(Species_ID),
    proteins_per_species_min    = min(n_proteins, na.rm = TRUE),
    proteins_per_species_max    = max(n_proteins, na.rm = TRUE),
    proteins_per_species_mean   = mean(n_proteins, na.rm = TRUE),
    proteins_per_species_sd     = sd(n_proteins, na.rm = TRUE),
    proteins_per_species_median = median(n_proteins, na.rm = TRUE),
    proteins_per_species_iqr    = IQR(n_proteins, na.rm = TRUE),
    dates_per_species_min       = min(n_dates, na.rm = TRUE),
    dates_per_species_max       = max(n_dates, na.rm = TRUE),
    dates_per_species_mean      = mean(n_dates, na.rm = TRUE),
    dates_per_species_sd        = sd(n_dates, na.rm = TRUE),
    dates_per_species_median    = median(n_dates, na.rm = TRUE),
    dates_per_species_iqr       = IQR(n_dates, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(section = "species_contribution_by_group") %>%
  relocate(section)

summary_csv <- bind_rows(
  rhodopsin_species_by_group %>% mutate(Category = NA_character_),
  protein_level_summary %>% mutate(Group2 = NA_character_),
  species_contribution_summary %>% mutate(Group2 = NA_character_),
  species_contribution_by_group
) %>%
  relocate(section, Group2, Category)

write_csv(summary_csv, summary_file_fitted)
write_csv(summary_csv, summary_file_spread)

# ============================================================
# 10) PLOT 1: FITTED ONLY
# ============================================================
p_oligo_fitted <- make_group_plot(
  group_name = "Oligo-like",
  ann_df = ann_df,
  grid = grid_fitted,
  spread_dat = NULL,
  mode = "fitted_only",
  show_x_axis = FALSE
)

p_rest_fitted <- make_group_plot(
  group_name = "Rest",
  ann_df = ann_df,
  grid = grid_fitted,
  spread_dat = NULL,
  mode = "fitted_only",
  show_x_axis = TRUE
)

p_combined_fitted <- p_oligo_fitted / p_rest_fitted +
  plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(
    legend.position = "top",
    legend.box = "horizontal"
  )

print(p_combined_fitted)

ggsave(
  plot_file_fitted,
  p_combined_fitted,
  width = 8,
  height = 9,
  device = pdf,
  bg = "white"
)

# ============================================================
# 11) PLOT 2: FITTED + EMPIRICAL SPREAD
# ============================================================
p_oligo_spread <- make_group_plot(
  group_name = "Oligo-like",
  ann_df = ann_df,
  grid = grid_fitted,
  spread_dat = spread_dat,
  mode = "empirical_spread",
  show_x_axis = FALSE
)

p_rest_spread <- make_group_plot(
  group_name = "Rest",
  ann_df = ann_df,
  grid = grid_fitted,
  spread_dat = spread_dat,
  mode = "empirical_spread",
  show_x_axis = TRUE
)

p_combined_spread <- p_oligo_spread / p_rest_spread +
  plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(
    legend.position = "top",
    legend.box = "horizontal"
  )

print(p_combined_spread)

ggsave(
  plot_file_spread,
  p_combined_spread,
  width = 8,
  height = 9,
  device = pdf,
  bg = "white"
)

# ============================================================
# 12) STATS FILES
# ============================================================
write_stats_file(
  out_file = stats_file_fitted,
  plot_file = plot_file_fitted,
  summary_file = summary_file_fitted,
  mode_label = "Fitted mean lines with 95% confidence ribbons",
  qc_counts = qc_counts,
  qc_summary = qc_summary,
  model_summary = model_summary,
  model_anova = model_anova,
  slopes_tbl = slopes_tbl,
  pairs_tbl = pairs_tbl,
  ann_df = ann_df,
  summary_csv = summary_csv,
  spread_bin_summary = NULL
)

write_stats_file(
  out_file = stats_file_spread,
  plot_file = plot_file_spread,
  summary_file = summary_file_spread,
  mode_label = "Fitted mean lines with empirical spread ribbons",
  qc_counts = qc_counts,
  qc_summary = qc_summary,
  model_summary = model_summary,
  model_anova = model_anova,
  slopes_tbl = slopes_tbl,
  pairs_tbl = pairs_tbl,
  ann_df = ann_df,
  summary_csv = summary_csv,
  spread_bin_summary = spread_bin_summary
)

# ============================================================
# 13) CONSOLE OUTPUT
# ============================================================
cat("\nDone.\n")
cat("Plot 1   :", plot_file_fitted, "\n")
cat("Summary 1:", summary_file_fitted, "\n")
cat("Stats 1  :", stats_file_fitted, "\n\n")

cat("Plot 2   :", plot_file_spread, "\n")
cat("Summary 2:", summary_file_spread, "\n")
cat("Stats 2  :", stats_file_spread, "\n")
```
## Statistical analysis and interpretation

### Data structure and quality control

- **Purpose**
  - Before fitting the statistical model, we quantified the structure of the dataset to ensure sufficient observations across protein classes and ecological groups.

- **Results**
  - Total observations: **29,160**
  - Number of species: **47**
  - Number of proteins: **1,744**
  - Number of sampling dates: **36**

- **Protein class distribution**

  | Group | Cytoplasmic | Transporter | Rhodopsin |
  |------|-------------|-------------|-----------|
  | Oligo-like | 7,281 | 4,954 | 466 |
  | Rest | 11,727 | 4,276 | 456 |

- **Interpretation**
  - All protein classes contain substantial numbers of observations.
  - Rhodopsins are less frequent but still sufficiently represented for slope estimation.
  - The dataset spans multiple species and sampling dates, requiring a model that accounts for repeated observations.

---

# Mixed-effects scaling model

### Model definition

We fitted a **linear mixed-effects model**:
log10(Protein abundance) ~ log10(Species abundance) × ProteinClass × Group2 + (1 | Species_ID) + (1 | Date)


- **Response variable**
  - `log10(Protein abundance)`

- **Predictor**
  - `log10(Species abundance)`

- **Factors**
  - `ProteinClass`: Cytoplasmic, Transporter, Rhodopsin
  - `Group2`: Oligo-like vs Rest

- **Random effects**
  - Species identity
  - Sampling date

---

# Why a log–log model is appropriate

The model operates on log-transformed variables.

This implies a **power-law relationship** in the original scale:

\[
Protein = k \cdot (Species\,abundance)^{\beta}
\]

- **β = slope of log-log regression**

Biological interpretation:

| β value | Meaning |
|-------|--------|
| β = 1 | protein abundance scales proportionally with species abundance |
| β < 1 | protein increases slower than species abundance |
| β > 1 | protein increases faster than species abundance |

### Why this is statistically appropriate

- Protein abundances arise from **multiplicative biological processes** (transcription, translation, cell abundance).
- Multiplicative processes become **linear after log transformation**.
- Log transformation typically **stabilizes variance** in abundance data.

This modelling approach is widely used in:

- metabolic scaling
- gene–protein scaling
- macroecological abundance relationships

---

# Random effects

The model includes: (1 | Species_ID) (1 | Date)


### Why random effects are necessary

#### Species random effect

- Each species contributes multiple observations.
- Without this term, observations from the same species would not be independent.
- The random intercept accounts for baseline expression differences among species.

#### Date random effect

- Measurements from the same sampling date may share environmental conditions.
- This term captures shared variation across samples collected at the same time.

### Estimated variance components

| Random effect | Variance | Interpretation |
|---------------|---------|---------------|
| Species | 0.029 | modest baseline differences among species |
| Date | 0.002 | very small date effect |
| Residual | 0.577 | most variation occurs at the protein observation level |

---

# Fixed-effect tests (Type III ANOVA)

The ANOVA evaluates whether each predictor contributes significantly to explaining protein abundance.

| Term | Interpretation |
|----|----|
| logA | overall scaling relationship between protein abundance and species abundance |
| ProteinClass | baseline expression differences among classes |
| Group2 | baseline differences between ecological groups |
| logA:ProteinClass | whether scaling slopes differ among protein classes |
| logA:Group2 | whether slopes differ between groups |
| ProteinClass:Group2 | whether class differences vary between groups |
| logA:ProteinClass:Group2 | whether class-specific slopes differ between groups |

### Results

Key significant effects:

- **ProteinClass** (p < 2e-16)
  - Strong baseline differences among protein classes.

- **logA × ProteinClass** (p < 2e-16)
  - Scaling relationships differ among protein classes.

- **ProteinClass × Group2** (p < 2e-16)
  - Class differences vary between ecological groups.

- **logA × ProteinClass × Group2** (p = 4.8e-16)
  - Class-specific scaling relationships differ between groups.

### Interpretation

The three-way interaction indicates that **the scaling of protein abundance with species abundance depends on both protein class and ecological group**.

---

# Estimation of scaling slopes

To directly estimate slopes for each class and group, we used: emtrends()


This extracts the derivative:

\[
\frac{d\log_{10}(Protein)}{d\log_{10}(Species)}
\]

which corresponds to the **scaling exponent**.

---

# Estimated slopes

## Oligo-like group

| Protein class | slope |
|---------------|------|
| Cytoplasmic | -0.019 |
| Transporter | 0.510 |
| Rhodopsin | 0.429 |

Interpretation:

- Cytoplasmic proteins show **no scaling with species abundance**.
- Transporters and rhodopsins **increase with species abundance**.

---

## Rest group

| Protein class | slope |
|---------------|------|
| Cytoplasmic | 0.204 |
| Transporter | 0.431 |
| Rhodopsin | 0.275 |

Interpretation:

- All classes increase with species abundance.
- Transporters show the strongest scaling.

---

# Pairwise slope comparisons

Slope differences were tested using **Tukey-adjusted contrasts**.

### Why this test is used

When multiple slopes are estimated from the same model, pairwise comparisons require adjustment to control the family-wise error rate.

Tukey correction is appropriate for comparing all pairs.

---

## Oligo-like group

| Comparison | Result |
|-----------|--------|
| Cytoplasmic vs Transporter | highly significant difference |
| Cytoplasmic vs Rhodopsin | significant difference |
| Transporter vs Rhodopsin | not significant |

Interpretation:

- Transporters and rhodopsins scale similarly.
- Both scale much more strongly than cytoplasmic proteins.

---

## Rest group

| Comparison | Result |
|-----------|--------|
| Cytoplasmic vs Transporter | significant difference |
| Cytoplasmic vs Rhodopsin | not significant |
| Transporter vs Rhodopsin | not significant |

Interpretation:

- Transporters scale more strongly than cytoplasmic proteins.
- Rhodopsin scaling is intermediate.

---

# Model assumptions

The relationship between:
log10(Protein abundance) and log10(Species abundance)


should be approximately linear.

Because abundance relationships often follow power laws, this assumption is typically reasonable.

---

### 2. Residual normality

Residuals should be approximately normally distributed.

Evaluated using:

- Q–Q plots
- residual histograms

---

### 3. Homoscedasticity

Residual variance should not systematically increase with fitted values.

Evaluated using:

- residual vs fitted plots
- residual vs logA plots

---

### 4. Independence

Observations are assumed independent **conditional on the random effects**.

This is addressed by including species and date random intercepts.

---

# Why the model is appropriate

Several factors support the validity of this approach:

- Large dataset (**29k observations**)
- Multiple species and sampling dates
- Log transformation stabilizes variance
- Mixed-effects structure accounts for repeated observations
- Scaling relationships are biologically meaningful
- Slopes are estimated and compared using established statistical methods

Therefore, fitting a **log–log linear mixed model** provides a robust framework for testing whether protein classes differ in how their abundances scale with species abundance.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(lme4)
  library(lmerTest)
  library(emmeans)
  library(readr)
  library(patchwork)
})

# ============================================================
# PARAMETERS
# ============================================================
rhodo_kos <- c("K04641", "K04643")

PSEUDO_A <- 1e-12
PSEUDO_P <- 1e-6

pdf_file_main <- "Protein_Expression_EmpiricalSpread_Overview_Transporter_Cytoplasmic_Rhodopsin.pdf"
stats_file    <- "Protein_Expression_EmpiricalSpread_Transporter_Cytoplasmic_Rhodopsin_stats.txt"
diag_pdf_file <- "Protein_Expression_EmpiricalSpread_Model_Diagnostics.pdf"

# empirical ribbon settings
n_bins          <- 24   # x-axis bins per Group2 x ProteinClass
min_points_bin  <- 5    # minimum observations needed per bin

# ============================================================
# HELPER FUNCTIONS
# ============================================================

num_stats <- function(x) {
  tibble(
    min    = min(x, na.rm = TRUE),
    max    = max(x, na.rm = TRUE),
    mean   = mean(x, na.rm = TRUE),
    sd     = sd(x, na.rm = TRUE),
    median = median(x, na.rm = TRUE),
    iqr    = IQR(x, na.rm = TRUE)
  )
}

safe_p_line <- function(pairs_tbl, group_name, labelA, labelB) {
  x1 <- pairs_tbl %>%
    filter(Group2 == group_name, contrast == paste(labelA, "-", labelB))

  x2 <- pairs_tbl %>%
    filter(Group2 == group_name, contrast == paste(labelB, "-", labelA))

  if (nrow(x1) == 1) {
    return(sprintf("%s vs %s: p = %s", labelA, labelB, signif(x1$p.value[1], 3)))
  }
  if (nrow(x2) == 1) {
    return(sprintf("%s vs %s: p = %s", labelA, labelB, signif(x2$p.value[1], 3)))
  }
  sprintf("%s vs %s: p = NA", labelA, labelB)
}

safe_slope_line <- function(slopes_tbl, group_name, class_name) {
  x <- slopes_tbl %>%
    filter(Group2 == group_name, ProteinClass == class_name)

  if (nrow(x) == 1) {
    sprintf("%s slope = %.3f", class_name, x$logA.trend[1])
  } else {
    sprintf("%s slope = NA", class_name)
  }
}

# ------------------------------------------------------------
# Fixed-effect prediction grid
# line = fitted mean from the mixed model
# no uncertainty ribbon here; spread is shown empirically
# ------------------------------------------------------------
make_prediction_grid <- function(model, dat, n = 200) {

  grid <- dat %>%
    group_by(Group2, ProteinClass) %>%
    summarise(
      logA_min = quantile(logA, 0.02, na.rm = TRUE),
      logA_max = quantile(logA, 0.98, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    rowwise() %>%
    do({
      tibble(
        Group2 = .$Group2,
        ProteinClass = .$ProteinClass,
        logA = seq(.$logA_min, .$logA_max, length.out = n)
      )
    }) %>%
    ungroup() %>%
    mutate(
      Group2 = factor(Group2, levels = levels(dat$Group2)),
      ProteinClass = factor(ProteinClass, levels = levels(dat$ProteinClass))
    )

  X <- model.matrix(~ logA * ProteinClass * Group2, data = grid)
  beta <- fixef(model)

  grid$fit <- as.numeric(X %*% beta)
  grid
}

# ------------------------------------------------------------
# Empirical spread ribbons
# These summarize observed data spread within x-bins
#
# outer ribbon: 10th to 90th percentile
# inner ribbon: 25th to 75th percentile
# median line : empirical median (optional)
# ------------------------------------------------------------
make_empirical_spread <- function(dat, n_bins = 24, min_points_bin = 5) {

  dat %>%
    group_by(Group2, ProteinClass) %>%
    mutate(
      x_bin = cut(logA, breaks = n_bins, include.lowest = TRUE)
    ) %>%
    group_by(Group2, ProteinClass, x_bin) %>%
    summarise(
      x_mid = mean(logA, na.rm = TRUE),
      x_min = min(logA, na.rm = TRUE),
      x_max = max(logA, na.rm = TRUE),
      y_q10 = quantile(logP, 0.10, na.rm = TRUE),
      y_q25 = quantile(logP, 0.25, na.rm = TRUE),
      y_med = quantile(logP, 0.50, na.rm = TRUE),
      y_q75 = quantile(logP, 0.75, na.rm = TRUE),
      y_q90 = quantile(logP, 0.90, na.rm = TRUE),
      n_bin = n(),
      .groups = "drop"
    ) %>%
    filter(n_bin >= min_points_bin) %>%
    arrange(Group2, ProteinClass, x_mid)
}

# ------------------------------------------------------------
# Diagnostic plots on transformed scale
# ------------------------------------------------------------
write_model_diagnostics <- function(model, dat, out_pdf) {
  res <- resid(model)
  fit <- fitted(model)

  pdf(out_pdf, width = 8, height = 8)
  par(mfrow = c(2, 2), mar = c(4.5, 4.5, 2, 1))

  plot(
    fit, res,
    pch = 16, cex = 0.6,
    xlab = "Fitted values",
    ylab = "Residuals",
    main = "Residuals vs fitted"
  )
  abline(h = 0, lty = 2)

  qqnorm(res, pch = 16, cex = 0.6, main = "Normal Q-Q")
  qqline(res, lty = 2)

  hist(
    res, breaks = 40,
    main = "Residual histogram",
    xlab = "Residuals"
  )

  plot(
    dat$logA, res,
    pch = 16, cex = 0.6,
    xlab = "log10(Species abundance)",
    ylab = "Residuals",
    main = "Residuals vs logA"
  )
  abline(h = 0, lty = 2)

  dev.off()
}

# ============================================================
# 1) SPECIES GROUPS
# ============================================================
species_groups2 <- species_meta_rhodopsin %>%
  select(Species_ID, Group_For_Plots) %>%
  distinct() %>%
  mutate(Group2 = ifelse(Group_For_Plots == "Oligo_like", "Oligo-like", "Rest")) %>%
  select(Species_ID, Group2)

# ============================================================
# 2) PROTEIN ANNOTATION MAP
# ============================================================
prot_annot <- gene_table_rhodopsin %>%
  select(Protein_ID, KEGG_ID, KEGG_Subcategory, PHOBIUS_Annotation) %>%
  distinct()

# ============================================================
# 3) MAIN ANALYSIS TABLE
#
# CLASS PRIORITY:
#   1. Rhodopsin
#   2. Transporter
#   3. Cytoplasmic
# ============================================================
dat <- protein_table_rhodopsin %>%
  left_join(prot_annot, by = c("Protein_ID", "KEGG_ID")) %>%
  left_join(
    species_abundance_rhodopsin %>%
      select(Species_ID, Date, Species_Abundance),
    by = c("Species_ID", "Date")
  ) %>%
  left_join(species_groups2, by = "Species_ID") %>%
  filter(
    Protein_Abundance_IBAQ_Norm > 0,
    Species_Abundance > 0,
    !is.na(KEGG_ID),
    KEGG_ID != "",
    KEGG_ID != "Unclassified",
    !is.na(Group2)
  ) %>%
  mutate(
    ProteinClass = case_when(
      KEGG_ID %in% rhodo_kos ~ "Rhodopsin",
      grepl("Transporter", KEGG_Subcategory, ignore.case = TRUE) ~ "Transporter",
      PHOBIUS_Annotation == "CytoplasmaticProtein" ~ "Cytoplasmic",
      TRUE ~ NA_character_
    ),
    ProteinClass = factor(
      ProteinClass,
      levels = c("Cytoplasmic", "Transporter", "Rhodopsin")
    ),
    Group2 = factor(Group2, levels = c("Oligo-like", "Rest")),
    logA = log10(Species_Abundance + PSEUDO_A),
    logP = log10(Protein_Abundance_IBAQ_Norm + PSEUDO_P)
  ) %>%
  filter(!is.na(ProteinClass))

# ============================================================
# 4) QC
# ============================================================
qc_counts <- dat %>%
  count(Group2, ProteinClass)

qc_summary <- tibble(
  metric = c("N_obs", "N_species", "N_proteins", "N_dates"),
  value = c(
    nrow(dat),
    n_distinct(dat$Species_ID),
    n_distinct(dat$Protein_ID),
    n_distinct(dat$Date)
  )
)

cat("\n===== QC =====\n")
print(qc_counts)
print(qc_summary)

# ============================================================
# 5) MIXED MODEL
# ============================================================
m <- lmer(
  logP ~ logA * ProteinClass * Group2 + (1 | Species_ID) + (1 | Date),
  data = dat,
  REML = FALSE
)

model_summary <- summary(m)
model_anova   <- anova(m)

cat("\n===== MIXED MODEL SUMMARY =====\n")
print(model_summary)

cat("\n===== MIXED MODEL ANOVA =====\n")
print(model_anova)

# ============================================================
# 6) SLOPES PER CLASS WITHIN GROUP
# ============================================================
slopes <- emtrends(m, ~ ProteinClass | Group2, var = "logA")
slopes_tbl <- as.data.frame(slopes)

pairs_tbl <- as.data.frame(
  pairs(slopes, by = "Group2", adjust = "tukey")
)

cat("\n===== SLOPES (d logP / d logA) =====\n")
print(slopes_tbl)

cat("\n===== SLOPE COMPARISONS (within Group2; Tukey) =====\n")
print(pairs_tbl)

# ============================================================
# 7) PANEL ANNOTATION TEXT
# ============================================================
ann_df <- tibble(
  Group2 = levels(dat$Group2),
  ann = c(
    paste(
      safe_slope_line(slopes_tbl, "Oligo-like", "Cytoplasmic"),
      safe_slope_line(slopes_tbl, "Oligo-like", "Transporter"),
      safe_slope_line(slopes_tbl, "Oligo-like", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Transporter", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Cytoplasmic", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Oligo-like", "Cytoplasmic", "Transporter"),
      sep = "\n"
    ),
    paste(
      safe_slope_line(slopes_tbl, "Rest", "Cytoplasmic"),
      safe_slope_line(slopes_tbl, "Rest", "Transporter"),
      safe_slope_line(slopes_tbl, "Rest", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Transporter", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Cytoplasmic", "Rhodopsin"),
      safe_p_line(pairs_tbl, "Rest", "Cytoplasmic", "Transporter"),
      sep = "\n"
    )
  )
)

# ============================================================
# 8) FITTED LINES + EMPIRICAL SPREAD RIBBONS
# ============================================================
grid <- make_prediction_grid(m, dat, n = 200)

spread_dat <- make_empirical_spread(
  dat = dat,
  n_bins = n_bins,
  min_points_bin = min_points_bin
)

# ribbon QC summary
spread_bin_summary <- spread_dat %>%
  group_by(Group2, ProteinClass) %>%
  summarise(
    n_bins_retained = n(),
    min_bin_n       = min(n_bin, na.rm = TRUE),
    max_bin_n       = max(n_bin, na.rm = TRUE),
    mean_bin_n      = mean(n_bin, na.rm = TRUE),
    median_bin_n    = median(n_bin, na.rm = TRUE),
    .groups = "drop"
  )

# ============================================================
# 9) GLOBAL AXIS LIMITS
# ============================================================
x_limits <- c(-1.2, 1.6)
y_limits <- c(2, 8)

x_breaks <- seq(-1, 1.5, by = 0.5)
y_breaks <- seq(2, 8, by = 2)

x_labels <- sprintf("%.1f", x_breaks)
x_labels[x_labels == "-0.0"] <- "0.0"

ann_x <- -1.18
ann_y <- 7.95

# ============================================================
# 10) BASE PLOT FUNCTION
#
# ribbons now show observed spread of the data:
# outer ribbon = 10th-90th percentile
# inner ribbon = 25th-75th percentile
# solid line   = fitted mixed-model mean
# dashed line  = empirical median within bins
# ============================================================
make_group_plot <- function(group_name, show_x_axis = TRUE) {

  ann_sub <- ann_df %>%
    filter(Group2 == group_name)

  ggplot() +
    # outer empirical spread
    geom_ribbon(
      data = spread_dat %>% filter(Group2 == group_name),
      aes(x = x_mid, ymin = y_q10, ymax = y_q90, fill = ProteinClass),
      alpha = 0.10
    ) +

    # inner empirical spread
    geom_ribbon(
      data = spread_dat %>% filter(Group2 == group_name),
      aes(x = x_mid, ymin = y_q25, ymax = y_q75, fill = ProteinClass),
      alpha = 0.22
    ) +

    # empirical median
    geom_line(
      data = spread_dat %>% filter(Group2 == group_name),
      aes(x = x_mid, y = y_med, color = ProteinClass),
      linewidth = 0.7,
      linetype = "22"
    ) +

    # fitted mixed-model line
    geom_line(
      data = grid %>% filter(Group2 == group_name),
      aes(x = logA, y = fit, color = ProteinClass),
      linewidth = 1.1
    ) +

    # annotation
    geom_text(
      data = ann_sub,
      aes(label = ann),
      x = ann_x,
      y = ann_y,
      inherit.aes = FALSE,
      hjust = 0,
      vjust = 1,
      size = 3.1
    ) +

    scale_color_manual(
      values = c(
        "Cytoplasmic" = "#1f77b4",
        "Transporter" = "#ff7f0e",
        "Rhodopsin"   = "#d62728"
      ),
      drop = FALSE
    ) +
    scale_fill_manual(
      values = c(
        "Cytoplasmic" = "#1f77b4",
        "Transporter" = "#ff7f0e",
        "Rhodopsin"   = "#d62728"
      ),
      drop = FALSE
    ) +
    scale_x_continuous(
      breaks = x_breaks,
      labels = x_labels
    ) +
    scale_y_continuous(
      breaks = y_breaks,
      labels = y_breaks
    ) +
    coord_cartesian(
      xlim = x_limits,
      ylim = y_limits,
      expand = FALSE
    ) +
    labs(
      title = group_name,
      x = if (show_x_axis) "log10(Species abundance)" else NULL,
      y = "log10(Protein abundance iBAQ)",
      color = "Protein class",
      fill  = "Protein class"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "top",
      legend.title = element_text(face = "bold"),
      panel.grid.major = element_line(color = "grey85", linewidth = 0.5),
      panel.grid.minor = element_blank(),
      axis.line = element_line(color = "grey50", linewidth = 0.4),
      axis.title.x = if (show_x_axis) element_text() else element_blank(),
      axis.text.x  = if (show_x_axis) element_text() else element_blank(),
      axis.ticks.x = if (show_x_axis) element_line(color = "grey50") else element_blank()
    )
}

# ============================================================
# 11) BUILD THE TWO ALIGNED PLOTS
# ============================================================
p_oligo <- make_group_plot("Oligo-like", show_x_axis = FALSE)
p_rest  <- make_group_plot("Rest", show_x_axis = TRUE)

# ============================================================
# 12) STACK THEM WITH A SHARED LEGEND
# ============================================================
p_combined <- p_oligo / p_rest +
  plot_layout(guides = "collect", heights = c(1, 1)) &
  theme(
    legend.position = "top",
    legend.box = "horizontal"
  )

print(p_combined)

ggsave(
  pdf_file_main,
  p_combined,
  width = 8,
  height = 9,
  device = pdf,
  bg = "white"
)

# ============================================================
# 13) MODEL DIAGNOSTICS
# ============================================================
write_model_diagnostics(m, dat, diag_pdf_file)

# ============================================================
# 14) WRITE STATS FILE
# ============================================================
sink(stats_file)

cat("============================================================\n")
cat("Protein expression mixed model statistics\n")
cat("Rhodopsin vs Transporter vs Cytoplasmic\n")
cat("Empirical spread ribbons + fitted mixed-model lines\n")
cat("============================================================\n\n")

cat("Files written by this script:\n")
cat(" -", pdf_file_main, "\n")
cat(" -", diag_pdf_file, "\n")
cat(" -", stats_file, "\n\n")

cat("============================================================\n")
cat("CLASSIFICATION RULES\n")
cat("============================================================\n\n")
cat("Priority order used for classification:\n")
cat("1. Rhodopsin   : KEGG_ID in rhodo_kos\n")
cat("2. Transporter : KEGG_Subcategory contains 'Transporter'\n")
cat("3. Cytoplasmic : PHOBIUS_Annotation == 'CytoplasmaticProtein'\n")
cat("Proteins are assigned once only, using this priority order.\n\n")

cat("============================================================\n")
cat("PLOT INTERPRETATION\n")
cat("============================================================\n\n")
cat("Solid line  : fitted mixed-model relationship on log10-log10 scale\n")
cat("Dashed line : empirical median log10(Protein abundance) within x-bins\n")
cat("Inner ribbon: empirical 25th-75th percentile of observed log10(Protein abundance)\n")
cat("Outer ribbon: empirical 10th-90th percentile of observed log10(Protein abundance)\n")
cat("Ribbons show observed data spread and are NOT confidence intervals of the fitted line.\n\n")

cat("Empirical ribbon construction parameters:\n")
cat(" - number of x-bins per Group2 x ProteinClass =", n_bins, "\n")
cat(" - minimum observations required per retained bin =", min_points_bin, "\n\n")

cat("============================================================\n")
cat("QC COUNTS\n")
cat("============================================================\n\n")
print(qc_counts)
cat("\n")
print(qc_summary)
cat("\n\n")

cat("============================================================\n")
cat("RIBBON BIN SUMMARY\n")
cat("============================================================\n\n")
print(spread_bin_summary)
cat("\n\n")

cat("============================================================\n")
cat("MIXED MODEL SUMMARY\n")
cat("============================================================\n\n")
print(model_summary)
cat("\n\n")

cat("============================================================\n")
cat("MIXED MODEL ANOVA\n")
cat("============================================================\n\n")
print(model_anova)
cat("\n\n")

cat("============================================================\n")
cat("SLOPES (EMTRENDS)\n")
cat("============================================================\n\n")
print(slopes_tbl)
cat("\n\n")

cat("============================================================\n")
cat("PAIRWISE SLOPE CONTRASTS WITHIN GROUP (TUKEY)\n")
cat("============================================================\n\n")
print(pairs_tbl)
cat("\n\n")

cat("============================================================\n")
cat("ANNOTATION TEXT USED IN FIGURE\n")
cat("============================================================\n\n")
print(ann_df)
cat("\n\n")

cat("============================================================\n")
cat("MODEL INTERPRETATION NOTES\n")
cat("============================================================\n\n")
cat(
  paste(
    "The mixed model was fitted on log10-transformed abundance variables.",
    "This implies a linear relationship on the transformed scale and a power-law relationship on the original scale.",
    "Slope differences were tested using emtrends(), which estimates class-specific trends of logP versus logA within each Group2 level.",
    "These slopes are scaling exponents on the log-log scale.",
    "The plotted ribbons do not come from model uncertainty; they summarize the empirical distribution of observed data within local x-bins.",
    sep = "\n"
  )
)
cat("\n\n")

cat("============================================================\n")
cat("MODEL ASSUMPTION CHECKLIST\n")
cat("============================================================\n\n")
cat(
  paste(
    "Inspect the diagnostics PDF and verify:",
    "1. Residuals vs fitted: no strong curvature or funnel pattern.",
    "2. Q-Q plot: no severe deviation from normality in residuals.",
    "3. Residuals vs logA: no remaining systematic trend.",
    "4. No single class/group combination with extreme leverage due to very small n.",
    "",
    "If these checks look acceptable, the log-log linear mixed model and slope contrasts are defensible.",
    "The empirical ribbons are descriptive visual summaries of observed spread and do not rely on model-distribution assumptions.",
    "",
    "If assumptions are poor, possible next steps include:",
    "- class-specific variance structures (e.g. nlme::lme with varIdent),",
    "- robust mixed modelling,",
    "- or restricting inference to the abundance range shared by all classes.",
    sep = "\n"
  )
)
cat("\n\n")

sink()

cat("\nDone.\n")
cat("Main plot    :", pdf_file_main, "\n")
cat("Diagnostics  :", diag_pdf_file, "\n")
cat("Stats        :", stats_file, "\n")
```

# Output fasta files with rhodopsins for alignment
```{r}
############################################################
# Export expressed rhodopsins as FASTA
# Header format:
# Species_ID_Group2_Phylum_Class_Order_Family_Genus_Species
#
# Assumes:
# - protein_table_rhodopsin contains expressed proteins
# - gene_table contains KEGG_ID and protein sequence
# - species_meta contains taxonomy
# - protein sequence column is named Protein_Sequence
#   (change below if your column has a different name)
############################################################

library(dplyr)

# Rhodopsin KO IDs
rhodo_kos <- c("K04643", "K04641")

############################################################
# 1) Identify expressed rhodopsins
############################################################

rhodo_expressed <- protein_table_rhodopsin %>%
  filter(KEGG_ID %in% rhodo_kos) %>%
  distinct(Species_ID, Protein_ID)

############################################################
# 2) Add sequence, taxonomy, and Oligo-like vs Rest category
############################################################

rhodo_fasta_df <- rhodo_expressed %>%
  left_join(
    gene_table %>% select(Protein_ID, Protein_Sequence),
    by = "Protein_ID"
  ) %>%
  left_join(
    species_meta %>%
      select(Species_ID, Group_For_Plots, Phylum, Class, Order, Family, Genus, Species),
    by = "Species_ID"
  ) %>%
  mutate(
    Group2 = ifelse(Group_For_Plots == "Oligo_like", "Oligo-like", "Rest"),
    fasta_header = paste(
      Species_ID,
      Group2,
      Phylum,
      Class,
      Order,
      Family,
      Genus,
      Species,
      sep = "_"
    )
  ) %>%
  filter(!is.na(Protein_Sequence))

############################################################
# 3) Write FASTA
############################################################

fasta_file <- "Expressed_Rhodopsins_Group2.fasta"
con <- file(fasta_file, "w")

for (i in seq_len(nrow(rhodo_fasta_df))) {
  writeLines(paste0(">", rhodo_fasta_df$fasta_header[i]), con)
  writeLines(rhodo_fasta_df$Protein_Sequence[i], con)
}

close(con)

cat("FASTA exported:", fasta_file, "\n")
```

