---
title: Protein Database Description
subtitle: SpringBloom2021 - Genome-resolved Metaproteomic Dataset
date: 09.04.2025
author: Cyrill Hofer
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    code_folding: hide
    thumbnails: true
    gallery: false
    fig_width: 15
    df_print: kable
    fig_caption: yes
    use_bookdown: true
    css: custom.css
---

# Data Preparation
- Note: This data does not contain and protein identifications from contaminants
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Clear R's Brain
rm(list = ls())

# Set the working directory
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_02/PeptidesProteinsRecovered/")

# Load the libraries
library(readr)
library(dplyr)
library(ggplot2)
library(tidyr)
library(scales)
library(forcats)
library(patchwork)

# Load the color palette (optional, keep if you want)
source("ColorPalette.R")

# ----------------------------------------------------------------------
# Load the NEW data (this replaces PeptideProteinDistribution.csv)
# ----------------------------------------------------------------------
dat <- read_csv("PeptideProteinDistribution_withResolution.csv", show_col_types = FALSE)
```

# publication
```{r}
# ----------------------------------------------------------------------
# 1) Recode & order Stage (same as your script)
# ----------------------------------------------------------------------
dat <- dat %>%
  mutate(
    Stage = recode(Stage,
      Raw      = "Unfiltered",
      `Q<1%`   = "1% Peptide FDR",
      `All<1%` = "1% Protein FDR",
      `2+ pep` = ">2 Peptides Per Protein"
    ),
    Stage = factor(Stage, levels = c(
      "Unfiltered",
      "1% Peptide FDR",
      "1% Protein FDR",
      ">2 Peptides Per Protein"
    ))
  ) %>%
  filter(Stage != "Unfiltered")

# ----------------------------------------------------------------------
# 2) Panel A+B (unchanged): Proteotypic vs Non-proteotypic
# ----------------------------------------------------------------------
bar_data_proteotypic <- bind_rows(
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Proteotypic",     Count = ProteotypicPeptides),
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Non-proteotypic", Count = Peptides - ProteotypicPeptides),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Proteotypic",     Count = ProteotypicProteins),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Non-proteotypic", Count = Proteins - ProteotypicProteins)
)

# ----------------------------------------------------------------------
# 3) NEW Panel C+D: Species-specific vs Not species-specific
#    (Requires columns from new file)
#    - SpeciesSpecificPeptides
#    - SpeciesSpecificProteinGroups  (or SpeciesSpecificProteins depending on your export)
# ----------------------------------------------------------------------

# Some earlier versions of the script used "SpeciesSpecificProteins".
# This makes your plotting robust to either naming.
if (!"SpeciesSpecificProteinGroups" %in% names(dat) && "SpeciesSpecificProteins" %in% names(dat)) {
  dat <- dat %>% rename(SpeciesSpecificProteinGroups = SpeciesSpecificProteins)
}

bar_data_specificity <- bind_rows(
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Species-specific",     Count = SpeciesSpecificPeptides),
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Not species-specific", Count = Peptides - SpeciesSpecificPeptides),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Species-specific",     Count = SpeciesSpecificProteinGroups),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Not species-specific", Count = Proteins - SpeciesSpecificProteinGroups)
)

# ----------------------------------------------------------------------
# 4) Colors
# ----------------------------------------------------------------------
fill_cols_proteotypic <- c(
  "Proteotypic"     = "#1f77b4",
  "Non-proteotypic" = "#ff7f0e"
)

fill_cols_specificity <- c(
  "Species-specific"      = "#2ca02c",
  "Not species-specific"  = "#7f7f7f"
)

# ----------------------------------------------------------------------
# 5) Helper to build each panel
# ----------------------------------------------------------------------
make_panel <- function(df, metric, breaks, limits, fill_cols) {
  ggplot(filter(df, Metric == metric), aes(Stage, Count, fill = Category)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(breaks = breaks, limits = limits, labels = comma) +
    labs(title = NULL, x = NULL, y = "Count") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_blank(),
      legend.title = element_blank()
    ) +
    scale_fill_manual(values = fill_cols)
}

# ----------------------------------------------------------------------
# 6) Make panels (adjust limits if your counts differ)
# ----------------------------------------------------------------------
# Proteotypic panels (original)
p_pep_prot <- make_panel(bar_data_proteotypic, "Peptides", seq(0, 90000, 10000), c(0, 95000), fill_cols_proteotypic)
p_prt_prot <- make_panel(bar_data_proteotypic, "Proteins", seq(0, 40000,  5000), c(0, 40000), fill_cols_proteotypic)

# Specificity panels (new)
p_pep_spec <- make_panel(bar_data_specificity, "Peptides", seq(0, 90000, 10000), c(0, 95000), fill_cols_specificity)
p_prt_spec <- make_panel(bar_data_specificity, "Proteins", seq(0, 40000,  5000), c(0, 40000), fill_cols_specificity)

# ----------------------------------------------------------------------
# 7) Combine into one figure (4 panels, 2 legends)
# ----------------------------------------------------------------------
p <- (p_pep_prot + p_prt_prot) / (p_pep_spec + p_prt_spec) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

print(p)
```
```{r}
# 1) Recode & order Stage (same as before)
dat <- dat %>%
  mutate(
    Stage = recode(Stage,
      Raw      = "Unfiltered",
      `Q<1%`   = "1% Peptide FDR",
      `All<1%` = "1% Protein FDR",
      `2+ pep` = ">2 Peptides Per Protein"
    ),
    Stage = factor(Stage, levels = c(
      "Unfiltered",
      "1% Peptide FDR",
      "1% Protein FDR",
      ">2 Peptides Per Protein"
    ))
  ) %>%
  filter(Stage != "Unfiltered")

# 2) Long-form table: species-specific vs not
bar_data <- bind_rows(
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Species-specific",     Count = SpeciesSpecificPeptides),
  dat %>% transmute(Stage, Metric = "Peptides", Category = "Not species-specific", Count = Peptides - SpeciesSpecificPeptides),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Species-specific",     Count = SpeciesSpecificProteinGroups),
  dat %>% transmute(Stage, Metric = "Proteins", Category = "Not species-specific", Count = Proteins - SpeciesSpecificProteinGroups)
)

# 3) Colors
fill_cols <- c(
  "Species-specific"     = "#2ca02c",
  "Not species-specific" = "#7f7f7f"
)

# 4) Helper to build each panel
make_panel <- function(metric, breaks, limits) {
  df <- filter(bar_data, Metric == metric)

  ggplot(df, aes(Stage, Count, fill = Category)) +
    geom_col(color = "black", linewidth = 0.2) +
    scale_y_continuous(breaks = breaks, limits = limits, labels = comma) +
    labs(title = NULL, x = NULL, y = "Count") +
    scale_fill_manual(values = fill_cols) +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x  = element_text(angle = 45, hjust = 1, face = "bold"),
      axis.title.y = element_blank(),
      legend.title = element_blank()
    )
}

# 5) Create panels
p_pep <- make_panel("Peptides", seq(0, 90000, 10000), c(0, 95000))
p_prt <- make_panel("Proteins", seq(0, 40000,  5000), c(0, 40000))

# 6) Combine and save
p <- (p_pep + p_prt) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

print(p)

ggsave(
  filename = "SpeciesSpecificity_Peptides_Proteins.pdf",
  plot     = p,
  width    = 12,
  height   = 20,
  units    = "cm",
  device   = pdf,
  dpi      = 300
)

```

## Species-specificity assignment for proteins and peptides

### Protein group specificity

Species specificity of protein identifications was assessed at the level of DIA-NN protein groups using a genome-resolved taxonomy reference.

- Protein groups reported by DIA-NN consist of one or multiple protein identifiers; multi-protein groups are represented as semicolon-separated lists.
- Each protein identifier encodes a species identifier as the prefix preceding the first hyphen.
- Species identifiers were mapped to a curated taxonomy table containing phylum, class, order, family, genus, and species assignments.
- For each protein group, all member proteins were compared across taxonomic ranks to determine the highest possible taxonomic resolution shared by **all** group members.
- Protein groups were assigned a single taxonomic resolution according to the following hierarchical logic:
  - **Species-resolved**: all group members belong to the same species.
  - **Genus-resolved**: group members differ at species level but share the same genus.
  - **Family-resolved**, **order-resolved**, **class-resolved**, or **phylum-resolved**: defined analogously.
  - **No match**: assigned if any group member lacked a valid taxonomy annotation or if no common taxonomic rank was shared.
- Protein groups classified as species-resolved were defined as **species-specific protein groups**.
- This approach provides a conservative estimate of taxonomic specificity by requiring complete agreement among all proteins within a group.

---

### Peptide specificity

Peptide-level specificity was derived directly from protein-group specificity assignments.

- Each unique peptide sequence (stripped sequence, without modifications) was associated with one or more protein groups as reported by DIA-NN.
- For each peptide, the taxonomic resolutions of all associated protein groups were collected.
- Peptide specificity was assigned conservatively as the **least specific** (highest-level) taxonomic resolution among its associated protein groups.
  - For example, a peptide mapping to both species-resolved and genus-resolved protein groups was classified as genus-resolved.
- Peptides were defined as **species-specific** only if all associated protein groups were species-resolved.
- This conservative propagation ensures that peptides are considered species-specific only when unambiguous taxonomic attribution is supported across all protein-group mappings.

```{r}
# ----------------------------------------------------------------------
# Summary table: species-specific vs non species-specific
# ----------------------------------------------------------------------

summary_table <- bind_rows(
  dat %>%
    transmute(
      Stage,
      Metric = "Peptides",
      Total = Peptides,
      Species_Specific = SpeciesSpecificPeptides
    ),
  dat %>%
    transmute(
      Stage,
      Metric = "Proteins",
      Total = Proteins,
      Species_Specific = SpeciesSpecificProteinGroups
    )
) %>%
  mutate(
    Not_Species_Specific = Total - Species_Specific,
    Fraction_Species_Specific = Species_Specific / Total
  ) %>%
  arrange(Metric, Stage)
summary_table
```

