---
title: Proteome Recovery Vs Species Abundance
subtitle: SpringBloom2021 - Genome-resolved Metaproteomic Dataset
date: 09.04.2025
author: Cyrill Hofer
output:
  rmdformats::readthedown:
    highlight: tango
    self_contained: true
    code_folding: hide
    thumbnails: true
    gallery: false
    fig_width: 15
    df_print: kable
    fig_caption: yes
    use_bookdown: true
    css: custom.css
---

# Data Preparation
```{r message=FALSE, warning=FALSE, paged.print=FALSE}
# Clear R's Brain
rm(list = ls())

# Set the working directory
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_02/ProteomeRecoveryVsSpeciesAbundance/")

# Load libraries
library(dplyr)
library(ggplot2)
library(ggeffects)
library(ggnewscale)
library(readr)
library(tibble)
library(car)
library(readr)


# Load the color palette
source("ColorPalette.R")

# Load the data
dat <- read_csv(file = "species_summary_peak_only.csv")
glimpse(dat)
```

```{r}
# -----------------------------
# 2) Filter to valid values for log-transform and correlations
#    (log10 requires >0)
# -----------------------------
plot_dat <- dat %>%
  filter(
    !is.na(Species_Abundance_at_Peak),
    !is.na(Proteome_Abundance_on_PeakDate),
    !is.na(Detected_Proteins_at_Peak),
    Species_Abundance_at_Peak > 0,
    Proteome_Abundance_on_PeakDate > 0,
    Detected_Proteins_at_Peak > 0
  ) %>%
  mutate(
    log10_species_abund  = log10(Species_Abundance_at_Peak),
    log10_proteome_abund = log10(Proteome_Abundance_on_PeakDate),
    log10_detected_prot  = log10(Detected_Proteins_at_Peak)
  )

cat("N species used after filtering:", nrow(plot_dat), "\n")

# -----------------------------
# 3) Normality assessment
#    IMPORTANT:
#    With large N, Shapiro-Wilk becomes overly sensitive.
#    We therefore use:
#      - Shapiro-Wilk (formal test; interpret cautiously)
#      - QQ plots (visual)
#    We test both raw and log10 versions.
# -----------------------------

# Helper: run Shapiro only if N is within test limits (R typically allows up to 5000)
run_shapiro <- function(x, label) {
  x <- x[is.finite(x)]
  if (length(x) < 3) {
    cat(label, ": too few values for Shapiro\n")
    return(invisible(NULL))
  }
  if (length(x) > 5000) {
    cat(label, ": N > 5000 (skipping Shapiro)\n")
    return(invisible(NULL))
  }
  out <- shapiro.test(x)
  cat(label, "Shapiro-Wilk: W =", round(out$statistic, 4), ", p =", signif(out$p.value, 3), "\n")
  invisible(out)
}

cat("\n--- Normality checks (Shapiro-Wilk; interpret with caution for large N) ---\n")
run_shapiro(plot_dat$Species_Abundance_at_Peak,      "Raw Species_Abundance_at_Peak")
run_shapiro(plot_dat$Proteome_Abundance_on_PeakDate, "Raw Proteome_Abundance_on_PeakDate")
run_shapiro(plot_dat$Detected_Proteins_at_Peak,      "Raw Detected_Proteins_at_Peak")

run_shapiro(plot_dat$log10_species_abund,            "Log10 Species abundance")
run_shapiro(plot_dat$log10_proteome_abund,           "Log10 Proteome abundance")
run_shapiro(plot_dat$log10_detected_prot,            "Log10 Detected proteins")

# QQ plots for visual assessment (log10 versions are usually most relevant here)
qq_plot <- function(x, title) {
  x <- x[is.finite(x)]
  ggplot(data.frame(x = x), aes(sample = x)) +
    stat_qq() +
    stat_qq_line() +
    theme_bw(base_size = 12) +
    labs(title = title, x = "Theoretical quantiles", y = "Sample quantiles")
}

print(qq_plot(plot_dat$log10_species_abund,  "QQ plot: log10 Species abundance"))
print(qq_plot(plot_dat$log10_proteome_abund, "QQ plot: log10 Proteome abundance"))
print(qq_plot(plot_dat$log10_detected_prot,  "QQ plot: log10 Detected proteins"))

# -----------------------------
# 4) Decide correlation method
#    Rule (reviewer-friendly):
#      - If both log10 variables are approximately normal -> Pearson on log10
#      - Otherwise -> Spearman (robust, monotonic)
#
# We use Shapiro p>0.05 as an automated heuristic, BUT still recommend Spearman
# for abundance data (heavy tails, compositional effects).
# -----------------------------
norm_ok_species  <- tryCatch(shapiro.test(plot_dat$log10_species_abund)$p.value  > 0.05, error = function(e) FALSE)
norm_ok_proteome <- tryCatch(shapiro.test(plot_dat$log10_proteome_abund)$p.value > 0.05, error = function(e) FALSE)

use_method <- if (norm_ok_species && norm_ok_proteome) "pearson" else "spearman"

cat("\nChosen correlation method (auto-heuristic):", use_method, "\n")
cat("Note: For abundance data, Spearman is typically the most robust and reviewer-safe.\n")

# -----------------------------
# 5) Correlation tests
#    A) Proteome vs Species abundance (log10)
#    B) Proteome vs Detected proteins (log10)
#    C) Species abundance vs Detected proteins (log10)  [optional but informative]
# -----------------------------
cat("\n--- Correlation tests on log10 scale ---\n")

test_A <- cor.test(plot_dat$log10_species_abund, plot_dat$log10_proteome_abund, method = use_method, exact = FALSE)
cat("\nA) log10 Proteome ~ log10 Species abundance\n")
print(test_A)

test_B <- cor.test(plot_dat$log10_detected_prot, plot_dat$log10_proteome_abund, method = use_method, exact = FALSE)
cat("\nB) log10 Proteome ~ log10 Detected proteins\n")
print(test_B)

test_C <- cor.test(plot_dat$log10_species_abund, plot_dat$log10_detected_prot, method = use_method, exact = FALSE)
cat("\nC) log10 Species abundance ~ log10 Detected proteins (informative)\n")
print(test_C)

# -----------------------------
# 6) "Does proteome correlate with species abundance independently of detection?"
#    Reviewer-friendly approach:
#      Fit a multiple regression on log10 scale:
#        log10_proteome ~ log10_species_abund + log10_detected_prot
#
# Interpretation:
#   - Coefficient for log10_species_abund tests association with proteome
#     AFTER accounting for detection depth.
#
# Notes:
#   - This is not a formal partial Spearman correlation,
#     but it is widely accepted, transparent, and easy to interpret.
# -----------------------------
cat("\n--- Multiple regression on log10 scale (controls for detection depth) ---\n")

fit_lm <- lm(log10_proteome_abund ~ log10_species_abund + log10_detected_prot, data = plot_dat)
print(summary(fit_lm))

# Optional: Robust regression as a sensitivity check (handles outliers better)
# Requires MASS (installed with base R on most systems)
suppressPackageStartupMessages(library(MASS))
fit_rlm <- MASS::rlm(log10_proteome_abund ~ log10_species_abund + log10_detected_prot, data = plot_dat)
cat("\nRobust regression (MASS::rlm) coefficients (sensitivity check):\n")
print(coef(summary(fit_rlm)))

# -----------------------------
# 7) Produce a reviewer-ready plot with annotation from test A
#    - Scatter in log10-log10 space
#    - Dot size = Detected_Proteins_at_Peak (raw counts for readability)
#    - Smooth line for visualization (LOESS); does NOT define the statistical test
# -----------------------------
# Build a compact annotation string from the correlation test
corr_label <- paste0(
  toupper(substr(use_method, 1, 1)), substr(use_method, 2, nchar(use_method)),
  " correlation: ",
  if (use_method == "pearson") "r = " else if (use_method == "spearman") "rho = " else "tau = ",
  round(unname(test_A$estimate), 3),
  "\nP = ", signif(test_A$p.value, 3),
  "\nN = ", nrow(plot_dat)
)

p <- ggplot(plot_dat, aes(x = log10_species_abund, y = log10_proteome_abund)) +
  geom_point(aes(size = Detected_Proteins_at_Peak), alpha = 0.60) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.9) +
  scale_size_continuous(name = "Detected proteins\n(at peak date)") +
  labs(
    x = "log10 Species abundance at peak date",
    y = "log10 Proteome abundance at peak date",
    title = "Proteome recovery vs species abundance (date-matched peak)"
  ) +
  annotate(
    "text",
    x = min(plot_dat$log10_species_abund, na.rm = TRUE),
    y = max(plot_dat$log10_proteome_abund, na.rm = TRUE),
    hjust = 0, vjust = 1,
    label = corr_label,
    size = 3.5
  ) +
  theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(face = "bold"),
    legend.key = element_blank()
  )

print(p)
```

```{r}
################################################################################
# Reviewer-proof stats & plots (PEAK-DATE ONLY; uses dat = species_summary_peak_only.csv)
# Author: Cyrill Hofer
#
# DATA ASSUMPTION:
#   dat has ONE ROW PER SPECIES and contains:
#     - Species_Abundance_at_Peak
#     - Proteome_Abundance_on_PeakDate
#     - Detected_Proteins_at_Peak
#     - taxonomy columns (Phylum, etc.)
#
# WHAT THIS SCRIPT DOES (reviewer-ready package):
#  1) QC + transformations (log10) and distribution checks (QQ + Shapiro)
#  2) Primary association: Spearman correlation (log10 abundance vs log10 proteome)
#     - bootstrap CI for rho (robust uncertainty estimate)
#  3) Confounding/mediation check:
#     - Spearman: proteome vs detected proteins; abundance vs detected proteins
#     - Multiple regression on log10 scale: proteome ~ abundance + detected proteins
#     - Robust regression sensitivity check (rlm)
#  4) Influence/outlier robustness:
#     - Leave-one-out slope stability (fast diagnostic)
#  5) Stratified sensitivity:
#     - Spearman correlation within major phyla (only if n>=20)
#  6) Publication-quality figure(s):
#     - Panel A: Proteome vs abundance (size = detected proteins) + LOESS (visual only)
#     - Panel B: Proteome vs abundance but color = phylum (optional)
#
# LIMITATION (explicit):
#  - With peak-date-only summary, you cannot model detection probability (needs zeros / all dates).
#  - "Proteome total signal" is mechanically linked to number of detected proteins.
#    We therefore provide a model controlling for detected proteins, and robust checks.
################################################################################

# -----------------------------
# 0) Libraries (keep to what you already loaded + MASS for rlm)
# -----------------------------
library(dplyr)
library(ggplot2)
library(car)
library(tibble)
library(MASS)

# -----------------------------
# 1) Minimal QC + log transforms
# -----------------------------
plot_dat <- dat %>%
  filter(
    !is.na(Species_Abundance_at_Peak),
    !is.na(Proteome_Abundance_on_PeakDate),
    !is.na(Detected_Proteins_at_Peak),
    Species_Abundance_at_Peak > 0,
    Proteome_Abundance_on_PeakDate > 0,
    Detected_Proteins_at_Peak > 0
  ) %>%
  mutate(
    log10_species_abund  = log10(Species_Abundance_at_Peak),
    log10_proteome_abund = log10(Proteome_Abundance_on_PeakDate),
    log10_detected_prot  = log10(Detected_Proteins_at_Peak)
  )

cat("N species used after filtering:", nrow(plot_dat), "\n")

# -----------------------------
# 2) Distribution checks (log scale is what we analyze/plot)
#    Shapiro is sensitive at N~500; interpret alongside QQ plots.
# -----------------------------
run_shapiro <- function(x, label) {
  x <- x[is.finite(x)]
  if (length(x) < 3) return(invisible(NULL))
  out <- shapiro.test(x)
  cat(label, "Shapiro-Wilk: W =", round(out$statistic, 4), ", p =", signif(out$p.value, 3), "\n")
  invisible(out)
}

cat("\n--- Normality checks (log10 scale; Shapiro is sensitive for large N) ---\n")
run_shapiro(plot_dat$log10_species_abund,  "log10 Species abundance")
run_shapiro(plot_dat$log10_proteome_abund, "log10 Proteome abundance")
run_shapiro(plot_dat$log10_detected_prot,  "log10 Detected proteins")

qq_plot <- function(x, title) {
  ggplot(data.frame(x = x), aes(sample = x)) +
    stat_qq() + stat_qq_line() +
    theme_bw(base_size = 12) +
    labs(title = title, x = "Theoretical quantiles", y = "Sample quantiles")
}
print(qq_plot(plot_dat$log10_species_abund,  "QQ plot: log10 Species abundance"))
print(qq_plot(plot_dat$log10_proteome_abund, "QQ plot: log10 Proteome abundance"))
print(qq_plot(plot_dat$log10_detected_prot,  "QQ plot: log10 Detected proteins"))

# -----------------------------
# 3) Primary correlation (reviewer-safe): Spearman on log10 scale
# -----------------------------
cat("\n--- Primary association (Spearman; log10 scale) ---\n")
cor_primary <- cor.test(plot_dat$log10_species_abund, plot_dat$log10_proteome_abund,
                        method = "spearman", exact = FALSE)
print(cor_primary)

# Bootstrap CI for Spearman rho (robust uncertainty; no distribution assumptions)
set.seed(1)
B <- 5000
n <- nrow(plot_dat)
rho_boot <- replicate(B, {
  idx <- sample.int(n, n, replace = TRUE)
  suppressWarnings(cor(plot_dat$log10_species_abund[idx],
                       plot_dat$log10_proteome_abund[idx],
                       method = "spearman"))
})
rho_ci <- quantile(rho_boot, c(0.025, 0.975), na.rm = TRUE)
cat("\nBootstrap 95% CI for Spearman rho:",
    paste0("[", round(rho_ci[1], 3), ", ", round(rho_ci[2], 3), "]"), "\n")

# -----------------------------
# 4) Confounding / coupling checks
#    (Total proteome signal is mechanically linked to # detected proteins)
# -----------------------------
cat("\n--- Coupling checks (Spearman; log10 scale) ---\n")

cor_prot_vs_detected <- cor.test(plot_dat$log10_detected_prot, plot_dat$log10_proteome_abund,
                                 method = "spearman", exact = FALSE)
cat("\nProteome vs detected proteins:\n")
print(cor_prot_vs_detected)

cor_abund_vs_detected <- cor.test(plot_dat$log10_species_abund, plot_dat$log10_detected_prot,
                                  method = "spearman", exact = FALSE)
cat("\nSpecies abundance vs detected proteins:\n")
print(cor_abund_vs_detected)

# -----------------------------
# 5) Multiple regression (log10 scale) controlling for detection depth
#    Interpretation:
#     - coefficient for log10_species_abund tests association with proteome
#       AFTER accounting for detection depth.
#     - NOT a causal adjustment (detected proteins may be mediator), but a
#       reviewer-required sensitivity check to show abundance effect persists.
# -----------------------------
cat("\n--- Multiple regression: log10 Proteome ~ log10 Abundance + log10 Detected proteins ---\n")
fit_lm <- lm(log10_proteome_abund ~ log10_species_abund + log10_detected_prot, data = plot_dat)
print(summary(fit_lm))

# Collinearity diagnostic (reviewers may ask)
cat("\nVIF (collinearity diagnostic; >5–10 may be concerning):\n")
print(car::vif(fit_lm))

# Robust regression sensitivity check (downweights outliers)
fit_rlm <- MASS::rlm(log10_proteome_abund ~ log10_species_abund + log10_detected_prot, data = plot_dat)
cat("\nRobust regression (rlm) coefficients (sensitivity check):\n")
print(coef(summary(fit_rlm)))

# -----------------------------
# 6) Influence diagnostic: leave-one-out slope stability (fast reviewer check)
#    We track how the abundance coefficient changes if we drop one species.
# -----------------------------
beta_full <- coef(fit_lm)["log10_species_abund"]
beta_loo <- vapply(seq_len(nrow(plot_dat)), function(i) {
  coef(lm(log10_proteome_abund ~ log10_species_abund + log10_detected_prot,
          data = plot_dat[-i, ]))["log10_species_abund"]
}, numeric(1))

cat("\nAbundance coefficient (full model):", round(beta_full, 3), "\n")
cat("LOO abundance coefficient range:",
    paste0("[", round(min(beta_loo), 3), ", ", round(max(beta_loo), 3), "]"), "\n")

# -----------------------------
# 7) Stratified sensitivity: Spearman within major phyla
#    (helps address taxon-structured biases; only for n>=20 per phylum)
# -----------------------------
cat("\n--- Stratified Spearman by Phylum (n>=20) ---\n")
phy_res <- plot_dat %>%
  filter(!is.na(Phylum)) %>%
  count(Phylum, name = "n") %>%
  filter(n >= 20) %>%
  pull(Phylum)

if (length(phy_res) == 0) {
  cat("No phyla with n>=20; skipping.\n")
} else {
  phylum_stats <- lapply(phy_res, function(ph) {
    sub <- plot_dat %>% filter(Phylum == ph)
    ct  <- cor.test(sub$log10_species_abund, sub$log10_proteome_abund,
                    method = "spearman", exact = FALSE)
    tibble(
      Phylum = ph,
      n = nrow(sub),
      rho = unname(ct$estimate),
      p_value = ct$p.value
    )
  }) %>% bind_rows() %>%
    arrange(desc(n))
  print(phylum_stats)
}

# -----------------------------
# 8) Reviewer-ready figure (primary)
#    - Spearman rho + bootstrap CI annotation
#    - LOESS smooth for visualization ONLY
#    - Point size = detected proteins
# -----------------------------
rho <- unname(cor_primary$estimate)
pval <- cor_primary$p.value
p_txt <- if (pval < 2.2e-16) "p < 2.2e-16" else paste0("p = ", signif(pval, 3))

anno <- paste0(
  "Spearman ρ = ", round(rho, 3),
  "\n95% CI (bootstrap) = [", round(rho_ci[1], 3), ", ", round(rho_ci[2], 3), "]",
  "\n", p_txt,
  "\nN = ", nrow(plot_dat)
)

p1 <- ggplot(plot_dat, aes(x = log10_species_abund, y = log10_proteome_abund)) +
  geom_point(aes(size = Detected_Proteins_at_Peak), alpha = 0.60) +
  geom_smooth(method = "loess", se = TRUE, color = "black", linewidth = 0.9) +
  scale_size_continuous(name = "Detected proteins\n(at peak date)") +
  labs(
    x = "log10 Species abundance at peak date",
    y = "log10 Proteome abundance at peak date",
    title = "Proteome recovery vs species abundance (date-matched peak)"
  ) +
  annotate(
    "text",
    x = min(plot_dat$log10_species_abund, na.rm = TRUE),
    y = max(plot_dat$log10_proteome_abund, na.rm = TRUE),
    hjust = 0, vjust = 1,
    label = anno,
    size = 3.5
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.key = element_blank())

print(p1)


# -----------------------------
# 9) OPTIONAL figure: color by phylum (addresses taxonomic bias visually)
#    If you have a custom palette, you can map it here; otherwise ggplot defaults.
# -----------------------------
p2 <- ggplot(plot_dat, aes(x = log10_species_abund, y = log10_proteome_abund, color = Phylum)) +
  geom_point(aes(size = Detected_Proteins_at_Peak), alpha = 0.60) +
  geom_smooth(aes(group = 1), method = "loess", se = TRUE, color = "black", linewidth = 0.9) +
  scale_size_continuous(name = "Detected proteins\n(at peak date)") +
  labs(
    x = "log10 Species abundance at peak date",
    y = "log10 Proteome abundance at peak date",
    title = "Proteome recovery vs species abundance (colored by phylum)"
  ) +
  theme_bw(base_size = 12) +
  theme(plot.title = element_text(face = "bold"),
        legend.key = element_blank())

print(p2)

# -----------------------------
# 10) Write a compact results table (for the manuscript)
# -----------------------------
results_tbl <- tibble(
  N = nrow(plot_dat),
  Spearman_rho_proteome_vs_abundance = unname(cor_primary$estimate),
  Spearman_p_proteome_vs_abundance   = cor_primary$p.value,
  Spearman_rho_proteome_vs_detected  = unname(cor_prot_vs_detected$estimate),
  Spearman_p_proteome_vs_detected    = cor_prot_vs_detected$p.value,
  Spearman_rho_abundance_vs_detected = unname(cor_abund_vs_detected$estimate),
  Spearman_p_abundance_vs_detected   = cor_abund_vs_detected$p.value,
  R2_lm = summary(fit_lm)$r.squared,
  beta_log10_abundance = coef(fit_lm)["log10_species_abund"],
  p_beta_log10_abundance = summary(fit_lm)$coefficients["log10_species_abund","Pr(>|t|)"],
  beta_log10_detected = coef(fit_lm)["log10_detected_prot"],
  p_beta_log10_detected = summary(fit_lm)$coefficients["log10_detected_prot","Pr(>|t|)"],
  rho_boot_ci_low = rho_ci[1],
  rho_boot_ci_high = rho_ci[2],
  beta_loo_min = min(beta_loo),
  beta_loo_max = max(beta_loo)
)

write.csv(results_tbl, "Stats_PeakDate_Correlation_Summary.csv", row.names = FALSE)
cat("\nWrote: Stats_PeakDate_Correlation_Summary.csv\n")

################################################################################
# INTERPRETATION GUIDE (for your manuscript; keep in text, not code)
#
# - Primary result: Spearman correlation supports monotonic association between
#   metagenome abundance and proteome recovery on the same date.
# - Strong coupling between proteome total signal and detected proteins is expected
#   because totals accumulate over detected features.
# - Multiple regression shows an abundance effect persists even after accounting
#   for detection depth (sensitivity, not causal adjustment).
# - Stratified correlations help rule out a single phylum driving the trend.
#
# WHAT THIS PEAK-DATE-ONLY DATASET CANNOT DO:
# - It cannot quantify detection probability vs abundance (needs zeros / all dates).
# - It cannot address temporal within-species variation (needs repeated dates).
################################################################################

```

## Proteome recovery vs metagenome species abundance (date-matched peak): statistics summary

### Question
Do species with higher **metagenome-derived abundance** on their **peak date** (Peak_Date) also show higher **proteome signal** measured on the **same date**?

### Expectation (a priori)
If proteome recovery reflects underlying biomass, then **proteome abundance** (summed, species-specific protein signal) should **increase monotonically** with **metagenome species abundance** when compared on the *same sampling date*.  
Because abundance data span orders of magnitude and are heavy-tailed, a **non-parametric** correlation is expected to be most appropriate.

---

### Data used
- Input table: `species_summary_peak_only.csv`
- One row per species (N = 481 after filtering)
- Key variables:
  - `Species_Abundance_at_Peak` (metagenome; peak date)
  - `Proteome_Abundance_on_PeakDate` (proteome; same date)
  - `Detected_Proteins_at_Peak` (proteome detection depth; same date)

---

### Pre-processing
- Remove invalid values for log-transform:
  - `Species_Abundance_at_Peak > 0`
  - `Proteome_Abundance_on_PeakDate > 0`
  - `Detected_Proteins_at_Peak > 0`
- Log10-transform abundance variables for analysis and visualization:
  - `log10_species_abund = log10(Species_Abundance_at_Peak)`
  - `log10_proteome_abund = log10(Proteome_Abundance_on_PeakDate)`
  - `log10_detected_prot = log10(Detected_Proteins_at_Peak)`

---

### Normality assessment (decision on parametric vs non-parametric)
**Tests performed**
- Shapiro–Wilk on log10-transformed variables (noting high sensitivity at N≈500)
- QQ-plots for visual assessment

**Result**
Even after log10 transform, Shapiro–Wilk remained significant (p < 0.01), consistent with minor tail deviations and/or ties.  
➡ Therefore, the primary inferential test uses **Spearman rank correlation** (non-parametric, robust).

---

### Statistical tests
1. **Primary test (main claim):**  
   Spearman correlation between `log10_species_abund` and `log10_proteome_abund`.

2. **Uncertainty (effect size precision):**  
   Non-parametric bootstrap confidence interval for Spearman’s ρ  
   - 5,000 bootstrap replicates  
   - each replicate resamples **N = 481 species (rows)** with replacement  
   - 95% CI = 2.5th–97.5th percentiles of bootstrap ρ values

3. **Detection-depth coupling (critical sensitivity checks):**
   - Spearman: `log10_proteome_abund` vs `log10_detected_prot`
   - Spearman: `log10_species_abund` vs `log10_detected_prot`

4. **Multiple regression sensitivity analysis (not causal, but reviewer-relevant):**
   - `log10_proteome_abund ~ log10_species_abund + log10_detected_prot`
   - plus robust regression (`MASS::rlm`) as an outlier-resistant check

---

### Results
**Primary association (Spearman; log10 scale)**
- Spearman ρ = **0.646**
- p-value **< 2.2×10⁻¹⁶**
- Bootstrap 95% CI for ρ ≈ **0.59–0.70**
- N = **481**

**Detection-depth coupling**
- Proteome vs detected proteins:
  - Spearman ρ = **0.845**, p < 2.2×10⁻¹⁶  
  (expected because total signal accumulates over detected proteins)
- Species abundance vs detected proteins:
  - Spearman ρ = **0.622**, p < 2.2×10⁻¹⁶  
  (more abundant species tend to yield more detected proteins)

**Multiple regression sensitivity (log10 scale)**
- Both predictors remain positive and significant:
  - `log10_species_abund` coefficient ≈ **0.254** (p ≈ 5.3×10⁻⁶)
  - `log10_detected_prot` coefficient ≈ **1.218** (p < 2.2×10⁻¹⁶)
- Model fit: R² ≈ **0.725**
- Robust regression produced similar coefficients → conclusions not driven by a small number of outliers

---

### Interpretation (reviewer-ready)
Across 481 species (one observation per species), **proteome recovery measured on the same date** shows a **moderately strong, statistically robust monotonic association** with **metagenome-derived species abundance** (Spearman ρ ≈ 0.65, bootstrap 95% CI ≈ 0.59–0.70).  

A strong correlation between total proteome signal and the **number of detected proteins** is also observed (ρ ≈ 0.85), consistent with both biological detectability and a mechanical coupling (total intensity sums over detected features). Importantly, species abundance retains a significant positive association with proteome signal even when accounting for detection depth in a sensitivity regression, supporting the conclusion that metagenome abundance is informative for proteome recovery under the date-matched peak design.

---

### Limitations to acknowledge explicitly
- Peak-date selection may emphasize “best-case detectability” per species and can inflate associations relative to a full time-series analysis.
- Total proteome signal is partially coupled to detection depth by definition; hence reporting both the primary Spearman test and the detection-depth sensitivity analyses is essential.
- This peak-date-only dataset cannot evaluate detection probability across zeros or temporal within-species variation.


```{r}
# ------------------------------------------------------------
# Publication-ready scatter plot
# Proteome recovery vs species abundance (date-matched peak)
# Final axis limits and binned protein detection
# ------------------------------------------------------------

# Prepare plotting data
clean_dat <- dat %>%
  filter(
    Species_Abundance_at_Peak > 0,
    Proteome_Abundance_on_PeakDate > 0,
    Detected_Proteins_at_Peak > 0
  ) %>%
  mutate(
    log10_species_abund  = log10(Species_Abundance_at_Peak),
    log10_proteome_abund = log10(Proteome_Abundance_on_PeakDate),

    # Bin detected proteins
    Detected_Protein_Class = cut(
      Detected_Proteins_at_Peak,
      breaks = c(0, 20, 50, 100, Inf),
      labels = c("0–20", "20–50", "50–100", "100+"),
      right  = TRUE
    )
  )

# Build plot
p <- ggplot(
  clean_dat,
  aes(x = log10_species_abund, y = log10_proteome_abund)
) +
  geom_point(
    aes(size = Detected_Protein_Class),
    shape  = 21,
    color  = "black",
    fill   = "#bfd3c1",
    alpha  = 0.8,
    stroke = 0.4
  ) +
  geom_smooth(
    method = "loess",
    se     = TRUE,
    color  = "black",
    linewidth = 0.9
  ) +
  # Only control the ticks shown (no limits!)
  scale_x_continuous(breaks = seq(-1, 1.5, by = 0.5)) +
  scale_y_continuous(breaks = seq(4, 9, by = 1)) +
  scale_size_manual(
    values = c("0–20" = 3, "20–50" = 5, "50–100" = 7, "100+" = 10),
    name = "Detected proteins\n(at peak date)"
  ) +
  labs(
    x = "log10 Species abundance at peak date",
    y = "log10 Proteome abundance at peak date"
  ) +
  theme_minimal() +
  theme(
    legend.position  = "right",
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 10),
    axis.title       = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

print(p)



```

Figure X | Proteome recovery reflects species abundance.
Each point represents one species. The x-axis shows log₁₀ metagenome-derived species abundance at the peak date, and the y-axis shows log₁₀ summed proteome abundance measured on the same date. Point size indicates the number of detected species-specific proteins (0–20, 20–50, 50–100, >100). The black curve represents a LOESS smoother for visualization only. Spearman’s rank correlation coefficient (ρ) with a 95% confidence interval estimated from 5,000 bootstrap resamples (species-level resampling, N = 481 per replicate) is shown.

```{r}
# ------------------------------------------------------------
# FINAL publication plot (no point loss) + stats annotation
# Adds: p-value stars (****) + numeric p format
# ------------------------------------------------------------

# Prepare data
clean_dat <- dat %>%
  filter(
    Species_Abundance_at_Peak > 0,
    Proteome_Abundance_on_PeakDate > 0,
    Detected_Proteins_at_Peak > 0
  ) %>%
  mutate(
    log10_species_abund  = log10(Species_Abundance_at_Peak),
    log10_proteome_abund = log10(Proteome_Abundance_on_PeakDate),
    Detected_Protein_Class = cut(
      Detected_Proteins_at_Peak,
      breaks = c(0, 20, 50, 100, Inf),
      labels = c("0–20", "20–50", "50–100", "100+"),
      right  = TRUE
    )
  )

# Spearman + bootstrap CI (self-contained, robust)
cor_test <- cor.test(clean_dat$log10_species_abund, clean_dat$log10_proteome_abund,
                     method = "spearman", exact = FALSE)

rho  <- unname(cor_test$estimate)
pval <- cor_test$p.value
N    <- nrow(clean_dat)

set.seed(1)
B <- 5000
n <- nrow(clean_dat)
rho_boot <- replicate(B, {
  idx <- sample.int(n, n, replace = TRUE)
  suppressWarnings(cor(clean_dat$log10_species_abund[idx],
                       clean_dat$log10_proteome_abund[idx],
                       method = "spearman"))
})
ci <- quantile(rho_boot, c(0.025, 0.975), na.rm = TRUE)
rho_low <- unname(ci[1])
rho_hi  <- unname(ci[2])

# p-value text and stars
p_txt <- if (is.na(pval)) {
  "p = NA"
} else if (pval < 2.2e-16) {
  "p < 2.2×10⁻¹⁶"
} else if (pval < 1e-3) {
  paste0("p = ", formatC(pval, format = "e", digits = 2))
} else {
  paste0("p = ", signif(pval, 3))
}

p_stars <- if (is.na(pval)) {
  ""
} else if (pval < 1e-4) {
  "****"
} else if (pval < 1e-3) {
  "***"
} else if (pval < 1e-2) {
  "**"
} else if (pval < 0.05) {
  "*"
} else {
  "ns"
}

# Build annotation label
stat_label <- paste0(
  "Spearman \u03C1 = ", round(rho, 2),
  "\n95% CI = [", round(rho_low, 2), ", ", round(rho_hi, 2), "]",
  "\n", p_txt, "  ", p_stars,
  "\nN = ", N
)

# Plot (NO limits -> no dropping points)
p <- ggplot(clean_dat, aes(x = log10_species_abund, y = log10_proteome_abund)) +
  geom_point(
    aes(size = Detected_Protein_Class),
  shape  = 21,
  color  = "black",                     # outline (opaque)
  fill   = scales::alpha("#bfd3c1", 0.4), # transparent fill
  stroke = 0.4
) +
  geom_smooth(
    method = "loess",
    se = TRUE,
    color = "black",
    linewidth = 0.9
  ) +
  # show the ticks you want (no data removal)
  scale_x_continuous(breaks = seq(-1, 1.5, by = 0.5)) +
  scale_y_continuous(breaks = seq(4, 9, by = 1)) +
  scale_size_manual(
    values = c("0–20" = 3, "20–50" = 5, "50–100" = 7, "100+" = 10),
    name = "Detected proteins\n(at peak date)"
  ) +
  labs(
    x = "log10 Species abundance at peak date",
    y = "log10 Proteome abundance at peak date"
  ) +
  annotate(
    "text",
    x = min(clean_dat$log10_species_abund, na.rm = TRUE),
    y = max(clean_dat$log10_proteome_abund, na.rm = TRUE),
    label = stat_label,
    hjust = 0,
    vjust = 1,
    size  = 3.5
  ) +
  theme_minimal() +
  theme(
    legend.position  = "right",
    axis.text.x      = element_text(size = 12),
    axis.text.y      = element_text(size = 10),
    axis.title       = element_text(size = 12),
    panel.grid.minor = element_blank()
  )

print(p)

ggsave(
  filename = "Fig_ProteomeRecovery_vs_SpeciesAbundance_FINAL.pdf",
  plot     = p,
  width    = 10,
  height   = 6,
  device   = pdf,
  bg       = "white"
)
```

