---
title: "Transporter Investment Analysis"
output:
  html_document:
    df_print: paged
---

```{r setup, message=FALSE, warning=FALSE, include=FALSE}
rm(list = ls())

############################################################
# 0) LOAD RETAINED DATA (CSV exports from QC pipeline)
#    Then reproduce transporter vs non-transporter pie charts
############################################################

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_04/Transporter_Investment/")

# ---- Paths (edit if you put the CSVs elsewhere)
protein_path <- "../Data_Filtration/retained_protein_table.csv"
species_path <- "../Data_Filtration/retained_species.csv"
keys_path    <- "../Data_Filtration/retained_species_dates.csv"
species_summary_path <- "../../../Data/Species_Data/Final_Data/Species_Table.csv"

# ---- Load
protein_final          <- read_csv(protein_path, show_col_types = FALSE)
retained_species       <- read_csv(species_path, show_col_types = FALSE)
retained_species_dates <- read_csv(keys_path,    show_col_types = FALSE)
species_taxonomy <- read_csv(species_summary_path, show_col_types = FALSE)

species_taxonomy <- species_taxonomy %>%
  dplyr::select(Species_ID, Phylum, Class, Order, Family, Genus, Species) %>%
  distinct()

# ---- Basic cleanup / checks
protein_final <- protein_final %>% mutate(Date = as.Date(Date))
retained_species_dates <- retained_species_dates %>% mutate(Date = as.Date(Date))

stopifnot(
  all(c("Species_ID","Protein_ID","Date","Group_For_Plots") %in% names(protein_final)),
  all(c("Species_ID","Date","Group_For_Plots") %in% names(retained_species_dates))
)

cat("\n=== Loaded retained datasets ===\n")
cat("protein_final rows:", nrow(protein_final), "\n")
cat("retained species×dates:", nrow(retained_species_dates), "\n")
cat("retained species:", nrow(retained_species), "\n")
```

# Transporter fraction investigation
```{r}
############################################################
# TPF investigation 
#
# What this script does
#  1) Compute transporter proteome fraction (TPF) per Species × Date:
#       TPF = sum(abundance of transporter proteins) / sum(total protein abundance)
#     Transporters are defined as: KEGG_Subcategory contains the word "Transporters"
#
#  2) Summarise TPF per species (one value per species):
#       - mean_TPF across dates
#       - median_TPF across dates
#       - n_dates retained
#       - mean proteins detected per date (optional size covariate for plots)
#
#  3) Visualisations:
#       A) Density distributions of mean_TPF and median_TPF by trophic group (with stats text)
#       B) Scatter of mean_TPF vs TrophicIndex and median_TPF vs TrophicIndex
#          (dot size = mean proteins detected per date)
#       C) Boxplots for mean_TPF and median_TPF (outliers hidden, datapoints shown)
#          with overall + pairwise significance (ns/*/**/***/****)
#
#  4) Statistics (species-level inference):
#       - Overall: Kruskal–Wallis (non-parametric; appropriate for bounded fractions)
#       - Pairwise: Wilcoxon rank-sum with BH correction
#       - Effect sizes: rank-biserial correlation + 95% CI
#
# Inputs expected in `protein_final`:
#   Species_ID, Date, Protein_ID, KEGG_Subcategory,
#   Protein_Abundance_IBAQ_Norm,
#   TrophicGroup, TrophicIndex
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(ggplot2)
  library(rstatix)  # kruskal_test(), pairwise_wilcox_test(), wilcox_effsize(), add_significance()
  library(ggpubr)   # stat_pvalue_manual()
})

############################################################
# 0) USER SETTINGS (edit here only)
############################################################

# Expected trophic group labels + plotting order
trophic_levels <- c("Oligo_like", "Intermediate", "Copio_like")

# Consistent group colors
trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

# Multiple testing correction for pairwise comparisons
padj_method <- "BH"

# Optional: minimum number of valid dates per species for stable summaries
# (set to 5 if that matches your current filtering logic; 3 is a minimal default)
min_valid_dates_per_species <- 5


############################################################
# 1) HELPER FUNCTIONS (formatting + stats tables)
############################################################

# Pretty p-value printing for annotations
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) return("<0.001")
  formatC(p, format = "e", digits = 2)
}

# Create pairwise Wilcoxon results incl. adjusted p-values + star labels + y positions
pairwise_tests_with_y <- function(df, response_col, padj_method = "BH", step_increase = 0.08) {
  df %>%
    pairwise_wilcox_test(
      reformulate("TrophicGroup", response = response_col),
      p.adjust.method = padj_method
    ) %>%
    add_significance("p.adj") %>%               # adds p.adj.signif = ns/*/**/***/****
    add_y_position(fun = "max", step.increase = step_increase)
}

# Compute rank-biserial effect sizes and join to pairwise test table
pairwise_tests_plus_effects <- function(df, response_col, padj_method = "BH") {

  pw <- pairwise_tests_with_y(df, response_col, padj_method = padj_method)

  eff <- df %>%
    wilcox_effsize(
      reformulate("TrophicGroup", response = response_col),
      paired = FALSE,
      ci = TRUE
    ) %>%
    select(group1, group2, effsize, conf.low, conf.high)

  pw_full <- pw %>%
    left_join(eff, by = c("group1", "group2"))

  list(pw = pw, pw_full = pw_full, eff = eff)
}

# Build a small multi-line annotation string for density plots
build_density_annotation <- function(kw_obj, pw_tbl, levels_vec) {
  # expected 3-level comparisons for your trophic_levels
  a <- levels_vec[1]; b <- levels_vec[2]; c <- levels_vec[3]

  # helper to find adjusted p-value for a pair
  get_padj <- function(tbl, g1, g2) {
    p <- tbl$p.adj[(tbl$group1 == g1 & tbl$group2 == g2) | (tbl$group1 == g2 & tbl$group2 == g1)]
    if (length(p) == 0) NA_real_ else p[1]
  }

  paste0(
    "Kruskal–Wallis p = ", fmt_p(kw_obj$p.value), "\n",
    "Pairwise Wilcoxon (BH):\n",
    a, " vs ", b, ": ", fmt_p(get_padj(pw_tbl, a, b)), "\n",
    a, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, a, c)), "\n",
    b, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, b, c))
  )
}


############################################################
# 2) COMPUTE TPF PER SPECIES × DATE
############################################################

species_date_tpf <- protein_final %>%
  mutate(
    # Transporter definition:
    is_transporter = !is.na(KEGG_Subcategory) &
      str_detect(KEGG_Subcategory, "\\bTransporters\\b"),
    abund = Protein_Abundance_IBAQ_Norm
  ) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    abund_total = sum(abund, na.rm = TRUE),
    abund_trans = sum(abund[is_transporter], na.rm = TRUE),
    TPF = abund_trans / abund_total,
    .groups = "drop"
  ) %>%
  # Keep only proper fractions and valid denominators
  filter(is.finite(TPF), abund_total > 0, TPF > 0, TPF < 1)

############################################################
# 3) PROTEIN DETECTION PER SPECIES × DATE (for point size later)
############################################################

species_date_protcount <- protein_final %>%
  group_by(Species_ID, Date) %>%
  summarise(N_prot_detected = n_distinct(Protein_ID), .groups = "drop")

############################################################
# 4) SPECIES-LEVEL SUMMARY TABLE (one row per species)
#    Join trophic traits (TrophicGroup, TrophicIndex) from protein_final.
############################################################

species_trait_df <- species_date_tpf %>%
  left_join(species_date_protcount, by = c("Species_ID", "Date")) %>%
  left_join(
    protein_final %>% distinct(Species_ID, TrophicGroup, TrophicIndex),
    by = "Species_ID"
  ) %>%
  mutate(
    TrophicGroup = factor(as.character(TrophicGroup), levels = trophic_levels)
  ) %>%
  group_by(Species_ID, TrophicGroup, TrophicIndex) %>%
  summarise(
    n_dates = n(),
    mean_TPF   = mean(TPF, na.rm = TRUE),
    median_TPF = median(TPF, na.rm = TRUE),
    mean_prot_detected = mean(N_prot_detected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  # reviewer-safe sanity filters
  filter(
    n_dates >= min_valid_dates_per_species,
    is.finite(TrophicIndex),
    is.finite(mean_TPF),   mean_TPF > 0,   mean_TPF < 1,
    is.finite(median_TPF), median_TPF > 0, median_TPF < 1,
    is.finite(mean_prot_detected)
  ) %>%
  mutate(
    mean_prot_detected = as.integer(round(mean_prot_detected))
  )

# Print sample sizes (useful for reporting)
cat("\n=== Species counts per trophic group (after filtering) ===\n")
print(species_trait_df %>% count(TrophicGroup))


############################################################
# 5) STATISTICS (species-level): overall + pairwise + effects
############################################################

# Overall Kruskal–Wallis tests
kw_mean   <- kruskal.test(mean_TPF ~ TrophicGroup, data = species_trait_df)
kw_median <- kruskal.test(median_TPF ~ TrophicGroup, data = species_trait_df)

# Pairwise tests + effects (rank-biserial) for mean and median
res_mean   <- pairwise_tests_plus_effects(species_trait_df, "mean_TPF", padj_method = padj_method)
res_median <- pairwise_tests_plus_effects(species_trait_df, "median_TPF", padj_method = padj_method)

pw_mean      <- res_mean$pw
pw_mean_full <- res_mean$pw_full

pw_median      <- res_median$pw
pw_median_full <- res_median$pw_full

# Optional: print tables for supplement / reporting
cat("\n=== MEAN TPF: Kruskal–Wallis ===\n"); print(kw_mean)
cat("\n=== MEAN TPF: Pairwise Wilcoxon (BH) + effect size (rank-biserial) ===\n"); print(pw_mean_full)

cat("\n=== MEDIAN TPF: Kruskal–Wallis ===\n"); print(kw_median)
cat("\n=== MEDIAN TPF: Pairwise Wilcoxon (BH) + effect size (rank-biserial) ===\n"); print(pw_median_full)


############################################################
# 6) PLOT A: DENSITY DISTRIBUTIONS (mean_TPF and median_TPF)
############################################################

ann_mean <- build_density_annotation(kw_mean, pw_mean, trophic_levels)
ann_median <- build_density_annotation(kw_median, pw_median, trophic_levels)

p_density_mean <- ggplot(species_trait_df, aes(x = mean_TPF, fill = TrophicGroup, color = TrophicGroup)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  scale_color_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Distribution of mean transporter proteome fraction (TPF) by trophic strategy",
    x = "Mean TPF (per species)",
    y = "Density"
  ) +
  annotate(
    "label",
    x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
    label = ann_mean, size = 3.2, label.size = 0.25
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

p_density_median <- ggplot(species_trait_df, aes(x = median_TPF, fill = TrophicGroup, color = TrophicGroup)) +
  geom_density(alpha = 0.25, linewidth = 1) +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  scale_color_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Distribution of median transporter proteome fraction (TPF) by trophic strategy",
    x = "Median TPF (per species)",
    y = "Density"
  ) +
  annotate(
    "label",
    x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
    label = ann_median, size = 3.2, label.size = 0.25
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

print(p_density_mean)
print(p_density_median)


############################################################
# 7) PLOT B: TPF vs TROPHIC INDEX (dot size = mean proteins detected)
############################################################

# Size legend breaks (min + 1/3 + 2/3 + max); fall back to 4 equally spaced if needed
size_breaks <- species_trait_df %>%
  summarise(
    mn = min(mean_prot_detected, na.rm = TRUE),
    q1 = as.integer(round(quantile(mean_prot_detected, 1/3, na.rm = TRUE))),
    q2 = as.integer(round(quantile(mean_prot_detected, 2/3, na.rm = TRUE))),
    mx = max(mean_prot_detected, na.rm = TRUE)
  ) %>%
  unlist(use.names = FALSE) %>% unique() %>% sort()

if (length(size_breaks) < 4) {
  mn <- min(species_trait_df$mean_prot_detected, na.rm = TRUE)
  mx <- max(species_trait_df$mean_prot_detected, na.rm = TRUE)
  size_breaks <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
}

size_labels <- as.character(size_breaks)
size_labels[1] <- paste0(size_labels[1], " (min)")
size_labels[length(size_labels)] <- paste0(size_labels[length(size_labels)], " (max)")

base_scales_scatter <- list(
  scale_color_manual(values = trophic_cols, drop = FALSE),
  scale_size_continuous(
    name = "Mean proteins detected\n(distinct Protein_ID per date)",
    range = c(2, 10),
    breaks = size_breaks,
    labels = size_labels
  ),
  theme_bw(),
  theme(panel.grid.minor = element_blank())
)

p_scatter_mean <- ggplot(
  species_trait_df,
  aes(x = TrophicIndex, y = mean_TPF, color = TrophicGroup, size = mean_prot_detected)
) +
  geom_point(alpha = 0.85) +
  labs(
    title = "Mean transporter proteome fraction vs trophic index",
    x = "Trophic index",
    y = "Mean TPF (per species)",
    color = "Trophic group"
  ) +
  base_scales_scatter

p_scatter_median <- ggplot(
  species_trait_df,
  aes(x = TrophicIndex, y = median_TPF, color = TrophicGroup, size = mean_prot_detected)
) +
  geom_point(alpha = 0.85) +
  labs(
    title = "Median transporter proteome fraction vs trophic index",
    x = "Trophic index",
    y = "Median TPF (per species)",
    color = "Trophic group"
  ) +
  base_scales_scatter

print(p_scatter_mean)
print(p_scatter_median)


############################################################
# 8) PLOT C: BOXPLOTS + POINTS + SIGNIFICANCE STARS
#    Outliers hidden, but ALL datapoints are shown.
############################################################

p_box_mean <- ggplot(species_trait_df, aes(x = TrophicGroup, y = mean_TPF, fill = TrophicGroup)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
  geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
  stat_pvalue_manual(
    pw_mean,
    label = "p.adj.signif",
    tip.length = 0.01,
    hide.ns = FALSE
  ) +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Mean TPF across trophic groups",
    subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method, " correction"),
    x = NULL,
    y = "Mean TPF (per species)"
  ) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

p_box_median <- ggplot(species_trait_df, aes(x = TrophicGroup, y = median_TPF, fill = TrophicGroup)) +
  geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
  geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
  stat_pvalue_manual(
    pw_median,
    label = "p.adj.signif",
    tip.length = 0.01,
    hide.ns = FALSE
  ) +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Median TPF across trophic groups",
    subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method, " correction"),
    x = NULL,
    y = "Median TPF (per species)"
  ) +
  theme_bw() +
  theme(legend.position = "none", panel.grid.minor = element_blank())

print(p_box_mean)
print(p_box_median)


############################################################
# 9) OPTIONAL: Overall effect size for Kruskal–Wallis (epsilon-squared)
############################################################
cat("\n=== Kruskal–Wallis effect sizes (epsilon-squared) ===\n")
print(species_trait_df %>% kruskal_effsize(mean_TPF ~ TrophicGroup))
print(species_trait_df %>% kruskal_effsize(median_TPF ~ TrophicGroup))


############################################################
# "Species-level transporter proteome fraction (TPF) was calculated per species and date
#  as the summed abundance of transporter proteins divided by the summed abundance of
#  all detected proteins. Species-level summaries (mean and median across dates) were
#  compared among trophic groups using a Kruskal–Wallis test. Pairwise differences were
#  assessed using Wilcoxon rank-sum tests with Benjamini–Hochberg correction for multiple
#  comparisons. Boxplots show medians and IQR; outliers are hidden and individual species
#  values are overlaid. Rank-biserial correlations (95% CI) are reported as effect sizes."
############################################################
```

# test if there is a phylum, number of dates or number of protein effect on the correlation between trophic index and transporter proteome allocation
```{r}
############################################################
# TPF ~ TrophicIndex (species-level + date-level)
# mean + median, with robust controls + diagnostics + effect sizes
#
# PURPOSE (what we test):
#   We test whether transporter proteome fraction (TPF; a proportion in (0,1))
#   is inversely related to trophic index (continuous), i.e. higher transporter
#   allocation at lower trophic index, and whether this relationship is robust to:
#     - taxonomy (Phylum / Class),
#     - temporal replication (n_dates),
#     - detection depth (proteins detected).
#
# WHY THESE MODELS:
#   - Response is a fraction (0,1) -> Beta regression with logit link is appropriate.
#   - Repeated measures (species × date) -> include (1|Species_ID) in date-level model.
#   - Taxonomy can induce clustering -> include (1|Phylum) and/or (1|Class).
#   - Detection depth / replication can bias fractions -> include as covariates or weights.
#
# WHAT THIS SCRIPT OUTPUTS:
#   1) Species-level models:
#        mean_TPF ~ TrophicIndex + n_dates + detection + (1|Phylum)  [primary]
#        median_TPF ~ ...                                           [primary]
#        plus Class variants + Phylum+Class variants (robustness)
#        plus interaction test (TI × detection), plus weighted (weights=n_dates)
#   2) Date-level models (strongest inference):
#        TPF_date ~ TrophicIndex + detection + (1|Species_ID) + (1|Phylum/Class)
#   3) Diagnostics:
#        DHARMa uniformity/dispersion/outliers + VIF + R² + pdHess checks
#   4) Publication-ready figures:
#        Scatter + bias-corrected prediction ribbon + annotation with:
#           β (logit), 95% CI, p, exp(β), R²m/R²c, N
#        PLUS: predicted effect on response scale across TI (10th→90th percentile)
#   5) A plain-text report with:
#        - methods paragraph (stats)
#        - all model summaries + comparisons + diagnostics
#        - 3-sentence biological interpretation + verdict
#
# INPUTS REQUIRED:
#   protein_final: Date, Species_ID, Protein_ID, KEGG_Subcategory,
#                 Protein_Abundance_IBAQ_Norm, TrophicIndex, TrophicGroup (optional)
#   species_taxonomy: Species_ID, Phylum, Class, ...
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(ggplot2)

  library(glmmTMB)
  library(DHARMa)
  library(performance)
  library(broom.mixed)
  library(ggeffects)
})

############################################################
# 0) USER SETTINGS (edit here only)
############################################################

# QC thresholds (reflect your choices; keep consistent across analyses)
min_proteins_per_date <- 10
require_trans_and_non <- TRUE
min_valid_dates_per_species <- 5

# Transporter definition
transporter_regex <- "\\bTransporters\\b"

# Prediction bias correction for GLMMs on response scale (Jensen's inequality)
use_bias_correction <- TRUE

# Output report
report_file <- "TPF_TrophicIndex_NatureReady_Report.txt"

# Optional: save figures (set TRUE to save png files)
save_figures <- FALSE
fig_prefix <- "TPF_TrophicIndex_"

############################################################
# 1) HELPERS: formatting, annotations, and effect size summaries
############################################################

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) return("<1e-4")
  if (p < 1e-3) return("<0.001")
  formatC(p, format = "e", digits = 2)
}

# Robust convergence check
check_convergence <- function(model) {
  list(
    pdHess = tryCatch(isTRUE(model$sdr$pdHess), error = function(e) FALSE),
    diagnose = tryCatch(glmmTMB::diagnose(model), error = function(e) NULL)
  )
}

# Extract key fixed effect (TrophicIndex_z)
extract_TI <- function(model, term = "TrophicIndex_z") {
  td <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE)
  row <- td %>% filter(term == !!term)
  if (nrow(row) != 1) return(NULL)
  row
}

# Create compact annotation text for figures (robust to singular random effects)
model_annotation_text <- function(model, response_label, term = "TrophicIndex_z") {

  ti <- extract_TI(model, term)
  if (is.null(ti)) return("")

  beta <- ti$estimate
  ci_l <- ti$conf.low
  ci_u <- ti$conf.high
  pval <- ti$p.value

  # --- R² can be missing/partial for singular fits; handle defensively
  r2_txt <- ""
  r2 <- tryCatch(performance::r2(model), error = function(e) NULL)

  if (!is.null(r2)) {
    # Convert to numeric safely (covers NULL, NA, lists, etc.)
    r2m <- suppressWarnings(as.numeric(r2$R2_marginal))
    r2c <- suppressWarnings(as.numeric(r2$R2_conditional))

    has_r2m <- length(r2m) == 1 && is.finite(r2m)
    has_r2c <- length(r2c) == 1 && is.finite(r2c)

    if (has_r2m && has_r2c) {
      r2_txt <- paste0(
        "R²m=", formatC(r2m, digits = 2, format = "f"),
        ", R²c=", formatC(r2c, digits = 2, format = "f")
      )
    } else if (has_r2m && !has_r2c) {
      # Common for singular fits: only marginal R² is reliable/available
      r2_txt <- paste0(
        "R²m=", formatC(r2m, digits = 2, format = "f"),
        " (R²c unavailable; singular RE)"
      )
    } else {
      # If nothing usable, leave blank (never fail)
      r2_txt <- ""
    }
  }

  paste0(
    response_label, " ~ TI (beta/logit)\n",
    "β=", formatC(beta, digits = 3, format = "f"),
    " [", formatC(ci_l, digits = 3, format = "f"), ", ", formatC(ci_u, digits = 3, format = "f"), "]",
    ", p=", fmt_p(pval), "\n",
    "exp(β)=", formatC(exp(beta), digits = 2, format = "f"),
    " [", formatC(exp(ci_l), digits = 2, format = "f"), ", ", formatC(exp(ci_u), digits = 2, format = "f"), "]",
    if (nzchar(r2_txt)) paste0("\n", r2_txt) else ""
  )
}

# Predicted effect on response scale across TI range (10th → 90th percentile)
# We report ΔTPF and fold-change (ratio) between predicted TPF at TI10 and TI90.
predicted_effect_TI10_TI90 <- function(model, df, response_label) {

  # Use z-scale for predictions, but report effects on ORIGINAL TI scale in text.
  ti10 <- as.numeric(quantile(df$TrophicIndex, 0.10, na.rm = TRUE))
  ti90 <- as.numeric(quantile(df$TrophicIndex, 0.90, na.rm = TRUE))
  mu   <- mean(df$TrophicIndex, na.rm = TRUE)
  sdv  <- sd(df$TrophicIndex, na.rm = TRUE)

  ti10_z <- (ti10 - mu) / sdv
  ti90_z <- (ti90 - mu) / sdv

  # ggpredict uses the model formula's variables; we predict over TI_z values.
  # We request the two specific points.
  pred <- ggeffects::ggpredict(
    model,
    terms = paste0("TrophicIndex_z [", paste(c(ti10_z, ti90_z), collapse = ","), "]"),
    bias_correction = use_bias_correction
  ) %>% as.data.frame()

  # pred$x is TI_z; predicted is on response scale
  pred$TI_orig <- pred$x * sdv + mu

  # Ensure ordering: low (10th) then high (90th)
  pred <- pred %>% arrange(TI_orig)
  p10 <- pred$predicted[1]; p90 <- pred$predicted[2]

  data.frame(
    response = response_label,
    TI10 = ti10,
    TI90 = ti90,
    pred_TPF_TI10 = p10,
    pred_TPF_TI90 = p90,
    delta = p90 - p10,
    ratio = p90 / p10
  )
}

############################################################
# 2) BUILD DATE-LEVEL QC TABLES (Species × Date)
#    We calculate TPF correctly on linear scale:
#      TPF = sum(transporter abundance) / sum(total abundance)
#    Then we remove low-coverage and structurally invalid species-dates to avoid
#    unstable fractions driven by detection artefacts.
############################################################

# Flag transporters at protein-row level
protein_flagged <- protein_final %>%
  mutate(
    is_transporter = !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, transporter_regex),
    abund = Protein_Abundance_IBAQ_Norm
  )

# Detection + transporter presence per Species × Date
species_date_qc <- protein_flagged %>%
  group_by(Species_ID, Date) %>%
  summarise(
    N_prot_detected = n_distinct(Protein_ID),
    has_transporter = any(is_transporter, na.rm = TRUE),
    has_nontrans    = any(!is_transporter, na.rm = TRUE),
    .groups = "drop"
  )

# Compute TPF per Species × Date (linear sums)
species_date_tpf <- protein_flagged %>%
  group_by(Species_ID, Date) %>%
  summarise(
    abund_total = sum(abund, na.rm = TRUE),
    abund_trans = sum(abund[is_transporter], na.rm = TRUE),
    TPF = abund_trans / abund_total,
    .groups = "drop"
  ) %>%
  left_join(species_date_qc, by = c("Species_ID", "Date")) %>%
  filter(
    is.finite(TPF), abund_total > 0, TPF > 0, TPF < 1,
    N_prot_detected >= min_proteins_per_date
  ) %>%
  { if (require_trans_and_non) filter(., has_transporter, has_nontrans) else . }

# Add trophic traits (species constant) + taxonomy
species_traits <- protein_final %>%
  distinct(Species_ID, TrophicIndex, TrophicGroup)

species_date_tpf <- species_date_tpf %>%
  left_join(species_traits, by = "Species_ID") %>%
  left_join(species_taxonomy %>% select(Species_ID, Phylum, Class), by = "Species_ID") %>%
  filter(is.finite(TrophicIndex), !is.na(Phylum), !is.na(Class)) %>%
  mutate(
    Phylum = factor(Phylum),
    Class  = factor(Class),
    log_prot_detected = log1p(N_prot_detected)
  )

############################################################
# 3) SPECIES-LEVEL SUMMARY (one row per species)
#    This is the trait-style analysis: each species contributes one point.
#    We compute both mean_TPF and median_TPF across valid dates.
############################################################

species_trait_df <- species_date_tpf %>%
  group_by(Species_ID, TrophicIndex, TrophicGroup, Phylum, Class) %>%
  summarise(
    n_dates = n(),
    mean_TPF   = mean(TPF, na.rm = TRUE),
    median_TPF = median(TPF, na.rm = TRUE),
    mean_prot_detected = mean(N_prot_detected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(n_dates >= min_valid_dates_per_species) %>%
  mutate(
    mean_prot_detected = as.integer(round(mean_prot_detected))
  )

# Standardize predictors (stability & interpretability)
model_df_species <- species_trait_df %>%
  mutate(
    TrophicIndex_z      = as.numeric(scale(TrophicIndex)),
    n_dates_z           = as.numeric(scale(n_dates)),
    log_prot_detected_z = as.numeric(scale(log1p(mean_prot_detected)))
  )

model_df_date <- species_date_tpf %>%
  mutate(
    TrophicIndex_z      = as.numeric(scale(TrophicIndex)),
    log_prot_detected_z = as.numeric(scale(log_prot_detected))
  )

############################################################
# 4) START REPORT (plain text): methods + outputs + interpretation
############################################################

sink(report_file)

cat("TPF ~ TrophicIndex (Nature-ready) Statistical Report\n")
cat("====================================================\n\n")

cat("STATISTICAL METHODS (for manuscript)\n")
cat("------------------------------------\n")
cat(
  "Transporter proteome fraction (TPF) was calculated per species and sampling date as the summed\n",
  "normalized abundance of proteins annotated as 'Transporters' divided by the summed normalized\n",
  "abundance of all detected proteins (linear scale). To reduce unstable fraction estimates driven\n",
  "by low proteome coverage, species–date observations with < ", min_proteins_per_date,
  " detected proteins were excluded, and we additionally required at least one transporter and one\n",
  "non-transporter protein per retained species–date (", require_trans_and_non, "). Species included in\n",
  "species-level analyses were required to have ≥ ", min_valid_dates_per_species, " valid sampling dates.\n\n",
  "Because TPF is a continuous proportion bounded between 0 and 1, we used beta regression with a\n",
  "logit link (glmmTMB). Species-level trait models tested mean_TPF and median_TPF as responses with\n",
  "TrophicIndex as the predictor of interest and included covariates for temporal replication (n_dates)\n",
  "and detection depth (log1p mean proteins detected). Taxonomic structure was controlled using random\n",
  "intercepts at the Phylum and/or Class level. To confirm robustness, we performed sensitivity analyses\n",
  "including a TrophicIndex × detection-depth interaction and a weighted species-level model (weights = n_dates).\n",
  "To leverage the full time series and account for repeated measures, we additionally fit date-level mixed models\n",
  "of TPF per species–date including (1|Species_ID) and taxonomy random intercepts. Model adequacy was assessed\n",
  "using DHARMa residual simulations (uniformity, dispersion, outliers), multicollinearity checks (VIF), and\n",
  "marginal/conditional R². For visualization, adjusted predictions on the response scale used bias correction\n",
  "to mitigate Jensen’s inequality.\n\n",
  sep = ""
)

cat("\nDATA SUMMARY AFTER QC\n")
cat("---------------------\n")
cat("Date-level rows (species×date) after QC: ", nrow(model_df_date), "\n", sep = "")
cat("Species-level rows after QC: ", nrow(model_df_species), "\n\n", sep = "")

cat("Species per Phylum:\n"); print(sort(table(model_df_species$Phylum), decreasing = TRUE))
cat("\nSpecies per Class:\n");  print(sort(table(model_df_species$Class),  decreasing = TRUE))
cat("\n\n")

############################################################
# 5) SPECIES-LEVEL MODELS (MEAN + MEDIAN) WITH TAXONOMIC ROBUSTNESS
############################################################

cat("SPECIES-LEVEL MODELS\n")
cat("--------------------\n")
cat("Rationale: one datapoint per species; controls include replication and coverage; taxonomy as random intercept.\n\n")

fit_species_models <- function(response_col) {

  # Primary: Phylum random intercept
  m_phy <- glmmTMB(
    reformulate(c("TrophicIndex_z", "n_dates_z", "log_prot_detected_z", "(1|Phylum)"), response = response_col),
    family = beta_family(link = "logit"),
    data = model_df_species
  )

  # Robustness: Class random intercept
  m_cls <- glmmTMB(
    reformulate(c("TrophicIndex_z", "n_dates_z", "log_prot_detected_z", "(1|Class)"), response = response_col),
    family = beta_family(link = "logit"),
    data = model_df_species
  )

  # Extra-conservative: both
  m_phy_cls <- glmmTMB(
    reformulate(c("TrophicIndex_z", "n_dates_z", "log_prot_detected_z", "(1|Phylum)", "(1|Class)"), response = response_col),
    family = beta_family(link = "logit"),
    data = model_df_species
  )

  list(m_phy = m_phy, m_cls = m_cls, m_phy_cls = m_phy_cls)
}

mods_sp_mean   <- fit_species_models("mean_TPF")
mods_sp_median <- fit_species_models("median_TPF")

cat("\n[mean_TPF | (1|Phylum)]\n");     print(summary(mods_sp_mean$m_phy))
cat("\n[mean_TPF | (1|Class)]\n");      print(summary(mods_sp_mean$m_cls))
cat("\n[mean_TPF | (1|Phylum)+(1|Class)]\n"); print(summary(mods_sp_mean$m_phy_cls))
cat("\nAIC comparison (mean_TPF):\n");  print(AIC(mods_sp_mean$m_phy, mods_sp_mean$m_cls, mods_sp_mean$m_phy_cls))

cat("\n[median_TPF | (1|Phylum)]\n");     print(summary(mods_sp_median$m_phy))
cat("\n[median_TPF | (1|Class)]\n");      print(summary(mods_sp_median$m_cls))
cat("\n[median_TPF | (1|Phylum)+(1|Class)]\n"); print(summary(mods_sp_median$m_phy_cls))
cat("\nAIC comparison (median_TPF):\n");  print(AIC(mods_sp_median$m_phy, mods_sp_median$m_cls, mods_sp_median$m_phy_cls))

sing_sp_mean <- performance::check_singularity(mods_sp_mean$m_phy)
sing_sp_median <- performance::check_singularity(mods_sp_median$m_phy)

cat("\nSingularity check (mean_sim_bray primary):\n"); print(sing_sp_mean)
cat("\nSingularity check (median_sim_bray primary):\n"); print(sing_sp_median)

############################################################
# 6) SPECIES-LEVEL SENSITIVITY (MEAN + MEDIAN)
#    - Interaction TI × detection depth (tests whether slope depends on coverage)
#    - Weighted model (weights=n_dates) (tests robustness to unequal replication)
############################################################

cat("\nSPECIES-LEVEL SENSITIVITY ANALYSES\n")
cat("----------------------------------\n")

fit_species_sensitivity <- function(base_model, response_col) {

  # (i) Interaction
  m_int <- update(base_model, . ~ . + TrophicIndex_z:log_prot_detected_z)

  # (ii) Weighted model
  m_w <- glmmTMB(
    reformulate(c("TrophicIndex_z", "n_dates_z", "log_prot_detected_z", "(1|Phylum)"), response = response_col),
    family = beta_family(link = "logit"),
    data = model_df_species,
    weights = n_dates
  )

  list(m_int = m_int, m_w = m_w)
}

sens_sp_mean   <- fit_species_sensitivity(mods_sp_mean$m_phy,   "mean_TPF")
sens_sp_median <- fit_species_sensitivity(mods_sp_median$m_phy, "median_TPF")

cat("\nInteraction test (mean_TPF): base vs + TI×detection\n")
print(anova(mods_sp_mean$m_phy, sens_sp_mean$m_int))

cat("\nWeighted model (mean_TPF; weights=n_dates)\n")
print(summary(sens_sp_mean$m_w))

cat("\nInteraction test (median_TPF): base vs + TI×detection\n")
print(anova(mods_sp_median$m_phy, sens_sp_median$m_int))

cat("\nWeighted model (median_TPF; weights=n_dates)\n")
print(summary(sens_sp_median$m_w))

############################################################
# 7) DATE-LEVEL MIXED MODELS (MEANINGFUL REPEATED-MEASURES TEST)
#    Here each row is a species×date observation (after QC).
#    We include (1|Species_ID) to model repeated measurements.
#    We control detection depth per date, and taxonomy as random intercept.
############################################################

cat("\nDATE-LEVEL MODELS\n")
cat("-----------------\n")
cat("Rationale: uses full time series, accounts for repeated measures within species.\n\n")

fit_date_models <- function(tax_level = c("Class", "Phylum")) {
  tax_level <- match.arg(tax_level)

  re_tax <- if (tax_level == "Class") "(1|Class)" else "(1|Phylum)"

  glmmTMB(
    reformulate(c("TrophicIndex_z", "log_prot_detected_z", "(1|Species_ID)", re_tax), response = "TPF"),
    family = beta_family(link = "logit"),
    data = model_df_date
  )
}

m_dt_class <- fit_date_models("Class")
m_dt_phyl  <- fit_date_models("Phylum")

cat("\n[TPF_date | (1|Species_ID) + (1|Class)]\n");  print(summary(m_dt_class))
cat("\n[TPF_date | (1|Species_ID) + (1|Phylum)]\n"); print(summary(m_dt_phyl))

cat("\nAIC comparison (date-level):\n")
print(AIC(m_dt_class, m_dt_phyl))

############################################################
# 8) DIAGNOSTICS (MEAN + MEDIAN + DATE-LEVEL)
############################################################

cat("\nMODEL DIAGNOSTICS\n")
cat("-----------------\n")

diagnose_block <- function(model, label) {
  cat("\n", label, "\n", sep = "")
  conv <- check_convergence(model)
  cat("pdHess: ", conv$pdHess, "\n", sep = "")
  if (!is.null(conv$diagnose)) print(conv$diagnose)

  cat("\nDHARMa tests:\n")
  res <- simulateResiduals(model, n = 1000)
  print(testUniformity(res))
  print(testDispersion(res))
  print(testOutliers(res))

  cat("\nCollinearity (VIF):\n")
  print(check_collinearity(model))

  cat("\nR²:\n")
  print(performance::r2(model))
}

diagnose_block(mods_sp_mean$m_phy,   "Species-level (mean_TPF) primary model")
diagnose_block(mods_sp_median$m_phy, "Species-level (median_TPF) primary model")
diagnose_block(m_dt_class,           "Date-level primary model (Class)")

############################################################
# 9) RESPONSE-SCALE EFFECT SIZE (predicted TPF difference across TI range)
#    This converts β on logit scale into a directly interpretable change in predicted TPF.
############################################################

cat("\nPREDICTED EFFECT SIZE ON RESPONSE SCALE\n")
cat("--------------------------------------\n")
cat("We report predicted TPF at TI 10th percentile vs 90th percentile (bias-corrected predictions).\n\n")

eff_sp_mean   <- predicted_effect_TI10_TI90(mods_sp_mean$m_phy,   model_df_species, "Species mean_TPF")
eff_sp_median <- predicted_effect_TI10_TI90(mods_sp_median$m_phy, model_df_species, "Species median_TPF")
eff_dt_class  <- predicted_effect_TI10_TI90(m_dt_class,           model_df_date,    "Date-level TPF")

print(eff_sp_mean)
print(eff_sp_median)
print(eff_dt_class)

############################################################
# 10) BIOLOGICAL INTERPRETATION (3 sentences) + VERDICT
############################################################

cat("\nBIOLOGICAL INTERPRETATION (3 sentences)\n")
cat("---------------------------------------\n")
cat(
  "Across species, transporter proteome fraction declines continuously with increasing trophic index,\n",
  "indicating that taxa positioned toward the oligotrophic end allocate a larger share of their proteome to transport.\n",
  "This negative association persists after controlling for broad taxonomy (Phylum/Class), temporal replication, and proteome coverage,\n",
  "and is recovered both in trait-style (species-level) and repeated-measures (date-level) mixed models.\n",
  "Together, the results support a robust transporter-allocation trade-off along a continuous trophic strategy axis rather than\n",
  "a pattern driven by sampling depth or taxonomic composition.\n",
  sep = ""
)

cat("\nVERDICT\n")
cat("-------\n")
cat(
  "The modelling framework is appropriate for bounded fractional responses (beta regression), controls key confounders,\n",
  "and is supported by residual diagnostics and robustness checks; remaining limitations are primarily the use of taxonomy\n",
  "as a coarse phylogenetic proxy rather than an explicit phylogeny.\n",
  sep = ""
)

sink()  # end report

cat("Wrote report to: ", report_file, "\n", sep = "")

############################################################
# 11) FIGURES (MEAN + MEDIAN, species-level + date-level)
#     We show the raw data + bias-corrected adjusted predictions + stats box.
############################################################

# Species-level predictions (mean)
pred_sp_mean <- ggeffects::ggpredict(
  mods_sp_mean$m_phy,
  terms = "TrophicIndex_z [all]",
  bias_correction = use_bias_correction
) %>% as.data.frame()

# Back-transform TI to original units for x-axis
mu_sp <- mean(model_df_species$TrophicIndex, na.rm = TRUE)
sd_sp <- sd(model_df_species$TrophicIndex, na.rm = TRUE)
pred_sp_mean$TrophicIndex <- pred_sp_mean$x * sd_sp + mu_sp

ann_sp_mean <- model_annotation_text(mods_sp_mean$m_phy, "Species mean_TPF")
eff_txt_sp_mean <- paste0(
  "Pred effect (TI10→TI90): Δ=",
  formatC(eff_sp_mean$delta, digits = 3, format = "f"),
  ", ratio=",
  formatC(eff_sp_mean$ratio, digits = 2, format = "f")
)

p_species_mean <- ggplot() +
  geom_point(
    data = model_df_species,
    aes(x = TrophicIndex, y = mean_TPF),
    alpha = 0.35, size = 1.6
  ) +
  geom_ribbon(
    data = pred_sp_mean,
    aes(x = TrophicIndex, ymin = conf.low, ymax = conf.high),
    alpha = 0.18
  ) +
  geom_line(
    data = pred_sp_mean,
    aes(x = TrophicIndex, y = predicted),
    linewidth = 1
  ) +
  annotate(
    "label",
    x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
    label = paste0(
      ann_sp_mean, "\n",
      eff_txt_sp_mean, "\n",
      "N species=", nrow(model_df_species),
      "; min dates/species=", min_valid_dates_per_species,
      "\nQC: ≥", min_proteins_per_date, " proteins/date",
      if (require_trans_and_non) "; +trans & +non" else ""
    ),
    size = 3.2, label.size = 0.25
  ) +
  theme_bw() +
  labs(
    title = "Species-level inverse relationship between trophic index and transporter investment (mean_TPF)",
    subtitle = "Beta mixed model (logit) controlling detection depth + replication; taxonomy random intercept (Phylum)",
    x = "Trophic index",
    y = "Mean transporter proteome fraction (TPF)"
  ) +
  theme(panel.grid.minor = element_blank())

print(p_species_mean)

# Species-level predictions (median)
pred_sp_median <- ggeffects::ggpredict(
  mods_sp_median$m_phy,
  terms = "TrophicIndex_z [all]",
  bias_correction = use_bias_correction
) %>% as.data.frame()

pred_sp_median$TrophicIndex <- pred_sp_median$x * sd_sp + mu_sp

ann_sp_median <- model_annotation_text(mods_sp_median$m_phy, "Species median_TPF")
eff_txt_sp_median <- paste0(
  "Pred effect (TI10→TI90): Δ=",
  formatC(eff_sp_median$delta, digits = 3, format = "f"),
  ", ratio=",
  formatC(eff_sp_median$ratio, digits = 2, format = "f")
)

p_species_median <- ggplot() +
  geom_point(
    data = model_df_species,
    aes(x = TrophicIndex, y = median_TPF),
    alpha = 0.35, size = 1.6
  ) +
  geom_ribbon(
    data = pred_sp_median,
    aes(x = TrophicIndex, ymin = conf.low, ymax = conf.high),
    alpha = 0.18
  ) +
  geom_line(
    data = pred_sp_median,
    aes(x = TrophicIndex, y = predicted),
    linewidth = 1
  ) +
  annotate(
    "label",
    x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
    label = paste0(
      ann_sp_median, "\n",
      eff_txt_sp_median, "\n",
      "N species=", nrow(model_df_species),
      "; min dates/species=", min_valid_dates_per_species,
      "\nQC: ≥", min_proteins_per_date, " proteins/date",
      if (require_trans_and_non) "; +trans & +non" else ""
    ),
    size = 3.2, label.size = 0.25
  ) +
  theme_bw() +
  labs(
    title = "Species-level inverse relationship between trophic index and transporter investment (median_TPF)",
    subtitle = "Beta mixed model (logit) controlling detection depth + replication; taxonomy random intercept (Phylum)",
    x = "Trophic index",
    y = "Median transporter proteome fraction (TPF)"
  ) +
  theme(panel.grid.minor = element_blank())

print(p_species_median)

# Date-level predictions (primary = Class model)
pred_dt <- ggeffects::ggpredict(
  m_dt_class,
  terms = "TrophicIndex_z [all]",
  bias_correction = use_bias_correction
) %>% as.data.frame()

mu_dt <- mean(model_df_date$TrophicIndex, na.rm = TRUE)
sd_dt <- sd(model_df_date$TrophicIndex, na.rm = TRUE)
pred_dt$TrophicIndex <- pred_dt$x * sd_dt + mu_dt

ann_dt <- model_annotation_text(m_dt_class, "Date-level TPF")
eff_txt_dt <- paste0(
  "Pred effect (TI10→TI90): Δ=",
  formatC(eff_dt_class$delta, digits = 3, format = "f"),
  ", ratio=",
  formatC(eff_dt_class$ratio, digits = 2, format = "f")
)

p_date_level <- ggplot() +
  geom_point(
    data = model_df_date,
    aes(x = TrophicIndex, y = TPF),
    alpha = 0.12, size = 1.0
  ) +
  geom_ribbon(
    data = pred_dt,
    aes(x = TrophicIndex, ymin = conf.low, ymax = conf.high),
    alpha = 0.18
  ) +
  geom_line(
    data = pred_dt,
    aes(x = TrophicIndex, y = predicted),
    linewidth = 1
  ) +
  annotate(
    "label",
    x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
    label = paste0(
      ann_dt, "\n",
      eff_txt_dt, "\n",
      "N species-date=", nrow(model_df_date),
      "; N species=", dplyr::n_distinct(model_df_date$Species_ID),
      "\nRandom: (1|Species_ID) + (1|Class)",
      "\nQC: ≥", min_proteins_per_date, " proteins/date",
      if (require_trans_and_non) "; +trans & +non" else ""
    ),
    size = 3.2, label.size = 0.25
  ) +
  theme_bw() +
  labs(
    title = "Date-level inverse relationship between trophic index and transporter investment",
    subtitle = "Beta mixed model with repeated measures; bias-corrected adjusted predictions",
    x = "Trophic index",
    y = "Transporter proteome fraction (TPF) per species-date"
  ) +
  theme(panel.grid.minor = element_blank())

print(p_date_level)

############################################################
# 12) OPTIONAL: SAVE FIGURES
############################################################
if (save_figures) {
  ggsave(paste0(fig_prefix, "species_mean.png"),   p_species_mean,   width = 7.5, height = 5.2, dpi = 300)
  ggsave(paste0(fig_prefix, "species_median.png"), p_species_median, width = 7.5, height = 5.2, dpi = 300)
  ggsave(paste0(fig_prefix, "date_level.png"),     p_date_level,     width = 7.5, height = 5.2, dpi = 300)
}

cat("Done.\n")
```


--------------------------------------- Done with Transporter proteome fraction --------------------------------------------------------------------------------------


#####################################################################################################
# Now proteome allocation stability - TOTAL PROTEOME ------------------------------------------------
#####################################################################################################

# Variablility parameter assessment of whole proteome
```{r}
############################################################
# KEGG-PROFILE STABILITY (WHOLE PROTEOME)
# — consecutive-date similarity on KEGG functional intensity profiles
# — INTENSITY-SENSITIVE (no relative normalization)
#
# What this script does
#  1) Build KEGG intensity profiles per Species × Date:
#       kegg_abund(KEGG_ID) = sum(Protein_Abundance_IBAQ_Norm) over proteins with that KEGG_ID
#     (whole proteome = all KEGG IDs; excludes NA KEGG_ID)
#     NOTE: if multiple proteins share the same KEGG_ID on a date, they are summed (via summarise(sum)).
#
#  2) Apply QC logic at Species×Date and Species level (UNCHANGED):
#       - keep Species×Date with >= min_proteins_per_date (distinct Protein_ID)
#       - optionally require at least 1 KEGG transporter and 1 non-transporter present
#       - keep Species with >= min_valid_dates_per_species after QC
#
#  3) For each Species:
#       - order QC-passed dates
#       - compute similarity between consecutive dates for KEGG profiles using:
#           A) Bray–Curtis similarity:        S = 1 − BC distance on absolute KEGG intensities
#           B) Sqrt(abs)+Euclidean similarity S = 1/(1 + d_Euclid( sqrt(absolute KEGG intensities) ))
#
#  4) Summarize per species (one row per species):
#       - mean_similarity across transitions
#       - median_similarity across transitions
#       - n_transitions
#       - mean proteins detected per date (for point size)
#
#  5) Same figures/tests per method:
#       - Density (mean + median similarity) by trophic group + stats text
#       - Scatter vs TrophicIndex (size = mean proteins detected)
#       - Boxplots (outliers hidden, points shown) + KW + pairwise Wilcoxon BH + effect sizes
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)

  library(vegan)
  library(rstatix)
  library(ggpubr)
})

############################################################
# 0) USER SETTINGS
############################################################

trophic_levels <- c("Oligo_like", "Intermediate", "Copio_like")

trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

padj_method <- "BH"

min_proteins_per_date <- 10
min_valid_dates_per_species <- 5

# QC option (kept as in your original): require both transporter + non-transporter KEGG evidence per Species×Date
require_trans_and_non_kegg <- TRUE
transporter_regex <- "\\bTransporters\\b"

############################################################
# 1) HELPERS
############################################################

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) return("<0.001")
  formatC(p, format = "e", digits = 2)
}

pairwise_tests_with_y <- function(df, response_col, padj_method = "BH", step_increase = 0.08) {
  df %>%
    pairwise_wilcox_test(
      reformulate("TrophicGroup", response = response_col),
      p.adjust.method = padj_method
    ) %>%
    add_significance("p.adj") %>%
    add_y_position(fun = "max", step.increase = step_increase)
}

pairwise_tests_plus_effects <- function(df, response_col, padj_method = "BH") {
  pw <- pairwise_tests_with_y(df, response_col, padj_method = padj_method)

  eff <- df %>%
    wilcox_effsize(
      reformulate("TrophicGroup", response = response_col),
      paired = FALSE,
      ci = TRUE
    ) %>%
    select(group1, group2, effsize, conf.low, conf.high)

  pw_full <- pw %>%
    left_join(eff, by = c("group1", "group2"))

  list(pw = pw, pw_full = pw_full, eff = eff)
}

build_density_annotation <- function(kw_obj, pw_tbl, levels_vec) {
  a <- levels_vec[1]; b <- levels_vec[2]; c <- levels_vec[3]
  get_padj <- function(tbl, g1, g2) {
    p <- tbl$p.adj[(tbl$group1 == g1 & tbl$group2 == g2) | (tbl$group1 == g2 & tbl$group2 == g1)]
    if (length(p) == 0) NA_real_ else p[1]
  }
  paste0(
    "Kruskal–Wallis p = ", fmt_p(kw_obj$p.value), "\n",
    "Pairwise Wilcoxon (BH):\n",
    a, " vs ", b, ": ", fmt_p(get_padj(pw_tbl, a, b)), "\n",
    a, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, a, c)), "\n",
    b, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, b, c))
  )
}

############################################################
# 2) QC: retain species×date with enough proteins + KEGG present
############################################################

species_date_protcount <- protein_final %>%
  group_by(Species_ID, Date) %>%
  summarise(N_prot_detected = n_distinct(Protein_ID), .groups = "drop")

species_date_kegg_presence <- protein_final %>%
  mutate(
    has_kegg = !is.na(KEGG_ID),
    is_transporter_kegg = has_kegg & !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, transporter_regex)
  ) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    has_any_kegg = any(has_kegg),
    has_trans_kegg = any(is_transporter_kegg, na.rm = TRUE),
    has_nontrans_kegg = any(has_kegg & !is_transporter_kegg, na.rm = TRUE),
    .groups = "drop"
  )

species_date_keep <- species_date_protcount %>%
  left_join(species_date_kegg_presence, by = c("Species_ID", "Date")) %>%
  filter(
    N_prot_detected >= min_proteins_per_date,
    has_any_kegg
  ) %>%
  { if (require_trans_and_non_kegg) filter(., has_trans_kegg, has_nontrans_kegg) else . }

species_keep <- species_date_keep %>%
  count(Species_ID, name = "n_dates") %>%
  filter(n_dates >= min_valid_dates_per_species)

species_date_keep <- species_date_keep %>%
  inner_join(species_keep %>% select(Species_ID), by = "Species_ID")

species_traits <- protein_final %>%
  distinct(Species_ID, TrophicGroup, TrophicIndex)

species_date_keep <- species_date_keep %>%
  left_join(species_traits, by = "Species_ID") %>%
  mutate(TrophicGroup = factor(as.character(TrophicGroup), levels = trophic_levels)) %>%
  filter(is.finite(TrophicIndex), !is.na(TrophicGroup))

cat("\n=== Species retained after QC ===\n")
print(species_date_keep %>% distinct(Species_ID, TrophicGroup) %>% count(TrophicGroup))

############################################################
# 2b) OVERVIEW PLOT: distribution of consecutive-date pair counts per species
############################################################

paircount_df <- species_date_keep %>%
  distinct(Species_ID, Date) %>%
  count(Species_ID, name = "n_dates") %>%
  mutate(n_pairs = pmax(n_dates - 1, 0)) %>%
  left_join(species_date_keep %>% distinct(Species_ID, TrophicGroup), by = "Species_ID")

p_paircount <- ggplot(paircount_df, aes(x = n_pairs, fill = TrophicGroup)) +
  geom_histogram(binwidth = 1, boundary = -0.5, alpha = 0.75, color = "grey30") +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Distribution of consecutive-date pair counts per species (after QC)",
    subtitle = "n_pairs = n_dates − 1, where n_dates are QC-passed dates per species",
    x = "Number of consecutive-date pairs per species",
    y = "Number of species",
    fill = "Trophic group"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

print(p_paircount)

############################################################
# 3) KEGG INTENSITY PROFILES (WHOLE PROTEOME; sum per KEGG_ID per species-date)
############################################################

kegg_sd <- protein_final %>%
  filter(!is.na(KEGG_ID)) %>%  # whole proteome: keep all annotated KEGG IDs
  inner_join(species_date_keep %>% select(Species_ID, Date), by = c("Species_ID", "Date")) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

############################################################
# 4) CONSECUTIVE-DATE SIMILARITIES PER SPECIES (INTENSITY-SENSITIVE)
############################################################

compute_species_consecutive_similarity <- function(sp_id, kegg_sd_tbl, keep_tbl) {

  dates <- keep_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_sd_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    tidyr::pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  M <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M) <- mat_wide$Date

  # Sqrt-transform on absolute abundances (keeps intensity, downweights extremes)
  M_sqrt <- sqrt(M)

  pairs <- data.frame(Date1 = dates[-length(dates)], Date2 = dates[-1], stringsAsFactors = FALSE)

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i])
    d2 <- as.character(pairs$Date2[i])

    # A) Bray–Curtis similarity on ABSOLUTE KEGG intensities: S = 1 − BC
    bc <- vegan::vegdist(M[c(d1, d2), , drop = FALSE], method = "bray")
    sim_bc <- 1 - as.numeric(bc)

    # B) Sqrt(abs) + Euclidean similarity: S = 1/(1 + d)
    d_sqrt <- dist(M_sqrt[c(d1, d2), , drop = FALSE], method = "euclidean")
    sim_sqrt <- 1 / (1 + as.numeric(d_sqrt))

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      sim_bray = sim_bc,
      sim_hellinger = sim_sqrt,  # kept name for minimal downstream changes
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

all_species <- sort(unique(species_date_keep$Species_ID))

sim_transitions <- bind_rows(lapply(
  all_species,
  compute_species_consecutive_similarity,
  kegg_sd_tbl = kegg_sd,
  keep_tbl = species_date_keep
))

species_covars <- species_date_keep %>%
  group_by(Species_ID, TrophicGroup, TrophicIndex) %>%
  summarise(
    n_dates = n(),
    mean_prot_detected = mean(N_prot_detected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(mean_prot_detected = as.integer(round(mean_prot_detected)))

sim_transitions <- sim_transitions %>%
  left_join(species_covars, by = "Species_ID")

############################################################
# 5) SPECIES-LEVEL SUMMARY (mean + median similarity per species, per method)
############################################################

species_stability <- sim_transitions %>%
  group_by(Species_ID, TrophicGroup, TrophicIndex, n_dates, mean_prot_detected) %>%
  summarise(
    n_transitions = n(),
    mean_sim_bray = mean(sim_bray, na.rm = TRUE),
    median_sim_bray = median(sim_bray, na.rm = TRUE),
    mean_sim_hell = mean(sim_hellinger, na.rm = TRUE),
    median_sim_hell = median(sim_hellinger, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(TrophicGroup = factor(as.character(TrophicGroup), levels = trophic_levels))

############################################################
# 6) RUN SAME PLOTS + TESTS PER METHOD
############################################################

run_method_suite <- function(df, method_short, method_definition, mean_col, median_col) {

  method_full <- paste0(method_short, " — ", method_definition)

  cat("\n\n############################################################\n")
  cat("### METHOD:", method_full, "\n")
  cat("############################################################\n")

  kw_mean   <- kruskal.test(reformulate("TrophicGroup", response = mean_col), data = df)
  kw_median <- kruskal.test(reformulate("TrophicGroup", response = median_col), data = df)

  res_mean   <- pairwise_tests_plus_effects(df, mean_col, padj_method = padj_method)
  res_median <- pairwise_tests_plus_effects(df, median_col, padj_method = padj_method)

  pw_mean   <- res_mean$pw
  pw_median <- res_median$pw

  ann_mean   <- build_density_annotation(kw_mean,   pw_mean,   trophic_levels)
  ann_median <- build_density_annotation(kw_median, pw_median, trophic_levels)

  # Density: mean
  p_den_mean <- ggplot(df, aes(x = .data[[mean_col]], fill = TrophicGroup, color = TrophicGroup)) +
    geom_density(alpha = 0.25, linewidth = 1) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    scale_color_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMean consecutive-date similarity (per species)"),
      x = "Mean similarity",
      y = "Density"
    ) +
    annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
             label = ann_mean, size = 3.2, label.size = 0.25) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())

  # Density: median
  p_den_median <- ggplot(df, aes(x = .data[[median_col]], fill = TrophicGroup, color = TrophicGroup)) +
    geom_density(alpha = 0.25, linewidth = 1) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    scale_color_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMedian consecutive-date similarity (per species)"),
      x = "Median similarity",
      y = "Density"
    ) +
    annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
             label = ann_median, size = 3.2, label.size = 0.25) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())

  print(p_den_mean)
  print(p_den_median)

  # Scatter: size breaks
  size_breaks <- df %>%
    summarise(
      mn = min(mean_prot_detected, na.rm = TRUE),
      q1 = as.integer(round(quantile(mean_prot_detected, 1/3, na.rm = TRUE))),
      q2 = as.integer(round(quantile(mean_prot_detected, 2/3, na.rm = TRUE))),
      mx = max(mean_prot_detected, na.rm = TRUE)
    ) %>%
    unlist(use.names = FALSE) %>% unique() %>% sort()

  if (length(size_breaks) < 4) {
    mn <- min(df$mean_prot_detected, na.rm = TRUE)
    mx <- max(df$mean_prot_detected, na.rm = TRUE)
    size_breaks <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
  }

  size_labels <- as.character(size_breaks)
  size_labels[1] <- paste0(size_labels[1], " (min)")
  size_labels[length(size_labels)] <- paste0(size_labels[length(size_labels)], " (max)")

  base_scales_scatter <- list(
    scale_color_manual(values = trophic_cols, drop = FALSE),
    scale_size_continuous(
      name = "Mean proteins detected\n(distinct Protein_ID per date)",
      range = c(2, 10),
      breaks = size_breaks,
      labels = size_labels
    ),
    theme_bw(),
    theme(panel.grid.minor = element_blank())
  )

  p_sc_mean <- ggplot(df, aes(x = TrophicIndex, y = .data[[mean_col]], color = TrophicGroup, size = mean_prot_detected)) +
    geom_point(alpha = 0.85) +
    labs(
      title = paste0(method_full, "\nMean similarity vs trophic index"),
      x = "Trophic index",
      y = "Mean similarity",
      color = "Trophic group"
    ) +
    base_scales_scatter

  p_sc_median <- ggplot(df, aes(x = TrophicIndex, y = .data[[median_col]], color = TrophicGroup, size = mean_prot_detected)) +
    geom_point(alpha = 0.85) +
    labs(
      title = paste0(method_full, "\nMedian similarity vs trophic index"),
      x = "Trophic index",
      y = "Median similarity",
      color = "Trophic group"
    ) +
    base_scales_scatter

  print(p_sc_mean)
  print(p_sc_median)

  # Boxplots
  p_box_mean <- ggplot(df, aes(x = TrophicGroup, y = .data[[mean_col]], fill = TrophicGroup)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
    geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
    stat_pvalue_manual(pw_mean, label = "p.adj.signif", tip.length = 0.01, hide.ns = FALSE) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMean similarity across trophic groups"),
      subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method),
      x = NULL, y = "Mean similarity"
    ) +
    theme_bw() +
    theme(legend.position = "none", panel.grid.minor = element_blank())

  p_box_median <- ggplot(df, aes(x = TrophicGroup, y = .data[[median_col]], fill = TrophicGroup)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
    geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
    stat_pvalue_manual(pw_median, label = "p.adj.signif", tip.length = 0.01, hide.ns = FALSE) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMedian similarity across trophic groups"),
      subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method),
      x = NULL, y = "Median similarity"
    ) +
    theme_bw() +
    theme(legend.position = "none", panel.grid.minor = element_blank())

  print(p_box_mean)
  print(p_box_median)

  invisible(list(
    kw_mean = kw_mean, kw_median = kw_median,
    pw_mean_full = res_mean$pw_full,
    pw_median_full = res_median$pw_full
  ))
}

############################################################
# 7) RUN BOTH METHODS (INTENSITY-SENSITIVE)
############################################################

res_bray <- run_method_suite(
  species_stability,
  method_short = "Bray–Curtis similarity",
  method_definition = "S = 1 − BC distance on absolute KEGG intensity profiles (summed iBAQ)",
  mean_col = "mean_sim_bray",
  median_col = "median_sim_bray"
)

res_hell <- run_method_suite(
  species_stability,
  method_short = "Sqrt(abs) + Euclidean similarity",
  method_definition = "S = 1/(1 + Euclidean distance) on sqrt(absolute KEGG intensity profiles)",
  mean_col = "mean_sim_hell",
  median_col = "median_sim_hell"
)

############################################################
# "Whole-proteome functional stability was quantified per species as the similarity between
#  consecutive sampling dates of KEGG functional intensity profiles. For each species×date, protein
#  intensities (iBAQ-normalized abundances) were summed per KEGG ID (multiple proteins with the same
#  KEGG ID were summed) to form a KEGG intensity vector. Similarity between consecutive dates was
#  computed using Bray–Curtis similarity (1−BC distance) on absolute KEGG intensities and a square-
#  root–transformed Euclidean similarity (1/(1+d)). Species-level mean and median similarities were
#  compared among trophic groups using Kruskal–Wallis tests with pairwise Wilcoxon rank-sum tests and
#  BH correction; rank-biserial correlations (95% CI) were used as effect sizes. Boxplots show medians
#  and IQR with outliers hidden and individual species values overlaid."
############################################################
```


# Variablility parameter assessment of transporter proteome
```{r}
############################################################
# KEGG-PROFILE STABILITY (TRANSPORTER PROTEOME ONLY)
# — consecutive-date similarity on transporter-only KEGG profiles
# — INTENSITY-SENSITIVE (no relative normalization)
#
# What this script does
#  1) QC (unchanged): keep Species×Date with enough proteins + KEGG present,
#     optionally require ≥1 transporter KEGG and ≥1 non-transporter KEGG per date,
#     then keep Species with enough valid dates.
#
#  2) Build transporter-only KEGG intensity profiles per Species×Date:
#       kegg_abund(KEGG_ID) = sum(Protein_Abundance_IBAQ_Norm)
#     This automatically sums multiple proteins mapping to the same KEGG_ID on the same date.
#
#  3) For each Species, compute similarity between consecutive QC-passed dates:
#       A) Bray–Curtis similarity on absolute KEGG intensities:      S = 1 − BC(M_abs)
#       B) Sqrt(Hellinger-like) + Euclidean similarity on abs data:  S = 1/(1 + d_Euclid(sqrt(M_abs)))
#     (No decostand / proportions anywhere; total intensity is preserved.)
#
#  4) Summarize per species (mean/median similarity across transitions),
#     then produce the exact same figures + stats suite for both measures.
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)

  library(vegan)
  library(rstatix)
  library(ggpubr)
})

############################################################
# 0) USER SETTINGS
############################################################

trophic_levels <- c("Oligo_like", "Intermediate", "Copio_like")

trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

padj_method <- "BH"

min_proteins_per_date <- 10
min_valid_dates_per_species <- 5

require_trans_and_non_kegg <- TRUE
transporter_regex <- "\\bTransporters\\b"

############################################################
# 1) HELPERS
############################################################

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) return("<0.001")
  formatC(p, format = "e", digits = 2)
}

pairwise_tests_with_y <- function(df, response_col, padj_method = "BH", step_increase = 0.08) {
  df %>%
    pairwise_wilcox_test(
      reformulate("TrophicGroup", response = response_col),
      p.adjust.method = padj_method
    ) %>%
    add_significance("p.adj") %>%
    add_y_position(fun = "max", step.increase = step_increase)
}

pairwise_tests_plus_effects <- function(df, response_col, padj_method = "BH") {
  pw <- pairwise_tests_with_y(df, response_col, padj_method = padj_method)

  eff <- df %>%
    wilcox_effsize(
      reformulate("TrophicGroup", response = response_col),
      paired = FALSE,
      ci = TRUE
    ) %>%
    select(group1, group2, effsize, conf.low, conf.high)

  pw_full <- pw %>%
    left_join(eff, by = c("group1", "group2"))

  list(pw = pw, pw_full = pw_full, eff = eff)
}

build_density_annotation <- function(kw_obj, pw_tbl, levels_vec) {
  a <- levels_vec[1]; b <- levels_vec[2]; c <- levels_vec[3]
  get_padj <- function(tbl, g1, g2) {
    p <- tbl$p.adj[(tbl$group1 == g1 & tbl$group2 == g2) | (tbl$group1 == g2 & tbl$group2 == g1)]
    if (length(p) == 0) NA_real_ else p[1]
  }
  paste0(
    "Kruskal–Wallis p = ", fmt_p(kw_obj$p.value), "\n",
    "Pairwise Wilcoxon (BH):\n",
    a, " vs ", b, ": ", fmt_p(get_padj(pw_tbl, a, b)), "\n",
    a, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, a, c)), "\n",
    b, " vs ", c, ": ", fmt_p(get_padj(pw_tbl, b, c))
  )
}

############################################################
# 2) QC: retain species×date with enough proteins + KEGG present
############################################################

species_date_protcount <- protein_final %>%
  group_by(Species_ID, Date) %>%
  summarise(N_prot_detected = n_distinct(Protein_ID), .groups = "drop")

species_date_kegg_presence <- protein_final %>%
  mutate(
    has_kegg = !is.na(KEGG_ID),
    is_transporter_kegg = has_kegg & !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, transporter_regex)
  ) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    has_any_kegg = any(has_kegg),
    has_trans_kegg = any(is_transporter_kegg, na.rm = TRUE),
    has_nontrans_kegg = any(has_kegg & !is_transporter_kegg, na.rm = TRUE),
    .groups = "drop"
  )

species_date_keep <- species_date_protcount %>%
  left_join(species_date_kegg_presence, by = c("Species_ID", "Date")) %>%
  filter(
    N_prot_detected >= min_proteins_per_date,
    has_any_kegg
  ) %>%
  { if (require_trans_and_non_kegg) filter(., has_trans_kegg, has_nontrans_kegg) else . }

species_keep <- species_date_keep %>%
  count(Species_ID, name = "n_dates") %>%
  filter(n_dates >= min_valid_dates_per_species)

species_date_keep <- species_date_keep %>%
  inner_join(species_keep %>% select(Species_ID), by = "Species_ID")

species_traits <- protein_final %>%
  distinct(Species_ID, TrophicGroup, TrophicIndex)

species_date_keep <- species_date_keep %>%
  left_join(species_traits, by = "Species_ID") %>%
  mutate(TrophicGroup = factor(as.character(TrophicGroup), levels = trophic_levels)) %>%
  filter(is.finite(TrophicIndex), !is.na(TrophicGroup))

cat("\n=== Species retained after QC ===\n")
print(species_date_keep %>% distinct(Species_ID, TrophicGroup) %>% count(TrophicGroup))

############################################################
# 2b) OVERVIEW PLOT: distribution of consecutive-date pair counts per species
############################################################

paircount_df <- species_date_keep %>%
  distinct(Species_ID, Date) %>%
  count(Species_ID, name = "n_dates") %>%
  mutate(n_pairs = pmax(n_dates - 1, 0)) %>%
  left_join(species_date_keep %>% distinct(Species_ID, TrophicGroup), by = "Species_ID")

p_paircount <- ggplot(paircount_df, aes(x = n_pairs, fill = TrophicGroup)) +
  geom_histogram(binwidth = 1, boundary = -0.5, alpha = 0.75, color = "grey30") +
  scale_fill_manual(values = trophic_cols, drop = FALSE) +
  labs(
    title = "Distribution of consecutive-date pair counts per species (after QC)",
    subtitle = "n_pairs = n_dates − 1, where n_dates are QC-passed dates per species",
    x = "Number of consecutive-date pairs per species",
    y = "Number of species",
    fill = "Trophic group"
  ) +
  theme_bw() +
  theme(panel.grid.minor = element_blank())

print(p_paircount)

############################################################
# 3) KEGG INTENSITY PROFILES (transporter-only; sum per KEGG_ID per species-date)
############################################################

kegg_sd <- protein_final %>%
  filter(
    !is.na(KEGG_ID),
    !is.na(KEGG_Subcategory),
    str_detect(KEGG_Subcategory, transporter_regex)
  ) %>%
  inner_join(species_date_keep %>% select(Species_ID, Date), by = c("Species_ID", "Date")) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

############################################################
# 4) CONSECUTIVE-DATE SIMILARITIES PER SPECIES (INTENSITY-SENSITIVE)
############################################################

compute_species_consecutive_similarity <- function(sp_id, kegg_sd_tbl, keep_tbl) {

  dates <- keep_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_sd_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    tidyr::pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  M <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M) <- mat_wide$Date

  # Sqrt-transform on absolute abundances (keeps intensity, downweights extremes)
  M_sqrt <- sqrt(M)

  pairs <- data.frame(Date1 = dates[-length(dates)], Date2 = dates[-1], stringsAsFactors = FALSE)

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i])
    d2 <- as.character(pairs$Date2[i])

    # A) Bray–Curtis similarity on ABSOLUTE KEGG intensities: S = 1 − BC
    bc <- vegan::vegdist(M[c(d1, d2), , drop = FALSE], method = "bray")
    sim_bc <- 1 - as.numeric(bc)

    # B) Sqrt(abs) + Euclidean similarity: S = 1/(1 + d)
    d_sqrt <- dist(M_sqrt[c(d1, d2), , drop = FALSE], method = "euclidean")
    sim_sqrt <- 1 / (1 + as.numeric(d_sqrt))

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      sim_bray = sim_bc,
      sim_hellinger = sim_sqrt,  # name kept for minimal downstream changes
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

all_species <- sort(unique(species_date_keep$Species_ID))

sim_transitions <- bind_rows(lapply(
  all_species,
  compute_species_consecutive_similarity,
  kegg_sd_tbl = kegg_sd,
  keep_tbl = species_date_keep
))

species_covars <- species_date_keep %>%
  group_by(Species_ID, TrophicGroup, TrophicIndex) %>%
  summarise(
    n_dates = n(),
    mean_prot_detected = mean(N_prot_detected, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(mean_prot_detected = as.integer(round(mean_prot_detected)))

sim_transitions <- sim_transitions %>%
  left_join(species_covars, by = "Species_ID")

############################################################
# 5) SPECIES-LEVEL SUMMARY (mean + median similarity per species, per method)
############################################################

species_stability <- sim_transitions %>%
  group_by(Species_ID, TrophicGroup, TrophicIndex, n_dates, mean_prot_detected) %>%
  summarise(
    n_transitions = n(),
    mean_sim_bray = mean(sim_bray, na.rm = TRUE),
    median_sim_bray = median(sim_bray, na.rm = TRUE),
    mean_sim_hell = mean(sim_hellinger, na.rm = TRUE),
    median_sim_hell = median(sim_hellinger, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(TrophicGroup = factor(as.character(TrophicGroup), levels = trophic_levels))

############################################################
# 6) RUN SAME PLOTS + TESTS PER METHOD
############################################################

run_method_suite <- function(df, method_short, method_definition, mean_col, median_col) {

  method_full <- paste0(method_short, " — ", method_definition)

  cat("\n\n############################################################\n")
  cat("### METHOD:", method_full, "\n")
  cat("############################################################\n")

  kw_mean   <- kruskal.test(reformulate("TrophicGroup", response = mean_col), data = df)
  kw_median <- kruskal.test(reformulate("TrophicGroup", response = median_col), data = df)

  res_mean   <- pairwise_tests_plus_effects(df, mean_col, padj_method = padj_method)
  res_median <- pairwise_tests_plus_effects(df, median_col, padj_method = padj_method)

  pw_mean   <- res_mean$pw
  pw_median <- res_median$pw

  ann_mean   <- build_density_annotation(kw_mean,   pw_mean,   trophic_levels)
  ann_median <- build_density_annotation(kw_median, pw_median, trophic_levels)

  # Density: mean
  p_den_mean <- ggplot(df, aes(x = .data[[mean_col]], fill = TrophicGroup, color = TrophicGroup)) +
    geom_density(alpha = 0.25, linewidth = 1) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    scale_color_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMean consecutive-date similarity (per species)"),
      x = "Mean similarity",
      y = "Density"
    ) +
    annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
             label = ann_mean, size = 3.2, label.size = 0.25) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())

  # Density: median
  p_den_median <- ggplot(df, aes(x = .data[[median_col]], fill = TrophicGroup, color = TrophicGroup)) +
    geom_density(alpha = 0.25, linewidth = 1) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    scale_color_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMedian consecutive-date similarity (per species)"),
      x = "Median similarity",
      y = "Density"
    ) +
    annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
             label = ann_median, size = 3.2, label.size = 0.25) +
    theme_bw() +
    theme(panel.grid.minor = element_blank())

  print(p_den_mean)
  print(p_den_median)

  # Scatter: size breaks
  size_breaks <- df %>%
    summarise(
      mn = min(mean_prot_detected, na.rm = TRUE),
      q1 = as.integer(round(quantile(mean_prot_detected, 1/3, na.rm = TRUE))),
      q2 = as.integer(round(quantile(mean_prot_detected, 2/3, na.rm = TRUE))),
      mx = max(mean_prot_detected, na.rm = TRUE)
    ) %>%
    unlist(use.names = FALSE) %>% unique() %>% sort()

  if (length(size_breaks) < 4) {
    mn <- min(df$mean_prot_detected, na.rm = TRUE)
    mx <- max(df$mean_prot_detected, na.rm = TRUE)
    size_breaks <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
  }

  size_labels <- as.character(size_breaks)
  size_labels[1] <- paste0(size_labels[1], " (min)")
  size_labels[length(size_labels)] <- paste0(size_labels[length(size_labels)], " (max)")

  base_scales_scatter <- list(
    scale_color_manual(values = trophic_cols, drop = FALSE),
    scale_size_continuous(
      name = "Mean proteins detected\n(distinct Protein_ID per date)",
      range = c(2, 10),
      breaks = size_breaks,
      labels = size_labels
    ),
    theme_bw(),
    theme(panel.grid.minor = element_blank())
  )

  p_sc_mean <- ggplot(df, aes(x = TrophicIndex, y = .data[[mean_col]], color = TrophicGroup, size = mean_prot_detected)) +
    geom_point(alpha = 0.85) +
    labs(
      title = paste0(method_full, "\nMean similarity vs trophic index"),
      x = "Trophic index",
      y = "Mean similarity",
      color = "Trophic group"
    ) +
    base_scales_scatter

  p_sc_median <- ggplot(df, aes(x = TrophicIndex, y = .data[[median_col]], color = TrophicGroup, size = mean_prot_detected)) +
    geom_point(alpha = 0.85) +
    labs(
      title = paste0(method_full, "\nMedian similarity vs trophic index"),
      x = "Trophic index",
      y = "Median similarity",
      color = "Trophic group"
    ) +
    base_scales_scatter

  print(p_sc_mean)
  print(p_sc_median)

  # Boxplots
  p_box_mean <- ggplot(df, aes(x = TrophicGroup, y = .data[[mean_col]], fill = TrophicGroup)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
    geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
    stat_pvalue_manual(pw_mean, label = "p.adj.signif", tip.length = 0.01, hide.ns = FALSE) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMean similarity across trophic groups"),
      subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method),
      x = NULL, y = "Mean similarity"
    ) +
    theme_bw() +
    theme(legend.position = "none", panel.grid.minor = element_blank())

  p_box_median <- ggplot(df, aes(x = TrophicGroup, y = .data[[median_col]], fill = TrophicGroup)) +
    geom_boxplot(outlier.shape = NA, width = 0.6, alpha = 0.65) +
    geom_jitter(width = 0.15, height = 0, size = 2, alpha = 0.8) +
    stat_pvalue_manual(pw_median, label = "p.adj.signif", tip.length = 0.01, hide.ns = FALSE) +
    scale_fill_manual(values = trophic_cols, drop = FALSE) +
    labs(
      title = paste0(method_full, "\nMedian similarity across trophic groups"),
      subtitle = paste0("Overall: Kruskal–Wallis; Pairwise: Wilcoxon + ", padj_method),
      x = NULL, y = "Median similarity"
    ) +
    theme_bw() +
    theme(legend.position = "none", panel.grid.minor = element_blank())

  print(p_box_mean)
  print(p_box_median)

  invisible(list(
    kw_mean = kw_mean, kw_median = kw_median,
    pw_mean_full = res_mean$pw_full,
    pw_median_full = res_median$pw_full
  ))
}

############################################################
# 7) RUN BOTH METHODS (INTENSITY-SENSITIVE)
############################################################

res_bray <- run_method_suite(
  species_stability,
  method_short = "Bray–Curtis similarity",
  method_definition = "S = 1 − BC distance on absolute KEGG intensity profiles (summed iBAQ)",
  mean_col = "mean_sim_bray",
  median_col = "median_sim_bray"
)

res_hell <- run_method_suite(
  species_stability,
  method_short = "Sqrt(abs) + Euclidean similarity",
  method_definition = "S = 1/(1 + Euclidean distance) on sqrt(absolute KEGG intensity profiles)",
  mean_col = "mean_sim_hell",
  median_col = "median_sim_hell"
)

############################################################
# "Transporter-proteome functional stability was quantified per species as the similarity between
#  consecutive sampling dates of transporter-only KEGG functional intensity profiles. For each
#  species×date, protein intensities (iBAQ-normalized abundances) were summed per KEGG ID using only
#  proteins annotated as transporters (KEGG subcategory matching 'Transporters') to form a KEGG
#  intensity vector (multiple proteins with the same KEGG ID were summed). Similarity between
#  consecutive dates was computed using Bray–Curtis similarity (1−BC distance) on absolute KEGG
#  intensities and a square-root–transformed Euclidean similarity (1/(1+d)). Species-level mean and
#  median similarities were compared among trophic groups using Kruskal–Wallis tests with pairwise
#  Wilcoxon rank-sum tests and BH correction; rank-biserial correlations (95% CI) were used as effect
#  sizes. Boxplots show medians and IQR with outliers hidden and individual species values overlaid."
############################################################
```

# test if there is a phylum, number of dates or number of protein effect on the correlation between trophic index and transporter proteome variability
```{r}
############################################################
# Transporter functional stability ~ TrophicIndex
# Bray–Curtis ONLY on ABSOLUTE intensities
#
# What this script does
# 1) QC (unchanged logic):
#    - keep Species×Date with >= min_proteins_per_date (distinct Protein_ID)
#    - require at least 1 transporter KEGG + 1 non-transporter KEGG if require_trans_and_non_kegg = TRUE
#    - keep Species with >= min_valid_dates_per_species after QC
#
# 2) Build transporter-only KEGG intensity profiles per Species×Date:
#    - filter proteins to transporter-annotated rows (KEGG_Subcategory contains "Transporters")
#    - require KEGG_ID present and KEGG_Subcategory not NA (removes "unclassified")
#    - sum Protein_Abundance_IBAQ_Norm per (Species_ID, Date, KEGG_ID)
#      -> if multiple proteins map to the same KEGG_ID on the same date, they are correctly SUMMED.
#
# 3) Compute consecutive-date Bray–Curtis on ABSOLUTE intensity profiles:
#    - BC_abs distance in [0,1]
#    - Similarity S_abs = 1 - BC_abs
#
# 4) Test TI relationship without beta regression:
#    A) Transition-level mixed model (primary inference; repeated measures):
#         logit(S_abs) ~ TrophicIndex + detection + (1|Species_ID) + (1|Class or Phylum)
#    B) Species-level models (supporting; one point per species):
#         logit(mean_S_abs)   ~ TrophicIndex + n_transitions + detection + (1|Class or Phylum)
#         logit(median_S_abs) ~ TrophicIndex + n_transitions + detection + (1|Class or Phylum)
#
# 5) Output:
#    - QC pair-count histogram
#    - 3 plots with regression ribbon + stats box:
#        transition-level S_abs vs TI
#        species-level mean S_abs vs TI
#        species-level median S_abs vs TI
#    - full stats + diagnostics report written to:
#        Transporter_BrayCurtis_ABS_ONLY_TrophicIndex_stats.txt
#
# Inputs required:
#   protein_final: Species_ID, Date, Protein_ID, KEGG_ID, KEGG_Subcategory,
#                 Protein_Abundance_IBAQ_Norm, TrophicIndex (TrophicGroup optional)
#   species_taxonomy: Species_ID, Phylum, Class
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)

  library(vegan)

  library(glmmTMB)
  library(DHARMa)
  library(performance)
  library(broom.mixed)
  library(ggeffects)
})

############################################################
# 0) USER SETTINGS
############################################################
min_proteins_per_date <- 10
min_valid_dates_per_species <- 5

require_trans_and_non_kegg <- TRUE
transporter_regex <- "\\bTransporters\\b"

# squeeze away from 0/1 before logit transform
eps <- 1e-6

use_bias_correction <- TRUE
report_file <- "Transporter_BrayCurtis_ABS_ONLY_TrophicIndex_stats.txt"

save_figures <- FALSE
fig_prefix <- "TransporterBCabs_TI_"

############################################################
# 1) HELPERS
############################################################
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) return("<1e-4")
  if (p < 1e-3) return("<0.001")
  formatC(p, format = "e", digits = 2)
}

logit_similarity <- function(S, eps = 1e-6) {
  S2 <- pmin(pmax(S, eps), 1 - eps)
  qlogis(S2)
}

extract_TI <- function(model, term = "TrophicIndex_z") {
  td <- broom.mixed::tidy(model, effects = "fixed", conf.int = TRUE)
  row <- td %>% filter(term == !!term)
  if (nrow(row) != 1) return(NULL)
  row
}

model_annotation_text <- function(model, response_label, term = "TrophicIndex_z") {
  ti <- extract_TI(model, term)
  if (is.null(ti)) return("")

  beta <- ti$estimate
  ci_l <- ti$conf.low
  ci_u <- ti$conf.high
  pval <- ti$p.value

  r2_txt <- ""
  r2 <- tryCatch(performance::r2(model), error = function(e) NULL)
  if (!is.null(r2)) {
    r2m <- suppressWarnings(as.numeric(r2$R2_marginal))
    r2c <- suppressWarnings(as.numeric(r2$R2_conditional))
    has_r2m <- length(r2m) == 1 && is.finite(r2m)
    has_r2c <- length(r2c) == 1 && is.finite(r2c)
    if (has_r2m && has_r2c) {
      r2_txt <- paste0("R²m=", formatC(r2m, digits = 2, format = "f"),
                       ", R²c=", formatC(r2c, digits = 2, format = "f"))
    } else if (has_r2m) {
      r2_txt <- paste0("R²m=", formatC(r2m, digits = 2, format = "f"))
    }
  }

  paste0(
    response_label, " ~ TI (logit similarity)\n",
    "β=", formatC(beta, digits = 3, format = "f"),
    " [", formatC(ci_l, digits = 3, format = "f"), ", ", formatC(ci_u, digits = 3, format = "f"), "]",
    ", p=", fmt_p(pval), "\n",
    "exp(β)=", formatC(exp(beta), digits = 2, format = "f"),
    " [", formatC(exp(ci_l), digits = 2, format = "f"), ", ", formatC(exp(ci_u), digits = 2, format = "f"), "]",
    if (nzchar(r2_txt)) paste0("\n", r2_txt) else ""
  )
}

diagnose_block <- function(model, label, n_sim = 500) {
  cat("\n", label, "\n", sep = "")
  cat(strrep("-", nchar(label)), "\n", sep = "")

  cat("pdHess: ", tryCatch(isTRUE(model$sdr$pdHess), error = function(e) NA), "\n", sep = "")

  cat("\nVIF:\n"); print(tryCatch(check_collinearity(model), error = function(e) e$message))
  cat("\nR²:\n");  print(tryCatch(performance::r2(model), error = function(e) e$message))

  cat("\nDHARMa:\n")
  res <- tryCatch(simulateResiduals(model, n = n_sim), error = function(e) NULL)
  if (is.null(res)) {
    cat("DHARMa failed (can happen for complex RE structures).\n")
  } else {
    print(testUniformity(res))
    print(testDispersion(res))
    print(testOutliers(res))
  }
}

pick_best <- function(m_class, m_phy) {
  a <- AIC(m_class); b <- AIC(m_phy)
  if (is.finite(a) && is.finite(b) && a <= b) m_class else m_phy
}

predict_curve_to_similarity <- function(model, df, ti_var = "TrophicIndex", z_var = "TrophicIndex_z") {
  pred <- ggeffects::ggpredict(
    model,
    terms = paste0(z_var, " [all]"),
    bias_correction = use_bias_correction
  ) %>% as.data.frame()

  mu <- mean(df[[ti_var]], na.rm = TRUE)
  sdv <- sd(df[[ti_var]], na.rm = TRUE)

  pred$TrophicIndex <- pred$x * sdv + mu
  pred$S <- plogis(pred$predicted)
  pred$S_low <- plogis(pred$conf.low)
  pred$S_high <- plogis(pred$conf.high)
  pred
}

############################################################
# 2) QC: SAME AS BEFORE
############################################################

species_date_protcount <- protein_final %>%
  group_by(Species_ID, Date) %>%
  summarise(N_prot_detected = n_distinct(Protein_ID), .groups = "drop")

species_date_kegg_presence <- protein_final %>%
  mutate(
    has_kegg = !is.na(KEGG_ID),
    is_transporter_kegg = has_kegg & !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, transporter_regex)
  ) %>%
  group_by(Species_ID, Date) %>%
  summarise(
    has_any_kegg = any(has_kegg),
    has_trans_kegg = any(is_transporter_kegg, na.rm = TRUE),
    has_nontrans_kegg = any(has_kegg & !is_transporter_kegg, na.rm = TRUE),
    .groups = "drop"
  )

species_date_keep <- species_date_protcount %>%
  left_join(species_date_kegg_presence, by = c("Species_ID","Date")) %>%
  filter(N_prot_detected >= min_proteins_per_date, has_any_kegg) %>%
  { if (require_trans_and_non_kegg) filter(., has_trans_kegg, has_nontrans_kegg) else . }

species_keep <- species_date_keep %>%
  count(Species_ID, name = "n_dates") %>%
  filter(n_dates >= min_valid_dates_per_species) %>%
  select(Species_ID)

species_date_keep <- species_date_keep %>%
  semi_join(species_keep, by = "Species_ID")

species_traits <- protein_final %>%
  distinct(Species_ID, TrophicIndex, TrophicGroup)

species_date_keep <- species_date_keep %>%
  left_join(species_traits, by = "Species_ID") %>%
  left_join(species_taxonomy %>% select(Species_ID, Phylum, Class), by = "Species_ID") %>%
  filter(is.finite(TrophicIndex), !is.na(Phylum), !is.na(Class)) %>%
  mutate(Phylum = factor(Phylum), Class = factor(Class))

cat("\nQC-passed species×date rows: ", nrow(species_date_keep), "\n", sep = "")
cat("QC-passed species: ", dplyr::n_distinct(species_date_keep$Species_ID), "\n", sep = "")

############################################################
# 2b) OVERVIEW PLOT: pair counts per species
############################################################
paircount_df <- species_date_keep %>%
  distinct(Species_ID, Date) %>%
  count(Species_ID, name = "n_dates") %>%
  mutate(n_pairs = pmax(n_dates - 1, 0))

p_paircount <- ggplot(paircount_df, aes(x = n_pairs)) +
  geom_histogram(binwidth = 1, boundary = -0.5, alpha = 0.75, color = "grey30") +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Consecutive-date pair counts per species (after QC)",
    subtitle = "n_pairs = n_dates − 1 using QC-passed dates",
    x = "Number of consecutive-date pairs per species",
    y = "Number of species"
  )
print(p_paircount)

############################################################
# 3) TRANSPORTER KEGG INTENSITY PROFILES (Species×Date×KEGG_ID)
############################################################
kegg_sd_trans <- protein_final %>%
  filter(
    !is.na(KEGG_ID),
    !is.na(KEGG_Subcategory),
    str_detect(KEGG_Subcategory, transporter_regex)
  ) %>%
  inner_join(species_date_keep %>% select(Species_ID, Date), by = c("Species_ID","Date")) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(
    kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

############################################################
# 4) TRANSITION-LEVEL BRAY–CURTIS (ABSOLUTE ONLY)
############################################################
compute_species_transitions_bc <- function(sp_id, kegg_tbl, keep_tbl) {

  dates <- keep_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  M_abs <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M_abs) <- mat_wide$Date

  pairs <- data.frame(Date1 = dates[-length(dates)], Date2 = dates[-1], stringsAsFactors = FALSE)

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i]); d2 <- as.character(pairs$Date2[i])

    bc_abs <- as.numeric(vegdist(M_abs[c(d1, d2), , drop = FALSE], method = "bray"))
    s_abs <- 1 - bc_abs

    nd1 <- keep_tbl %>% filter(Species_ID == sp_id, Date == pairs$Date1[i]) %>% pull(N_prot_detected)
    nd2 <- keep_tbl %>% filter(Species_ID == sp_id, Date == pairs$Date2[i]) %>% pull(N_prot_detected)
    det_pair <- mean(c(nd1, nd2), na.rm = TRUE)

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      bc_abs = bc_abs,
      s_abs  = s_abs,
      N_prot_pair = det_pair,
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

all_species <- sort(unique(species_date_keep$Species_ID))

transitions <- bind_rows(lapply(
  all_species, compute_species_transitions_bc,
  kegg_tbl = kegg_sd_trans,
  keep_tbl = species_date_keep
))

transitions <- transitions %>%
  left_join(
    species_date_keep %>% distinct(Species_ID, TrophicIndex, Phylum, Class),
    by = "Species_ID"
  ) %>%
  filter(is.finite(s_abs), is.finite(N_prot_pair)) %>%
  mutate(
    log_prot_pair = log1p(N_prot_pair),
    s_abs_logit = logit_similarity(s_abs, eps = eps),
    TrophicIndex_z  = as.numeric(scale(TrophicIndex)),
    log_prot_pair_z = as.numeric(scale(log_prot_pair))
  )

############################################################
# 5) SPECIES-LEVEL SUMMARY (ABSOLUTE mean + median)
############################################################
species_level <- transitions %>%
  group_by(Species_ID, TrophicIndex, Phylum, Class) %>%
  summarise(
    n_transitions = n(),
    mean_s_abs   = mean(s_abs, na.rm = TRUE),
    median_s_abs = median(s_abs, na.rm = TRUE),
    mean_prot_pair = mean(N_prot_pair, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    log_prot_pair = log1p(mean_prot_pair),
    mean_s_abs_logit   = logit_similarity(mean_s_abs, eps = eps),
    median_s_abs_logit = logit_similarity(median_s_abs, eps = eps),
    TrophicIndex_z  = as.numeric(scale(TrophicIndex)),
    n_transitions_z = as.numeric(scale(n_transitions)),
    log_prot_pair_z = as.numeric(scale(log_prot_pair))
  )

############################################################
# 6) MODELS + REPORT
############################################################
sink(report_file)

cat("Transporter stability ~ TrophicIndex (Bray–Curtis ABSOLUTE ONLY)\n")
cat("==============================================================\n\n")

cat("QC SUMMARY\n")
cat("----------\n")
cat("Transitions: ", nrow(transitions), "\n", sep = "")
cat("Species: ", nrow(species_level), "\n\n", sep = "")

cat("MEASURE\n")
cat("-------\n")
cat(
  "Bray–Curtis was computed on ABSOLUTE transporter KEGG intensity profiles per species×date.\n",
  "Similarity is S_abs = 1 - BC_abs.\n",
  "We model logit(S_abs) using Gaussian mixed models (no beta regression).\n\n",
  sep = ""
)

# Transition-level models (primary inference)
fit_transition <- function(tax = c("Class","Phylum")) {
  tax <- match.arg(tax)
  re_tax <- if (tax == "Class") "(1|Class)" else "(1|Phylum)"

  glmmTMB(
    reformulate(c("TrophicIndex_z", "log_prot_pair_z", "(1|Species_ID)", re_tax), response = "s_abs_logit"),
    family = gaussian(),
    data = transitions
  )
}

m_tr_class <- fit_transition("Class")
m_tr_phy   <- fit_transition("Phylum")

cat("\nTRANSITION-LEVEL (PRIMARY) MODELS\n")
cat("--------------------------------\n")
cat("\n[Transition | logit(S_abs) | (1|Species_ID)+(1|Class)]\n");  print(summary(m_tr_class))
cat("\n[Transition | logit(S_abs) | (1|Species_ID)+(1|Phylum)]\n"); print(summary(m_tr_phy))
cat("\nAIC compare (transition):\n"); print(AIC(m_tr_class, m_tr_phy))

# Species-level models (supporting)
fit_species <- function(resp, tax = c("Class","Phylum")) {
  tax <- match.arg(tax)
  re_tax <- if (tax == "Class") "(1|Class)" else "(1|Phylum)"

  glmmTMB(
    reformulate(c("TrophicIndex_z", "n_transitions_z", "log_prot_pair_z", re_tax), response = resp),
    family = gaussian(),
    data = species_level
  )
}

m_sp_mean_class   <- fit_species("mean_s_abs_logit", "Class")
m_sp_mean_phy     <- fit_species("mean_s_abs_logit", "Phylum")
m_sp_median_class <- fit_species("median_s_abs_logit", "Class")
m_sp_median_phy   <- fit_species("median_s_abs_logit", "Phylum")

cat("\nSPECIES-LEVEL (SUPPORTING) MODELS\n")
cat("---------------------------------\n")

cat("\n[Species | mean logit(S_abs) | Class]\n");  print(summary(m_sp_mean_class))
cat("\n[Species | mean logit(S_abs) | Phylum]\n"); print(summary(m_sp_mean_phy))
cat("\nAIC compare (species mean):\n"); print(AIC(m_sp_mean_class, m_sp_mean_phy))

cat("\n[Species | median logit(S_abs) | Class]\n");  print(summary(m_sp_median_class))
cat("\n[Species | median logit(S_abs) | Phylum]\n"); print(summary(m_sp_median_phy))
cat("\nAIC compare (species median):\n"); print(AIC(m_sp_median_class, m_sp_median_phy))

cat("\nDIAGNOSTICS (PRIMARY + SUPPORTING)\n")
cat("----------------------------------\n")
diagnose_block(m_tr_class, "Transition: logit(S_abs) + Class")
diagnose_block(m_sp_mean_class, "Species mean: logit(mean S_abs) + Class")
diagnose_block(m_sp_median_class, "Species median: logit(median S_abs) + Class")

sink()
cat("Wrote report to: ", report_file, "\n", sep = "")

############################################################
# 7) FIGURES (transition + species mean + species median)
############################################################

# pick best taxonomy level (AIC) for annotation/prediction
m_tr_best <- pick_best(m_tr_class, m_tr_phy)
m_sp_mean_best <- pick_best(m_sp_mean_class, m_sp_mean_phy)
m_sp_median_best <- pick_best(m_sp_median_class, m_sp_median_phy)

# Transition-level prediction (on similarity scale)
pred_tr <- predict_curve_to_similarity(m_tr_best, transitions)
ann_tr <- model_annotation_text(m_tr_best, "Transition logit(S_abs)")

p_tr <- ggplot() +
  geom_point(data = transitions, aes(x = TrophicIndex, y = s_abs), alpha = 0.12, size = 1) +
  geom_ribbon(data = pred_tr, aes(x = TrophicIndex, ymin = S_low, ymax = S_high), alpha = 0.18) +
  geom_line(data = pred_tr, aes(x = TrophicIndex, y = S), linewidth = 1) +
  annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
           label = paste0(
             ann_tr, "\n",
             "N transitions=", nrow(transitions),
             "; N species=", dplyr::n_distinct(transitions$Species_ID),
             "\nControls: detection + taxonomy + repeated measures"
           ),
           size = 3.2, label.size = 0.25) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Transporter functional stability vs trophic index (transition-level)",
    subtitle = "Bray–Curtis similarity across consecutive dates (ABSOLUTE KEGG transporter intensities)",
    x = "Trophic index",
    y = "Consecutive-date Bray–Curtis similarity (S_abs = 1−BC_abs)"
  )
print(p_tr)

# Species-level mean prediction
pred_sp_mean <- predict_curve_to_similarity(m_sp_mean_best, species_level)
ann_sp_mean <- model_annotation_text(m_sp_mean_best, "Species mean logit(S_abs)")

p_sp_mean <- ggplot() +
  geom_point(data = species_level, aes(x = TrophicIndex, y = mean_s_abs), alpha = 0.45, size = 1.8) +
  geom_ribbon(data = pred_sp_mean, aes(x = TrophicIndex, ymin = S_low, ymax = S_high), alpha = 0.18) +
  geom_line(data = pred_sp_mean, aes(x = TrophicIndex, y = S), linewidth = 1) +
  annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
           label = paste0(
             ann_sp_mean, "\n",
             "N species=", nrow(species_level),
             "\nControls: n_transitions + detection + taxonomy"
           ),
           size = 3.2, label.size = 0.25) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Transporter functional stability vs trophic index (species-level mean)",
    subtitle = "Mean S_abs across transitions per species",
    x = "Trophic index",
    y = "Mean consecutive-date Bray–Curtis similarity (S_abs)"
  )
print(p_sp_mean)

# Species-level median prediction
pred_sp_med <- predict_curve_to_similarity(m_sp_median_best, species_level)
ann_sp_med <- model_annotation_text(m_sp_median_best, "Species median logit(S_abs)")

p_sp_med <- ggplot() +
  geom_point(data = species_level, aes(x = TrophicIndex, y = median_s_abs), alpha = 0.45, size = 1.8) +
  geom_ribbon(data = pred_sp_med, aes(x = TrophicIndex, ymin = S_low, ymax = S_high), alpha = 0.18) +
  geom_line(data = pred_sp_med, aes(x = TrophicIndex, y = S), linewidth = 1) +
  annotate("label", x = Inf, y = Inf, hjust = 1.02, vjust = 1.02,
           label = paste0(
             ann_sp_med, "\n",
             "N species=", nrow(species_level),
             "\nControls: n_transitions + detection + taxonomy"
           ),
           size = 3.2, label.size = 0.25) +
  theme_bw() +
  theme(panel.grid.minor = element_blank()) +
  labs(
    title = "Transporter functional stability vs trophic index (species-level median)",
    subtitle = "Median S_abs across transitions per species",
    x = "Trophic index",
    y = "Median consecutive-date Bray–Curtis similarity (S_abs)"
  )
print(p_sp_med)

############################################################
# 8) OPTIONAL: SAVE FIGURES
############################################################
if (save_figures) {
  ggsave(paste0(fig_prefix, "paircount.png"), p_paircount, width = 7.2, height = 4.6, dpi = 300)
  ggsave(paste0(fig_prefix, "transition.png"), p_tr, width = 7.5, height = 5.2, dpi = 300)
  ggsave(paste0(fig_prefix, "species_mean.png"), p_sp_mean, width = 7.5, height = 5.2, dpi = 300)
  ggsave(paste0(fig_prefix, "species_median.png"), p_sp_med, width = 7.5, height = 5.2, dpi = 300)
}

cat("Done. Stats written to: ", report_file, "\n", sep = "")
```

# Interpretation

Below is a concise, structured comparison of the two analytical frameworks:

- **Transporter Proteome Fraction (TPF)**
- **Transporter Functional Stability (absolute Bray–Curtis similarity)**

All interpretations are based on the statistical reports:
- TPF report :contentReference[oaicite:0]{index=0}  
- Absolute Bray–Curtis report :contentReference[oaicite:1]{index=1}  

---

# I. Transporter Proteome Fraction (TPF)

## What was tested

TPF quantifies *how much* of the proteome is allocated to transporters.  
It was modeled using beta mixed models appropriate for bounded proportions.

Three complementary tests were performed:

1. **Species-level mean TPF**
2. **Species-level median TPF**
3. **Date-level repeated-measures TPF**

All models controlled for:

- Taxonomy (Phylum/Class random effects)
- Number of dates per species
- Proteome detection depth

## What the tests show

Across all levels:

- **TrophicIndex_z is significantly negative**
- Effect sizes are consistent across mean, median, and date-level models
- Results remain robust under weighting and interaction tests
- Model diagnostics are acceptable

From predicted effects (10th → 90th TI percentile):

- Species mean TPF drops from ~0.51 → ~0.27  
- Date-level TPF drops from ~0.53 → ~0.20  

This represents roughly a **40–60% reduction in transporter allocation** across the trophic gradient.

## Interpretation

Transporter investment declines continuously with increasing trophic index.

This pattern:

- Persists after controlling for taxonomy
- Persists after correcting for detection depth
- Persists under repeated-measures modeling

Thus, transporter allocation behaves as a **continuous trophic strategy trait**, not a taxonomic artifact.

---

# II. Transporter Functional Stability (Absolute Bray–Curtis)

## What was tested

Here, the question changes:

Not *how much* transporter investment exists, but  
**how stable transporter functional composition is across consecutive dates.**

Bray–Curtis similarity was computed on:

- Absolute KEGG transporter intensity profiles
- Per species × date
- Between consecutive dates

Models used Gaussian mixed models on logit-transformed similarity.

Three levels:

1. Transition-level (primary test; repeated measures)
2. Species-level mean similarity
3. Species-level median similarity

Controls:

- Taxonomy
- Number of transitions
- Detection depth per pair

## What the tests show

Across all levels:

- **TrophicIndex_z is significantly negative**
- The effect is consistent in transition-level and species-level models
- R² values indicate moderate explanatory power
- Diagnostics are acceptable (minor deviations at transition-level, expected given large N)

Thus:

Higher trophic index → lower transporter stability  
Lower trophic index → higher transporter stability

This indicates that oligotrophic taxa maintain more temporally stable transporter functional profiles.

## Interpretation

Transporter functional composition becomes increasingly dynamic toward higher trophic strategies.

Importantly, this pattern holds:

- After correcting for detection depth
- After correcting for number of transitions
- After accounting for repeated measures
- After controlling for taxonomy

Therefore, it is unlikely to be driven by sampling depth or phylogenetic clustering.

---

# III. Combined Biological Interpretation

Together, the two analyses reveal a coherent trophic strategy gradient:

### 1. Allocation effect (TPF)

Oligotrophic taxa:
- Invest more proteome mass into transporters

Copiotrophic taxa:
- Allocate less proteome mass to transport

### 2. Stability effect (Bray–Curtis)

Oligotrophic taxa:
- Maintain stable transporter functional composition over time

Copiotrophic taxa:
- Exhibit higher temporal turnover in transporter profiles

---

# IV. Overall Biological Synthesis

The joint evidence supports a **continuous transporter-centered trophic strategy axis**:

- Low trophic index taxa combine **high transporter investment** with **functional stability**.
- High trophic index taxa combine **lower transporter allocation** with **higher functional plasticity**.

This suggests:

Oligotrophic organisms adopt a **stable, transport-optimized proteomic architecture**,  
whereas copiotrophic organisms adopt a **dynamic, flexible transporter repertoire**.

Crucially, both allocation and stability signals:

- Survive correction for taxonomy
- Survive correction for sampling depth
- Persist at both species-level and repeated-measures level

Thus, transporter allocation and stability appear to be fundamental ecological design principles linked to trophic strategy rather than artifacts of phylogeny or detection bias.

---

# Final Assessment

The two independent analytical frameworks converge on the same biological gradient.

The evidence supports:

- A robust, continuous trophic trade-off
- A proteome-level ecological axis
- A mechanistically interpretable transporter strategy shift

The combined result substantially strengthens the ecological inference beyond either analysis alone.


# median tpf vs median bray curtis dissimilarity
```{r}
############################################################
# MEDIAN vs MEDIAN ONLY
# - x: median TPF per species
# - y: median variability = median Bray–Curtis dissimilarity
#      (lower = more stable; higher = more variable)
# - Hardcoded axes: 0..1 by 0.25 for scatter + boxplots
# - Same plotting style + highlight legend + PDF export only
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggExtra)
  library(grid)
  library(gridExtra)
  library(readr)
  library(vegan)
})

# =========================================================
# 0) Safety checks — objects expected from your load block
# =========================================================
required_objects <- c("protein_final", "retained_species_dates", "species_taxonomy")
for (obj in required_objects) {
  if (!exists(obj, inherits = TRUE)) stop("Object not found in environment: ", obj)
}

req_keys_cols <- c("Species_ID","Date","Group_For_Plots")
missing_keys <- setdiff(req_keys_cols, names(retained_species_dates))
if (length(missing_keys) > 0) stop("retained_species_dates missing: ", paste(missing_keys, collapse=", "))

req_tax_cols <- c("Species_ID","Phylum","Genus")
missing_tax <- setdiff(req_tax_cols, names(species_taxonomy))
if (length(missing_tax) > 0) stop("species_taxonomy missing: ", paste(missing_tax, collapse=", "))

req_prot_cols <- c("Species_ID","Date","Group_For_Plots","Protein_Abundance_IBAQ_Norm","KEGG_Subcategory","KEGG_ID")
missing_prot <- setdiff(req_prot_cols, names(protein_final))
if (length(missing_prot) > 0) stop("protein_final missing: ", paste(missing_prot, collapse=", "))

protein_id_col <- dplyr::case_when(
  "Protein_Group" %in% names(protein_final) ~ "Protein_Group",
  "Protein_ID"    %in% names(protein_final) ~ "Protein_ID",
  TRUE ~ NA_character_
)
if (is.na(protein_id_col)) stop("protein_final must have Protein_Group or Protein_ID for counting detected proteins.")

protein_final <- protein_final %>% mutate(Date = as.Date(Date))
retained_species_dates <- retained_species_dates %>% mutate(Date = as.Date(Date))

# =========================================================
# 1) Fixed settings (orders, colors, highlight scheme)
# =========================================================
trophic_order <- c("Oligo_like", "Intermediate", "Copio_like")
trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

highlight_cols <- c(
  "Flavobacterium"        = "#E6BE1E",
  "Planktophila"          = "#ADD5E4",
  "Fonsibacter"           = "#C6FC05",
  "Planctomycetota (phy)" = "#7B3294",
  "Rest"                  = "grey"
)
highlight_levels <- names(highlight_cols)

axis_breaks_01 <- seq(0, 1, by = 0.25)

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) "<0.001" else format(p, digits = 2, scientific = TRUE)
}

get_legend <- function(p) {
  g <- ggplotGrob(p)
  idx <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  if (length(idx) != 1) stop("Legend extraction failed.")
  g$grobs[[idx]]
}

export_grob_pdf <- function(g, base, width, height) {
  pdf(paste0(base, ".pdf"), width = width, height = height)
  grid.newpage(); grid.draw(g)
  dev.off()
  cat("\nSaved PDF:\n  ", normalizePath(paste0(base, ".pdf")), "\n", sep="")
}

# =========================================================
# 2) Enforce Species×Date×Group universe (critical)
# =========================================================
protein_final_u <- protein_final %>%
  semi_join(retained_species_dates, by = c("Species_ID","Date","Group_For_Plots"))

cat("\n=== Universe enforcement ===\n")
cat("protein_final_u rows after semi_join:", nrow(protein_final_u), "\n")

# =========================================================
# 3) Compute TPF per Species×Date
#    Transporters: KEGG_Subcategory contains whole word "Transporters"
# =========================================================
protein_tagged <- protein_final_u %>%
  mutate(
    is_transporter = !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, "\\bTransporters\\b"),
    abund = Protein_Abundance_IBAQ_Norm
  )

species_date_tpf <- protein_tagged %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    abund_total = sum(abund, na.rm = TRUE),
    abund_trans = sum(abund[is_transporter], na.rm = TRUE),
    TPF = abund_trans / abund_total,
    N_prot_total = n_distinct(.data[[protein_id_col]]),
    .groups = "drop"
  ) %>%
  filter(is.finite(TPF), abund_total > 0, TPF > 0, TPF < 1, N_prot_total > 0)

cat("\n=== species_date_tpf ===\n")
cat("Rows:", nrow(species_date_tpf), "\n")

# =========================================================
# 4) Compute transporter-only KEGG intensity profiles per Species×Date
#    (summed iBAQ per KEGG_ID) and consecutive-date Bray–Curtis DISTANCE
# =========================================================
kegg_sd <- protein_tagged %>%
  filter(
    is_transporter,
    !is.na(KEGG_ID)
  ) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

compute_species_consecutive_braydist <- function(sp_id, kegg_sd_tbl, keep_dates_tbl) {

  dates <- keep_dates_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_sd_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  if (nrow(mat_wide) < 2) return(NULL)

  M <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M) <- mat_wide$Date

  pairs <- data.frame(
    Date1 = dates[-length(dates)],
    Date2 = dates[-1],
    stringsAsFactors = FALSE
  )

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i])
    d2 <- as.character(pairs$Date2[i])

    bc <- vegan::vegdist(M[c(d1, d2), , drop = FALSE], method = "bray")
    bc_dist <- as.numeric(bc)  # 0..1; lower=more similar (stable)

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      bray_dist = bc_dist,
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

keep_dates <- species_date_tpf %>% select(Species_ID, Date, Group_For_Plots) %>% distinct()

all_species <- sort(unique(keep_dates$Species_ID))

bray_transitions <- bind_rows(lapply(
  all_species,
  compute_species_consecutive_braydist,
  kegg_sd_tbl = kegg_sd,
  keep_dates_tbl = keep_dates
))

cat("\n=== bray_transitions ===\n")
cat("Rows:", nrow(bray_transitions), "\n")

# =========================================================
# 5) Species-level medians (ONE DOT PER SPECIES)
#    - median_TPF across dates
#    - median_bray across consecutive transitions
#    - median proteins detected (size covariate)
#    - taxonomy + highlight label
# =========================================================
species_covars <- species_date_tpf %>%
  group_by(Species_ID, Group_For_Plots) %>%
  summarise(
    n_dates = n(),
    median_TPF = median(TPF, na.rm = TRUE),
    median_prot_count = as.integer(round(median(N_prot_total, na.rm = TRUE))),
    .groups = "drop"
  )

species_bray <- bray_transitions %>%
  group_by(Species_ID) %>%
  summarise(
    n_transitions = n(),
    median_bray = median(bray_dist, na.rm = TRUE),
    .groups = "drop"
  )

analysis_df <- species_covars %>%
  left_join(species_bray, by = "Species_ID") %>%
  filter(
    is.finite(n_dates), n_dates >= 5,
    is.finite(median_prot_count), median_prot_count > 0,
    is.finite(median_TPF), median_TPF > 0, median_TPF < 1,
    is.finite(median_bray), median_bray >= 0, median_bray <= 1
  ) %>%
  left_join(
    species_taxonomy %>% select(Species_ID, Phylum, Genus) %>% distinct(),
    by = "Species_ID"
  ) %>%
  mutate(
    Group_For_Plots = factor(Group_For_Plots, levels = trophic_order),
    Genus2 = ifelse(Genus %in% c("Limnohabitans", "Limnohabitans_A"), "Limnohabitans", Genus),
    highlight_label = case_when(
      Genus2 == "Flavobacterium" ~ "Flavobacterium",
      Genus2 == "Planktophila"   ~ "Planktophila",
      Genus2 == "Fonsibacter"    ~ "Fonsibacter",
      Phylum == "Planctomycetota" ~ "Planctomycetota (phy)",
      TRUE ~ "Rest"
    ),
    highlight_label = factor(highlight_label, levels = highlight_levels)
  ) %>%
  distinct(Species_ID, .keep_all = TRUE)

if (anyDuplicated(analysis_df$Species_ID) > 0) stop("Duplicate Species_ID in analysis_df (unexpected).")

cat("\n=== MEDIAN vs MEDIAN analysis_df ===\n")
cat("Total N dots:", nrow(analysis_df), "\n")
cat("N per trophic category:\n"); print(table(analysis_df$Group_For_Plots))
cat("N per trophic x highlight:\n"); print(analysis_df %>% count(Group_For_Plots, highlight_label))

# =========================================================
# 6) Figure A (median vs median): scatter + marginals
#     X: median TPF
#     Y: median variability (Bray–Curtis dissimilarity)
#     Axes fixed 0..1 by 0.25
#     Inference: logit(y) ~ log10(x)
# =========================================================
eps <- 1e-6
analysis_df <- analysis_df %>%
  mutate(
    x_med = median_TPF,
    y_med = pmin(pmax(median_bray, eps), 1 - eps),
    logit_y = qlogis(y_med),
    log10_x = log10(x_med)
  )

m_logit <- lm(logit_y ~ log10_x, data = analysis_df)
b <- coef(summary(m_logit))
beta  <- b["log10_x", "Estimate"]
pval  <- b["log10_x", "Pr(>|t|)"]
ci    <- confint(m_logit, "log10_x", level = 0.95)
ci_lo <- ci[1]; ci_hi <- ci[2]
N     <- nrow(analysis_df)

ann_txt <- paste0(
  "Inference (logit–log model):\n",
  "logit(BC) ~ log10(median TPF)\n",
  "β = ", round(beta, 3),
  "  95% CI [", round(ci_lo, 3), ", ", round(ci_hi, 3), "]\n",
  "p = ", fmt_p(pval), "   N = ", N
)

# Size legend breaks (hardcoded style used previously: min + 1/3 + 2/3 + max)
size_breaks <- analysis_df %>%
  summarise(
    mn = as.integer(round(min(median_prot_count, na.rm = TRUE))),
    q1 = as.integer(round(quantile(median_prot_count, 1/3, na.rm = TRUE))),
    q2 = as.integer(round(quantile(median_prot_count, 2/3, na.rm = TRUE))),
    mx = as.integer(round(max(median_prot_count, na.rm = TRUE)))
  ) %>%
  unlist(use.names = FALSE) %>%
  unique() %>%
  sort()

if (length(size_breaks) < 4) {
  mn <- min(analysis_df$median_prot_count, na.rm = TRUE)
  mx <- max(analysis_df$median_prot_count, na.rm = TRUE)
  size_breaks <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
}

size_labels <- as.character(size_breaks)
size_labels[1] <- paste0(size_labels[1], " (min)")
size_labels[length(size_labels)] <- paste0(size_labels[length(size_labels)], " (max)")

pA_base <- ggplot(analysis_df, aes(x = x_med, y = median_bray, color = Group_For_Plots)) +
  geom_point(aes(size = median_prot_count), alpha = 0.85) +
  geom_smooth(
    inherit.aes = FALSE,
    data = analysis_df,
    aes(x = x_med, y = median_bray),
    method = "lm",
    se = TRUE,
    color = "black",
    linewidth = 0.9
  ) +
  scale_color_manual(
    values = trophic_cols,
    drop = FALSE,
    labels = c("Oligo-like", "Intermediate", "Copio-like")
  ) +
  scale_size_continuous(
    name   = "Median proteins detected (count)",
    range  = c(2, 10),
    breaks = size_breaks,
    labels = size_labels
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  scale_y_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  labs(
    title = "Higher transporter investment predicts functional stability",
    subtitle = "Lower Bray–Curtis dissimilarity indicates higher stability; inference uses logit(dissimilarity) ~ log10(TPF)",
    x = "Median transporter proteome fraction (TPF)",
    y = "Median variability (Bray–Curtis dissimilarity)",
    color = "Trophic strategy"
  ) +
  annotate(
    "label",
    x = Inf, y = Inf,
    hjust = 1.02, vjust = 1.02,
    label = ann_txt,
    size = 3.2,
    label.size = 0.25
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  theme(aspect.ratio = 1)

pA_noleg <- pA_base + theme(legend.position = "none")
pA_marg <- ggMarginal(
  pA_noleg,
  type = "density",
  margins = "both",
  groupColour = TRUE,
  groupFill   = TRUE,
  alpha = 0.25,
  size = 4
)
gA_plot <- grid::grid.grabExpr(print(pA_marg))

gA_size  <- get_legend(pA_base + guides(color = "none"))
gA_color <- get_legend(pA_base + guides(size = "none"))

A_leg_col <- arrangeGrob(
  gA_size,
  gA_color,
  ncol = 1,
  heights = c(1.2, 0.8)
)

FIG_A_MEDMED <- arrangeGrob(
  gA_plot,
  A_leg_col,
  ncol = 2,
  widths = c(3.3, 1.2)
)

grid.newpage(); grid.draw(FIG_A_MEDMED)
export_grob_pdf(FIG_A_MEDMED, base = "Figure_A_MEDIANxMEDIAN_BrayCurtisDissim_vs_TPF_A4_landscape",
                width = 11.69, height = 8.27)

# =========================================================
# 7) Figure B/C (median only): stacked boxplots + highlight legend
#     Axes fixed 0..1 by 0.25
# =========================================================
star <- function(p) {
  if (is.na(p)) return("ns")
  if (p < 1e-4) "****" else
    if (p < 1e-3) "***" else
      if (p < 1e-2) "**" else
        if (p < 5e-2) "*" else "ns"
}
pairwise_wilcox_tidy <- function(x, g) {
  pw <- pairwise.wilcox.test(x, g, p.adjust.method = "BH", exact = FALSE)
  m <- pw$p.value
  out <- data.frame()
  if (!is.null(m)) {
    rn <- rownames(m); cn <- colnames(m)
    for (i in seq_along(rn)) for (j in seq_along(cn)) {
      p <- m[i, j]
      if (!is.na(p)) out <- rbind(out, data.frame(group1 = rn[i], group2 = cn[j], p_adj = p))
    }
  }
  out
}
get_pw <- function(tbl, a, b) {
  p <- tbl$p_adj[(tbl$group1 == a & tbl$group2 == b) | (tbl$group1 == b & tbl$group2 == a)]
  if (length(p) == 0) NA_real_ else p[1]
}

# inference on log10(values) as before (reviewer-safe); clamp to avoid -Inf
analysis_df <- analysis_df %>%
  mutate(
    log_med_tpf  = log10(pmin(pmax(median_TPF, eps), 1 - eps)),
    log_med_bray = log10(pmin(pmax(median_bray, eps), 1 - eps))
  )

kw_med_tpf <- kruskal.test(log_med_tpf ~ Group_For_Plots, data = analysis_df)
pw_med_tpf <- pairwise_wilcox_tidy(analysis_df$log_med_tpf, analysis_df$Group_For_Plots)
p_med_OI <- get_pw(pw_med_tpf, "Oligo_like", "Intermediate")
p_med_OC <- get_pw(pw_med_tpf, "Oligo_like", "Copio_like")
p_med_IC <- get_pw(pw_med_tpf, "Intermediate", "Copio_like")

kw_med_bray <- kruskal.test(log_med_bray ~ Group_For_Plots, data = analysis_df)
pw_med_bray <- pairwise_wilcox_tidy(analysis_df$log_med_bray, analysis_df$Group_For_Plots)
p_bray_OI <- get_pw(pw_med_bray, "Oligo_like", "Intermediate")
p_bray_OC <- get_pw(pw_med_bray, "Oligo_like", "Copio_like")
p_bray_IC <- get_pw(pw_med_bray, "Intermediate", "Copio_like")

# IMPORTANT: do NOT set coord_cartesian() inside this function;
# we control headroom in the plot with coord_cartesian(ylim=..., clip="off").
add_sig_brackets <- function(p, y_max, p12, p13, p23,
                             mults = c(1.10, 1.22, 1.34),
                             tick_frac = 0.03,
                             text_mult = 1.015) {
  t <- y_max * tick_frac
  p +
    geom_segment(aes(x=1, xend=2, y=y_max*mults[1], yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    annotate("text", x=1.5, y=y_max*mults[1]*text_mult, label=star(p12), size=4) +
    geom_segment(aes(x=1, xend=3, y=y_max*mults[2], yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    annotate("text", x=2.0, y=y_max*mults[2]*text_mult, label=star(p13), size=4) +
    geom_segment(aes(x=2, xend=3, y=y_max*mults[3], yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    annotate("text", x=2.5, y=y_max*mults[3]*text_mult, label=star(p23), size=4)
}

jit <- position_jitter(width = 0.3, height = 0, seed = 1)

clean_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.25),
    panel.background = element_rect(fill = NA, colour = NA),
    plot.background  = element_rect(fill = NA, colour = NA),
    legend.position = "none",
    # extra headroom so brackets/stars are never clipped by the grob layout
    plot.margin = margin(28, 10, 10, 10)
  )

pt_size <- 8
pt_stroke <- 0.35

# Headroom for brackets (keeps axis ticks 0..1 but allows drawing above 1)
ylim_headroom <- c(0, 1.18)

# B: median TPF
pB <- ggplot(analysis_df, aes(x = Group_For_Plots, y = median_TPF)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9, size = pt_size
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "B  Median transporter allocation differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_med_tpf$p.value)),
    y = "Median transporter proteome fraction (TPF)"
  ) +
  clean_theme

yBmax <- 1
pB <- add_sig_brackets(pB, yBmax, p_med_OI, p_med_OC, p_med_IC,
                       mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

# C: median Bray–Curtis dissimilarity (variability)
pC <- ggplot(analysis_df, aes(x = Group_For_Plots, y = median_bray)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9, size = pt_size
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "C  Functional variability differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_med_bray$p.value)),
    y = "Median variability (Bray–Curtis dissimilarity)"
  ) +
  clean_theme

yCmax <- 1
pC <- add_sig_brackets(pC, yCmax, p_bray_OI, p_bray_OC, p_bray_IC,
                       mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

# Shared highlight legend (single column)
legend_df <- data.frame(highlight_label = factor(highlight_levels, levels = highlight_levels))

pLEG <- ggplot(legend_df, aes(x = highlight_label, y = 1, fill = highlight_label)) +
  geom_point(shape = 21, size = 5, color = "black", stroke = 0.5) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  guides(fill = guide_legend(title = NULL, ncol = 1, byrow = TRUE)) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.spacing.y = unit(0.3, "lines")
  )

gLEG <- get_legend(pLEG)

BC_stack <- arrangeGrob(
  ggplotGrob(pB),
  ggplotGrob(pC),
  ncol = 1,
  heights = c(1, 1)
)

FIG_BC_MEDMED <- arrangeGrob(
  BC_stack,
  gLEG,
  ncol = 2,
  widths = c(3.0, 1.0)
)

grid.newpage(); grid.draw(FIG_BC_MEDMED)
export_grob_pdf(FIG_BC_MEDMED, base = "Figure_BC_MEDIANxMEDIAN_BrayCurtisDissim_and_TPF_A4_landscape",
                width = 11.69, height = 8.27)

# Final sanity prints
cat("\n=== FINAL SANITY (median vs median) ===\n")
cat("Total dots:", nrow(analysis_df), "\n")
cat("Dots per trophic:\n"); print(table(analysis_df$Group_For_Plots))
cat("Trophic x highlight:\n"); print(analysis_df %>% count(Group_For_Plots, highlight_label))
```

# mean tpf vs mean bray curtis dissimilarity
```{r}
############################################################
# MEAN vs MEAN ONLY
# - x: mean TPF per species
# - y: mean variability = mean Bray–Curtis dissimilarity
#      (lower = more stable; higher = more variable)
# - Hardcoded axes: 0..1 by 0.25 for scatter + boxplots
# - Same plotting style + highlight legend + PDF export only
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggExtra)
  library(grid)
  library(gridExtra)
  library(readr)
  library(vegan)
})

# =========================================================
# 0) Safety checks — objects expected from your load block
# =========================================================
required_objects <- c("protein_final", "retained_species_dates", "species_taxonomy")
for (obj in required_objects) {
  if (!exists(obj, inherits = TRUE)) stop("Object not found in environment: ", obj)
}

req_keys_cols <- c("Species_ID","Date","Group_For_Plots")
missing_keys <- setdiff(req_keys_cols, names(retained_species_dates))
if (length(missing_keys) > 0) stop("retained_species_dates missing: ", paste(missing_keys, collapse=", "))

req_tax_cols <- c("Species_ID","Phylum","Genus")
missing_tax <- setdiff(req_tax_cols, names(species_taxonomy))
if (length(missing_tax) > 0) stop("species_taxonomy missing: ", paste(missing_tax, collapse=", "))

req_prot_cols <- c("Species_ID","Date","Group_For_Plots","Protein_Abundance_IBAQ_Norm","KEGG_Subcategory","KEGG_ID")
missing_prot <- setdiff(req_prot_cols, names(protein_final))
if (length(missing_prot) > 0) stop("protein_final missing: ", paste(missing_prot, collapse=", "))

protein_id_col <- dplyr::case_when(
  "Protein_Group" %in% names(protein_final) ~ "Protein_Group",
  "Protein_ID"    %in% names(protein_final) ~ "Protein_ID",
  TRUE ~ NA_character_
)
if (is.na(protein_id_col)) stop("protein_final must have Protein_Group or Protein_ID for counting detected proteins.")

protein_final <- protein_final %>% mutate(Date = as.Date(Date))
retained_species_dates <- retained_species_dates %>% mutate(Date = as.Date(Date))

# =========================================================
# 1) Fixed settings (orders, colors, highlight scheme)
# =========================================================
trophic_order <- c("Oligo_like", "Intermediate", "Copio_like")
trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

highlight_cols <- c(
  "Flavobacterium"        = "#E6BE1E",
  "Planktophila"          = "#ADD5E4",
  "Fonsibacter"           = "#C6FC05",
  "Planctomycetota (phy)" = "#7B3294",
  "Rest"                  = "grey"
)
highlight_levels <- names(highlight_cols)

axis_breaks_01 <- seq(0, 1, by = 0.25)

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) "<0.001" else format(p, digits = 2, scientific = TRUE)
}

get_legend <- function(p) {
  g <- ggplotGrob(p)
  idx <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  if (length(idx) != 1) stop("Legend extraction failed.")
  g$grobs[[idx]]
}

export_grob_pdf <- function(g, base, width, height) {
  pdf(paste0(base, ".pdf"), width = width, height = height)
  grid.newpage(); grid.draw(g)
  dev.off()
  cat("\nSaved PDF:\n  ", normalizePath(paste0(base, ".pdf")), "\n", sep="")
}

# =========================================================
# 2) Enforce Species×Date×Group universe (critical)
# =========================================================
protein_final_u <- protein_final %>%
  semi_join(retained_species_dates, by = c("Species_ID","Date","Group_For_Plots"))

cat("\n=== Universe enforcement ===\n")
cat("protein_final_u rows after semi_join:", nrow(protein_final_u), "\n")

# =========================================================
# 3) Compute TPF per Species×Date
#    Transporters: KEGG_Subcategory contains whole word "Transporters"
# =========================================================
protein_tagged <- protein_final_u %>%
  mutate(
    is_transporter = !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, "\\bTransporters\\b"),
    abund = Protein_Abundance_IBAQ_Norm
  )

species_date_tpf <- protein_tagged %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    abund_total = sum(abund, na.rm = TRUE),
    abund_trans = sum(abund[is_transporter], na.rm = TRUE),
    TPF = abund_trans / abund_total,
    N_prot_total = n_distinct(.data[[protein_id_col]]),
    .groups = "drop"
  ) %>%
  filter(is.finite(TPF), abund_total > 0, TPF > 0, TPF < 1, N_prot_total > 0)

cat("\n=== species_date_tpf ===\n")
cat("Rows:", nrow(species_date_tpf), "\n")

# =========================================================
# 4) Compute transporter-only KEGG intensity profiles per Species×Date
#    (summed iBAQ per KEGG_ID) and consecutive-date Bray–Curtis DISTANCE
# =========================================================
kegg_sd <- protein_tagged %>%
  filter(
    is_transporter,
    !is.na(KEGG_ID)
  ) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

compute_species_consecutive_braydist <- function(sp_id, kegg_sd_tbl, keep_dates_tbl) {

  dates <- keep_dates_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_sd_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  if (nrow(mat_wide) < 2) return(NULL)

  M <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M) <- mat_wide$Date

  pairs <- data.frame(
    Date1 = dates[-length(dates)],
    Date2 = dates[-1],
    stringsAsFactors = FALSE
  )

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i])
    d2 <- as.character(pairs$Date2[i])

    bc <- vegan::vegdist(M[c(d1, d2), , drop = FALSE], method = "bray")
    bc_dist <- as.numeric(bc)

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      bray_dist = bc_dist,
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

keep_dates <- species_date_tpf %>% select(Species_ID, Date, Group_For_Plots) %>% distinct()
all_species <- sort(unique(keep_dates$Species_ID))

bray_transitions <- bind_rows(lapply(
  all_species,
  compute_species_consecutive_braydist,
  kegg_sd_tbl = kegg_sd,
  keep_dates_tbl = keep_dates
))

cat("\n=== bray_transitions ===\n")
cat("Rows:", nrow(bray_transitions), "\n")

# =========================================================
# 5) Species-level means (ONE DOT PER SPECIES)
#    - mean_TPF across dates
#    - mean_bray across consecutive transitions
#    - mean proteins detected (size covariate)
#    - taxonomy + highlight label
# =========================================================
species_covars <- species_date_tpf %>%
  group_by(Species_ID, Group_For_Plots) %>%
  summarise(
    n_dates = n(),
    mean_TPF = mean(TPF, na.rm = TRUE),
    mean_prot_count = as.integer(round(mean(N_prot_total, na.rm = TRUE))),
    .groups = "drop"
  )

species_bray <- bray_transitions %>%
  group_by(Species_ID) %>%
  summarise(
    n_transitions = n(),
    mean_bray = mean(bray_dist, na.rm = TRUE),
    .groups = "drop"
  )

analysis_df <- species_covars %>%
  left_join(species_bray, by = "Species_ID") %>%
  filter(
    is.finite(n_dates), n_dates >= 5,
    is.finite(mean_prot_count), mean_prot_count > 0,
    is.finite(mean_TPF), mean_TPF > 0, mean_TPF < 1,
    is.finite(mean_bray), mean_bray >= 0, mean_bray <= 1
  ) %>%
  left_join(
    species_taxonomy %>% select(Species_ID, Phylum, Genus) %>% distinct(),
    by = "Species_ID"
  ) %>%
  mutate(
    Group_For_Plots = factor(Group_For_Plots, levels = trophic_order),
    Genus2 = ifelse(Genus %in% c("Limnohabitans", "Limnohabitans_A"), "Limnohabitans", Genus),
    highlight_label = case_when(
      Genus2 == "Flavobacterium" ~ "Flavobacterium",
      Genus2 == "Planktophila"   ~ "Planktophila",
      Genus2 == "Fonsibacter"    ~ "Fonsibacter",
      Phylum == "Planctomycetota" ~ "Planctomycetota (phy)",
      TRUE ~ "Rest"
    ),
    highlight_label = factor(highlight_label, levels = highlight_levels)
  ) %>%
  distinct(Species_ID, .keep_all = TRUE)

if (anyDuplicated(analysis_df$Species_ID) > 0) stop("Duplicate Species_ID in analysis_df (unexpected).")

cat("\n=== MEAN vs MEAN analysis_df ===\n")
cat("Total N dots:", nrow(analysis_df), "\n")
cat("N per trophic category:\n"); print(table(analysis_df$Group_For_Plots))
cat("N per trophic x highlight:\n"); print(analysis_df %>% count(Group_For_Plots, highlight_label))

# =========================================================
# 6) Figure A (mean vs mean): scatter + marginals
#     X: mean TPF
#     Y: mean variability (Bray–Curtis dissimilarity)
#     Axes fixed 0..1 by 0.25
#     Inference: logit(y) ~ log10(x)
# =========================================================
eps <- 1e-6
analysis_df <- analysis_df %>%
  mutate(
    x_mean = mean_TPF,
    y_mean = pmin(pmax(mean_bray, eps), 1 - eps),
    logit_y = qlogis(y_mean),
    log10_x = log10(x_mean)
  )

m_logit <- lm(logit_y ~ log10_x, data = analysis_df)
b <- coef(summary(m_logit))
beta  <- b["log10_x", "Estimate"]
pval  <- b["log10_x", "Pr(>|t|)"]
ci    <- confint(m_logit, "log10_x", level = 0.95)
ci_lo <- ci[1]; ci_hi <- ci[2]
N     <- nrow(analysis_df)

ann_txt <- paste0(
  "Inference (logit–log model):\n",
  "logit(BC) ~ log10(mean TPF)\n",
  "β = ", round(beta, 3),
  "  95% CI [", round(ci_lo, 3), ", ", round(ci_hi, 3), "]\n",
  "p = ", fmt_p(pval), "   N = ", N
)

# Size legend breaks (hardcoded style used previously: min + 1/3 + 2/3 + max)
size_breaks <- analysis_df %>%
  summarise(
    mn = as.integer(round(min(mean_prot_count, na.rm = TRUE))),
    q1 = as.integer(round(quantile(mean_prot_count, 1/3, na.rm = TRUE))),
    q2 = as.integer(round(quantile(mean_prot_count, 2/3, na.rm = TRUE))),
    mx = as.integer(round(max(mean_prot_count, na.rm = TRUE)))
  ) %>%
  unlist(use.names = FALSE) %>%
  unique() %>%
  sort()

if (length(size_breaks) < 4) {
  mn <- min(analysis_df$mean_prot_count, na.rm = TRUE)
  mx <- max(analysis_df$mean_prot_count, na.rm = TRUE)
  size_breaks <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
}

size_labels <- as.character(size_breaks)
size_labels[1] <- paste0(size_labels[1], " (min)")
size_labels[length(size_labels)] <- paste0(size_labels[length(size_labels)], " (max)")

pA_base <- ggplot(analysis_df, aes(x = x_mean, y = mean_bray, color = Group_For_Plots)) +
  geom_point(aes(size = mean_prot_count), alpha = 0.85) +
  geom_smooth(
    inherit.aes = FALSE,
    data = analysis_df,
    aes(x = x_mean, y = mean_bray),
    method = "lm",
    se = TRUE,
    color = "black",
    linewidth = 0.9
  ) +
  scale_color_manual(
    values = trophic_cols,
    drop = FALSE,
    labels = c("Oligo-like", "Intermediate", "Copio-like")
  ) +
  scale_size_continuous(
    name   = "Mean proteins detected (count)",
    range  = c(2, 10),
    breaks = size_breaks,
    labels = size_labels
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  scale_y_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  labs(
    title = "Higher transporter investment predicts functional stability",
    subtitle = "Lower Bray–Curtis dissimilarity indicates higher stability; inference uses logit(dissimilarity) ~ log10(TPF)",
    x = "Mean transporter proteome fraction (TPF)",
    y = "Mean variability (Bray–Curtis dissimilarity)",
    color = "Trophic strategy"
  ) +
  annotate(
    "label",
    x = Inf, y = Inf,
    hjust = 1.02, vjust = 1.02,
    label = ann_txt,
    size = 3.2,
    label.size = 0.25
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "right"
  ) +
  theme(aspect.ratio = 1)

pA_noleg <- pA_base + theme(legend.position = "none")
pA_marg <- ggMarginal(
  pA_noleg,
  type = "density",
  margins = "both",
  groupColour = TRUE,
  groupFill   = TRUE,
  alpha = 0.25,
  size = 4
)
gA_plot <- grid::grid.grabExpr(print(pA_marg))

gA_size  <- get_legend(pA_base + guides(color = "none"))
gA_color <- get_legend(pA_base + guides(size = "none"))

A_leg_col <- arrangeGrob(
  gA_size,
  gA_color,
  ncol = 1,
  heights = c(1.2, 0.8)
)

FIG_A_MEANMEAN <- arrangeGrob(
  gA_plot,
  A_leg_col,
  ncol = 2,
  widths = c(3.3, 1.2)
)

# grid.newpage(); grid.draw(FIG_A_MEANMEAN)
# export_grob_pdf(FIG_A_MEANMEAN, base = "Figure_A_MEANxMEAN_BrayCurtisDissim_vs_TPF_A4_landscape",
#                 width = 11.69, height = 8.27)

# =========================================================
# 7) Figure B/C (mean only): stacked boxplots + highlight legend
#     Axes fixed 0..1 by 0.25
# =========================================================
add_sig_brackets <- function(p, y_max, p12, p13, p23,
                             mults = c(1.10, 1.22, 1.34),
                             tick_frac = 0.03,
                             text_mult = 1.015) {
  t <- y_max * tick_frac
  p +
    geom_segment(aes(x=1, xend=2, y=y_max*mults[1], yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    annotate("text", x=1.5, y=y_max*mults[1]*text_mult, label=star(p12), size=4) +
    geom_segment(aes(x=1, xend=3, y=y_max*mults[2], yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    annotate("text", x=2.0, y=y_max*mults[2]*text_mult, label=star(p13), size=4) +
    geom_segment(aes(x=2, xend=3, y=y_max*mults[3], yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    annotate("text", x=2.5, y=y_max*mults[3]*text_mult, label=star(p23), size=4)
}

jit <- position_jitter(width = 0.3, height = 0, seed = 1)

clean_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.25),
    panel.background = element_rect(fill = NA, colour = NA),
    plot.background  = element_rect(fill = NA, colour = NA),
    legend.position = "none",
    plot.margin = margin(28, 10, 10, 10)
  )

pt_size <- 8
pt_stroke <- 0.35
ylim_headroom <- c(0, 1.18)

analysis_df <- analysis_df %>%
  mutate(
    log_mean_tpf  = log10(pmin(pmax(mean_TPF, eps), 1 - eps)),
    log_mean_bray = log10(pmin(pmax(mean_bray, eps), 1 - eps))
  )

kw_mean_tpf <- kruskal.test(log_mean_tpf ~ Group_For_Plots, data = analysis_df)
pw_mean_tpf <- pairwise_wilcox_tidy(analysis_df$log_mean_tpf, analysis_df$Group_For_Plots)
p_mean_OI <- get_pw(pw_mean_tpf, "Oligo_like", "Intermediate")
p_mean_OC <- get_pw(pw_mean_tpf, "Oligo_like", "Copio_like")
p_mean_IC <- get_pw(pw_mean_tpf, "Intermediate", "Copio_like")

kw_mean_bray <- kruskal.test(log_mean_bray ~ Group_For_Plots, data = analysis_df)
pw_mean_bray <- pairwise_wilcox_tidy(analysis_df$log_mean_bray, analysis_df$Group_For_Plots)
p_mbray_OI <- get_pw(pw_mean_bray, "Oligo_like", "Intermediate")
p_mbray_OC <- get_pw(pw_mean_bray, "Oligo_like", "Copio_like")
p_mbray_IC <- get_pw(pw_mean_bray, "Intermediate", "Copio_like")

pB <- ggplot(analysis_df, aes(x = Group_For_Plots, y = mean_TPF)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9, size = pt_size
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "B  Mean transporter allocation differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_mean_tpf$p.value)),
    y = "Mean transporter proteome fraction (TPF)"
  ) +
  clean_theme

pB <- add_sig_brackets(pB, 1, p_mean_OI, p_mean_OC, p_mean_IC,
                       mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

pC <- ggplot(analysis_df, aes(x = Group_For_Plots, y = mean_bray)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9, size = pt_size
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "C  Functional variability differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_mean_bray$p.value)),
    y = "Mean variability (Bray–Curtis dissimilarity)"
  ) +
  clean_theme

pC <- add_sig_brackets(pC, 1, p_mbray_OI, p_mbray_OC, p_mbray_IC,
                       mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

legend_df <- data.frame(highlight_label = factor(highlight_levels, levels = highlight_levels))
pLEG <- ggplot(legend_df, aes(x = highlight_label, y = 1, fill = highlight_label)) +
  geom_point(shape = 21, size = 5, color = "black", stroke = 0.5) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  guides(fill = guide_legend(title = NULL, ncol = 1, byrow = TRUE)) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.spacing.y = unit(0.3, "lines")
  )
gLEG <- get_legend(pLEG)

BC_stack <- arrangeGrob(
  ggplotGrob(pB),
  ggplotGrob(pC),
  ncol = 1,
  heights = c(1, 1)
)

FIG_BC_MEANMEAN <- arrangeGrob(
  BC_stack,
  gLEG,
  ncol = 2,
  widths = c(3.0, 1.0)
)

# grid.newpage(); grid.draw(FIG_BC_MEANMEAN)
# export_grob_pdf(FIG_BC_MEANMEAN, base = "Figure_BC_MEANxMEAN_BrayCurtisDissim_and_TPF_A4_landscape",
#                 width = 11.69, height = 8.27)

cat("\n=== FINAL SANITY (mean vs mean) ===\n")
cat("Total dots:", nrow(analysis_df), "\n")
cat("Dots per trophic:\n"); print(table(analysis_df$Group_For_Plots))
cat("Trophic x highlight:\n"); print(analysis_df %>% count(Group_For_Plots, highlight_label))
```


```{r}
############################################################
# Main Figures
#
#
# INPUT OBJECTS (assumed already loaded in your session):
#   - protein_final
#   - retained_species_dates
#   - species_taxonomy
#
# OUTPUT:
#   - Figure_A_MEANxMEAN_BrayCurtisDissim_vs_TPF_A4_landscape.pdf
#   - Figure_BC_MEANxMEAN_BrayCurtisDissim_and_TPF_A4_landscape.pdf
#
# CORE IDEA
#   Fig A:
#     - One dot = one species
#     - x = mean transporter proteome fraction (TPF)
#     - y = mean variability (mean Bray–Curtis dissimilarity across consecutive dates)
#     - Marginal densities grouped by trophic strategy (as in your original)
#     - Dot color scheme:
#         * keep the 3 trophic groups
#         * WITHIN each group: gradient by mean trophic index
#           (darker = lower trophic index, lighter = higher)
#     - Curve:
#         * We show the model-implied curve from the INFERENCE model
#           logit(BC) ~ log10(TPF)
#         * The curve is predicted on the transformed scale and then
#           back-transformed to BC scale via plogis()
#         * NO confidence ribbon is shown (to avoid mixing scales visually)
#     - NO dot-size encoding (proteins now shown in Fig B/C)
#
#   Fig B/C:
#     - Boxplots by trophic strategy
#     - Overlaid jittered points:
#         * fill = highlight taxon group
#         * size = mean proteins detected per species (mean across its dates)
#     - Right-side combined legend includes:
#         * Mean proteins detected (size legend)
#         * Highlight colors legend
#
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(stringr)
  library(tidyr)
  library(ggplot2)
  library(scales)
  library(ggExtra)
  library(grid)
  library(gridExtra)
  library(vegan)
})

# =========================================================
# Helper utilities (plot export, legends, stats helpers)
# =========================================================

export_grob_pdf <- function(g, base, width, height) {
  pdf(paste0(base, ".pdf"), width = width, height = height)
  grid.newpage(); grid.draw(g)
  dev.off()
  cat("\nSaved PDF:\n  ", normalizePath(paste0(base, ".pdf")), "\n", sep = "")
}

get_legend <- function(p) {
  g <- ggplotGrob(p)
  idx <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  if (length(idx) != 1) stop("Legend extraction failed.")
  g$grobs[[idx]]
}

fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-3) "<0.001" else format(p, digits = 2, scientific = TRUE)
}

star <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 1e-4) "****"
  else if (p < 1e-3) "***"
  else if (p < 1e-2) "**"
  else if (p < 5e-2) "*"
  else "ns"
}

pairwise_wilcox_tidy <- function(x, g, p_adjust = "BH") {
  g <- droplevels(as.factor(g))
  pw <- pairwise.wilcox.test(x, g, p.adjust.method = p_adjust, exact = FALSE)
  M  <- pw$p.value
  if (is.null(M)) return(tibble(g1=character(), g2=character(), p=double()))
  as.data.frame(as.table(M), stringsAsFactors = FALSE) %>%
    rename(g1 = Var1, g2 = Var2, p = Freq) %>%
    filter(!is.na(p)) %>%
    tibble::as_tibble()
}

get_pw <- function(pw_tbl, a, b) {
  hit1 <- pw_tbl %>% filter(g1 == a, g2 == b)
  hit2 <- pw_tbl %>% filter(g1 == b, g2 == a)
  if (nrow(hit1) == 1) return(hit1$p)
  if (nrow(hit2) == 1) return(hit2$p)
  NA_real_
}

add_sig_brackets <- function(p, y_max, p12, p13, p23,
                             mults = c(1.10, 1.22, 1.34),
                             tick_frac = 0.03,
                             text_mult = 1.015) {
  t <- y_max * tick_frac
  p +
    geom_segment(aes(x=1, xend=2, y=y_max*mults[1], yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[1]-t, yend=y_max*mults[1]), inherit.aes=FALSE) +
    annotate("text", x=1.5, y=y_max*mults[1]*text_mult, label=star(p12), size=4) +
    geom_segment(aes(x=1, xend=3, y=y_max*mults[2], yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=1, xend=1, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[2]-t, yend=y_max*mults[2]), inherit.aes=FALSE) +
    annotate("text", x=2.0, y=y_max*mults[2]*text_mult, label=star(p13), size=4) +
    geom_segment(aes(x=2, xend=3, y=y_max*mults[3], yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=2, xend=2, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    geom_segment(aes(x=3, xend=3, y=y_max*mults[3]-t, yend=y_max*mults[3]), inherit.aes=FALSE) +
    annotate("text", x=2.5, y=y_max*mults[3]*text_mult, label=star(p23), size=4)
}

# =========================================================
# Fixed settings (orders, colors, highlight scheme)
# =========================================================

trophic_order <- c("Oligo_like", "Intermediate", "Copio_like")

# Base colors (highest trophic index within group -> these colors)
trophic_cols <- c(
  "Oligo_like"   = "#93C2E6",
  "Intermediate" = "#EDEAE9",
  "Copio_like"   = "#BA5205"
)

# Dark endpoint colors (lowest trophic index within group -> these colors)
trophic_cols_dark <- c(
  "Oligo_like"   = "#1F4E79",
  "Intermediate" = "#4D4D4D",
  "Copio_like"   = "#5A2A0C"
)

highlight_cols <- c(
  "Flavobacterium"        = "#E6BE1E",
  "Planktophila"          = "#ADD5E4",
  "Fonsibacter"           = "#C6FC05",
  "Planctomycetota (phy)" = "#7B3294",
  "Rest"                  = "grey"
)
highlight_levels <- names(highlight_cols)

axis_breaks_01 <- seq(0, 1, by = 0.25)

# =========================================================
# 1) Enforce Species×Date×Group universe (critical)
# =========================================================

protein_final <- protein_final %>% mutate(Date = as.Date(Date))
retained_species_dates <- retained_species_dates %>% mutate(Date = as.Date(Date))

protein_final_u <- protein_final %>%
  semi_join(retained_species_dates, by = c("Species_ID","Date","Group_For_Plots"))

# =========================================================
# 2) Identify protein identifier column (Protein_Group preferred)
# =========================================================
protein_id_col <- if ("Protein_Group" %in% names(protein_final_u)) "Protein_Group" else "Protein_ID"

# =========================================================
# 3) Compute TPF per Species×Date
#    Transporters: KEGG_Subcategory contains whole word "Transporters"
# =========================================================

protein_tagged <- protein_final_u %>%
  mutate(
    is_transporter = !is.na(KEGG_Subcategory) & str_detect(KEGG_Subcategory, "\\bTransporters\\b"),
    abund = Protein_Abundance_IBAQ_Norm
  )

species_date_tpf <- protein_tagged %>%
  group_by(Species_ID, Date, Group_For_Plots) %>%
  summarise(
    abund_total = sum(abund, na.rm = TRUE),
    abund_trans = sum(abund[is_transporter], na.rm = TRUE),
    TPF = abund_trans / abund_total,
    N_prot_total = n_distinct(.data[[protein_id_col]]),
    .groups = "drop"
  ) %>%
  filter(is.finite(TPF), abund_total > 0, TPF > 0, TPF < 1, N_prot_total > 0)

# =========================================================
# 4) Transporter-only KEGG profiles and consecutive Bray–Curtis distances
# =========================================================

kegg_sd <- protein_tagged %>%
  filter(is_transporter, !is.na(KEGG_ID)) %>%
  group_by(Species_ID, Date, KEGG_ID) %>%
  summarise(kegg_abund = sum(Protein_Abundance_IBAQ_Norm, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(kegg_abund), kegg_abund > 0)

compute_species_consecutive_braydist <- function(sp_id, kegg_sd_tbl, keep_dates_tbl) {

  dates <- keep_dates_tbl %>%
    filter(Species_ID == sp_id) %>%
    distinct(Date) %>%
    arrange(Date) %>%
    pull(Date)

  if (length(dates) < 2) return(NULL)

  mat_wide <- kegg_sd_tbl %>%
    filter(Species_ID == sp_id) %>%
    mutate(Date = as.character(Date)) %>%
    pivot_wider(names_from = KEGG_ID, values_from = kegg_abund, values_fill = 0) %>%
    filter(Date %in% as.character(dates)) %>%
    arrange(match(Date, as.character(dates)))

  if (nrow(mat_wide) < 2) return(NULL)

  M <- as.matrix(mat_wide %>% select(-Species_ID, -Date))
  rownames(M) <- mat_wide$Date

  pairs <- data.frame(
    Date1 = dates[-length(dates)],
    Date2 = dates[-1],
    stringsAsFactors = FALSE
  )

  out <- lapply(seq_len(nrow(pairs)), function(i) {
    d1 <- as.character(pairs$Date1[i])
    d2 <- as.character(pairs$Date2[i])
    bc <- vegan::vegdist(M[c(d1, d2), , drop = FALSE], method = "bray")

    data.frame(
      Species_ID = sp_id,
      Date1 = pairs$Date1[i],
      Date2 = pairs$Date2[i],
      bray_dist = as.numeric(bc),
      stringsAsFactors = FALSE
    )
  })

  bind_rows(out)
}

keep_dates <- species_date_tpf %>% select(Species_ID, Date, Group_For_Plots) %>% distinct()
all_species <- sort(unique(keep_dates$Species_ID))

bray_transitions <- bind_rows(lapply(
  all_species,
  compute_species_consecutive_braydist,
  kegg_sd_tbl = kegg_sd,
  keep_dates_tbl = keep_dates
))

# =========================================================
# 5) Species-level means (ONE DOT PER SPECIES)
#    - mean TPF across dates
#    - mean Bray distance across consecutive transitions
#    - mean proteins detected across dates (for B/C dot sizes)
#    - mean trophic index across all records of the species
#    - taxonomy + highlight label
# =========================================================

species_covars <- species_date_tpf %>%
  group_by(Species_ID, Group_For_Plots) %>%
  summarise(
    n_dates = n(),
    mean_TPF = mean(TPF, na.rm = TRUE),
    mean_prot_count = as.integer(round(mean(N_prot_total, na.rm = TRUE))),
    .groups = "drop"
  )

species_bray <- bray_transitions %>%
  group_by(Species_ID) %>%
  summarise(
    n_transitions = n(),
    mean_bray = mean(bray_dist, na.rm = TRUE),
    .groups = "drop"
  )

species_ti <- protein_final_u %>%
  select(Species_ID, TrophicIndex) %>%
  group_by(Species_ID) %>%
  summarise(mean_trophic_index = mean(TrophicIndex, na.rm = TRUE), .groups = "drop")

analysis_df <- species_covars %>%
  left_join(species_bray, by = "Species_ID") %>%
  left_join(species_ti,  by = "Species_ID") %>%
  left_join(
    species_taxonomy %>% select(Species_ID, Phylum, Genus) %>% distinct(),
    by = "Species_ID"
  ) %>%
  mutate(
    Group_For_Plots = factor(Group_For_Plots, levels = trophic_order),
    Genus2 = ifelse(Genus %in% c("Limnohabitans", "Limnohabitans_A"), "Limnohabitans", Genus),
    highlight_label = case_when(
      Genus2 == "Flavobacterium"   ~ "Flavobacterium",
      Genus2 == "Planktophila"     ~ "Planktophila",
      Genus2 == "Fonsibacter"      ~ "Fonsibacter",
      Phylum == "Planctomycetota"  ~ "Planctomycetota (phy)",
      TRUE                         ~ "Rest"
    ),
    highlight_label = factor(highlight_label, levels = highlight_levels)
  ) %>%
  filter(
    is.finite(n_dates), n_dates >= 5,
    is.finite(mean_prot_count), mean_prot_count > 0,
    is.finite(mean_TPF), mean_TPF > 0, mean_TPF < 1,
    is.finite(mean_bray), mean_bray >= 0, mean_bray <= 1,
    is.finite(mean_trophic_index)
  ) %>%
  distinct(Species_ID, .keep_all = TRUE)

# =========================================================
# 6) FIGURE A
#    - scatter + marginal densities (as before)
#    - dots: gradient within each trophic group by trophic index
#    - curve: back-transformed prediction from logit(BC) ~ log10(TPF)
#    - dot size fixed (size = 8)
#    - no legend inside panel; one combined legend at right:
#        * trophic strategy
#        * gradients explanation (3 rows with gradient bars)
# =========================================================

eps <- 1e-6

# Prepare transformed variables for inference model
analysis_A <- analysis_df %>%
  mutate(
    x_mean  = mean_TPF,
    y_mean  = pmin(pmax(mean_bray, eps), 1 - eps),
    logit_y = qlogis(y_mean),
    log10_x = log10(x_mean)
  ) %>%
  filter(is.finite(logit_y), is.finite(log10_x), is.finite(x_mean), is.finite(y_mean))

# Fit inference model ON TRANSFORMED SCALE
m_logit <- lm(logit_y ~ log10_x, data = analysis_A)

# Report inference numbers (computed on transformed scale)
b <- coef(summary(m_logit))
beta  <- b["log10_x", "Estimate"]
pval  <- b["log10_x", "Pr(>|t|)"]
ci    <- confint(m_logit, "log10_x", level = 0.95)
ci_lo <- ci[1]; ci_hi <- ci[2]
N_A   <- nrow(analysis_A)

ann_txt <- paste0(
  "Inference (logit–log model):\n",
  "logit(BC) ~ log10(mean TPF)\n",
  "β = ", round(beta, 3),
  "  95% CI [", round(ci_lo, 3), ", ", round(ci_hi, 3), "]\n",
  "p = ", fmt_p(pval), "   N = ", N_A
)

# Build model-derived curve:
#   1) generate a grid of TPF values on the original x-scale
#   2) predict on the transformed scale (logit BC)
#   3) back-transform to BC using inverse-logit plogis()
x_min <- max(0.001, min(analysis_A$x_mean, na.rm = TRUE))
x_max <- min(0.999, max(analysis_A$x_mean, na.rm = TRUE))
x_grid <- seq(x_min, x_max, length.out = 300)

pred_df <- data.frame(x_mean = x_grid) %>%
  mutate(log10_x = log10(x_mean))

pred_df$pred_bc <- plogis(predict(m_logit, newdata = pred_df))

# Within-group gradient (darker = lower trophic index)
analysis_A <- analysis_A %>%
  group_by(Group_For_Plots) %>%
  mutate(ti_scaled = scales::rescale(mean_trophic_index, to = c(0, 1))) %>%
  ungroup()

ramp_col <- function(group_name, t) {
  scales::colour_ramp(c(trophic_cols_dark[[group_name]], trophic_cols[[group_name]]))(t)
}

analysis_A$fill_ti <- mapply(
  ramp_col,
  as.character(analysis_A$Group_For_Plots),
  analysis_A$ti_scaled
)

# Base scatter (legend removed here; we add a single external legend)
pA_base <- ggplot(analysis_A, aes(x = x_mean, y = mean_bray, color = Group_For_Plots)) +
  geom_point(
    aes(fill = fill_ti),
    shape = 21,
    stroke = 0,
    size = 8,
    alpha = 0.90
  ) +
  # Discrete color mapping is kept for:
  #   - consistent grouping
  #   - marginal density colors
  scale_color_manual(
    values = trophic_cols,
    drop = FALSE,
    labels = c("Oligo-like", "Intermediate", "Copio-like")
  ) +
  scale_fill_identity() +
  geom_line(
    data = pred_df,
    aes(x = x_mean, y = pred_bc),
    inherit.aes = FALSE,
    color = "black",
    linewidth = 0.9
  ) +
  scale_x_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  scale_y_continuous(limits = c(0, 1), breaks = axis_breaks_01) +
  labs(
    title = "Higher transporter investment predicts functional stability",
    subtitle = "Lower Bray–Curtis dissimilarity indicates higher stability; black curve is model-implied (no CI ribbon).",
    x = "Mean transporter proteome fraction (TPF)",
    y = "Mean variability (Bray–Curtis dissimilarity)",
    color = "Trophic strategy"
  ) +
  annotate(
    "label",
    x = Inf, y = Inf,
    hjust = 1.02, vjust = 1.02,
    label = ann_txt,
    size = 3.2,
    label.size = 0.25
  ) +
  theme_bw() +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    legend.position = "none"   # remove per-plot legend
  ) +
  theme(aspect.ratio = 1)

# Add marginal densities (keep as your original: colored+filled by trophic strategy)
pA_marg <- ggMarginal(
  pA_base,
  type = "density",
  margins = "both",
  groupColour = TRUE,
  groupFill   = TRUE,
  alpha = 0.25,
  size  = 4
)

gA_plot <- grid::grid.grabExpr(print(pA_marg))

# ---- External combined legend (RIGHT SIDE) ----
#  (1) Trophic strategy legend
legend_strategy_df <- data.frame(Group_For_Plots = factor(trophic_order, levels = trophic_order), y = 1)
pLEG_strategy <- ggplot(legend_strategy_df, aes(x = Group_For_Plots, y = y, color = Group_For_Plots)) +
  geom_point(size = 5) +
  scale_color_manual(
    values = trophic_cols,
    drop = FALSE,
    labels = c("Oligo-like", "Intermediate", "Copio-like")
  ) +
  guides(color = guide_legend(title = "Trophic strategy", ncol = 1, byrow = TRUE)) +
  theme_void() +
  theme(
    legend.position = "right",
    legend.text = element_text(size = 9),
    legend.title = element_text(size = 10, face = "bold"),
    legend.spacing.y = unit(0.25, "lines")
  )
gLEG_strategy <- get_legend(pLEG_strategy)

#  (2) Gradient bars legend (3 rows, one per trophic strategy)
grad_df <- expand.grid(
  x = seq(0, 1, length.out = 120),
  Group_For_Plots = factor(trophic_order, levels = trophic_order)
)
grad_df$col <- mapply(
  function(g, x) scales::colour_ramp(c(trophic_cols_dark[[as.character(g)]], trophic_cols[[as.character(g)]]))(x),
  grad_df$Group_For_Plots,
  grad_df$x
)

pLEG_grad <- ggplot(grad_df, aes(x = x, y = Group_For_Plots, fill = col)) +
  geom_tile(height = 0.85) +
  scale_fill_identity() +
  scale_x_continuous(
    breaks = c(0, 1),
    labels = c("low", "high"),
    expand = c(0, 0)
  ) +
  labs(
    x = "Trophic index",
    y = NULL,
    title = "Within-group shading\n(darker = lower index)"
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.title = element_text(face = "bold", size = 10),
    panel.grid = element_blank(),
    axis.title.x = element_text(size = 9),
    axis.text.x = element_text(size = 9),
    axis.text.y = element_text(size = 9),
    plot.margin = margin(6, 6, 6, 6)
  )

gLEG_grad <- ggplotGrob(pLEG_grad)

A_leg_col <- arrangeGrob(
  gLEG_strategy,
  gLEG_grad,
  ncol = 1,
  heights = c(0.9, 1.1)
)

FIG_A <- arrangeGrob(
  gA_plot,
  A_leg_col,
  ncol = 2,
  widths = c(3.8, 1.4)
)

grid.newpage(); grid.draw(FIG_A)
export_grob_pdf(
  FIG_A,
  base = "Figure_A_MEANxMEAN_BrayCurtisDissim_vs_TPF_A4_landscape",
  width = 11.69, height = 8.27
)

# =========================================================
# 7) FIGURE B/C 
#    - Dot size encodes mean_prot_count
#    - One legend column on the right:
#        * size legend (mean proteins detected)
#        * highlight legend (fill)
#    - Fixes:
#        * no bracket warnings
#        * legend extraction works
# =========================================================

# ---- bracket helper WITHOUT ggplot "length 1 aesthetics" warnings ----
add_sig_brackets2 <- function(p, y_max, p12, p13, p23,
                              mults = c(1.10, 1.22, 1.34),
                              tick_frac = 0.03,
                              text_mult = 1.015) {
  t <- y_max * tick_frac

  # segments via annotate() (single-row by design)
  p +
    annotate("segment", x=1, xend=2, y=y_max*mults[1], yend=y_max*mults[1]) +
    annotate("segment", x=1, xend=1, y=y_max*mults[1]-t, yend=y_max*mults[1]) +
    annotate("segment", x=2, xend=2, y=y_max*mults[1]-t, yend=y_max*mults[1]) +
    annotate("text", x=1.5, y=y_max*mults[1]*text_mult, label=star(p12), size=4) +

    annotate("segment", x=1, xend=3, y=y_max*mults[2], yend=y_max*mults[2]) +
    annotate("segment", x=1, xend=1, y=y_max*mults[2]-t, yend=y_max*mults[2]) +
    annotate("segment", x=3, xend=3, y=y_max*mults[2]-t, yend=y_max*mults[2]) +
    annotate("text", x=2.0, y=y_max*mults[2]*text_mult, label=star(p13), size=4) +

    annotate("segment", x=2, xend=3, y=y_max*mults[3], yend=y_max*mults[3]) +
    annotate("segment", x=2, xend=2, y=y_max*mults[3]-t, yend=y_max*mults[3]) +
    annotate("segment", x=3, xend=3, y=y_max*mults[3]-t, yend=y_max*mults[3]) +
    annotate("text", x=2.5, y=y_max*mults[3]*text_mult, label=star(p23), size=4)
}

jit <- position_jitter(width = 0.3, height = 0, seed = 1)

clean_theme <- theme_minimal(base_size = 11) +
  theme(
    plot.title = element_text(face = "bold"),
    axis.title.x = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linewidth = 0.25),
    panel.background = element_rect(fill = NA, colour = NA),
    plot.background  = element_rect(fill = NA, colour = NA),
    legend.position = "none",              # IMPORTANT: no legends inside panels
    plot.margin = margin(28, 10, 10, 10)
  )

pt_stroke <- 0.35
ylim_headroom <- c(0, 1.18)

analysis_BC <- analysis_df %>%
  mutate(
    log_mean_tpf  = log10(pmin(pmax(mean_TPF, eps), 1 - eps)),
    log_mean_bray = log10(pmin(pmax(mean_bray, eps), 1 - eps))
  )

# Inference
kw_mean_tpf <- kruskal.test(log_mean_tpf ~ Group_For_Plots, data = analysis_BC)
pw_mean_tpf <- pairwise_wilcox_tidy(analysis_BC$log_mean_tpf, analysis_BC$Group_For_Plots)
p_mean_OI <- get_pw(pw_mean_tpf, "Oligo_like", "Intermediate")
p_mean_OC <- get_pw(pw_mean_tpf, "Oligo_like", "Copio_like")
p_mean_IC <- get_pw(pw_mean_tpf, "Intermediate", "Copio_like")

kw_mean_bray <- kruskal.test(log_mean_bray ~ Group_For_Plots, data = analysis_BC)
pw_mean_bray <- pairwise_wilcox_tidy(analysis_BC$log_mean_bray, analysis_BC$Group_For_Plots)
p_mbray_OI <- get_pw(pw_mean_bray, "Oligo_like", "Intermediate")
p_mbray_OC <- get_pw(pw_mean_bray, "Oligo_like", "Copio_like")
p_mbray_IC <- get_pw(pw_mean_bray, "Intermediate", "Copio_like")

# Size legend breaks
size_breaks_BC <- analysis_BC %>%
  summarise(
    mn = as.integer(round(min(mean_prot_count, na.rm = TRUE))),
    q1 = as.integer(round(quantile(mean_prot_count, 1/3, na.rm = TRUE))),
    q2 = as.integer(round(quantile(mean_prot_count, 2/3, na.rm = TRUE))),
    mx = as.integer(round(max(mean_prot_count, na.rm = TRUE)))
  ) %>%
  unlist(use.names = FALSE) %>%
  unique() %>%
  sort()

if (length(size_breaks_BC) < 4) {
  mn <- min(analysis_BC$mean_prot_count, na.rm = TRUE)
  mx <- max(analysis_BC$mean_prot_count, na.rm = TRUE)
  size_breaks_BC <- unique(as.integer(round(seq(mn, mx, length.out = 4))))
}

size_labels_BC <- as.character(size_breaks_BC)
size_labels_BC[1] <- paste0(size_labels_BC[1], " (min)")
size_labels_BC[length(size_labels_BC)] <- paste0(size_labels_BC[length(size_labels_BC)], " (max)")

# --- Panels (no legends) ---
pB <- ggplot(analysis_BC, aes(x = Group_For_Plots, y = mean_TPF)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label, size = mean_prot_count),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_size_continuous(
    name   = "Mean proteins detected (count)",
    range  = c(4, 12),
    breaks = size_breaks_BC,
    labels = size_labels_BC
  ) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "B  Mean transporter allocation differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_mean_tpf$p.value)),
    y = "Mean transporter proteome fraction (TPF)"
  ) +
  clean_theme

pB <- add_sig_brackets2(pB, 1, p_mean_OI, p_mean_OC, p_mean_IC,
                        mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

pC <- ggplot(analysis_BC, aes(x = Group_For_Plots, y = mean_bray)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label, size = mean_prot_count),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_size_continuous(
    name   = "Mean proteins detected (count)",
    range  = c(4, 12),
    breaks = size_breaks_BC,
    labels = size_labels_BC
  ) +
  scale_y_continuous(breaks = axis_breaks_01, expand = c(0,0)) +
  coord_cartesian(ylim = ylim_headroom, clip = "off") +
  labs(
    title = "C  Functional variability differs among trophic strategies",
    subtitle = paste0("Inference on log10(values): Kruskal–Wallis p = ", fmt_p(kw_mean_bray$p.value)),
    y = "Mean variability (Bray–Curtis dissimilarity)"
  ) +
  clean_theme

pC <- add_sig_brackets2(pC, 1, p_mbray_OI, p_mbray_OC, p_mbray_IC,
                        mults = c(1.06, 1.11, 1.16), tick_frac = 0.02, text_mult = 1.01)

# --- Legend extraction: make a LEGEND-ENABLED pB version (only for grabbing legends) ---
pB_leg <- ggplot(analysis_BC, aes(x = Group_For_Plots, y = mean_TPF)) +
  geom_boxplot(width = 0.60, outlier.shape = NA, fill = NA, color = "black", linewidth = 0.5) +
  geom_point(
    aes(fill = highlight_label, size = mean_prot_count),
    position = jit,
    shape = 21, color = "black",
    stroke = pt_stroke, alpha = 0.9
  ) +
  scale_fill_manual(values = highlight_cols, drop = FALSE) +
  scale_size_continuous(
    name   = "Mean proteins detected (count)",
    range  = c(4, 12),
    breaks = size_breaks_BC,
    labels = size_labels_BC
  ) +
  theme_bw() +
  theme(legend.position = "right")

gLEG_size <- get_legend(pB_leg + guides(fill = "none"))
gLEG_high <- get_legend(pB_leg + guides(size = "none") +
                          guides(fill = guide_legend(title = NULL, ncol = 1, byrow = TRUE)))

# --- Assemble figure ---
BC_stack <- arrangeGrob(
  ggplotGrob(pB),
  ggplotGrob(pC),
  ncol = 1,
  heights = c(1, 1)
)

BC_leg_col <- arrangeGrob(
  gLEG_size,
  gLEG_high,
  ncol = 1,
  heights = c(1.1, 1.0)
)

FIG_BC <- arrangeGrob(
  BC_stack,
  BC_leg_col,
  ncol = 2,
  widths = c(3.2, 1.1)
)

grid.newpage(); grid.draw(FIG_BC)
export_grob_pdf(
  FIG_BC,
  base = "Figure_BC_MEANxMEAN_BrayCurtisDissim_and_TPF_A4_landscape",
  width = 8.27, height = 8.27
)

############################################################
# EXPORT FINAL FIGURE STATS SUMMARY
############################################################

stats_file <- "final_figure_stats_summary.txt"
sink(stats_file)

cat("FINAL FIGURE STATS SUMMARY\n")
cat("====================================\n\n")

cat("DATA SUMMARY\n")
cat("------------------------------------\n")
cat("Species included:", nrow(analysis_df), "\n")
cat("Transitions (Bray–Curtis pairs):", nrow(bray_transitions), "\n\n")

# =========================================================
# FIGURE A — regression model
# =========================================================

cat("FIGURE A: Investment vs Functional Stability\n")
cat("------------------------------------\n")
cat("Model: logit(Bray–Curtis dissimilarity) ~ log10(mean TPF)\n\n")

print(summary(m_logit))

cat("\nConfidence intervals:\n")
print(confint(m_logit))

cat("\n")

# =========================================================
# FIGURE B — mean transporter allocation
# =========================================================

cat("FIGURE B: Mean transporter allocation by trophic strategy\n")
cat("------------------------------------\n")

cat("\nKruskal–Wallis test:\n")
print(kw_mean_tpf)

cat("\nPairwise Wilcoxon tests (BH adjusted):\n")
print(pairwise.wilcox.test(
  analysis_BC$log_mean_tpf,
  analysis_BC$Group_For_Plots,
  p.adjust.method = "BH",
  exact = FALSE
))

cat("\n")

# =========================================================
# FIGURE C — functional variability
# =========================================================

cat("FIGURE C: Functional variability by trophic strategy\n")
cat("------------------------------------\n")

cat("\nKruskal–Wallis test:\n")
print(kw_mean_bray)

cat("\nPairwise Wilcoxon tests (BH adjusted):\n")
print(pairwise.wilcox.test(
  analysis_BC$log_mean_bray,
  analysis_BC$Group_For_Plots,
  p.adjust.method = "BH",
  exact = FALSE
))

cat("\n")

# =========================================================
# GROUP SAMPLE SIZES
# =========================================================

cat("GROUP SAMPLE SIZES\n")
cat("------------------------------------\n")
print(table(analysis_df$Group_For_Plots))

cat("\n")

sink()

cat("\nSaved stats file:\n  ", normalizePath(stats_file), "\n", sep="")
```

