---
title: "ABC Transporters Analysis Script"
output:
  html_document:
    df_print: paged
---

```{r}
############################################################
# 0) CLEAN START + LOAD PACKAGES + LOAD DATA
#    (copy/paste as the first cell in your R notebook)
############################################################

rm(list = ls())

# Avoid legacy plyr masking tidyverse verbs
if ("package:dplyr" %in% search()) {
  detach("package:dplyr", unload = TRUE, character.only = TRUE)
}

suppressPackageStartupMessages({
  library(readr)
  library(dplyr)
  library(tidyr)
  library(stringr)
  library(ggplot2)
  library(scales)
})

# ---- Working directory (project root for this analysis) ----
setwd("/SynologyDrive/PhD/Projects/SpringBloom2021/DataAnalysis_SpringBloom2021/Figure_04/ABC_Transporters/")

# ---- Data inputs (one line each) ----
protein_final          <- read_csv("../Data_Filtration/retained_protein_table.csv", show_col_types = FALSE)
retained_species       <- read_csv("../Data_Filtration/retained_species.csv", show_col_types = FALSE)
retained_species_dates <- read_csv("../Data_Filtration/retained_species_dates.csv", show_col_types = FALSE)

gene_dat <- read_csv("../../../Data/Gene_Data/Final_Data/Gene_Table.csv", show_col_types = FALSE)

transporter_table_raw <- read_csv(
  "../../../Data/KEGG_Category_Data/Final_Data/KEGG_Transporter_Categories_Curated.csv",
  show_col_types = FALSE
)

# ---- (Optional) quick sanity prints ----
glimpse(protein_final)
glimpse(retained_species)
glimpse(retained_species_dates)
glimpse(gene_dat)
glimpse(transporter_table_raw)

```


```{r}
############################################################
# 1) DATA MANIPULATION + CONSTRUCTION OF THE TWO MAIN TABLES
#
# OUTPUT TABLES:
#   (A) gene_retained_annot
#       - genes from retained species only
#       - transporter ontology added by KEGG_ID (KO)
#       - Group_For_Plots added
#       - KEGG_Subcategory added (mapped from proteomics annotations)
#
#   (B) protein_retained_annot
#       - proteins restricted to the retained species×date universe
#       - transporter ontology added by KEGG_ID (KO)
#       - already contains Group_For_Plots + KEGG_Subcategory from protein_final
#
#   - strict enforcement of retained species×date universe for proteomics
#   - one-row-per-KO transporter ontology to prevent double counting
#   - explicit joins by Species_ID / Date / Protein_ID / KEGG_ID
############################################################

# ------------------------------------------------------------
# A) Collapse transporter ontology to ONE ROW PER KO (KEGG_ID)
#
# Rationale:
#   Some KOs appear multiple times in the curated transporter table
#   (e.g. due to alternative hierarchical placements).
#   To avoid double counting, we collapse to one row per KO and
#   merge duplicated annotations by concatenating unique values
#   with ";" for each annotation category.
# ------------------------------------------------------------

transporter_table_final <- transporter_table_raw %>%
  mutate(across(everything(), as.character)) %>%
  group_by(KEGG_ID) %>%
  summarise(
    Transporter_Family = paste(sort(unique(na.omit(Transporter_Family))), collapse = ";"),
    major_group        = paste(sort(unique(na.omit(major_group))), collapse = ";"),
    brite_id           = paste(sort(unique(na.omit(brite_id))), collapse = ";"),
    brite_name         = paste(sort(unique(na.omit(brite_name))), collapse = ";"),
    Path_L1            = paste(sort(unique(na.omit(Path_L1))), collapse = ";"),
    Path_L2            = paste(sort(unique(na.omit(Path_L2))), collapse = ";"),
    Path_L3            = paste(sort(unique(na.omit(Path_L3))), collapse = ";"),
    Path_L4            = paste(sort(unique(na.omit(Path_L4))), collapse = ";"),
    Substrate_Group   = paste(sort(unique(na.omit(Substrate_Group))), collapse = ";"),
    Substrate_Category = paste(sort(unique(na.omit(Substrate_Category))), collapse = ";"),
    Path_L5            = paste(sort(unique(na.omit(Path_L5))), collapse = ";"),
    path               = paste(sort(unique(na.omit(path))), collapse = ";"),
    .groups = "drop"
  )

# ------------------------------------------------------------
# B) Build the *retained* proteomics table (species×date enforced)
# ------------------------------------------------------------
protein_retained_annot <- protein_final %>%
  semi_join(retained_species_dates %>% select(Species_ID, Date), by = c("Species_ID", "Date")) %>%
  left_join(gene_dat %>% select(Protein_ID, KEGG_ID) %>% distinct()) %>%
  left_join(transporter_table_final, by = c("KEGG_ID","Substrate_Category","Transporter_Family"))

# ------------------------------------------------------------
# C) Build the *retained* gene table (species enforced)
# ------------------------------------------------------------
gene_retained_annot <- gene_dat %>%
  semi_join(retained_species %>% select(Species_ID), by = "Species_ID") %>%
  left_join(transporter_table_final, by = "KEGG_ID") %>%
  left_join(retained_species %>% select(Species_ID, Group_For_Plots), by = c("Species_ID"))

# ------------------------------------------------------------
# D) Add KEGG_Subcategory to genes using a KEGG_ID -> KEGG_Subcategory map
#    Source-of-truth: protein_retained_annot (observed, retained, annotated proteome)
#    If a KEGG_ID maps to multiple KEGG_Subcategory strings, collapse with ";"
# ------------------------------------------------------------
kegg_subcat_map <- protein_retained_annot %>%
  select(KEGG_ID, KEGG_Subcategory) %>%
  filter(!is.na(KEGG_ID), !is.na(KEGG_Subcategory), KEGG_ID != "Unclassified") %>%
  distinct() %>%
  group_by(KEGG_ID) %>%
  summarise(
    KEGG_Subcategory = paste(sort(unique(KEGG_Subcategory)), collapse = ";"),
    .groups = "drop"
  )

gene_retained_annot <- gene_retained_annot %>%
  left_join(kegg_subcat_map, by = "KEGG_ID")

# ------------------------------------------------------------
# E) Minimal sanity checks (recommended)
# ------------------------------------------------------------

# 1) Proteomics: every row must be within retained species×date universe
stopifnot(
  nrow(anti_join(
    protein_retained_annot %>% distinct(Species_ID, Date),
    retained_species_dates %>% distinct(Species_ID, Date),
    by = c("Species_ID", "Date")
  )) == 0
)

# 2) Genes: every species must be in retained_species
stopifnot(
  nrow(anti_join(
    gene_retained_annot %>% distinct(Species_ID),
    retained_species %>% distinct(Species_ID),
    by = "Species_ID"
  )) == 0
)

# 3) Quick coverage diagnostics
message("Proteins retained: ", nrow(protein_retained_annot))
message("Species in proteomics retained: ", n_distinct(protein_retained_annot$Species_ID))
message("Genes retained: ", nrow(gene_retained_annot))
message("Species in genes retained: ", n_distinct(gene_retained_annot$Species_ID))

# 4) Confirm KEGG_Subcategory was added to genes (not necessarily complete)
gene_retained_annot %>%
  summarise(
    n_genes = n(),
    n_with_kegg_id = sum(!is.na(KEGG_ID) & KEGG_ID != "Unclassified"),
    n_with_subcat  = sum(!is.na(KEGG_Subcategory)),
    prop_subcat_among_kegg = n_with_subcat / n_with_kegg_id
  )

# Optional: inspect final tables
glimpse(protein_retained_annot)
glimpse(gene_retained_annot)

```

```{r}
############################################################
# NESTED OVERVIEW (3 BARS PER GROUP): TOTAL vs ABC GENES vs ABC PROTEINS
#
# What this shows (per Group_For_Plots):
#   1) Total species with proteome coverage (baseline universe)
#   2) Species with ≥1 ABC gene in the genome (genomic potential)
#   3) Species with ≥1 detected ABC protein (expression)
#
# Labels: n (percent of Total species in that group)
#
# Inputs needed:
#   - gene_retained_annot
#   - protein_retained_annot
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(ggplot2)
  library(scales)
})

# ---- group order requested ----
group_levels <- c("Oligo_like", "Intermediate", "Copio_like")

# ------------------------------------------------------------
# 1) Total species per group (proteome-covered universe)
# ------------------------------------------------------------
total_species <- protein_retained_annot %>%
  distinct(Species_ID, Group_For_Plots) %>%
  count(Group_For_Plots, name = "n_species") %>%
  mutate(layer = "Total species")

# ------------------------------------------------------------
# 2) Species with ≥1 ABC gene (genome)
# ------------------------------------------------------------
abc_species_genes <- gene_retained_annot %>%
  semi_join(protein_retained_annot %>% distinct(Species_ID), by = "Species_ID") %>% # ensure same universe
  filter(Transporter_Family == "ABC") %>%
  distinct(Species_ID, Group_For_Plots) %>%
  count(Group_For_Plots, name = "n_species") %>%
  mutate(layer = "Genome (ABC genes)")

# ------------------------------------------------------------
# 3) Species with ≥1 detected ABC protein (proteome)
# ------------------------------------------------------------
abc_species_proteins <- protein_retained_annot %>%
  filter(Transporter_Family == "ABC") %>%
  distinct(Species_ID, Group_For_Plots) %>%
  count(Group_For_Plots, name = "n_species") %>%
  mutate(layer = "Proteome (ABC proteins)")

# ------------------------------------------------------------
# Combine + add percent relative to "Total species" in each group
# ------------------------------------------------------------
nested_overview <- bind_rows(total_species, abc_species_genes, abc_species_proteins) %>%
  group_by(Group_For_Plots) %>%
  mutate(
    n_total = n_species[layer == "Total species"],
    percent = n_species / n_total,
    label = paste0(n_species, " (", percent(percent, accuracy = 1), ")")
  ) %>%
  ungroup() %>%
  mutate(
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels),
    layer = factor(layer, levels = c("Total species", "Genome (ABC genes)", "Proteome (ABC proteins)"))
  )

# ------------------------------------------------------------
# Plot: 3 bars per group, with labels
# ------------------------------------------------------------
ggplot(nested_overview, aes(x = Group_For_Plots, y = n_species, fill = layer)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(
    aes(label = label),
    position = position_dodge(width = 0.8),
    vjust = -0.3,
    size = 3
  ) +
  labs(
    x = NULL,
    y = "Number of species",
    fill = NULL,
    title = "ABC transporters: genomic potential vs detected expression",
    subtitle = "Per group: total species vs species encoding ABC genes vs species with detected ABC proteins"
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )
```



```{r}
# ============================================================
# ONE SUMMARY TABLE (Value + Percent + n_species) + ABS/REL PLOTS
#
# Output: summary_tbl_final
#   One row per Panel × Group_For_Plots × Class containing:
#     - Value    : genes count / protein count / summed mean abundance
#     - Total    : total per Panel×Group (sums across classes)
#     - Percent  : Value / Total
#     - n_species: number of species contributing ≥1 item to that class
#
# Exclusive classes (no double counting; stacked bars are valid):
#   1) ABC transporters        (Transporter_Family == "ABC")
#   2) Other transporters      (KEGG_Subcategory contains "Transporters" but NOT ABC)
#   3) Non-transporters        (everything else)
#
# Species universe:
#   Genes are restricted to species present in protein_retained_annot (proteome-covered taxa),
#   ensuring fair comparison across panels.
#
# Proteomics duplication:
#   protein_retained_annot can have multiple rows per Protein_ID (KEGG_ID splitting).
#   We collapse per Species×Protein_ID for protein counts and per Species×Date×Protein_ID for abundance.
# ============================================================

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(scales)
})

# ---- ordering requested ----
group_levels <- c("Oligo_like", "Intermediate", "Copio_like")
class_levels <- c("Non-transporters", "Other transporters", "ABC transporters")
panel_levels <- c("Genes", "Detected proteins", "Protein abundance")

# ---- helper: KEGG_Subcategory membership flag ----
is_transporter_subcat <- function(x) !is.na(x) & grepl("Transporters", x, fixed = TRUE)

# ---- species universe: proteome-covered taxa ----
species_with_proteome <- protein_retained_annot %>% distinct(Species_ID)

# ============================================================
# PANEL 1: GENES (counts)
# ============================================================
genes_items <- gene_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  mutate(
    Panel = "Genes",
    Class = case_when(
      Transporter_Family == "ABC" ~ "ABC transporters",
      is_transporter_subcat(KEGG_Subcategory) ~ "Other transporters",
      TRUE ~ "Non-transporters"
    )
  )

genes_totals <- genes_items %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

genes_species <- genes_items %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

# ============================================================
# PANEL 2: DETECTED PROTEINS (counts of distinct Protein_ID)
#   Collapse KEGG-split rows: one record per Species×Protein_ID
# ============================================================
proteins_items <- protein_retained_annot %>%
  distinct(Species_ID, Group_For_Plots, Protein_ID, KEGG_Subcategory, Transporter_Family) %>%
  group_by(Species_ID, Group_For_Plots, Protein_ID) %>%
  summarise(
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_trans = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Detected proteins",
    Class = case_when(
      is_abc ~ "ABC transporters",
      is_trans ~ "Other transporters",
      TRUE ~ "Non-transporters"
    )
  )

proteins_totals <- proteins_items %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

proteins_species <- proteins_items %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

# ============================================================
# PANEL 3: PROTEIN ABUNDANCE
#   Collapse KEGG-split rows: one record per Species×Date×Protein_ID
#   Aggregate abundance as:
#     Species×Date×Class sum -> Species×Class mean across dates -> Group×Class sum across species
# ============================================================
abund_items <- protein_retained_annot %>%
  group_by(Species_ID, Group_For_Plots, Date, Protein_ID) %>%
  summarise(
    abundance = first(Protein_Abundance_IBAQ_Norm),
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_trans = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Protein abundance",
    Class = case_when(
      is_abc ~ "ABC transporters",
      is_trans ~ "Other transporters",
      TRUE ~ "Non-transporters"
    )
  )

abundance_totals <- abund_items %>%
  group_by(Panel, Group_For_Plots, Class, Species_ID, Date) %>%
  summarise(Value = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(Panel, Group_For_Plots, Class, Species_ID) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  group_by(Panel, Group_For_Plots, Class) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop")

abundance_species <- abund_items %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

# ============================================================
# FINAL: One summary table (Value + Percent + n_species)
# ============================================================
totals_tbl <- bind_rows(genes_totals, proteins_totals, abundance_totals) %>%
  mutate(
    Panel = factor(Panel, levels = panel_levels),
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels),
    Class = factor(Class, levels = class_levels)
  ) %>%
  group_by(Panel, Group_For_Plots) %>%
  mutate(
    Total = sum(Value, na.rm = TRUE),
    Percent = Value / Total
  ) %>%
  ungroup()

species_tbl <- bind_rows(genes_species, proteins_species, abundance_species) %>%
  mutate(
    Panel = factor(Panel, levels = panel_levels),
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels),
    Class = factor(Class, levels = class_levels)
  )

summary_tbl_final <- totals_tbl %>%
  left_join(species_tbl, by = c("Panel", "Group_For_Plots", "Class")) %>%
  arrange(Panel, Group_For_Plots, Class)

# ---- Optional: absolute + relative stacked plots from the same table ----
p_abs <- ggplot(summary_tbl_final, aes(x = Group_For_Plots, y = Value, fill = Class)) +
  geom_col() +
  facet_wrap(~ Panel, scales = "free_y", ncol = 1) +
  labs(x = NULL, y = NULL, fill = NULL, title = "Transporter classes across groups (absolute)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p_rel <- ggplot(summary_tbl_final, aes(x = Group_For_Plots, y = Percent, fill = Class)) +
  geom_col() +
  facet_wrap(~ Panel, scales = "free_y", ncol = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(x = NULL, y = NULL, fill = NULL, title = "Transporter classes across groups (relative; 100%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        strip.background = element_blank(),
        strip.text = element_text(face = "bold"))

p_abs
p_rel

# ---- Print table (optional) ----
summary_tbl_final
```

### What is shown
The figure summarizes **genomic potential (genes)**, **detected expression (proteins)**, and **investment (protein abundance)** of transporter classes across trophic strategies (oligo-like, intermediate, copio-like).  
Each panel uses **exclusive functional classes** (Non-transporters, Other transporters, ABC transporters), ensuring that absolute counts and relative proportions are directly comparable and free of double counting despite multi-annotated KEGG categories.

---

### Key observations

1. **Genomic potential (Genes)**  
   Across all trophic groups, **non-transporters dominate gene content**, as expected for complete bacterial genomes.  
   ABC transporters represent a **small but consistent fraction** of the gene repertoire in all groups, with only modest differences among oligo-like, intermediate, and copio-like taxa.  
   → This indicates that **ABC transport capacity is broadly encoded**, but not strongly differentiated at the genomic level.

2. **Detected proteins (Expression)**  
   In contrast to gene counts, the **relative contribution of transporters increases markedly at the protein level**, especially in oligo-like taxa.  
   ABC transporters form a **substantial fraction of detected proteins** in oligotrophs, while copio-like taxa show comparatively fewer detected ABC proteins.  
   → This suggests **selective expression** of transporter machinery rather than simple reflection of genome content.

3. **Protein abundance (Investment)**  
   The strongest signal emerges at the abundance level:  
   oligo-like taxa allocate a **disproportionately large share of proteomic investment to ABC transporters**, far exceeding their genomic representation.  
   Intermediate taxa show mixed investment across transporter classes, while copio-like taxa invest relatively little in ABC systems.  
   → This pattern is consistent with **high-affinity uptake strategies in oligotrophs**, where energetic investment in ABC transporters is favored under nutrient limitation.

---

### Critical interpretation

- The **decoupling between genomic potential and proteomic investment** indicates that transporter strategies are primarily regulated at the level of expression and allocation, not gene presence.
- The progressive enrichment of ABC transporters from genes → proteins → abundance in oligo-like taxa provides strong evidence for **functional specialization**, rather than annotation or detection bias.
- Because all summaries are restricted to **species with proteome coverage** and use **exclusive classification rules**, the observed trends cannot be explained by differential genome sizes, multi-category KEGG annotations, or uneven sampling across dates.

---

### Conclusion
Together, these data support a model in which **ABC transporters are a key functional axis distinguishing trophic strategies**, with oligotrophic bacteria investing heavily in high-affinity uptake systems at the proteomic level despite similar genomic repertoires across groups.


```{r}
############################################################
# Left: absolute transporter pool per group (ABC + Other) + ABC% label
# Right: group contribution to ABC and to Other (100% bars)
#
# Uses: gene_retained_annot, protein_retained_annot
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(forcats)
  library(ggplot2)
  library(scales)
})

# -----------------------------
# Settings
# -----------------------------
group_levels <- c("Oligo_like", "Intermediate", "Copio_like")
panel_levels <- c("Genes", "Detected proteins", "Protein abundance")
class_levels <- c("Other", "ABC")

is_transporter_subcat <- function(x) !is.na(x) & grepl("Transporters", x, fixed = TRUE)

# "transporter pool" = ABC OR transporter subcategory
is_any_transporter <- function(tf, subcat) (tf == "ABC") | is_transporter_subcat(subcat)

# Species universe: proteome-covered taxa (reviewer-safe comparison)
species_with_proteome <- protein_retained_annot %>% distinct(Species_ID)

# ============================================================
# Build transporter-only totals for each panel
# ============================================================

# --- Genes (counts) ---
genes_totals_trans <- gene_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  mutate(
    is_abc = Transporter_Family == "ABC",
    is_trans = is_transporter_subcat(KEGG_Subcategory),
    is_transporter = is_any_transporter(Transporter_Family, KEGG_Subcategory),
    Class = case_when(
      is_abc ~ "ABC",
      is_trans ~ "Other",
      TRUE ~ NA_character_
    ),
    Panel = "Genes"
  ) %>%
  filter(is_transporter) %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

# --- Detected proteins (distinct Protein_ID counts; collapse KEGG splits) ---
proteins_totals_trans <- protein_retained_annot %>%
  distinct(Species_ID, Group_For_Plots, Protein_ID, KEGG_Subcategory, Transporter_Family) %>%
  group_by(Species_ID, Group_For_Plots, Protein_ID) %>%
  summarise(
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_trans = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    is_transporter = is_abc | is_trans,
    Class = case_when(
      is_abc ~ "ABC",
      is_trans ~ "Other",
      TRUE ~ NA_character_
    ),
    Panel = "Detected proteins"
  ) %>%
  filter(is_transporter) %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

# --- Protein abundance (species mean across dates; then sum across species) ---
abund_totals_trans <- protein_retained_annot %>%
  group_by(Species_ID, Group_For_Plots, Date, Protein_ID) %>%
  summarise(
    abundance = first(Protein_Abundance_IBAQ_Norm),
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_trans = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    is_transporter = is_abc | is_trans,
    Class = case_when(
      is_abc ~ "ABC",
      is_trans ~ "Other",
      TRUE ~ NA_character_
    ),
    Panel = "Protein abundance"
  ) %>%
  filter(is_transporter) %>%
  # Species×Date×Class sum
  group_by(Panel, Group_For_Plots, Species_ID, Date, Class) %>%
  summarise(Value = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  # Species×Class mean across dates
  group_by(Panel, Group_For_Plots, Species_ID, Class) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  # Group×Class sum across species
  group_by(Panel, Group_For_Plots, Class) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop")

# Combine
trans_totals_all <- bind_rows(genes_totals_trans, proteins_totals_trans, abund_totals_trans) %>%
  mutate(
    Panel = factor(Panel, levels = panel_levels),
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels),
    Class = factor(Class, levels = class_levels)
  )

# ============================================================
# Derive labels: "ABC = xx%" within transporter pool per group
# ============================================================
abc_labels <- trans_totals_all %>%
  group_by(Panel, Group_For_Plots) %>%
  summarise(
    TotalTransporters = sum(Value, na.rm = TRUE),
    ABCValue = sum(Value[Class == "ABC"], na.rm = TRUE),
    ABCFrac = ifelse(TotalTransporters > 0, ABCValue / TotalTransporters, NA_real_),
    .groups = "drop"
  ) %>%
  mutate(label = paste0("ABC ", percent(ABCFrac, accuracy = 1)))

# ============================================================
# COLUMN 1: Absolute stacked bars (ABC + Other) per group
# + ABC% label above bar
# ============================================================
# total per bar for label placement
bar_tops <- trans_totals_all %>%
  group_by(Panel, Group_For_Plots) %>%
  summarise(y_top = sum(Value, na.rm = TRUE), .groups = "drop") %>%
  left_join(abc_labels, by = c("Panel", "Group_For_Plots"))

p_abs <- ggplot(trans_totals_all, aes(x = Group_For_Plots, y = Value, fill = Class)) +
  geom_col(width = 0.75) +
  facet_wrap(~ Panel, ncol = 1, scales = "free_y") +
  geom_text(
    data = bar_tops,
    aes(x = Group_For_Plots, y = y_top, label = label),
    inherit.aes = FALSE,
    vjust = -0.4,
    size = 3
  ) +
  labs(
    x = NULL, y = NULL, fill = NULL,
    title = "Absolute transporter signal by trophic group",
    subtitle = "Bars: transporter pool (ABC + other). Labels: ABC fraction within transporters."
  ) +
  theme_classic() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold")
  )

p_abs
```

```{r}
############################################################
# Alluvial transporter-pool figure (3 panels) + summary table
# ----------------------------------------------------------
# Goal:
#   Visualize how the "transporter pool" is partitioned between
#   trophic groups (Oligo / Intermediate / Copio) and transporter
#   classes (ABC vs Other transporters), using three complementary
#   measures:
#
#   A) Genes                = counts of transporter genes
#   B) Detected proteins    = counts of distinct transporter proteins
#                             (collapsed per Protein_ID to avoid double-
#                             counting from multiple annotations)
#   C) Protein abundance    = quantitative transporter signal:
#                             for each species, compute mean (over dates)
#                             of total transporter abundance, then sum
#                             species means within each trophic group
#
# Transporter pool definition:
#   A gene/protein belongs to transporter pool if:
#     - Transporter_Family == "ABC"
#       OR
#     - KEGG_Subcategory contains the exact string "Transporters"
#
# Class assignment (mutually exclusive):
#   - "ABC"   if Transporter_Family == "ABC"
#   - "Other" if NOT ABC but KEGG_Subcategory contains "Transporters"
#
# Species universe harmonization:
#   All three panels are restricted to species that are present in
#   the proteomics table (protein_retained_annot), ensuring that gene
#   and protein comparisons are made across the same species set.
#
# Outputs:
#   - FIG_alluvial_transporter_pool_proportions.pdf
#   - TABLE_alluvial_transporter_pool_summary_with_species.csv
#
# Required input data frames:
#   gene_retained_annot:
#     Species_ID, Group_For_Plots, Transporter_Family, KEGG_Subcategory, ...
#   protein_retained_annot:
#     Species_ID, Group_For_Plots, Protein_ID, Date,
#     Protein_Abundance_IBAQ_Norm, Transporter_Family, KEGG_Subcategory, ...
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggalluvial)
  library(scales)
  library(gridExtra)
  library(grid)
})

# -----------------------------
# 0) Settings and aesthetics
# -----------------------------
group_levels <- c("Oligo_like", "Intermediate", "Copio_like")
class_levels <- c("ABC", "Other")
panel_levels <- c("Genes", "Detected proteins", "Protein abundance")

grp_cols <- c(
  Oligo_like   = "#2C7FB8",  # blue
  Intermediate = "grey60",   # grey
  Copio_like   = "#D7301F"   # red
)

# -----------------------------
# 1) Helper functions (core definitions)
# -----------------------------

# Transporter-like subcategory: KEGG_Subcategory contains the string "Transporters"
is_transporter_subcat <- function(x) !is.na(x) & grepl("Transporters", x, fixed = TRUE)

# Transporter pool membership: ABC OR KEGG transporter subcategory
is_any_transporter <- function(tf, subcat) (tf == "ABC") | is_transporter_subcat(subcat)

# Mutually exclusive class assignment (ABC has priority)
assign_transporter_class <- function(tf, subcat) {
  is_abc  <- tf == "ABC"
  is_tr   <- is_transporter_subcat(subcat)
  case_when(
    is_abc ~ "ABC",
    is_tr  ~ "Other",
    TRUE   ~ NA_character_
  )
}

# -----------------------------
# 2) Define the harmonized species universe
# -----------------------------
# We restrict all panels to species covered by proteomics so that gene/protein
# panels are directly comparable across the same taxa.
species_with_proteome <- protein_retained_annot %>% distinct(Species_ID)

# -----------------------------
# 3) Panel A: transporter genes (counts)
# -----------------------------
# Count transporter genes per trophic group and class within the harmonized species set.
genes_totals_trans <- gene_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  mutate(
    Panel = "Genes",
    is_transporter = is_any_transporter(Transporter_Family, KEGG_Subcategory),
    Class = assign_transporter_class(Transporter_Family, KEGG_Subcategory)
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

# -----------------------------
# 4) Panel B: detected transporter proteins (distinct Protein_ID counts)
# -----------------------------
# Proteomics annotations can appear multiple times per protein (e.g. multiple KEGG hits).
# To avoid double-counting, we collapse to one record per Species × Protein_ID and
# then classify the protein as transporter if ANY annotation indicates ABC or transporter subcategory.
proteins_totals_trans <- protein_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  distinct(Species_ID, Group_For_Plots, Protein_ID, KEGG_Subcategory, Transporter_Family) %>%
  group_by(Species_ID, Group_For_Plots, Protein_ID) %>%
  summarise(
    is_abc  = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_tr   = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Detected proteins",
    is_transporter = is_abc | is_tr,
    Class = case_when(
      is_abc ~ "ABC",
      is_tr  ~ "Other",
      TRUE   ~ NA_character_
    )
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  count(Panel, Group_For_Plots, Class, name = "Value")

# -----------------------------
# 5) Panel C: transporter protein abundance (species-mean over time, summed within group)
# -----------------------------
# Rationale: We want abundance to reflect a typical (mean-over-time) species signal,
# and then ask how trophic groups differ in the summed contribution of their species.
#
# Steps:
#   (i)  Species × Date × Protein_ID: take one abundance value per protein (first()),
#        classify transporter class based on ANY annotation for that protein instance.
#   (ii) Species × Date × Class: sum abundance across proteins (total transporter abundance per date)
#   (iii)Species × Class: mean across dates (typical species abundance)
#   (iv) Group × Class: sum species means (group-level total)
abund_totals_trans <- protein_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  group_by(Species_ID, Group_For_Plots, Date, Protein_ID) %>%
  summarise(
    abundance = first(Protein_Abundance_IBAQ_Norm),
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_tr  = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Protein abundance",
    is_transporter = is_abc | is_tr,
    Class = case_when(
      is_abc ~ "ABC",
      is_tr  ~ "Other",
      TRUE   ~ NA_character_
    )
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  group_by(Panel, Group_For_Plots, Species_ID, Date, Class) %>%
  summarise(Value = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(Panel, Group_For_Plots, Species_ID, Class) %>%
  summarise(Value = mean(Value, na.rm = TRUE), .groups = "drop") %>%
  group_by(Panel, Group_For_Plots, Class) %>%
  summarise(Value = sum(Value, na.rm = TRUE), .groups = "drop")

# -----------------------------
# 6) Combine panels + compute panel-normalized proportions
# -----------------------------
totals_all <- bind_rows(genes_totals_trans, proteins_totals_trans, abund_totals_trans) %>%
  mutate(
    Panel = factor(Panel, levels = panel_levels),
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels),
    Class = factor(Class, levels = class_levels)
  )

# Each panel is normalized independently to sum to 1.0 (100% of transporter pool in that panel).
totals_prop <- totals_all %>%
  group_by(Panel) %>%
  mutate(Value_prop = Value / sum(Value, na.rm = TRUE)) %>%
  ungroup()

# -----------------------------
# 7) Plotting (alluvial panels + shared legend)
# -----------------------------
make_alluvial_panel <- function(df_prop, panel_name, panel_letter) {
  d <- df_prop %>% filter(Panel == panel_name)

  ggplot(d, aes(axis1 = Group_For_Plots, axis2 = Class, y = Value_prop)) +
    geom_alluvium(aes(fill = Group_For_Plots),
                  alpha = 0.90, color = "grey25", linewidth = 0.25) +
    geom_stratum(fill = "grey92", color = "black", linewidth = 0.35) +
    geom_text(stat = "stratum", aes(label = after_stat(stratum)), size = 3.0) +
    scale_fill_manual(values = grp_cols, drop = FALSE) +
    scale_y_continuous(labels = percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.03))) +
    labs(
      title = paste0(panel_letter, "  ", panel_name),
      x = NULL,
      y = "Proportion of transporter pool",
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      legend.position = "none",
      plot.margin = margin(4, 6, 4, 6)
    )
}

get_legend_grob <- function(p) {
  g <- ggplotGrob(p + theme(legend.position = "right"))
  idx <- which(sapply(g$grobs, function(x) x$name) == "guide-box")
  if (length(idx) != 1) stop("Legend extraction failed.")
  g$grobs[[idx]]
}

pA <- make_alluvial_panel(totals_prop, "Genes", "A")
pB <- make_alluvial_panel(totals_prop, "Detected proteins", "B")
pC <- make_alluvial_panel(totals_prop, "Protein abundance", "C")

p_leg_src <- ggplot(totals_prop, aes(x = Group_For_Plots, y = Value_prop, fill = Group_For_Plots)) +
  geom_col() +
  scale_fill_manual(values = grp_cols, drop = FALSE) +
  theme_void(base_size = 10) +
  theme(legend.position = "right")

gLEG <- get_legend_grob(p_leg_src)

FIG <- arrangeGrob(
  arrangeGrob(pA, pB, pC, ncol = 3),
  gLEG,
  ncol = 2,
  widths = c(4.6, 1.1)
)

# -----------------------------
# 8) Summary table (absolute + fractions + contributing species)
# -----------------------------
# n_species counts how many unique Species_IDs contributed to each Panel × Group × Class.
# For abundance, we count species with a strictly positive mean transporter abundance in that class.
genes_species <- gene_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  mutate(
    Panel = "Genes",
    is_transporter = is_any_transporter(Transporter_Family, KEGG_Subcategory),
    Class = assign_transporter_class(Transporter_Family, KEGG_Subcategory)
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

proteins_species <- protein_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  distinct(Species_ID, Group_For_Plots, Protein_ID, KEGG_Subcategory, Transporter_Family) %>%
  group_by(Species_ID, Group_For_Plots, Protein_ID) %>%
  summarise(
    is_abc  = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_tr   = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Detected proteins",
    is_transporter = is_abc | is_tr,
    Class = case_when(is_abc ~ "ABC", is_tr ~ "Other", TRUE ~ NA_character_)
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

abund_species <- protein_retained_annot %>%
  semi_join(species_with_proteome, by = "Species_ID") %>%
  group_by(Species_ID, Group_For_Plots, Date, Protein_ID) %>%
  summarise(
    abundance = first(Protein_Abundance_IBAQ_Norm),
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_tr  = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    Panel = "Protein abundance",
    is_transporter = is_abc | is_tr,
    Class = case_when(is_abc ~ "ABC", is_tr ~ "Other", TRUE ~ NA_character_)
  ) %>%
  filter(is_transporter, !is.na(Class)) %>%
  group_by(Panel, Group_For_Plots, Species_ID, Date, Class) %>%
  summarise(Value_tmp = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(Panel, Group_For_Plots, Species_ID, Class) %>%
  summarise(Value_tmp = mean(Value_tmp, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(Value_tmp) & Value_tmp > 0) %>%
  distinct(Panel, Group_For_Plots, Class, Species_ID) %>%
  count(Panel, Group_For_Plots, Class, name = "n_species")

species_counts_all <- bind_rows(genes_species, proteins_species, abund_species)

summary_table_final <- totals_all %>%
  rename(Value_abs = Value) %>%
  group_by(Panel) %>%
  mutate(
    Value_prop = Value_abs / sum(Value_abs, na.rm = TRUE),
    Value_pct  = 100 * Value_prop
  ) %>%
  ungroup() %>%
  left_join(species_counts_all, by = c("Panel", "Group_For_Plots", "Class")) %>%
  mutate(n_species = replace_na(n_species, 0L)) %>%
  rename(
    Trophic_group = Group_For_Plots,
    Transporter_class = Class
  ) %>%
  arrange(Panel, Trophic_group, Transporter_class)

# Optional sanity check: each panel should sum to ~1.0
# summary_table_final %>% group_by(Panel) %>% summarise(sum_fraction = sum(Value_prop))

# -----------------------------
# 9) Export (table + PDF figure)
# -----------------------------
write.csv(
  summary_table_final,
  file = "TABLE_alluvial_transporter_pool_summary_with_species.csv",
  row.names = FALSE
)

ggsave(
  filename = "FIG_alluvial_transporter_pool_proportions.pdf",
  plot     = FIG,
  width    = 190,
  height   = 75,
  units    = "mm",
  device   = cairo_pdf,
  bg       = "white"
)

# Display in viewer
grid.newpage(); grid.draw(FIG)


# ============================================================
# Absolute (not proportion) version of the same alluvial figure
# Uses totals_all (absolute Value) instead of totals_prop
# ============================================================
# Panel-specific y-axis breaks (absolute scale)
y_breaks_abs <- function(panel) {
  switch(panel,
    "Genes"             = scales::breaks_width(5000),
    "Detected proteins" = scales::breaks_width(500),
    "Protein abundance" = scales::breaks_width(2e8)   # 200,000,000
  )
}

y_labels_abs <- function(panel) {
  if (panel == "Protein abundance") {
    scales::label_scientific(digits = 2)
  } else {
    scales::label_number(big.mark = "'")
  }
}

make_alluvial_panel_abs <- function(df_abs, panel_name, panel_letter) {

  d <- df_abs %>% filter(Panel == panel_name)

  ggplot(
    d,
    aes(
      axis1 = Group_For_Plots,
      axis2 = Class,
      y = Value
    )
  ) +
    geom_alluvium(
      aes(fill = Group_For_Plots),
      alpha = 0.90,
      color = "grey25",
      linewidth = 0.25
    ) +
    geom_stratum(
      fill = "grey92",
      color = "black",
      linewidth = 0.35
    ) +
    geom_text(
      stat = "stratum",
      aes(label = after_stat(stratum)),
      size = 3.0
    ) +
    scale_fill_manual(values = grp_cols, drop = FALSE) +
    scale_y_continuous(
      breaks = y_breaks_abs(panel_name),
      labels = y_labels_abs(panel_name),
      expand = expansion(mult = c(0, 0.03))
    ) +
    labs(
      title = paste0(panel_letter, "  ", panel_name),
      x = NULL,
      y = "Absolute transporter pool",
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      legend.position = "none",
      plot.margin = margin(4, 6, 4, 6)
    )
}


pA_abs <- make_alluvial_panel_abs(totals_all, "Genes", "A")
pB_abs <- make_alluvial_panel_abs(totals_all, "Detected proteins", "B")
pC_abs <- make_alluvial_panel_abs(totals_all, "Protein abundance", "C")

FIG_abs <- arrangeGrob(
  arrangeGrob(pA_abs, pB_abs, pC_abs, ncol = 3),
  gLEG,
  ncol = 2,
  widths = c(4.6, 1.1)
)

ggsave(
  "FIG_alluvial_transporter_pool_absolute.pdf",
  plot   = FIG_abs,
  width  = 180,
  height = 60,
  units  = "mm",
  device = pdf)

# Display in viewer
grid.newpage(); grid.draw(FIG_abs)
```


```{r}
############################################################
# Alluvial plots: transporter abundance → substrate bins
# + Summary table (absolute, proportion, percent, n_species)
# ----------------------------------------------------------
# Two panels:
#   A) ABC transporters
#   B) Other transporters
#
# Left stratum: trophic group (Oligo / Intermediate / Copio)
# Right stratum: substrate bin (reduced categories)
# Flow height: summed species-mean protein abundance (iBAQ_norm)
#
# Summary table reports, for each Class × Trophic group × Substrate bin:
#   - Value_abs  : summed species-mean abundance (iBAQ_norm)
#   - Value_prop : within-Class proportion (sums to 1 within ABC, within Other)
#   - Value_pct  : 100 * Value_prop
#   - n_species  : number of unique species contributing (>0 mean signal) to that cell
#
# Proteins annotated to multiple substrate groups (e.g. "Sugars;Alcohols")
# are split and contribute to multiple bins.
############################################################

suppressPackageStartupMessages({
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  library(ggalluvial)
  library(scales)
  library(gridExtra)
  library(grid)
})

# -----------------------------
# 1) Settings and helpers
# -----------------------------
group_levels <- c("Oligo_like", "Intermediate", "Copio_like")

grp_cols <- c(
  Oligo_like   = "#2C7FB8",
  Intermediate = "grey60",
  Copio_like   = "#D7301F"
)

is_transporter_subcat <- function(x) {
  !is.na(x) & grepl("Transporters", x, fixed = TRUE)
}

# -----------------------------
# 2) Protein-level table: Class + Substrate + abundance (iBAQ_norm)
# -----------------------------
prot_abund_substrate <- protein_retained_annot %>%
  group_by(Species_ID, Group_For_Plots, Date, Protein_ID) %>%
  summarise(
    abundance = first(Protein_Abundance_IBAQ_Norm),
    is_abc = any(Transporter_Family == "ABC", na.rm = TRUE),
    is_tr  = any(is_transporter_subcat(KEGG_Subcategory), na.rm = TRUE),
    Substrate_Group = paste(unique(na.omit(Substrate_Group)), collapse = ";"),
    .groups = "drop"
  ) %>%
  mutate(
    Class = case_when(
      is_abc ~ "ABC",
      is_tr  ~ "Other",
      TRUE   ~ NA_character_
    )
  ) %>%
  filter(!is.na(Class)) %>%
  separate_rows(Substrate_Group, sep = ";") %>%
  mutate(
    Substrate_Bin = case_when(
      Substrate_Group %in% c("Sugars", "Alcohols_Polyols") ~ "Sugars / polyols",
      Substrate_Group %in% c("AminoAcids_Peptides", "Polyamines") ~ "Amino acids / peptides",
      Substrate_Group == "Metals_TraceElements" ~ "Metals / trace elements",
      Substrate_Group == "Inorganic_Nutrients" ~ "Inorganic nutrients",
      Substrate_Group == "Diverse_Substrates" ~ "Diverse substrates",
      TRUE ~ "Rest / other"
    ),
    Substrate_Bin = factor(
      Substrate_Bin,
      levels = c(
        "Sugars / polyols",
        "Amino acids / peptides",
        "Metals / trace elements",
        "Inorganic nutrients",
        "Diverse substrates",
        "Rest / other"
      )
    ),
    Group_For_Plots = factor(Group_For_Plots, levels = group_levels)
  )

# -----------------------------
# 3) Aggregate: summed species-mean abundance (iBAQ_norm)
# -----------------------------
# (i) Species × Date × (Class, Bin): sum proteins
# (ii) Species × (Class, Bin): mean across dates
# (iii) Group × (Class, Bin): sum species means
flow_abund_bin <- prot_abund_substrate %>%
  group_by(Group_For_Plots, Species_ID, Date, Class, Substrate_Bin) %>%
  summarise(Value_abs = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(Group_For_Plots, Species_ID, Class, Substrate_Bin) %>%
  summarise(Value_abs = mean(Value_abs, na.rm = TRUE), .groups = "drop") %>%
  group_by(Group_For_Plots, Class, Substrate_Bin) %>%
  summarise(Value_abs = sum(Value_abs, na.rm = TRUE), .groups = "drop")

# -----------------------------
# 4) Summary table: n_species + proportions within Class
# -----------------------------
# Species contributors: species with >0 mean signal in that Class × Bin
species_counts <- prot_abund_substrate %>%
  group_by(Group_For_Plots, Species_ID, Date, Class, Substrate_Bin) %>%
  summarise(Value_tmp = sum(abundance, na.rm = TRUE), .groups = "drop") %>%
  group_by(Group_For_Plots, Species_ID, Class, Substrate_Bin) %>%
  summarise(Value_tmp = mean(Value_tmp, na.rm = TRUE), .groups = "drop") %>%
  filter(!is.na(Value_tmp) & Value_tmp > 0) %>%
  distinct(Group_For_Plots, Class, Substrate_Bin, Species_ID) %>%
  count(Group_For_Plots, Class, Substrate_Bin, name = "n_species")

summary_table_substrate_abund <- flow_abund_bin %>%
  left_join(species_counts, by = c("Group_For_Plots", "Class", "Substrate_Bin")) %>%
  mutate(n_species = replace_na(n_species, 0L)) %>%
  group_by(Class) %>%
  mutate(
    Value_prop = Value_abs / sum(Value_abs, na.rm = TRUE),
    Value_pct  = 100 * Value_prop
  ) %>%
  ungroup() %>%
  rename(Trophic_group = Group_For_Plots) %>%
  arrange(Class, Trophic_group, Substrate_Bin)

# Export summary table
write.csv(
  summary_table_substrate_abund,
  file = "TABLE_transporter_abundance_substrate_bins_summary.csv",
  row.names = FALSE
)

# -----------------------------
# 5) Plotting function
# -----------------------------
make_substrate_alluvial_abund <- function(df, class_name, panel_letter) {

  d <- df %>% filter(Class == class_name)

  ggplot(d, aes(axis1 = Group_For_Plots, axis2 = Substrate_Bin, y = Value_abs)) +
    geom_alluvium(
      aes(fill = Group_For_Plots),
      alpha = 0.90, color = "grey25", linewidth = 0.25
    ) +
    geom_stratum(
      fill = "grey92", color = "black", linewidth = 0.35
    ) +
    geom_text(
      stat = "stratum",
      aes(label = after_stat(stratum)),
      size = 3
    ) +
    scale_fill_manual(values = grp_cols, drop = FALSE) +
    scale_y_continuous(
      labels = label_scientific(digits = 2),
      expand = expansion(mult = c(0, 0.03))
    ) +
    labs(
      title = paste0(panel_letter, "  ", class_name, " transporters"),
      x = NULL,
      y = "Summed species-mean abundance (iBAQ_norm)",
      fill = NULL
    ) +
    theme_bw(base_size = 10) +
    theme(
      plot.title = element_text(face = "bold", size = 11),
      axis.text.x = element_text(size = 9),
      axis.title.y = element_text(size = 9),
      legend.position = "none",
      plot.margin = margin(4, 6, 4, 6)
    )
}

# -----------------------------
# 6) Build figure (ABC vs Other) + export PDF
# -----------------------------
p_ABC   <- make_substrate_alluvial_abund(flow_abund_bin, "ABC", "A")
p_Other <- make_substrate_alluvial_abund(flow_abund_bin, "Other", "B")

FIG_substrate_abund <- arrangeGrob(p_ABC, p_Other, ncol = 1)

ggsave(
  "FIG_alluvial_transporter_abundance_substrate_bins_ABC_vs_Other.pdf",
  plot   = FIG_substrate_abund,
  width  = 60,
  height = 60,
  units  = "mm",
  device = pdf
)

# Optional display:
grid.newpage(); grid.draw(FIG_substrate_abund)
```


